### 2023-03-01 manuscript/main.tex
xml file plus new manuscript folder

% !TeX document-id = {d42653b9-9ea4-48c2-860b-3d6e6e9d8ab0}
% !TeX TXS-program:compile = txs:///pdflatex/[--shell-escape]
\documentclass[12pt]{article} % For LaTeX2e
%\usepackage{neurips_2021}
\usepackage[colorlinks, citecolor={blue}]{hyperref}
\usepackage{url}
\usepackage{amsfonts,amscd,amssymb}
\usepackage{amsthm,amsmath,natbib}
\usepackage{algorithm,algorithmicx,algpseudocode}
\usepackage{bm}
\usepackage{bbm} %bb font numbers
\usepackage[table]{xcolor}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
%\usepackage[nolists]{endfloat}
\usepackage{listings}
\usepackage[textsize=tiny]{todonotes}
\usepackage{tikz}
\usetikzlibrary{shapes.misc}
\usepackage{etoolbox}
\usepackage{appendix}
\usepackage[format=plain,
labelfont={it},
textfont=it]{caption}
\usepackage{subcaption}
\usepackage{wrapfig}
\usepackage{xr}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{authblk}
\usepackage{mathbbol}
\usepackage{thmtools}
\usepackage{amsfonts}

\usetikzlibrary{matrix}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{arrows,shapes}
\usetikzlibrary{decorations}
\usetikzlibrary{decorations.pathmorphing}
\usetikzlibrary{fit}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.geometric}

\newtoggle{quickdraw}
%\toggletrue{quickdraw} % Uncomment this to render more quickly (non-random)

\definecolor{lightgrey}{rgb}{0.9,0.9,0.9}
\definecolor{darkgreen}{rgb}{0,0.3,0}
%\definecolor{darkred}{rgb}{0.3,0,0}

\definecolorset{rgb}{}{}{darkred,0.8,0,0;darkgreen,0,0.5,0;darkblue,0,0,0.5}

%\doublespacing

\newtheorem{thm}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{prop}{Proposition}
\newtheorem{cor}{Corollary}
\newtheorem{remark}{Remark}
\newtheorem{example}{Example}
\newtheorem{mydef}{Definition}
\newtheorem*{assumption}{Assumption}
\newtheorem{clm}{Claim}

\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand*{\fplus}{\genfrac{}{}{0pt}{}{}{+}}
\newcommand*{\fdots}{\genfrac{}{}{0pt}{}{}{\cdots}}
\newcommand{\mb}{\mathbf}
\newcommand{\mc}{\mathcal}
\newcommand{\dx}{\mbox{d}}

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\numTaxa}{N}
\newcommand{\numTraits}{D}
\newcommand{\numDatasets}{M}
%\newcommand{\numLatent}{D}
\newcommand{\taxonIndex}{i}
\newcommand{\traitIndex}{j}
\newcommand{\traitData}{\vec{Y}}
\newcommand{\traitDatum}{y}
\newcommand{\datasetIndex}{m}
\newcommand{\exemplar}{\text{e}}

\newcommand{\sequences}{\vec{S}}
\newcommand{\latentData}{\vec{X}}
\newcommand{\latentdata}{\vec{x}}
\newcommand{\latentDatum}{x}
\newcommand{\phylogeneticParameters}{\boldsymbol{\phi}}
\newcommand{\phylogeny}{{\cal G}}
\newcommand{\tree}{\phylogeny}
%\newcommand{\otherParameters}{\boldsymbol{\
\newcommand{\transpose}{^{t}}

\newcommand{\distanceMatrix}{\mathbf{Y}}
\newcommand{\distance}{y}
\newcommand{\summant}{r}

\newcommand{\cdensity}[2]{\ensuremath{p(#1 \,|\,#2)}}
\newcommand{\density}[1]{\ensuremath{p(#1 )}}

\newcommand{\treeNode}{\nu}

\newcommand{\traitVariance}{\mathbf{\Sigma}}
\newcommand{\nodeIndex}{c}

%\newcommand{\parent}[1]{\mbox{\tiny pa}(#1)}
\newcommand{\parentBig}[1]{\mbox{pa}(#1)}

\newcommand{\sibling}[1]{\mbox{\tiny sib}(#1)}
\newcommand{\siblingBig}[1]{\mbox{sib}(#1)}

\newcommand{\rootMean}{\boldsymbol{\mu}_0}
\newcommand{\rootVarianceScalar}{\tau_0}
\newcommand{\unsequencedVarianceScalar}{\tau_{\exemplar}}
\newcommand{\treeVariance}{\vec{V}_{\tree}}
\newcommand{\hatTreeVariance}{\hat{\vec{V}}_{\tree}}
\newcommand{\mdsSD}{\sigma}
\newcommand{\mdsVariance}{\mdsSD^2}
\newcommand{\residual}{\hat{\traitDatum}}
\newcommand{\modelDistance}{\delta}
\newcommand{\cdf}{\phi}
\newcommand{\normalCDF}[1]{\Phi \left( #1 \right)}

\newcommand{\order}[1]{{\cal O}\hspace{-0.2em}\left( #1 \right)}

\newcommand{\rootNode}{\nu^{\datasetIndex}_{2 \numTaxa_{\datasetIndex} -1 }}
\newcommand{\pathLength}[1]{d(F, #1 )}
\newcommand{\pathLengthNew}[2]{
d_{F}
(
{#1}, {#2}
)
}
\newcommand{\J}{\vec{J}}
\newcommand{\pprime}{^{\prime}}
\newcommand{\otherIndex}{i \pprime}
\def\kronecker{\raisebox{1pt}{\ensuremath{\:\otimes\:}}}

\definecolor{trevorblue}{rgb}{0.330, 0.484, 0.828}
\definecolor{trevoryellow}{rgb}{0.829, 0.680, 0.306}

\allowdisplaybreaks

%\makeatletter
%\def\title@font{\Huge}
%\let\ltx@maketitle\@maketitle
%\def\@maketitle{\bgroup%
%	\let\ltx@title\@title%
%	\def\@title{\resizebox{\textwidth}{!}{%
%			\mbox{\title@font\ltx@title}%
%	}}%
%	\ltx@maketitle%
%	\egroup}
%\makeatother

\title{Approximating the matrix exponential derivative with respect to a structured matrix}

%\date{}

\author{Andrew J.~Holbrook}

\affil{UCLA Biostatistics}

\renewcommand\Authands{ and }

\graphicspath{{figures/}}

\begin{document}

\maketitle

\begin{abstract}

We investigate the accuracy of a first-order approximation to the directional derivative of the matrix exponential evaluated on continuous-time Markov chain (CTMC) generator matrices.

\end{abstract}

\section{Introduction}\label{sec:intro}

%\newcommand{\x}{\mathbf{x}}
%\renewcommand{\v}{\mathbf{v}}

%\newcommand{\TTTheta}{\boldsymbol{\Theta}}
%\newcommand{\tr}{\mbox{tr}}
%\newcommand{\X}{\mathbf{X}}
%\renewcommand{\u}{\mathbf{u}}
\newcommand{\QQ}{\mathbf{Q}}
\newcommand{\DD}{\mathbf{D}}
\newcommand{\MM}{\mathbf{M}}
\newcommand{\JJ}{\mathbf{J}}
\newcommand{\II}{\mathbf{I}}
\newcommand{\RR}{\mathbf{R}}
%\newcommand{\haar}{\mathcal{H}}
%\newcommand{\orthog}{\mathcal{O}}
%\newcommand{\SSigma}{\boldsymbol{\Sigma}}
\newcommand{\Zero}{\boldsymbol{0}}
\newcommand{\ttheta}{\boldsymbol{\theta}}
\newcommand{\Llambda}{\boldsymbol{\Lambda}}
\newcommand{\vecc}{\mbox{vec}}
\newcommand{\CC}{\mathbf{C}}
\newcommand{\BB}{\mathbf{B}}
\renewcommand{\AA}{\mathbf{A}}
\newcommand{\GG}{\mathbf{G}}

Let $\QQ=[q_{ij}]$ be the rate matrix for an $S$-dimensional continuous time Markov chain, where $q_{ij}>0$ for $i\neq j$, and $q_{ii} = -\sum_{j\neq i} q_{ij}$.  In our application of interest, $\QQ$ is almost surely diagonalizable, i.e., $\QQ=\MM \DD \MM^{-1}$ for 
\begin{align*}
	\DD = \begin{pmatrix}
		d_1& 0 &\dots & & \\
		0  & d_2 &0  & \dots& \\
		\vdots &0 & \ddots & \\
		& \vdots& & d_{S-1} &0 \\
		&&&0&0
	\end{pmatrix} \, ,
\end{align*}
where the structure of $\QQ$ implies eigenvalues
\begin{enumerate}
	\item $d_s\in \mathbb{C}$ with $\mbox{Re}(d_s)<0$ for $s\in\{1,\dots,S-1\}$, and
	\item $d_S =0$ corresponding to the eigenvector $\boldsymbol{1}=(1,\dots,1)^T$. 
\end{enumerate} 
We may further define a generalized inverse $\QQ^+=\MM \DD^{+} \MM^{-1}$ of $\QQ$ using $\DD^+$, the Moore-Penrose pseudoinverse obtained by inverting the non-zero elements of $\DD$.   Finally, we note that diagonalizability implies $\DD$ and $\DD^+$ commute, i.e., $\DD^+\DD=\DD\DD^+$.

 We are interested in efficiently approximating the directional derivative of the matrix exponential
\begin{align*}
	e^{t\QQ} = \sum_{k=0}^\infty \frac{t^k\QQ^k}{k!}
\end{align*}
for arbitrary $t>0$ using the formula \citep{najfeld1995derivatives}
\begin{align}\label{eq:deriv}
	\nabla_{\JJ} e^{t\QQ}  =  e^{t\QQ}  \sum_{k=0}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \, .
\end{align}
Here, $\JJ$ gives the direction and $\{\JJ,\QQ^k\}$ is the matrix commutator series satisfying the recursion
\begin{align*}
	\{\JJ,\QQ^0\} &= \JJ \, , \\ 
	\{\JJ,\QQ^k\} &= [\{\JJ,\QQ^{k-1}\} , \QQ]  = \{\JJ,\QQ^{k-1}\} \QQ - \QQ \{\JJ,\QQ^{k-1}\} \, .
\end{align*}
Thus, the first-order approximation to derivative \eqref{eq:deriv} is 
\begin{align}\label{eq:firstOrder}
	\widetilde{\nabla}_{\JJ} e^{t\QQ}= t e^{t\QQ}\JJ \, .
\end{align}
We want to bound the truncation error
\begin{align*}
\mbox{err}(t,\QQ,\JJ)	:=\lVert \nabla_{\JJ} e^{t\QQ} - \widetilde{\nabla}_{\JJ} e^{t\QQ} \rVert \, ,
\end{align*}
 where $\lVert \cdot \rVert$ denotes an appropriate matrix norm.  In this context, we find that naive bounds (Appendix \ref{sec:loose}) that do not exploit the structure of $\QQ$  are rather loose (Figure \ref{fig:frob}).  
 In our efforts to fully leverage the structure of $\QQ$, we will make use of the general formula for the matrix commutator series  \citep{volkin1968iterated}
\begin{align}\label{eq:commSeries}
	\{ \JJ, \QQ^k\} = \sum_{r=0}^k (-1)^{k+r}  {k \choose r} \QQ^{k-r} \JJ \QQ^r  \, .
\end{align}

\subsection{An empirically tight asymptotic estimate}

We begin by noting
\begin{align}\label{eq:ub1}
	\mbox{err}(t,\QQ,\JJ) &=  \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \rVert \\ \nonumber
	&\leq  \lVert  e^{t\QQ}\JJ  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^k \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
		&=  \lVert  e^{t\QQ}\JJ \QQ^+  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^{k+1} \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
		&=  \lVert  e^{t\QQ}\JJ \QQ^+ (e^{t\QQ} - t\QQ -\II) \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
			&\leq \lVert  e^{t\QQ}\JJ \QQ^+ ( t\QQ +\II) \rVert  + \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert  +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
			&=: \mathfrak{a} + \mathfrak{b} + \mathfrak{c} \, .
\end{align}
Empirical experiments using the Frobenius and infinity norms (Figures \ref{fig:frob} and \ref{fig:inf}, respectively) show that the first term
\begin{align}
	 \mathfrak{a} = \left\Vert e^{t\QQ} \JJ \QQ^+(t\QQ+\II)  \right\Vert \, ,
\end{align}
although not a strict upper bound, is asymptotically tight for the true error as $t\rightarrow \infty$. We would, therefore, like to show that the terms $\mathfrak{b}$ and $\mathfrak{c}$ go to zero as $t$ diverges.

\section{Using the structure of $\QQ$}

We use the fact that
\begin{align*}
	e^{t\QQ}\QQ = \MM \exp(t\DD) \DD \MM^{-1} =  \MM \begin{pmatrix}
	e^{td_1}	d_1& 0 &\dots & & \\
		0  & e^{td_2}d_2 &0  & \dots& \\
		\vdots &0 & \ddots & \\
		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
		&&&0&0
	\end{pmatrix} \MM^{-1} \, .
\end{align*}
Thus, assuming use of a submultiplicative norm leads to the result
\begin{align*}
	\lVert e^{t\QQ}\QQ \rVert &\leq \left\Vert  \begin{pmatrix}
		e^{td_1}	d_1& 0 &\dots & & \\
		0  & e^{td_2}d_2 &0  & \dots& \\
		\vdots &0 & \ddots & \\
		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
		&&&0&0
	\end{pmatrix}  \right\Vert  \lVert \MM\rVert   \lVert \MM^{-1}\rVert  \, .
\end{align*}
For, e.g., the Frobenius norm, we have
\begin{align*}
\lVert e^{t\QQ}\QQ \rVert_F	\leq \sqrt{S-1} \max_s \{|e^{td_{s}} d_{s}| \}  \lVert \MM\rVert_F   \lVert \MM^{-1}\rVert_F  \, ,
\end{align*}
where we  use the fact that $\lVert \cdot \rVert_F \leq \sqrt{r} \lVert \cdot \rVert_2$ for any rank $r$ matrix and $\lVert \cdot \rVert_2$ the spectral norm.  We therefore write
\begin{align}
	\lVert e^{t\QQ}\QQ \rVert_F = \order{\max_s \left\{|e^{td_{s}} d_{s}| \right\}} \,,  \quad 	\lVert e^{t\QQ}\QQ^+ \rVert_F = \order{\max_s \left\{ \left| e^{td_{s}} \frac{1}{d_{s}}\right| \right\}} \, .
\end{align}
Because $\mbox{Re}(d_s)<0$ for $s\in \{1,\dots,S-1\}$, we have
\begin{align}
	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert_F = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert_F = 0\, ,
\end{align}
and, by equivalence of norms, we have
\begin{align}
	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert = 0\, ,
\end{align}
in general for any matrix norm $\lVert \cdot \rVert$. Moreover, the matrix $e^{t\QQ}$ is a stochastic matrix with rows summing to unity. Therefore $\lVert e^{t\QQ}\rVert_\infty = 1$, and, by norm equivalence, $\lVert e^{t\QQ}\rVert = \mathcal{O}(1)$.
 These results lead to the following fact for term $\mathfrak{b}$ in the upper bound of \eqref{eq:ub1}:
\begin{align*}
	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert \leq  \lim_{t\rightarrow \infty} \lVert  e^{t\QQ} \rVert \lVert \JJ  \rVert \lVert \QQ^+ e^{t\QQ} \rVert = 0 \, .
\end{align*}
Thus, we obtain the intermediary result for $t$ large:
\begin{align*}
\lim_{t\rightarrow \infty}\mbox{err}(t,\QQ,\JJ) \leq \lim_{t\rightarrow \infty}  \mathfrak{a}+ \mathfrak{c} \, .
\end{align*}
It remains to show that $\mathfrak{c}$ goes to zero as $t$ diverges, i.e., 
\begin{align}
	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert = 0\, .
\end{align}

\section{Random thoughts}

\subsection{Another form}

\newcommand{\PPsi}{\boldsymbol{\Psi}}

\citet{najfeld1995derivatives} use the integral representation to prove the interesting result
\begin{align*}
	\nabla_{\JJ} e^{t\QQ} = \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ \PPsi(t)  \right) \MM^{-1} \, ,
\end{align*}
where $\PPsi(t)$ has elements
\begin{align*}
		[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
		t e^{td_i}\, ,	&  d_i = d_j \\
		\frac{e^{td_i}-e^{td_j}}{d_i-d_j} \, ,& d_i \neq d_j 
	\end{array} \, .
\end{align*}
For our application, the eigenvalues equal one another with measure zero.  Taking $t$ large, we have
\begin{align*}
	[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
		t e^{td_i}\, ,	&  i = j \\
		0 \, ,& i \neq j 
	\end{array} \, .
\end{align*}
It follows that, again for $t$ large,
\begin{align*}
	\mbox{err}(t,\QQ,\JJ) &\approx \lVert   \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ t e^{t\DD}  \right) \MM^{-1}       -  t e^{t\QQ} \JJ            \rVert \\
	&= t\lVert   \MM \rVert \lVert \left(  \left(\MM^{-1} \JJ \MM\right) \circ  e^{t\DD}  \right) \MM^{-1}       -   e^{t\DD}\MM^{-1} \JJ            \rVert \, .
\end{align*}

\appendix

\section{Loose bounds}\label{sec:loose}

The following loose bounds do not leverage the structure of the CTMC generator matrix $\QQ$.

\subsection{Upper bound 1}

For a submultiplicative matrix norm, we have
\begin{align*}
	\Vert \{\MM_1, \MM_2^k \}  \Vert \leq 2^k \Vert \MM_1\Vert \Vert\MM_2 \Vert ^k \, .
\end{align*}
We use this to obtain our first upper bound on $err_{ij}(\QQ)$:
\begin{align*}
	err_{ij}(\QQ) &= \left\Vert e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
	&\leq \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \left\Vert \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
	&=  \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \frac{t}{k+1} \frac{(2t \Vert \QQ \Vert)^k}{k!}\\
	&= \Vert e^{t\QQ} \Vert   \frac{\exp(2t\Vert \QQ \Vert)-2t\Vert\QQ\Vert-1}{2\Vert \QQ \Vert} 
\end{align*}

\subsection{Upper bound 2}

We start with another representation of the derivative \citep{al2009computing}:
\begin{align*}
	\nabla_{\JJ} e^{t\QQ} = \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} 
\end{align*}
and the fact that
the fact that, for the Frobenius norm $\Vert \cdot \Vert_F$, we have \citep{wu2010short}
\begin{align*}
	\Vert [\MM_1,\MM_2] \Vert_F \leq \sqrt{2} \Vert \MM_1 \Vert_F \Vert \MM_2\Vert_F \, .
\end{align*}
Using these results, the truncation error may be rewritten
\begin{align*}
	err_{ij}(\QQ)  &=\left\Vert - te^{t\QQ} \JJ  +  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}\right\Vert_F \\
	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \left( \left( \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} \right) - k \QQ^{k-1} \JJ \right)  \right\Vert_F\\
	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ \right\Vert_F \\
	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ\right\Vert_F \\
	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}\left( \JJ \QQ^{k-p}  -  \QQ^{k-p} \JJ \right)\right\Vert_F \\
	&= \left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}[\JJ, \QQ^{k-p}] \right\Vert_F  \\
	&\leq  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{p-1} \Vert_F \Vert[\JJ, \QQ^{k-p}]\Vert_F 
	\leq \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{k-1} \Vert_F  \\
	&= \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} (k-1) \Vert \QQ \Vert_F^{k-1} = \sqrt{2} \frac{e^{t\Vert \QQ\Vert_F}(t\Vert\QQ\Vert_F - 1) + 1}{\Vert \QQ \Vert_F} \, .
\end{align*}

\begin{figure}[!p]
	\centering
	\includegraphics[width=\linewidth]{frobeniusScaling.pdf}
	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using Frobenius norm.}\label{fig:frob}
\end{figure}

\begin{figure}[!p]
	\centering
	\includegraphics[width=\linewidth]{infinityScaling.pdf}
	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using infinity norm.}\label{fig:inf}
\end{figure}

\bibliographystyle{sysbio}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2023-03-01 code/angles.R

@@ -115,6 +115,48 @@ gg2 <- ggplot(df,aes(x=t,y=Error,color=Dimension)) +
   theme_bw()
 gg2
 
+# df      <- data.frame(Error=numeric(),Dimension=character(),t=numeric())
+# 
+# 
+# for(n in 3:50) {
+#   for (t in seq(from=10^(-10),to=10,length.out=100)) {
+#     
+#     Q       <- matrix(rexp(n^2),n,n)
+#     diag(Q) <- 0
+#     diag(Q) <- - rowSums(Q)
+#     E       <- matrix(0,n,n)
+#     d <- sample(1:n,size=1)
+#     dprime <- sample((1:n)[-d],size=1)
+#     E[d,dprime]   <- 1
+#     E[d,d]  <- -1
+#     #E[2,2] <- 1
+#     drv <- blockwise(t=t,Q,E)
+#     frstRdr <- expm::expm(t*Q) %*% E *t # correct approx
+#     
+#     df <- rbind(df,
+#                 c(acos(sum(diag(frstRdr%*%drv))/
+#                          (norm(frstRdr,type="F")*norm(drv,type="F")))*(360/(2*pi)),
+#                   n,
+#                   t))
+#     
+#   }
+# }
+# 
+# colnames(df) <- c("Error","Dimension","t")
+# df$Dimension <- factor(df$Dimension)
+# df$Error <- as.numeric(df$Error)
+# df$t     <- as.numeric(df$t)
+# 
+# gg3 <- ggplot(df,aes(x=t,y=Error,color=Dimension)) +
+#   geom_line(size=1.5) +
+#   geom_hline(yintercept = 90) +
+#   # scale_y_log10() +
+#   ylab(" ") +
+#   coord_cartesian(ylim=c(0,100)) +
+#   ggtitle("Off/on-diagonal directional derivative vs 1st order approx.") +
+#   theme_bw()
+# gg3
+
 library(ggpubr)
 ggsave(ggarrange(gg,gg2,
                  ncol=2,nrow=1,

### 2023-03-01 code/run_beast.sh
proper paths

@@ -1,0 +1,1 @@
+java -jar -Djava.library.path=/usr/lib/jvm /home/andrew/beast-mcmc/build/dist/beast.jar -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml

### 2023-03-01 code/run_beast.sh
detect beagle maybe

@@ -1,1 +1,1 @@
-java -jar -Djava.library.path=/usr/lib/jvm /home/andrew/beast-mcmc/build/dist/beast.jar -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml
+java -jar -Djava.library.path=/usr/lib/jvm /home/andrew/beast-mcmc/build/dist/beast.jar -beagle_order 1 -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml

### 2023-03-01 code/run_beast.sh
ok now

@@ -1,1 +1,1 @@
-java -jar -Djava.library.path=/usr/lib/jvm /home/andrew/beast-mcmc/build/dist/beast.jar -beagle_order 1 -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml
+java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -beagle_order 1 -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml

### 2023-03-01 code/run_beast.sh
beagle stuff

@@ -1,1 +1,1 @@
-java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -beagle_order 1 -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml
+java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -beagle_gpu -beagle_order 1 -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml

### 2023-03-01 code/run_beast.sh
trying not to use beagle

@@ -1,1 +1,1 @@
-java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -beagle_gpu -beagle_order 1 -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml
+java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -java -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml

### 2023-03-01 code/run_beast.sh
finally running

@@ -1,1 +1,1 @@
-java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -java -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml
+java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.Ed.xml

### 2023-03-03 code/run_beast_noRandom.sh
ready for run without random effects

@@ -1,0 +1,1 @@
+java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.noRandom.xml

### 2023-03-07 code/angles.R
interesting angles result

@@ -44,14 +44,14 @@ for (t in seq(from=10^(-10),to=10,length.out=100)) {
   diag(Q) <- 0
   diag(Q) <- - rowSums(Q)
   E       <- matrix(0,n,n)
-  #E[sample(1:n,size=1),sample(1:n,size=1)]   <- 1
-  d <- sample(1:n,size=1)
-  E[d,d] <- 1
+  E[sample(1:n,size=1),sample(1:n,size=1)]   <- 1
+  #d <- sample(1:n,size=1)
+  #E[d,d] <- 1
   drv <- blockwise(t=t,Q,E)
   frstRdr <- expm::expm(t*Q) %*% E *t # correct approx
 
   df <- rbind(df,
-              c(acos(sum(diag(frstRdr%*%drv))/
+              c(acos(sum(diag(t(frstRdr)%*%drv))/
                   (norm(frstRdr,type="F")*norm(drv,type="F")))*(360/(2*pi)),
                 n,
                 t))
@@ -70,50 +70,9 @@ gg <- ggplot(df,aes(x=t,y=Error,color=Dimension)) +
   # scale_y_log10() +
   ylab("Angle (deg.)") +
   coord_cartesian(ylim=c(0,100)) +
-  ggtitle("Diagonal directional derivative vs 1st order approx.") +
+  ggtitle("Directional derivative vs 1st order approx.") +
   theme_bw()
 gg
-
-df      <- data.frame(Error=numeric(),Dimension=character(),t=numeric())
-
-
-for(n in 3:50) {
-  for (t in seq(from=10^(-10),to=10,length.out=100)) {
-    
-    Q       <- matrix(rexp(n^2),n,n)
-    diag(Q) <- 0
-    diag(Q) <- - rowSums(Q)
-    E       <- matrix(0,n,n)
-    d <- sample(1:n,size=1)
-    dprime <- sample((1:n)[-d],size=1)
-    E[d,dprime]   <- 1
-    #E[2,2] <- 1
-    drv <- blockwise(t=t,Q,E)
-    frstRdr <- expm::expm(t*Q) %*% E *t # correct approx
-    
-    df <- rbind(df,
-                c(acos(sum(diag(frstRdr%*%drv))/
-                         (norm(frstRdr,type="F")*norm(drv,type="F")))*(360/(2*pi)),
-                  n,
-                  t))
-    
-  }
-}
-
-colnames(df) <- c("Error","Dimension","t")
[…]

### 2023-03-08 code/relativeErrorsNormalized.R
ready for short runs

@@ -1,0 +1,335 @@
+setwd("~/expmDerivative/")
+library(ggplot2)
+
+blockwise <- function(t,Q,E) {
+  B1  <- cbind(Q,E)
+  B2  <- cbind(Q*0,Q)
+  B   <- rbind(B1,B2) 
+  etB <- expm::expm(t*B)
+  return(etB[1:(dim(etB)[1]/2),(dim(etB)[1]/2+1):dim(etB)[1]])
+}
+
+cnvrgncBnd <- function(t, Q, Jij) {
+  n <- dim(Q)[1]
+  eig.obj <- eigen(Q)
+  inv.spctrm <- 1/eig.obj$values
+  inv.spctrm[length(inv.spctrm)] <- 0
+  pinv    <- Re(eig.obj$vectors %*% diag(inv.spctrm) %*% solve(eig.obj$vectors))
+  mat <- expm::expm(t*Q)%*%Jij%*%pinv%*%(Q*t+diag(n))
+  return(norm(mat,type="F"))
+}
+
+nvBnd1 <- function(t, Q, Jij) {
+  nrmXptQ <- norm(expm::expm(t*Q),type="F")
+  nrmQ    <- norm(Q,type="F")
+  out     <- nrmXptQ / (2*nrmQ) * (exp(2*t*nrmQ) - 2*t*nrmQ - 1)
+  return(out)
+}
+
+nvBnd2 <- function(t, Q, Jij) {
+  nrmQ    <- norm(Q,type="F")
+  out     <- sqrt(2)/nrmQ * (exp(t*nrmQ)*(t*nrmQ - 1) + 1)
+  return(out)
+}
+
+################################################################################
+
+df      <- data.frame(Error=numeric(),Type=character(),t=numeric())
+n       <- 10
+Q       <- matrix(rexp(n^2),n,n)
+diag(Q) <- 0
+diag(Q) <- - rowSums(Q)
+E       <- matrix(0,n,n)
+d <- sample(1:n,size=1)
+dprime <- sample((1:n)[-d],size=1)
+E[d,dprime]   <- 1
+E[d,d]  <- -1
+
+for (t in seq(from=10^(-10),to=10,length.out=100)) {
+
+  drv <- blockwise(t=t,Q,E)
+  frstRdr <- expm::expm(t*Q) %*% E *t # correct approx
+
+  df <- rbind(df,
+              c(norm(frstRdr-drv,type = "F"),
+                "Error",
+                t))
+  df <- rbind(df,
[…]

### 2023-03-08 code/run_beast_noRandomyesHMC.sh

@@ -1,1 +1,0 @@
-java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -seed 666 -overwrite /home/andrew/expmDerivative/code/covid10Mar_285plusNYC.noRandomyesHMC.xml

### 2023-03-08 code/run_beast_shortEdnoHmc.sh

@@ -1,0 +1,1 @@
+java -jar -Djava.library.path=/home/andrew/lib /home/andrew/beast-mcmc/build/dist/beast.jar -seed 666 -overwrite /home/andrew/expmDerivative/code/short.EdnoHmc.xml

### 2023-03-09 manuscript/main.tex
need to start manuscript

@@
 
-\title{Approximating the matrix exponential derivative with respect to a structured matrix}
+\title{Approximating the matrix exponential derivative with respect to the CTMC generator matrix, with application to global SARS-CoV-2}
 
@@
 
-\author{Andrew J.~Holbrook}
+\author[1]{Andrew J.~Holbrook}
+\author[1]{Andrew F.~Magee}
+\author[2]{Nathan E.~Glatt-Holtz}
+\author[1,3,4]{Marc A.~Suchard}
 
-\affil{UCLA Biostatistics}
+ \affil[1]{Department of Biostatistics, University of California, Los Angeles}
+ \affil[2]{Department of Mathematics, Tulane University}
+ \affil[3]{Department of Biomathematics, University of California, Los Angeles}
+ \affil[4]{Department of Human Genetics, University of California, Los Angeles}
 
@@
 
-Let $\QQ=[q_{ij}]$ be the rate matrix for an $S$-dimensional continuous time Markov chain, where $q_{ij}>0$ for $i\neq j$, and $q_{ii} = -\sum_{j\neq i} q_{ij}$.
+\section{Global spread of SARS-CoV-2}
+
+%Let $\QQ=[q_{ij}]$ be the rate matrix for an $S$-dimensional continuous time Markov chain, where $q_{ij}>0$ for $i\neq j$, and $q_{ii} = -\sum_{j\neq i} q_{ij}$.
 In our application of interest, $\QQ$ is almost surely diagonalizable, i.e., $\QQ=\MM \DD \MM^{-1}$ for
-\begin{align*}
-	\DD = \begin{pmatrix}
-		d_1& 0 &\dots & & \\
-		0  & d_2 &0  & \dots& \\
-		\vdots &0 & \ddots & \\
-		& \vdots& & d_{S-1} &0 \\
-		&&&0&0
-	\end{pmatrix} \, ,
-\end{align*}
-where the structure of $\QQ$ implies eigenvalues
-\begin{enumerate}
-	\item $d_s\in \mathbb{C}$ with $\mbox{Re}(d_s)<0$ for $s\in\{1,\dots,S-1\}$, and
-	\item $d_S =0$ corresponding to the eigenvector $\boldsymbol{1}=(1,\dots,1)^T$. 
-\end{enumerate} 
-We may further define a generalized inverse $\QQ^+=\MM \DD^{+} \MM^{-1}$ of $\QQ$ using $\DD^+$, the Moore-Penrose pseudoinverse obtained by inverting the non-zero elements of $\DD$.
+%\begin{align*}
+%	\DD = \begin{pmatrix}
+%		d_1& 0 &\dots & & \\
+%		0  & d_2 &0  & \dots& \\
+%		\vdots &0 & \ddots & \\
+%		& \vdots& & d_{S-1} &0 \\
+%		&&&0&0
+%	\end{pmatrix} \, ,
+%\end{align*}
+%where the structure of $\QQ$ implies eigenvalues
+%\begin{enumerate}
+%	\item $d_s\in \mathbb{C}$ with $\mbox{Re}(d_s)<0$ for $s\in\{1,\dots,S-1\}$, and
+%	\item $d_S =0$ corresponding to the eigenvector $\boldsymbol{1}=(1,\dots,1)^T$. 
+%\end{enumerate} 
+%We may further define a generalized inverse $\QQ^+=\MM \DD^{+} \MM^{-1}$ of $\QQ$ using $\DD^+$, the Moore-Penrose pseudoinverse obtained by inverting the non-zero elements of $\DD$.
 Finally, we note that diagonalizability implies $\DD$ and $\DD^+$ commute, i.e., $\DD^+\DD=\DD\DD^+$.
-
- We are interested in efficiently approximating the directional derivative of the matrix exponential
-\begin{align*}
-	e^{t\QQ} = \sum_{k=0}^\infty \frac{t^k\QQ^k}{k!}
-\end{align*}
-for arbitrary $t>0$ using the formula \citep{najfeld1995derivatives}
-\begin{align}\label{eq:deriv}
-	\nabla_{\JJ} e^{t\QQ}  =  e^{t\QQ}  \sum_{k=0}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \, .
-\end{align}
-Here, $\JJ$ gives the direction and $\{\JJ,\QQ^k\}$ is the matrix commutator series satisfying the recursion
-\begin{align*}
-	\{\JJ,\QQ^0\} &= \JJ \, , \\ 
-	\{\JJ,\QQ^k\} &= [\{\JJ,\QQ^{k-1}\} , \QQ]  = \{\JJ,\QQ^{k-1}\} \QQ - \QQ \{\JJ,\QQ^{k-1}\} \, .
-\end{align*}
-Thus, the first-order approximation to derivative \eqref{eq:deriv} is 
-\begin{align}\label{eq:firstOrder}
-	\widetilde{\nabla}_{\JJ} e^{t\QQ}= t e^{t\QQ}\JJ \, .
-\end{align}
-We want to bound the truncation error
-\begin{align*}
-\mbox{err}(t,\QQ,\JJ)	:=\lVert \nabla_{\JJ} e^{t\QQ} - \widetilde{\nabla}_{\JJ} e^{t\QQ} \rVert \, ,
-\end{align*}
- where $\lVert \cdot \rVert$ denotes an appropriate matrix norm.
+%
+%
+% We are interested in efficiently approximating the directional derivative of the matrix exponential
+%\begin{align*}
+%	e^{t\QQ} = \sum_{k=0}^\infty \frac{t^k\QQ^k}{k!}
+%\end{align*}
+%for arbitrary $t>0$ using the formula \citep{najfeld1995derivatives}
+%\begin{align}\label{eq:deriv}
+%	\nabla_{\JJ} e^{t\QQ}  =  e^{t\QQ}  \sum_{k=0}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \, .
+%\end{align}
+%Here, $\JJ$ gives the direction and $\{\JJ,\QQ^k\}$ is the matrix commutator series satisfying the recursion
+%\begin{align*}
+%	\{\JJ,\QQ^0\} &= \JJ \, , \\ 
+%	\{\JJ,\QQ^k\} &= [\{\JJ,\QQ^{k-1}\} , \QQ]  = \{\JJ,\QQ^{k-1}\} \QQ - \QQ \{\JJ,\QQ^{k-1}\} \, .
+%\end{align*}
+%Thus, the first-order approximation to derivative \eqref{eq:deriv} is 
+%\begin{align}\label{eq:firstOrder}
+%	\widetilde{\nabla}_{\JJ} e^{t\QQ}= t e^{t\QQ}\JJ \, .
+%\end{align}
+%We want to bound the truncation error
+%\begin{align*}
+%\mbox{err}(t,\QQ,\JJ)	:=\lVert \nabla_{\JJ} e^{t\QQ} - \widetilde{\nabla}_{\JJ} e^{t\QQ} \rVert \, ,
+%\end{align*}
+% where $\lVert \cdot \rVert$ denotes an appropriate matrix norm.
 In this context, we find that naive bounds (Appendix \ref{sec:loose}) that do not exploit the structure of $\QQ$  are rather loose (Figure \ref{fig:frob}).
- In our efforts to fully leverage the structure of $\QQ$, we will make use of the general formula for the matrix commutator series  \citep{volkin1968iterated}
-\begin{align}\label{eq:commSeries}
-	\{ \JJ, \QQ^k\} = \sum_{r=0}^k (-1)^{k+r}  {k \choose r} \QQ^{k-r} \JJ \QQ^r  \, .
-\end{align}
-
-\subsection{An empirically tight asymptotic estimate}
-
-We begin by noting
-\begin{align}\label{eq:ub1}
-	\mbox{err}(t,\QQ,\JJ) &=  \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \rVert \\ \nonumber
-	&\leq  \lVert  e^{t\QQ}\JJ  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^k \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
-		&=  \lVert  e^{t\QQ}\JJ \QQ^+  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^{k+1} \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
-		&=  \lVert  e^{t\QQ}\JJ \QQ^+ (e^{t\QQ} - t\QQ -\II) \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
-			&\leq \lVert  e^{t\QQ}\JJ \QQ^+ ( t\QQ +\II) \rVert  + \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert  +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
-			&=: \mathfrak{a} + \mathfrak{b} + \mathfrak{c} \, .
-\end{align}
-Empirical experiments using the Frobenius and infinity norms (Figures \ref{fig:frob} and \ref{fig:inf}, respectively) show that the first term
-\begin{align}
-	 \mathfrak{a} = \left\Vert e^{t\QQ} \JJ \QQ^+(t\QQ+\II)  \right\Vert \, ,
-\end{align}
-although not a strict upper bound, is asymptotically tight for the true error as $t\rightarrow \infty$.
+% In our efforts to fully leverage the structure of $\QQ$, we will make use of the general formula for the matrix commutator series  \citep{volkin1968iterated}
+%\begin{align}\label{eq:commSeries}
+%	\{ \JJ, \QQ^k\} = \sum_{r=0}^k (-1)^{k+r}  {k \choose r} \QQ^{k-r} \JJ \QQ^r  \, .
+%\end{align}
+%
+%\subsection{An empirically tight asymptotic estimate}
+%
+%We begin by noting
+%\begin{align}\label{eq:ub1}
+%	\mbox{err}(t,\QQ,\JJ) &=  \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \rVert \\ \nonumber
+%	&\leq  \lVert  e^{t\QQ}\JJ  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^k \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
+%		&=  \lVert  e^{t\QQ}\JJ \QQ^+  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^{k+1} \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
+%		&=  \lVert  e^{t\QQ}\JJ \QQ^+ (e^{t\QQ} - t\QQ -\II) \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
+%			&\leq \lVert  e^{t\QQ}\JJ \QQ^+ ( t\QQ +\II) \rVert  + \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert  +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
+%			&=: \mathfrak{a} + \mathfrak{b} + \mathfrak{c} \, .
+%\end{align}
+%Empirical experiments using the Frobenius and infinity norms (Figures \ref{fig:frob} and \ref{fig:inf}, respectively) show that the first term
+%\begin{align}
+%	 \mathfrak{a} = \left\Vert e^{t\QQ} \JJ \QQ^+(t\QQ+\II)  \right\Vert \, ,
+%\end{align}
+%although not a strict upper bound, is asymptotically tight for the true error as $t\rightarrow \infty$.
 We would, therefore, like to show that the terms $\mathfrak{b}$ and $\mathfrak{c}$ go to zero as $t$ diverges.
-
-\section{Using the structure of $\QQ$}
-
-We use the fact that
-\begin{align*}
-	e^{t\QQ}\QQ = \MM \exp(t\DD) \DD \MM^{-1} =  \MM \begin{pmatrix}
-	e^{td_1}	d_1& 0 &\dots & & \\
-		0  & e^{td_2}d_2 &0  & \dots& \\
-		\vdots &0 & \ddots & \\
-		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
-		&&&0&0
-	\end{pmatrix} \MM^{-1} \, .
-\end{align*}
-Thus, assuming use of a submultiplicative norm leads to the result
-\begin{align*}
-	\lVert e^{t\QQ}\QQ \rVert &\leq \left\Vert  \begin{pmatrix}
-		e^{td_1}	d_1& 0 &\dots & & \\
-		0  & e^{td_2}d_2 &0  & \dots& \\
-		\vdots &0 & \ddots & \\
-		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
-		&&&0&0
-	\end{pmatrix}  \right\Vert  \lVert \MM\rVert   \lVert \MM^{-1}\rVert  \, .
-\end{align*}
-For, e.g., the Frobenius norm, we have
-\begin{align*}
-\lVert e^{t\QQ}\QQ \rVert_F	\leq \sqrt{S-1} \max_s \{|e^{td_{s}} d_{s}| \}  \lVert \MM\rVert_F   \lVert \MM^{-1}\rVert_F  \, ,
-\end{align*}
-where we  use the fact that $\lVert \cdot \rVert_F \leq \sqrt{r} \lVert \cdot \rVert_2$ for any rank $r$ matrix and $\lVert \cdot \rVert_2$ the spectral norm.  We therefore write
-\begin{align}
-	\lVert e^{t\QQ}\QQ \rVert_F = \order{\max_s \left\{|e^{td_{s}} d_{s}| \right\}} \,,  \quad 	\lVert e^{t\QQ}\QQ^+ \rVert_F = \order{\max_s \left\{ \left| e^{td_{s}} \frac{1}{d_{s}}\right| \right\}} \, .
-\end{align}
-Because $\mbox{Re}(d_s)<0$ for $s\in \{1,\dots,S-1\}$, we have
-\begin{align}
-	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert_F = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert_F = 0\, ,
-\end{align}
-and, by equivalence of norms, we have
-\begin{align}
-	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert = 0\, ,
-\end{align}
-in general for any matrix norm $\lVert \cdot \rVert$.
+%
+%
+%
+%
+%
+%
+%
+%
+%\section{Using the structure of $\QQ$}
+%
+%We use the fact that
+%\begin{align*}
+%	e^{t\QQ}\QQ = \MM \exp(t\DD) \DD \MM^{-1} =  \MM \begin{pmatrix}
+%	e^{td_1}	d_1& 0 &\dots & & \\
+%		0  & e^{td_2}d_2 &0  & \dots& \\
+%		\vdots &0 & \ddots & \\
+%		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
+%		&&&0&0
+%	\end{pmatrix} \MM^{-1} \, .
+%\end{align*}
+%Thus, assuming use of a submultiplicative norm leads to the result
+%\begin{align*}
+%	\lVert e^{t\QQ}\QQ \rVert &\leq \left\Vert  \begin{pmatrix}
+%		e^{td_1}	d_1& 0 &\dots & & \\
+%		0  & e^{td_2}d_2 &0  & \dots& \\
+%		\vdots &0 & \ddots & \\
+%		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
+%		&&&0&0
+%	\end{pmatrix}  \right\Vert  \lVert \MM\rVert   \lVert \MM^{-1}\rVert  \, .
+%\end{align*}
+%For, e.g., the Frobenius norm, we have
+%\begin{align*}
+%\lVert e^{t\QQ}\QQ \rVert_F	\leq \sqrt{S-1} \max_s \{|e^{td_{s}} d_{s}| \}  \lVert \MM\rVert_F   \lVert \MM^{-1}\rVert_F  \, ,
+%\end{align*}
+%where we  use the fact that $\lVert \cdot \rVert_F \leq \sqrt{r} \lVert \cdot \rVert_2$ for any rank $r$ matrix and $\lVert \cdot \rVert_2$ the spectral norm.  We therefore write
+%\begin{align}
+%	\lVert e^{t\QQ}\QQ \rVert_F = \order{\max_s \left\{|e^{td_{s}} d_{s}| \right\}} \,,  \quad 	\lVert e^{t\QQ}\QQ^+ \rVert_F = \order{\max_s \left\{ \left| e^{td_{s}} \frac{1}{d_{s}}\right| \right\}} \, .
+%\end{align}
+%Because $\mbox{Re}(d_s)<0$ for $s\in \{1,\dots,S-1\}$, we have
+%\begin{align}
+%	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert_F = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert_F = 0\, ,
+%\end{align}
+%and, by equivalence of norms, we have
+%\begin{align}
+%	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert = 0\, ,
+%\end{align}
+%in general for any matrix norm $\lVert \cdot \rVert$.
 Moreover, the matrix $e^{t\QQ}$ is a stochastic matrix with rows summing to unity.
 Therefore $\lVert e^{t\QQ}\rVert_\infty = 1$, and, by norm equivalence, $\lVert e^{t\QQ}\rVert = \mathcal{O}(1)$.
- These results lead to the following fact for term $\mathfrak{b}$ in the upper bound of \eqref{eq:ub1}:
-\begin{align*}
-	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert \leq  \lim_{t\rightarrow \infty} \lVert  e^{t\QQ} \rVert \lVert \JJ  \rVert \lVert \QQ^+ e^{t\QQ} \rVert = 0 \, .
-\end{align*}
-Thus, we obtain the intermediary result for $t$ large:
-\begin{align*}
-\lim_{t\rightarrow \infty}\mbox{err}(t,\QQ,\JJ) \leq \lim_{t\rightarrow \infty}  \mathfrak{a}+ \mathfrak{c} \, .
-\end{align*}
-It remains to show that $\mathfrak{c}$ goes to zero as $t$ diverges, i.e., 
-\begin{align}
-	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert = 0\, .
-\end{align}
+% These results lead to the following fact for term $\mathfrak{b}$ in the upper bound of \eqref{eq:ub1}:
+%\begin{align*}
+%	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert \leq  \lim_{t\rightarrow \infty} \lVert  e^{t\QQ} \rVert \lVert \JJ  \rVert \lVert \QQ^+ e^{t\QQ} \rVert = 0 \, .
+%\end{align*}
+%Thus, we obtain the intermediary result for $t$ large:
+%\begin{align*}
+%\lim_{t\rightarrow \infty}\mbox{err}(t,\QQ,\JJ) \leq \lim_{t\rightarrow \infty}  \mathfrak{a}+ \mathfrak{c} \, .
+%\end{align*}
+%It remains to show that $\mathfrak{c}$ goes to zero as $t$ diverges, i.e., 
+%\begin{align}
+%	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert = 0\, .
+%\end{align}
+%
+%
+%
+%\section{Random thoughts}
+%
+%\subsection{Another form}
+%
+%\newcommand{\PPsi}{\boldsymbol{\Psi}}
+%
+%\citet{najfeld1995derivatives} use the integral representation to prove the interesting result
+%\begin{align*}
+%	\nabla_{\JJ} e^{t\QQ} = \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ \PPsi(t)  \right) \MM^{-1} \, ,
+%\end{align*}
+%where $\PPsi(t)$ has elements
+%\begin{align*}
+%		[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
+%		t e^{td_i}\, ,	&  d_i = d_j \\
+%		\frac{e^{td_i}-e^{td_j}}{d_i-d_j} \, ,& d_i \neq d_j 
+%	\end{array} \, .
+%\end{align*}
+%For our application, the eigenvalues equal one another with measure zero.  Taking $t$ large, we have
+%\begin{align*}
+%	[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
+%		t e^{td_i}\, ,	&  i = j \\
+%		0 \, ,& i \neq j 
+%	\end{array} \, .
+%\end{align*}
+%It follows that, again for $t$ large,
+%\begin{align*}
+%	\mbox{err}(t,\QQ,\JJ) &\approx \lVert   \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ t e^{t\DD}  \right) \MM^{-1}       -  t e^{t\QQ} \JJ            \rVert \\
+%	&= t\lVert   \MM \rVert \lVert \left(  \left(\MM^{-1} \JJ \MM\right) \circ  e^{t\DD}  \right) \MM^{-1}       -   e^{t\DD}\MM^{-1} \JJ            \rVert \, .
+%\end{align*}
+%
+%\appendix
+%
+%\section{Loose bounds}\label{sec:loose}
+%
+%The following loose bounds do not leverage the structure of the CTMC generator matrix $\QQ$.
+%
+%\subsection{Upper bound 1}
+%
+%
+%
+%For a submultiplicative matrix norm, we have
+%\begin{align*}
+%	\Vert \{\MM_1, \MM_2^k \}  \Vert \leq 2^k \Vert \MM_1\Vert \Vert\MM_2 \Vert ^k \, .
+%\end{align*}
+%We use this to obtain our first upper bound on $err_{ij}(\QQ)$:
+%\begin{align*}
+%	err_{ij}(\QQ) &= \left\Vert e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
+%	&\leq \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \left\Vert \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
+%	&=  \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \frac{t}{k+1} \frac{(2t \Vert \QQ \Vert)^k}{k!}\\
+%	&= \Vert e^{t\QQ} \Vert   \frac{\exp(2t\Vert \QQ \Vert)-2t\Vert\QQ\Vert-1}{2\Vert \QQ \Vert} 
+%\end{align*}
+%
+%
+%\subsection{Upper bound 2}
+%
+%We start with another representation of the derivative \citep{al2009computing}:
+%\begin{align*}
+%	\nabla_{\JJ} e^{t\QQ} = \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} 
+%\end{align*}
+%and the fact that
+%the fact that, for the Frobenius norm $\Vert \cdot \Vert_F$, we have \citep{wu2010short}
+%\begin{align*}
+%	\Vert [\MM_1,\MM_2] \Vert_F \leq \sqrt{2} \Vert \MM_1 \Vert_F \Vert \MM_2\Vert_F \, .
+%\end{align*}
+%Using these results, the truncation error may be rewritten
+%\begin{align*}
+%	err_{ij}(\QQ)  &=\left\Vert - te^{t\QQ} \JJ  +  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}\right\Vert_F \\
+%	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \left( \left( \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} \right) - k \QQ^{k-1} \JJ \right)  \right\Vert_F\\
+%	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ \right\Vert_F \\
+%	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ\right\Vert_F \\
+%	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}\left( \JJ \QQ^{k-p}  -  \QQ^{k-p} \JJ \right)\right\Vert_F \\
+%	&= \left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}[\JJ, \QQ^{k-p}] \right\Vert_F  \\
+%	&\leq  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{p-1} \Vert_F \Vert[\JJ, \QQ^{k-p}]\Vert_F 
+%	\leq \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{k-1} \Vert_F  \\
+%	&= \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} (k-1) \Vert \QQ \Vert_F^{k-1} = \sqrt{2} \frac{e^{t\Vert \QQ\Vert_F}(t\Vert\QQ\Vert_F - 1) + 1}{\Vert \QQ \Vert_F} \, .
+%\end{align*}
 
-\section{Random thoughts}
-
-\subsection{Another form}
-
-\newcommand{\PPsi}{\boldsymbol{\Psi}}
-
-\citet{najfeld1995derivatives} use the integral representation to prove the interesting result
-\begin{align*}
-	\nabla_{\JJ} e^{t\QQ} = \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ \PPsi(t)  \right) \MM^{-1} \, ,
-\end{align*}
-where $\PPsi(t)$ has elements
-\begin{align*}
-		[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
-		t e^{td_i}\, ,	&  d_i = d_j \\
-		\frac{e^{td_i}-e^{td_j}}{d_i-d_j} \, ,& d_i \neq d_j 
-	\end{array} \, .
-\end{align*}
-For our application, the eigenvalues equal one another with measure zero.  Taking $t$ large, we have
-\begin{align*}
-	[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
-		t e^{td_i}\, ,	&  i = j \\
-		0 \, ,& i \neq j 
-	\end{array} \, .
-\end{align*}
-It follows that, again for $t$ large,
-\begin{align*}
-	\mbox{err}(t,\QQ,\JJ) &\approx \lVert   \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ t e^{t\DD}  \right) \MM^{-1}       -  t e^{t\QQ} \JJ            \rVert \\
-	&= t\lVert   \MM \rVert \lVert \left(  \left(\MM^{-1} \JJ \MM\right) \circ  e^{t\DD}  \right) \MM^{-1}       -   e^{t\DD}\MM^{-1} \JJ            \rVert \, .
-\end{align*}
-
-\appendix
-
-\section{Loose bounds}\label{sec:loose}
-
-The following loose bounds do not leverage the structure of the CTMC generator matrix $\QQ$.
-
-\subsection{Upper bound 1}
-
-For a submultiplicative matrix norm, we have
-\begin{align*}
-	\Vert \{\MM_1, \MM_2^k \}  \Vert \leq 2^k \Vert \MM_1\Vert \Vert\MM_2 \Vert ^k \, .
-\end{align*}
-We use this to obtain our first upper bound on $err_{ij}(\QQ)$:
-\begin{align*}
-	err_{ij}(\QQ) &= \left\Vert e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
-	&\leq \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \left\Vert \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
-	&=  \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \frac{t}{k+1} \frac{(2t \Vert \QQ \Vert)^k}{k!}\\
-	&= \Vert e^{t\QQ} \Vert   \frac{\exp(2t\Vert \QQ \Vert)-2t\Vert\QQ\Vert-1}{2\Vert \QQ \Vert} 
-\end{align*}
-
-\subsection{Upper bound 2}
-
-We start with another representation of the derivative \citep{al2009computing}:
-\begin{align*}
-	\nabla_{\JJ} e^{t\QQ} = \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} 
-\end{align*}
-and the fact that
-the fact that, for the Frobenius norm $\Vert \cdot \Vert_F$, we have \citep{wu2010short}
-\begin{align*}
-	\Vert [\MM_1,\MM_2] \Vert_F \leq \sqrt{2} \Vert \MM_1 \Vert_F \Vert \MM_2\Vert_F \, .
-\end{align*}
-Using these results, the truncation error may be rewritten
-\begin{align*}
-	err_{ij}(\QQ)  &=\left\Vert - te^{t\QQ} \JJ  +  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}\right\Vert_F \\
-	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \left( \left( \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} \right) - k \QQ^{k-1} \JJ \right)  \right\Vert_F\\
-	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ \right\Vert_F \\
-	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ\right\Vert_F \\
-	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}\left( \JJ \QQ^{k-p}  -  \QQ^{k-p} \JJ \right)\right\Vert_F \\
-	&= \left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}[\JJ, \QQ^{k-p}] \right\Vert_F  \\
-	&\leq  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{p-1} \Vert_F \Vert[\JJ, \QQ^{k-p}]\Vert_F 
-	\leq \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{k-1} \Vert_F  \\
-	&= \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} (k-1) \Vert \QQ \Vert_F^{k-1} = \sqrt{2} \frac{e^{t\Vert \QQ\Vert_F}(t\Vert\QQ\Vert_F - 1) + 1}{\Vert \QQ \Vert_F} \, .
-\end{align*}
-
-\begin{figure}[!p]
-	\centering
-	\includegraphics[width=\linewidth]{frobeniusScaling.pdf}
-	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using Frobenius norm.}\label{fig:frob}
-\end{figure}
-
-\begin{figure}[!p]
-	\centering
-	\includegraphics[width=\linewidth]{infinityScaling.pdf}
-	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using infinity norm.}\label{fig:inf}
-\end{figure}
+%\begin{figure}[!p]
+%	\centering
+%	\includegraphics[width=\linewidth]{frobeniusScaling.pdf}
+%	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using Frobenius norm.}\label{fig:frob}
+%\end{figure}
+%
+%\begin{figure}[!p]
+%	\centering
+%	\includegraphics[width=\linewidth]{infinityScaling.pdf}
+%	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using infinity norm.}\label{fig:inf}
+%\end{figure}
 

### 2023-03-09 code/analyzeSARS2.R
analysis begun

@@ -1,0 +1,43 @@
+setwd("~/expmDerivative/")
+library(ggplot2)
+library(readr)
+
+df <- read_table2("data/airTravel.txt", col_names = FALSE)
+airTravs <- unlist(df)
+atMat <- matrix(0,44,44)
+k <- 1
+for(i in 1:44) {
+  for(j in 1:44) {
+    if(i!=j){
+      atMat[i,j] <- airTravs[k]
+      k <- k+1
+    }
+  }
+}
+
+df <- read_table2("data/continentalDists.txt", col_names = FALSE)
+airTravs <- unlist(df)
+contMat <- matrix(0,44,44)
+k <- 1
+for(i in 1:44) {
+  for(j in 1:44) {
+    if(i!=j){
+      contMat[i,j] <- airTravs[k]
+      k <- k+1
+    }
+  }
+}
+
+df <- read_table2("data/hubeiAsymm.txt", col_names = FALSE)
+airTravs <- unlist(df)
+hubMat <- matrix(0,44,44)
+k <- 1
+for(i in 1:44) {
+  for(j in 1:44) {
+    if(i!=j){
+      hubMat[i,j] <- airTravs[k]
+      k <- k+1
+    }
+  }
+}
+

### 2023-03-09 code/get_locations.py

@@ -1,0 +1,16 @@
+# This file generate the xml files to run BEAST analyses
+# Xiang Ji
+import argparse, os, re
+
+def main():
+	with open("expmDerivative/code/covid10Mar_285plusNYC.Ed.xml",'r') as infile:
+		with open("expmDerivative/data/locations.txt", 'w') as outfile:
+			for s in infile:
+				if re.match("(.*)<state code=\"(.*)", s):
+					result = re.search('<state code=\"(.*)\"/>', s)
+					outfile.write(str(result.group(1)) + "\n")
+
+if __name__ == '__main__':
+	main()
+
+# python get_locations_times.py

### 2023-03-10 code/analyzeSARS2.R
toward visualizations

@@ -2,6 +2,11 @@ setwd("~/expmDerivative/")
 library(ggplot2)
 library(readr)
 
+#
+####
+####### get predictors
+####
+#
 df <- read_table2("data/airTravel.txt", col_names = FALSE)
 airTravs <- unlist(df)
 atMat <- matrix(0,44,44)
@@ -41,3 +46,78 @@ for(i in 1:44) {
   }
 }
 
+#
+###
+###### get locations (incorrect)
+###
+#
+locations <- read_table2("data/locations.txt", col_names = FALSE)
+
+colnames(atMat) <- unlist(locations)
+rownames(atMat) <- unlist(locations)
+
+colnames(contMat) <- unlist(locations)
+rownames(contMat) <- unlist(locations)
+
+colnames(hubMat) <- unlist(locations)
+rownames(hubMat) <- unlist(locations)
+
+#
+###
+###### visualize posts
+###
+#
+df <- read_table2("output/covid10Mar_285plusNYC.Ed.log", skip = 3)
+
+df2 <- df[,9:11]
+df2 <- reshape2::melt(df2)
+df2$Predictor <- "cat"
+df2$Predictor[which(df2$variable=="glmCoefficients1")] <- "IntraContinental"
+df2$Predictor[which(df2$variable=="glmCoefficients2")] <- "AirTraffic"
+df2$Predictor[which(df2$variable=="glmCoefficients3")] <- "HubeiAsymmetry"
+
+gg <- ggplot(df2,aes(x=value,color=Predictor)) +
+  geom_density(adjust=2) +
+  xlab("Coefficient value") +
+  ylab("Density") +
+  ggtitle("Fixed-effects for infinitesimal rates between locations") +
+  theme_bw()
+gg
+
+#
+####
+####### get random coeffs matrices
+####
+#
+df2 <- df[,12:1903]
+nSamps <- dim(df2)[1]
+randCoeffs <- list()
+
+
+for(n in 1:nSamps) {
+  k <- 1
+  rcMat <- matrix(0,44,44)
+  for(i in 1:44) {
+    for(j in 1:44) {
+      if(i!=j){
+        rcMat[i,j] <- unlist(df2[n,k])
+        k <- k+1
+      }
+    }
+  }
+  randCoeffs[[n]] <- rcMat
+}
+
+totalCoeffs <- list()
+for (n in 1:nSamps) {
[…]

### 2023-03-10 code/analyzeSARS2.R
fixed effects

@@ -1,6 +1,9 @@
 setwd("~/expmDerivative/")
 library(ggplot2)
 library(readr)
+library(ggthemes)
+library(scales)
+
 
 #
 ####
@@ -11,12 +14,16 @@ df <- read_table2("data/airTravel.txt", col_names = FALSE)
 airTravs <- unlist(df)
 atMat <- matrix(0,44,44)
 k <- 1
-for(i in 1:44) {
-  for(j in 1:44) {
-    if(i!=j){
-      atMat[i,j] <- airTravs[k]
-      k <- k+1
-    }
+for(i in 1:43) {
+  for(j in (i+1):44) {
+    atMat[i,j] <- airTravs[k]
+    k <- k+1
+  }
+}
+for(i in 1:43) {
+  for(j in (i+1):44) {
+    atMat[j,i] <- airTravs[k]
+    k <- k+1
   }
 }
 
@@ -24,12 +31,16 @@ df <- read_table2("data/continentalDists.txt", col_names = FALSE)
 airTravs <- unlist(df)
 contMat <- matrix(0,44,44)
 k <- 1
-for(i in 1:44) {
-  for(j in 1:44) {
-    if(i!=j){
+for(i in 1:43) {
+  for(j in (i+1):44) {
       contMat[i,j] <- airTravs[k]
       k <- k+1
-    }
+  }
+}
+for(i in 1:43) {
+  for(j in (i+1):44) {
+    contMat[j,i] <- airTravs[k]
+    k <- k+1
   }
 }
 
@@ -37,21 +48,41 @@ df <- read_table2("data/hubeiAsymm.txt", col_names = FALSE)
 airTravs <- unlist(df)
 hubMat <- matrix(0,44,44)
 k <- 1
-for(i in 1:44) {
-  for(j in 1:44) {
-    if(i!=j){
-      hubMat[i,j] <- airTravs[k]
-      k <- k+1
-    }
+for(i in 1:43) {
+  for(j in (i+1):44) {
+    hubMat[i,j] <- airTravs[k]
+    k <- k+1
+  }
+}
+for(i in 1:43) {
+  for(j in (i+1):44) {
+    hubMat[j,i] <- airTravs[k]
+    k <- k+1
   }
 }
 
 #
 ###
-###### get locations (incorrect)
+###### get locations 
 ###
 #
 locations <- read_table2("data/locations.txt", col_names = FALSE)
[…]

### 2023-03-10 code/analyzeSARS2.R
standardized diagonals

@@ -98,11 +98,14 @@ rownames(hubMat) <- unlist(locations)
 ###### visualize fixed-effects
 ###
 #
+atMat <- atMat- (max(atMat)-abs(min(atMat)))/2
+atMat <- atMat/max(atMat)
+diag(atMat) <- 0
 df <- reshape2::melt(atMat)
 #df$value <- exp(df$value)
 colnames(df) <- c("Var1","Var2","Values")
-df$Values <- df$Values - (max(df$Values)-abs(min(df$Values)))/2
-df$Values <- df$Values / max(df$Values)
+# df$Values <- df$Values - (max(df$Values)-abs(min(df$Values)))/2
+# df$Values <- df$Values / max(df$Values)
 
 gg2 <- ggplot(data = df, aes(x=Var1, y=Var2, fill=Values)) + 
   geom_tile() +
@@ -117,10 +120,12 @@ gg2 <- ggplot(data = df, aes(x=Var1, y=Var2, fill=Values)) +
         legend.text = element_text())
 gg2
 
+
+contMat <- contMat- (max(contMat)-abs(min(contMat)))/2
+contMat <- contMat/max(contMat)
+diag(contMat) <- 0
 df <- reshape2::melt(contMat)
 colnames(df) <- c("Var1","Var2","Values")
-df$Values <- df$Values - (max(df$Values)-abs(min(df$Values)))/2
-df$Values <- df$Values / max(df$Values)
 gg3 <- ggplot(data = df, aes(x=Var1, y=Var2, fill=`Values`)) + 
   geom_tile() +
   guides(x =  guide_axis(angle = 90)) +
@@ -194,16 +199,19 @@ df2 <- df[,12:1903]
 nSamps <- dim(df2)[1]
 randCoeffs <- list()
 
-
 for(n in 1:nSamps) {
   k <- 1
   rcMat <- matrix(0,44,44)
-  for(i in 1:44) {
-    for(j in 1:44) {
-      if(i!=j){
+  for(i in 1:43) {
+    for(j in (i+1):44) {
         rcMat[i,j] <- unlist(df2[n,k])
         k <- k+1
-      }
+    }
+  }
+  for(i in 1:43) {
+    for(j in (i+1):44) {
+      rcMat[j,i] <- unlist(df2[n,k])
+      k <- k+1
     }
   }
   randCoeffs[[n]] <- rcMat
[…]

### 2023-03-10 code/analyzeSARS2.R
figures done for now

@@ -145,7 +145,7 @@ gg4 <- ggplot(data = df, aes(x=Var1, y=Var2, fill=`Values`)) +
   geom_tile() +
   guides(x =  guide_axis(angle = 90)) +
   xlab(NULL) + ylab(NULL) +
-  ggtitle("Hubei asymmetric proximity") +
+  ggtitle("Hubei asymmetry") +
   scale_fill_gradientn(breaks=c(-1,0,1),
     colours = tableau_div_gradient_pal("Temperature Diverging")(seq(0, 1, length = 25))) +
   theme(plot.title = element_text(vjust = -1),
@@ -176,18 +176,24 @@ system2(command = "pdfcrop",
 df <- read_table2("output/covid10Mar_285plusNYC.Ed.log", skip = 3)
 
 df2 <- df[,9:11]
-df2 <- reshape2::melt(df2)
-df2$Predictor <- "cat"
-df2$Predictor[which(df2$variable=="glmCoefficients1")] <- "IntraContinental"
-df2$Predictor[which(df2$variable=="glmCoefficients2")] <- "AirTraffic"
-df2$Predictor[which(df2$variable=="glmCoefficients3")] <- "HubeiAsymmetry"
+nSamps <- dim(df2)[1]
+df2 <- df2[ceiling(nSamps/10):nSamps,]
 
-gg <- ggplot(df2,aes(x=value,color=Predictor)) +
-  geom_density(adjust=2) +
+df2 <- reshape2::melt(df2)
+df2$`Fixed effects` <- "cat"
+df2$`Fixed effects`[which(df2$variable=="glmCoefficients1")] <- "Intracontinental\nproximity"
+df2$`Fixed effects`[which(df2$variable=="glmCoefficients2")] <- "Air traffic\nproximity"
+df2$`Fixed effects`[which(df2$variable=="glmCoefficients3")] <- "Hubei asymmetry"
+
+gg <- ggplot(df2,aes(x=value,color=`Fixed effects`)) +
+  geom_density(adjust=2,size=1.5) +
   xlab("Coefficient value") +
   ylab("Density") +
-  ggtitle("Fixed-effects for infinitesimal rates between locations") +
-  theme_bw()
+  scale_color_manual(values = stata_pal("s2color")(15)[c(4,7,3)]) +
+  ggtitle("Posteriors for infinitesimal rates regression") +
+  theme_bw() +
+  theme(legend.margin=margin(0,0,0,0),
[…]

### 2023-03-13 manuscript/main.tex
fixed effects
figures done for now
started writing

@@
 
-\title{Approximating the matrix exponential derivative with respect to the CTMC generator matrix, with application to global SARS-CoV-2}
+\title{Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2}
 
@@
 \section{Global spread of SARS-CoV-2}
+
+\begin{figure}[!p]
+	\centering
+	\includegraphics[width=\linewidth]{fixedEffects.pdf}
+	\caption{Fixed effects.}\label{fig:FE}
+\end{figure}
+
+\begin{figure}[!p]
+	\centering
+	\includegraphics[width=\linewidth]{postViz.pdf}
+	\caption{Posteriors.}\label{fig:FE}
+\end{figure}
+
+Hamiltonian Monte Carlo \citep{neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate proposals far away from the current Markov chain state.
+On the one hand, this approach to proposal generation helps reduce autocorrelation between chain states and is particularly helpful within higher-dimensional state spaces.
+On the other hand, the numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.
 

### 2023-03-13 manuscript/main.tex
more writing

@@
 %\renewcommand{\u}{\mathbf{u}}
+\newcommand{\Y}{\mathbf{Y}}
 \newcommand{\QQ}{\mathbf{Q}}
@@
 \newcommand{\GG}{\mathbf{G}}
+\newcommand{\mom}{\boldsymbol{\xi}}
 
@@
 
-\begin{figure}[!p]
+\begin{figure}[!t]
 	\centering
@@
 
-Hamiltonian Monte Carlo \citep{neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate proposals far away from the current Markov chain state.
+Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.
 On the one hand, this approach to proposal generation helps reduce autocorrelation between chain states and is particularly helpful within higher-dimensional state spaces.
-On the other hand, the numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.
+On the other hand, numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.
+Using HMC for Bayesian inference, the potential energy takes the form $u(\ttheta)=-\log \pi(\ttheta)$, where $\pi(\cdot)$ is the posterior density function $\pi(\ttheta):=p(\ttheta|\Y)$ of parameter $\ttheta\in\mathbb{R}^D$ given observed data $\Y$.
+In this context, a single iteration of the the St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator \citep{leimkuhler2004simulating}, beginning at time $\tau$ and with $\epsilon>0$ small, takes the form
+\begin{align}\label{eq:leap}
+	\mom\left(\tau + \frac{\epsilon}{2}\right) &= \mom(\tau) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau))  \\  \nonumber
+	\ttheta(\tau+\epsilon) &= \ttheta(\tau) + \epsilon \, \MM^{-1}\mom(\tau+\frac{\epsilon}{2}) \\ \nonumber
+	\mom(\tau + \epsilon) &= \mom\left(\tau+\frac{\epsilon}{2}\right) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau+\epsilon))  \, ,
+\end{align}
+where $\mom \sim \mbox{N}_D(\Zero,\MM)$ is the Hamiltonian system's Gaussian-distributed momentum variable.  
+Different strategies aim to speed up the leapfrog integrator's many log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$ in different ways, as these numerical routines often represent the algorithm's computational bottleneck.
+In model-specific contexts, \citet{holbrook2021massive,holbrook2022bayesian,holbrook2022viral} develop parallelization strategies, and \citet{ji2020gradients} develop dynamic programming techniques, to accelerate $\nabla \log \pi (\ttheta)$.
+
+A small body of literature considers another approach by recognizing
 

### 2023-03-13 manuscript/main.tex
more writing

@@
 
-Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.
+Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid,neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.
 On the one hand, this approach to proposal generation helps reduce autocorrelation between chain states and is particularly helpful within higher-dimensional state spaces.
@@
 
-A small body of literature considers another approach by recognizing
+A small body of work considers another approach by replacing $\nabla \log \pi (\ttheta)$ with a suitable approximation $\widehat{\nabla} \log \pi (\ttheta)$ and recognizing that the analogue of \eqref{eq:leap} continues to satisfy path reversibility and volume preservation, two essential ingredients for well-specified HMC.
+It follows immediately that the resulting`surrogate trajectory HMC' continues to sample the correct target distribution $\pi(\cdot)$.
+The majority of surrogate HMC methods first obtain a small sample of exact gradient evaluations and then use some model to interpolate: \citet{zhang2017precomputing} assume the approximate gradient follows a piecewise constant form across a grid; \citet{rasmussen2003gaussian,lan2016emulation} find approximations using Gaussian processes; and \citet{zhang2017hamiltonian,li2019neural} use neural networks.
+Recently \textcolor{red}{MAGEE}
 

### 2023-03-14 manuscript/main.tex
another paragraph

--- 2023-03-13 manuscript/main.tex
@@
 
+ We are interested in efficiently approximating the directional derivative of the matrix exponential
+\begin{align*}
+	e^{t\QQ} = \sum_{k=0}^\infty \frac{t^k\QQ^k}{k!}
+\end{align*}
+for arbitrary $t>0$ using the formula \citep{najfeld1995derivatives}
+\begin{align}\label{eq:deriv}
+	\nabla_{\JJ} e^{t\QQ}  =  e^{t\QQ}  \sum_{k=0}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \, .
+\end{align}
+Here, $\JJ$ gives the direction and $\{\JJ,\QQ^k\}$ is the matrix commutator series satisfying the recursion
+\begin{align*}
+	\{\JJ,\QQ^0\} &= \JJ \, , \\ 
+	\{\JJ,\QQ^k\} &= [\{\JJ,\QQ^{k-1}\} , \QQ]  = \{\JJ,\QQ^{k-1}\} \QQ - \QQ \{\JJ,\QQ^{k-1}\} \, .
+\end{align*}
+Thus, the first-order approximation to derivative \eqref{eq:deriv} is 
+\begin{align}\label{eq:firstOrder}
+	\widetilde{\nabla}_{\JJ} e^{t\QQ}= t e^{t\QQ}\JJ \, .
+\end{align}
+
 \section{Global spread of SARS-CoV-2}
@@
 
-A small body of literature considers another approach by recognizing
+A small body of work considers another approach by replacing $\nabla \log \pi (\ttheta)$ with a suitable approximation $\widehat{\nabla} \log \pi (\ttheta)$ and recognizing that the modified \eqref{eq:leap} continues to satisfy path reversibility and volume preservation, two essential ingredients for well-specified HMC.
+It follows immediately that the resulting `surrogate trajectory HMC' continues to sample the correct target distribution $\pi(\cdot)$.
+Nonetheless, the acceptance rates and overall efficiency of such samplers may suffer when approximations are poor.
+The majority of surrogate HMC methods first obtain a small sample of exact gradient evaluations and then use some model to interpolate: \citet{zhang2017precomputing} assume the approximate gradient follows a piecewise constant form across a grid; \citet{rasmussen2003gaussian,lan2016emulation} construct approximations using Gaussian processes; and \citet{zhang2017hamiltonian,li2019neural} do the same using neural networks.
+But an even simpler approach to surrogate HMC may be appropriate when gradients have series representations.
+\textcolor{red}{MAGEE} consider phylogenetic models that involve CTMC priors and show that the first-order approximation \eqref{eq:firstOrder} of the matrix exponential derivative \eqref{eq:deriv} performs remarkably well within surrogate HMC, achieving an over 30-fold efficiency gain compared to random-walk Metropolis for a 14-state model with over 180 model parameters.
+Here, we demonstrate similar strong performance of the first-order approximation within surrogate HMC for a 44-state model with almost 1900 model parameters.
 

### 2023-03-14 manuscript/main.tex
money
more writing

@@
 \newcommand{\mom}{\boldsymbol{\xi}}
+\newcommand{\ppi}{\boldsymbol{\pi}}
+\renewcommand{\P}{\mathbf{P}}
 
@@
 
+\subsection{Surrogate trajectory HMC}
+
 Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid,neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.
@@
 \textcolor{red}{MAGEE} consider phylogenetic models that involve CTMC priors and show that the first-order approximation \eqref{eq:firstOrder} of the matrix exponential derivative \eqref{eq:deriv} performs remarkably well within surrogate HMC, achieving an over 30-fold efficiency gain compared to random-walk Metropolis for a 14-state model with over 180 model parameters.
-Here, we demonstrate similar strong performance of the first-order approximation within surrogate HMC for a 44-state model with almost 1900 model parameters.
+Here, we demonstrate similar strong performance of the first-order approximation within surrogate HMC for a 44-state model with almost 1,900 model parameters.
+
+\subsection{Phylogeographic CTMC model}
+
+Start with a (possibly unknown) rooted and bifurcating phylogenetic tree $\phylogeny$ consisting of $N$ leaf nodes that correspond to observed biological specimens and $N-1$ internal nodes that correspond to unobserved ancestors.
+The tree also contains $2N-2$ branches of length $t_v$ connecting each child node $v$ to its parent $u$.
+ Given $\phylogeny$, we model the evolution of $S$ characters along each branch of the tree according to a CTMC model with $S\times S$ generator matrix $\QQ$ and stationary distribution $\ppi=(\pi_1,\dots,\pi_S)=\lim_{t\rightarrow \infty}e^{t\QQ}\widetilde{\ppi}$, for $\widetilde{\ppi}$ for any arbitrary probability vector.
+Examples of characters are the $S=4$ nucleotides within a set of aligned genome sequences  \citep{jukes1969evolution} and the set of $S=15$ geographic regions visited by an influenza subtype \citep{lemey2014unifying}.
+Depending on the given problem, we may scale $t_v$ to be raw time (e.g., years in Section \ref{sec:sars}) or the expected number of substitutions with respect to $\ppi$.
+In the former case, one may augment the model with a rate scalar $\gamma$ that modulates the expected number of substitutions across all branches, and the finite-time transition probability matrix along branch $v$ becomes $\P_v:=e^{\gamma t_v \QQ}$.
+
+\subsection{Bayesian analysis of SARS-CoV-2 contagion}\label{sec:sars}
+
+\section{Discussion}
+
+\section*{Acknowledgments}
+
+AJH is supported by a gift from the Karen Toffler Charitable Trust and by grants NIH K25 AI153816, NSF DMS 2152774 and NSF DMS 2236854.
 

### 2023-03-15 manuscript/main.tex
phylo ctmc done

@@
 %\renewcommand{\u}{\mathbf{u}}
-\newcommand{\Y}{\mathbf{Y}}
+\newcommand{\y}{\mathbf{y}}
 \newcommand{\QQ}{\mathbf{Q}}
@@
 \renewcommand{\P}{\mathbf{P}}
+\newcommand{\p}{\mathbf{p}}
+\newcommand{\q}{\mathbf{q}}
+\newcommand{\B}{\mathbf{B}}
 
@@
 
-\subsection{Surrogate trajectory HMC}
+\subsection{Surrogate trajectory HMC}\label{sec:surrHmc}
 
@@
 On the other hand, numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.
-Using HMC for Bayesian inference, the potential energy takes the form $u(\ttheta)=-\log \pi(\ttheta)$, where $\pi(\cdot)$ is the posterior density function $\pi(\ttheta):=p(\ttheta|\Y)$ of parameter $\ttheta\in\mathbb{R}^D$ given observed data $\Y$.
+Using HMC for Bayesian inference, the potential energy takes the form $u(\ttheta)=-\log \pi(\ttheta)$, where $\pi(\cdot)$ is the posterior density function $\pi(\ttheta):=p(\ttheta|\y)$ of parameter $\ttheta\in\mathbb{R}^D$ given observed data $\y$.
 In this context, a single iteration of the the St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator \citep{leimkuhler2004simulating}, beginning at time $\tau$ and with $\epsilon>0$ small, takes the form
@@
 
-\subsection{Phylogeographic CTMC model}
+\subsection{Phylogenetic CTMC model}
 
@@
 Examples of characters are the $S=4$ nucleotides within a set of aligned genome sequences  \citep{jukes1969evolution} and the set of $S=15$ geographic regions visited by an influenza subtype \citep{lemey2014unifying}.
-Depending on the given problem, we may scale $t_v$ to be raw time (e.g., years in Section \ref{sec:sars}) or the expected number of substitutions with respect to $\ppi$.
+We may scale $t_v$ to be raw time (e.g., years in Section \ref{sec:sars}) or the expected number of substitutions with respect to $\ppi$ depending on the given problem.
 In the former case, one may augment the model with a rate scalar $\gamma$ that modulates the expected number of substitutions across all branches, and the finite-time transition probability matrix along branch $v$ becomes $\P_v:=e^{\gamma t_v \QQ}$.
+In the following, we further posit $\QQ=\QQ(\ttheta)$ for $\ttheta$ a vector of parameters.
+
+ Let data $\y$ be the $N$-vector with elements $y_n\in \{1,\dots,S\}$, the observed states of each observed biological specimen $n \in \{1,\dots,N\}$.
+At any node $v$, the likelihood takes the form \citep{ji2020gradients}
+ \begin{align}
+p(\y|\ttheta) = \p_v^T \q_v \, ,
+ \end{align}
+where $\p_v$ and $\q_v$ are the post-order and pre-order partial likelihood vectors, respectively.
+The former describes the probability of the observed states for all observed specimens (i.e., leaf nodes) that descend from node $v$, conditioned on the state at node $v$.
+The latter describes analogous probabilities for all observed specimens not descending from node $v$.
+If we suppose that node $u$ gives rise to two child nodes, $v$ and $w$, then
+\begin{align}\label{eq:recursions}
+	\p_u = \P_v \p_v \circ \P_w \p_w\, ,  \quad \q_v= \P_v^T \left( \q_u \circ  \P_w \p_w\right) \, .
+\end{align}
+Using the chain rule and the fact that $\p_v$ does not depend on $\P_v$, one may write the likelihood's derivative with respect to a single parameter $\theta_d$ thus:
+\begin{align}\label{eq:phyloGrad}
+	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta)  &=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial p(\y|\ttheta)}{\partial \P_v} \frac{\partial \P_v}{\partial \theta_d} \right) = \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial p(\y|\ttheta)}{\partial \P_v} \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d} \right)  \\
+	&=\sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial p(\y|\ttheta)}{\partial \P_v} \sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \right)  \, .
+\end{align}
+Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.
+These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
+\begin{align}
+\log q_{ss'} = b_{ss'} + \epsilon_{ss'} \,, \quad s\neq s'\, .
+\end{align} 
+Here, the fixed-effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random-effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.
+The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.
+In this context, \textcolor{red}{MAGEE} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.
+In the following section, we use this method to analyze the global spread of SARS-CoV-2 and show that the first-order approximation maintains its performance for an even higher-dimensional problem than previously considered.
 

### 2023-03-16 manuscript/main.tex
polished gradient equation
beginning analysis section

@@
 \newcommand{\B}{\mathbf{B}}
+\newcommand{\X}{\mathbf{X}}
 
@@
 \section{Global spread of SARS-CoV-2}
-
-\begin{figure}[!t]
-	\centering
-	\includegraphics[width=\linewidth]{fixedEffects.pdf}
-	\caption{Fixed effects.}\label{fig:FE}
-\end{figure}
-
-\begin{figure}[!t]
-	\centering
-	\includegraphics[width=\linewidth]{postViz.pdf}
-	\caption{Posteriors.}\label{fig:FE}
-\end{figure}
 
@@
 \begin{align}\label{eq:phyloGrad}
-	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta)  &=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial p(\y|\ttheta)}{\partial \P_v} \frac{\partial \P_v}{\partial \theta_d} \right) = \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial p(\y|\ttheta)}{\partial \P_v} \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d} \right)  \\
-	&=\sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial p(\y|\ttheta)}{\partial \P_v} \sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \right)  \, .
+	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta) 
+	&=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v} \frac{\partial \P^T_v}{\partial \theta_d} \right)
+	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T \frac{\partial e^{\gamma t_v \QQ^T}}{\partial \theta_d}\right)  \\ \nonumber
+	&=\sum_{v=1}^{2N-2} \p_v^T \left(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ^T}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \right) (\q_u\circ \P_w\p_w)  \, .
 \end{align}
@@
 These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
-\begin{align}
+\begin{align}\label{eq:mixed}
 \log q_{ss'} = b_{ss'} + \epsilon_{ss'} \,, \quad s\neq s'\, .
@@
 \subsection{Bayesian analysis of SARS-CoV-2 contagion}\label{sec:sars}
+
+\begin{figure}[!t]
+	\centering
+	\includegraphics[width=\linewidth]{fixedEffects.pdf}
+	\caption{Fixed effects.}\label{fig:FE}
+\end{figure}
+
+\begin{figure}[!t]
+	\centering
+	\includegraphics[width=\linewidth]{postViz.pdf}
+	\caption{Posteriors.}\label{fig:post}
+\end{figure}
+
+We use the phylogenetic CTMC framework to model the early spread of SARS-CoV-2---the virus responsible for the ongoing COVID-19 pandemic---based on 285 observed viral samples collected from $S=44$ regions worldwide.
+These regions comprise 15 provinces within China and 29 countries without.
+Understanding the manner in which viruses travel between human populations is an object of ongoing study, and phylogeographic analyses point to the central role of travel networks including those measured by airline passenger counts \citep{holbrook2021massive} or google mobility data \citep{worobey2020emergence}.
+Here, we include three such predictors of travel in our CTMC model by expanding the fixed-effects in regression model \eqref{eq:mixed} to take the form
+\begin{align*}
+	\B = \theta_1 \X_1 + \theta_2  \X_2 + \theta_3 \X_3 \, ,
+\end{align*} 
+for each $\X$ a fixed $44\times 44$ matrix predictor and each $\theta$ a real-valued regression coefficient.
+Figure \ref{fig:FE} presents the three predictors of interest: $\X_1$ contains air travel proximities between locations as measured by annualized air passenger counts between airports contained within a region \citep{holbrook2021massive};  $\X_2$ contains intracontinental proximities arising from physical distances when two regions inhabit the same continent and fixed at the minimum otherwise; finally, $\X_3$ describes the Hubei asymmetry, i.e., the fact that people could enter, but not leave, the Chinese province of Hubei during the early pandemic.
 

### 2023-03-16 manuscript/main.tex
stopping 4 now

@@
 	&=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v} \frac{\partial \P^T_v}{\partial \theta_d} \right)
-	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T \frac{\partial e^{\gamma t_v \QQ^T}}{\partial \theta_d}\right)  \\ \nonumber
-	&=\sum_{v=1}^{2N-2} \p_v^T \left(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ^T}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \right) (\q_u\circ \P_w\p_w)  \, .
+	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T\left( \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d}\right)^T\right)  \\ \nonumber
+	&=\sum_{v=1}^{2N-2} \p_v^T \left(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \right)^T (\q_u\circ \P_w\p_w)  \, .
 \end{align}
@@
 Figure \ref{fig:FE} presents the three predictors of interest: $\X_1$ contains air travel proximities between locations as measured by annualized air passenger counts between airports contained within a region \citep{holbrook2021massive};  $\X_2$ contains intracontinental proximities arising from physical distances when two regions inhabit the same continent and fixed at the minimum otherwise; finally, $\X_3$ describes the Hubei asymmetry, i.e., the fact that people could enter, but not leave, the Chinese province of Hubei during the early pandemic.
+
+In the context of a Bayesian analysis, we specify independent normal priors on $\theta_1$, $\theta_2$ and $\theta_3$ with means of 0 and variances of 2.
+We also assume that the 1,892 random-effects $\epsilon_{ss'}$ follow sparsity inducing Bayesian bridge priors with global scale parameter $\tau$ and exponent $\alpha=0.25$.
+Here, we follow \citet{nishimura2022shrinkage} and specify a Gamma prior on $\tau^{-\alpha}$ with a shape parameter of 1 and a rate parameter of 2.
+Finally, we place a flat prior on the rate scalar $\gamma$.
+Inferring the posterior distribution of all 1,897 model parameters requires an advanced MCMC strategy.
+Namely, we adopt an HMC-within-Gibbs approach, updating the scalars $\gamma$ and $\tau$ independently but updating all 1,895 regression parameters (both fixed- and random-effects) using surrogate HMC accomplished with the first-order approximation
+\begin{align*}
+  \gamma t_v  e^{\gamma t_v \QQ} \JJ_{ss'}    \approx \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} 
+\end{align*}
+within \eqref{eq:phyloGrad}.
+
+We generate 10 million MCMC samples in this manner, saving 1 in 10,000 Markov chain states in order to conserve memory.  
+\begin{itemize}
+	\item ESS
+	\item Figure 2
+	\item Tree
+\end{itemize}
+
+In addition to these scientific questions of interest, we are interested in the performance of the first-order approximation as a surrogate gradient for HMC in such a high-dimensional setting.
+To do isolate this performance, we temporarily fix the Bayesian bridge global-scale $\tau$ at 2.5$\times10^{-5}$.
+We generate
 

### 2023-03-16 manuscript/main.tex

@@
 In addition to these scientific questions of interest, we are interested in the performance of the first-order approximation as a surrogate gradient for HMC in such a high-dimensional setting.
-To do isolate this performance, we temporarily fix the Bayesian bridge global-scale $\tau$ at 2.5$\times10^{-5}$.
-We generate
+To isolate this performance, we fix the Bayesian bridge global-scale $\tau$ at 2.5$\times10^{-5}$.
+We generate a Markov chain with 80,000 states, saving every tenth state and removing the first 1000 states as burn-in.
+Despite the relatively small number of iterations, we observe large ESS that suggest satisfactory accuracy of the first-order approximation within the context of high-dimensional surrogate HMC.
+The ESS for the three fixed-effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343 and 498, respectively.
+The median and minimum ESS for the 1,892 random-effects are 1,514 and 1,161, respectively.
 

### 2023-03-16 manuscript/main.tex
post viz figs visualized

@@
 
-We generate 10 million MCMC samples in this manner, saving 1 in 10,000 Markov chain states in order to conserve memory.  
+We generate 8 million MCMC samples in this manner, saving 1 in 10,000 Markov chain states, in order to guarantee a minimum effective sample size (ESS) greater than 100.
+By far, the worst-mixing parameter  is the global scale $\tau$, which obtains an ESS of 185.
+The three fixed-effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ obtain ESS's of 721, 721 and 644, respectively.
+The median and minimum ESS for the 1,892 random-effects $\epsilon_{ss'}$ are 721 and 190, respectively.
+We note that, after thinning and removing burn-in, the sample only consists of 721 states, so ESS of 721 implies negligible autocorrelation between samples.
+Figure \ref{fig:post} presents
+
 \begin{itemize}
-	\item ESS
 	\item Figure 2
@@
 The ESS for the three fixed-effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343 and 498, respectively.
-The median and minimum ESS for the 1,892 random-effects are 1,514 and 1,161, respectively.
+The median and minimum ESS for the 1,892 random-effects $\epsilon_{ss'}$ are 1,514 and 1,161, respectively.
 

### 2023-03-16 code/postDiagnostics.R

@@ -9,3 +9,20 @@ effectiveSize(df[,9:11]) #        1053.3909        1342.5106         498.3175
 
 # random effects
 summary(effectiveSize(df[,12:1903]))
+
+
+# full run
+df <- read_table2("output/covid10Mar_285plusNYC.Ed.log", skip = 3)
+df <- df[df$state>=800000,]
+
+# global scale
+effectiveSize(df[,1904])
+
+# fixed effects
+effectiveSize(df[,9:11])
+
+# random effects
+summary(effectiveSize(df[,12:1903]))
+
+
+

### 2023-03-16 manuscript/main.tex
need tree then done

@@
 We note that, after thinning and removing burn-in, the sample only consists of 721 states, so ESS of 721 implies negligible autocorrelation between samples.
-Figure \ref{fig:post} presents
+The left plot of Figure \ref{fig:post} presents posterior densities for the fixed-effect coefficients from regression \eqref{eq:mixed}.
+The air traffic proximity, intracontinental proximity and Hubei asymmetry coefficients have posterior means and 95\% credible intervals of 0.76 (0.50, 1.03), 0.04 (-0.04, 0.12) and 0.27 (-0.39, 0.78), respectively.
+While these results suggest a positive association between each predictor matrix and the infinitesimal generator matrix $\QQ$, the only predictor with a statistically significant association is air traffic proximity.
+This result agrees with a previous phylogeographic analysis of the global spread of influenza \citep{holbrook2021massive}.
+The right plot of Figure \ref{fig:post} presents the posterior mean for generator $\QQ$.
+As one may expect, the matrix looks similar to that of the air traffic predictor matrix in Figure \ref{fig:FE}, but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.
 
 \begin{itemize}
-	\item Figure 2
-	\item Tree
+	\item tree
 \end{itemize}

### 2023-03-16 code/postDiagnostics.R

@@ -24,5 +24,6 @@ effectiveSize(df[,9:11])
 # random effects
 summary(effectiveSize(df[,12:1903]))
 
-
+colMeans(df[,9:11])
+coda::HPDinterval(coda::as.mcmc(df[,9:11]))
 

### 2023-03-18 manuscript/main.tex
improved math typesetting
a tree
application done

@@
 	&=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v} \frac{\partial \P^T_v}{\partial \theta_d} \right)
-	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T\left( \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d}\right)^T\right)  \\ \nonumber
-	&=\sum_{v=1}^{2N-2} \p_v^T \left(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \right)^T (\q_u\circ \P_w\p_w)  \, .
+	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T\bigg( \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d}\bigg)^{\!T}\right)  \\ \nonumber
+	&=\sum_{v=1}^{2N-2} \p_v^T \Biggl(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \Biggr)^{\!T} (\q_u\circ \P_w\p_w)  \, .
 \end{align}
 Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.
-These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
+These computations become particularly onerous when one opts for a gold-standard mixed effect model \textcolor{red}{(MAGEE)} and specifies
 \begin{align}\label{eq:mixed}
@@
 \end{align} 
-Here, the fixed-effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random-effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.
+Here, the fixed-effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.
 The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.
@@
 	\includegraphics[width=\linewidth]{fixedEffects.pdf}
-	\caption{Fixed effects.}\label{fig:FE}
+	\vspace{-2em}
+	\caption{Matrix predictors described in \eqref{eq:linear} combine in a linear manner to form the fixed effect matrix $\B$ featured within the mixed effects regression model \eqref{eq:mixed}.
+Air traffic proximities \emph{(\textbf{a})} are proportional the number of air passengers exchanged between airports within respective regions \citep{holbrook2021massive}.
+Intracontinental proximities \emph{(\textbf{b})} take values between -1 for regions on different continents to 1 for adjacent regions.
+The Hubei asymmetry \emph{(\textbf{c})} roughly characterizes the Hubei quarantine of early 2020.}\label{fig:FE}
 \end{figure}
+
+We use the phylogenetic CTMC framework to model the early spread of SARS-CoV-2---the virus responsible for the ongoing COVID-19 pandemic---based on 285 observed viral samples collected from $S=31$ regions worldwide between December 24, 2019 and March 19, 2020.
+These regions comprise 13 provinces within China and 18 countries without.
+Understanding the manner in which viruses travel between human populations is an object of ongoing study, and phylogeographic analyses point to the central role of travel networks including those measured by airline passenger counts \citep{holbrook2021massive} or google mobility data \citep{worobey2020emergence}.
+Here, we include three such predictors of travel in our CTMC model by expanding the fixed effects in regression model \eqref{eq:mixed} to take the form
+\begin{align}\label{eq:linear}
+	\B = \theta_1 \X_1 + \theta_2  \X_2 + \theta_3 \X_3 \, ,
+\end{align} 
+for each $\X$ a fixed $44\times 44$ matrix predictor and each $\theta$ a real-valued regression coefficient, where we have expanded the number of regions between which viruses may travel by including two additional Chinese provinces and 11 additional countries.
+Figure \ref{fig:FE} presents the three predictors of interest: $\X_1$ contains air travel proximities between locations as measured by annualized air passenger counts between airports contained within a region \citep{holbrook2021massive};  $\X_2$ contains intracontinental proximities arising from physical distances when two regions inhabit the same continent and fixed at the minimum otherwise; finally, $\X_3$ describes the Hubei asymmetry, i.e., the non-existence of human travel out of the Hubei province in early 2020.
 
@@
 	\includegraphics[width=\linewidth]{postViz.pdf}
-	\caption{Posteriors.}\label{fig:post}
+	\vspace{-2em}
+	\caption{Posterior inference.
+Posterior densities \emph{(\textbf{a})} for the three fixed-effect regression coefficients corresponding to the predictor matrices of Figure \ref{fig:FE} reflect largely positive associations between predictors and infinitesimal rate matrix, although air traffic proximity has the only statistically significant coefficient with posterior mean of 0.76 and 95\% credible interval of (0.50, 1.03).
+The posterior mean \emph{(\textbf{b})} infinitesimal rate matrix closely resembles the air traffic predictor, while reflecting the Hubei asymmetry to a lesser extent.}\label{fig:post}
 \end{figure}
 
-We use the phylogenetic CTMC framework to model the early spread of SARS-CoV-2---the virus responsible for the ongoing COVID-19 pandemic---based on 285 observed viral samples collected from $S=44$ regions worldwide.
-These regions comprise 15 provinces within China and 29 countries without.
-Understanding the manner in which viruses travel between human populations is an object of ongoing study, and phylogeographic analyses point to the central role of travel networks including those measured by airline passenger counts \citep{holbrook2021massive} or google mobility data \citep{worobey2020emergence}.
-Here, we include three such predictors of travel in our CTMC model by expanding the fixed-effects in regression model \eqref{eq:mixed} to take the form
-\begin{align*}
-	\B = \theta_1 \X_1 + \theta_2  \X_2 + \theta_3 \X_3 \, ,
-\end{align*} 
-for each $\X$ a fixed $44\times 44$ matrix predictor and each $\theta$ a real-valued regression coefficient.
-Figure \ref{fig:FE} presents the three predictors of interest: $\X_1$ contains air travel proximities between locations as measured by annualized air passenger counts between airports contained within a region \citep{holbrook2021massive};  $\X_2$ contains intracontinental proximities arising from physical distances when two regions inhabit the same continent and fixed at the minimum otherwise; finally, $\X_3$ describes the Hubei asymmetry, i.e., the fact that people could enter, but not leave, the Chinese province of Hubei during the early pandemic.
-
 In the context of a Bayesian analysis, we specify independent normal priors on $\theta_1$, $\theta_2$ and $\theta_3$ with means of 0 and variances of 2.
-We also assume that the 1,892 random-effects $\epsilon_{ss'}$ follow sparsity inducing Bayesian bridge priors with global scale parameter $\tau$ and exponent $\alpha=0.25$.
+We also assume that the 1,892 random effects $\epsilon_{ss'}$ follow sparsity inducing Bayesian bridge priors with global scale parameter $\tau$ and exponent $\alpha=0.25$.
 Here, we follow \citet{nishimura2022shrinkage} and specify a Gamma prior on $\tau^{-\alpha}$ with a shape parameter of 1 and a rate parameter of 2.
@@
 Inferring the posterior distribution of all 1,897 model parameters requires an advanced MCMC strategy.
-Namely, we adopt an HMC-within-Gibbs approach, updating the scalars $\gamma$ and $\tau$ independently but updating all 1,895 regression parameters (both fixed- and random-effects) using surrogate HMC accomplished with the first-order approximation
+Namely, we adopt an HMC-within-Gibbs approach, updating the scalars $\gamma$ and $\tau$ independently but updating all 1,895 regression parameters (both fixed and random effects) using surrogate HMC accomplished with the first-order approximation
 \begin{align*}
@@
 The three fixed-effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ obtain ESS's of 721, 721 and 644, respectively.
-The median and minimum ESS for the 1,892 random-effects $\epsilon_{ss'}$ are 721 and 190, respectively.
+The median and minimum ESS for the 1,892 random effects $\epsilon_{ss'}$ are 721 and 190, respectively.
 We note that, after thinning and removing burn-in, the sample only consists of 721 states, so ESS of 721 implies negligible autocorrelation between samples.
-The left plot of Figure \ref{fig:post} presents posterior densities for the fixed-effect coefficients from regression \eqref{eq:mixed}.
+The left plot of Figure \ref{fig:post} presents posterior densities for the fixed effect coefficients from regression \eqref{eq:mixed}.
 The air traffic proximity, intracontinental proximity and Hubei asymmetry coefficients have posterior means and 95\% credible intervals of 0.76 (0.50, 1.03), 0.04 (-0.04, 0.12) and 0.27 (-0.39, 0.78), respectively.
@@
 As one may expect, the matrix looks similar to that of the air traffic predictor matrix in Figure \ref{fig:FE}, but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.
+Finally, we randomly generate regions of unobserved ancestors from their posterior predictive distributions at every 100-thousandth MCMC iteration.
+Figure \ref{fig:tree} projects the empirical posterior predictive mode of these locations onto the phylogenetic tree $\phylogeny$.
 
-\begin{itemize}
-	\item tree
-\end{itemize}
+\begin{figure}[!t]
+	\centering
+	\includegraphics[width=0.8\linewidth]{summary.tree.pdf}
+	\vspace{-1em}
+	\caption{Posterior predictive modes for (unobserved) ancestral locations color a phylogenetic tree that describes the shared evolutionary history of 285 observed SARS-CoV-2 viruses.}\label{fig:tree}
+\end{figure}
 
@@
 Despite the relatively small number of iterations, we observe large ESS that suggest satisfactory accuracy of the first-order approximation within the context of high-dimensional surrogate HMC.
-The ESS for the three fixed-effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343 and 498, respectively.
-The median and minimum ESS for the 1,892 random-effects $\epsilon_{ss'}$ are 1,514 and 1,161, respectively.
+The ESS for the three fixed effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343 and 498, respectively.
+The median and minimum ESS for the 1,892 random effects $\epsilon_{ss'}$ are 1,514 and 1,161, respectively.
 

### 2023-03-18 manuscript/main.tex
hyphen when adjective

@@
 Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.
-These computations become particularly onerous when one opts for a gold-standard mixed effect model \textcolor{red}{(MAGEE)} and specifies
+These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
 \begin{align}\label{eq:mixed}
@@
 \end{align} 
-Here, the fixed-effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.
+Here, the fixed effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.
 The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.
@@
 	\vspace{-2em}
-	\caption{Matrix predictors described in \eqref{eq:linear} combine in a linear manner to form the fixed effect matrix $\B$ featured within the mixed effects regression model \eqref{eq:mixed}.
+	\caption{Matrix predictors described in \eqref{eq:linear} combine in a linear manner to form the fixed-effect matrix $\B$ featured within the mixed-effects regression model \eqref{eq:mixed}.
 Air traffic proximities \emph{(\textbf{a})} are proportional the number of air passengers exchanged between airports within respective regions \citep{holbrook2021massive}.
@@
 \begin{align}\label{eq:linear}
-	\B = \theta_1 \X_1 + \theta_2  \X_2 + \theta_3 \X_3 \, ,
+	\B = \theta_1 \X_1 + \theta_2  \X_2 + \theta_3 \X_3 
 \end{align} 
@@
 We note that, after thinning and removing burn-in, the sample only consists of 721 states, so ESS of 721 implies negligible autocorrelation between samples.
-The left plot of Figure \ref{fig:post} presents posterior densities for the fixed effect coefficients from regression \eqref{eq:mixed}.
+The left plot of Figure \ref{fig:post} presents posterior densities for the fixed-effect coefficients from regression \eqref{eq:mixed}.
 The air traffic proximity, intracontinental proximity and Hubei asymmetry coefficients have posterior means and 95\% credible intervals of 0.76 (0.50, 1.03), 0.04 (-0.04, 0.12) and 0.27 (-0.39, 0.78), respectively.
@@
 Despite the relatively small number of iterations, we observe large ESS that suggest satisfactory accuracy of the first-order approximation within the context of high-dimensional surrogate HMC.
-The ESS for the three fixed effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343 and 498, respectively.
+The ESS for the three fixed-effect regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343 and 498, respectively.
 The median and minimum ESS for the 1,892 random effects $\epsilon_{ss'}$ are 1,514 and 1,161, respectively.

### 2023-04-04 manuscript/main.tex
small titile page edits

@@
 
-\author[1]{Andrew J.~Holbrook}
-\author[1]{Andrew F.~Magee}
-\author[2]{Nathan E.~Glatt-Holtz}
-\author[1,3,4]{Marc A.~Suchard}
+\author[1]{Gustavo Didier}
+\author[1]{Nathan E.~Glatt-Holtz}
+\author[2]{Andrew J.~Holbrook\thanks{Corresponding author: \url{aholbroo@g.ucla.edu}.}}
+\author[2]{\\Andrew F.~Magee}
+\author[2,3,4]{Marc A.~Suchard}
 
- \affil[1]{Department of Biostatistics, University of California, Los Angeles}
- \affil[2]{Department of Mathematics, Tulane University}
+ \affil[1]{Department of Mathematics, Tulane University}
+  \affil[2]{Department of Biostatistics, University of California, Los Angeles}
  \affil[3]{Department of Biomathematics, University of California, Los Angeles}
@@
 But an even simpler approach to surrogate HMC may be appropriate when gradients have series representations.
-\textcolor{red}{MAGEE} consider phylogenetic models that involve CTMC priors and show that the first-order approximation \eqref{eq:firstOrder} of the matrix exponential derivative \eqref{eq:deriv} performs remarkably well within surrogate HMC, achieving an over 30-fold efficiency gain compared to random-walk Metropolis for a 14-state model with over 180 model parameters.
+\citet{magee2023random} consider phylogenetic models that involve CTMC priors and show that the first-order approximation \eqref{eq:firstOrder} of the matrix exponential derivative \eqref{eq:deriv} performs remarkably well within surrogate HMC, achieving an over 30-fold efficiency gain compared to random-walk Metropolis for a 14-state model with over 180 model parameters.
 Here, we demonstrate similar strong performance of the first-order approximation within surrogate HMC for a 44-state model with almost 1,900 model parameters.
@@
 Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.
-These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
+These computations become particularly onerous when one opts for a gold-standard mixed-effect model \citep{magee2023random} and specifies
 \begin{align}\label{eq:mixed}
@@
 The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.
-In this context, \textcolor{red}{MAGEE} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.
+In this context, \citet{magee2023random} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.
 In the following section, we use this method to analyze the global spread of SARS-CoV-2 and show that the first-order approximation maintains its performance for an even higher-dimensional problem than previously considered.

### 2023-04-04 code/derivativeExperiment.R

@@ -52,12 +52,14 @@ blockwise <- function(t,Q,E) {
 #
 n       <- 50
 Q       <- matrix(rexp(n^2),n,n)
+Q       <- (Q + t(Q))/2
 diag(Q) <- 0
 diag(Q) <- - rowSums(Q)
 eigs    <- eigen(Q)$values
 
 for(n in 100*1:10) {
   Q       <- matrix(rexp(n^2),n,n)
+  Q       <- (Q + t(Q))/2
   diag(Q) <- 0
   diag(Q) <- - rowSums(Q)
   eigs    <- c(eigs,eigen(Q)$values)

### 2023-04-05 manuscript/main.tex
abstract

@@
 
-\title{Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2}
+\title{Approximating the matrix exponential derivative with respect to the CTMC generator matrix, with application to global SARS-CoV-2}
 
-%\date{}
+\date{}
 
@@
 
-We investigate the accuracy of a first-order approximation to the directional derivative of the matrix exponential evaluated on continuous-time Markov chain (CTMC) generator matrices.
+The continuous-time Markov chain (CTMC) is the workhorse model of evolutionary biology.
+Learning CTMC model parameters using modern, gradient-based methods requires the derivative of the matrix exponential evaluated at the CTMC's infinitesimal generator (rate) matrix.
+Motivated by the derivative's extreme computational complexity as a function of state space cardinality, recent work demonstrates the surprising effectiveness of a fast, first-order approximation for a host of problems in computational biology.
+We obtain rigorous bounds for the error accrued by such an approach and establish a ``blessing of dimensionality'' result stemming from the novel insight that, in a certain sense, high-dimensional CTMC quickly converge to stationarity.
+Finally, we apply the first-order approximation within surrogate-trajectory Hamiltonian Monte Carlo to the analysis of the early spread of SARS-CoV-2 across 44 geographic regions that comprise a state space of unprecedented dimensionality within evolutionary biology.
 

### 2023-09-12 manuscript/pnas_revision.tex
revisions commit 1

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\usepackage{amsthm}
\usepackage{bbm}
\usepackage[capitalize,nameinlink,noabbrev]{cleveref}
\usepackage{color}

\newtheorem{Theorem}{Theorem}
\newtheorem{Proposition}[Theorem]{Proposition}
\newtheorem{Lemma}[Theorem]{Lemma}
\newtheorem{Corollary}[Theorem]{Corollary}
\newtheorem{Example}[Theorem]{Example}
\newtheorem{Definition}[Theorem]{Definition}
\newtheorem{Remark}[Theorem]{Remark}
\newtheorem{Question}[Theorem]{Question}
\newtheorem{Assumption}[Theorem]{Assumption}

\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand*{\fplus}{\genfrac{}{}{0pt}{}{}{+}}
\newcommand*{\fdots}{\genfrac{}{}{0pt}{}{}{\cdots}}
\newcommand{\mb}{\mathbf}
\newcommand{\mc}{\mathcal}
\newcommand{\dx}{\mbox{d}}

\newcommand{\phylogeny}{{\cal T}}
\newcommand{\tree}{\phylogeny}
\newcommand{\order}[1]{{\cal O}\hspace{-0.2em}\left( #1 \right)}

\newcommand{\bbZ}{{\mathbb Z}}
\newcommand{\bbR}{{\mathbb R}}
\newcommand{\bbN}{{\mathbb N}}
\newcommand{\bbC}{{\mathbb C}}
\newcommand{\bbL}{{\mathbb L}}
\newcommand{\bbP}{{\mathbb P}}
\newcommand{\bbQ}{{\mathbb Q}}
\newcommand{\bbF}{{\mathbb F}}
\newcommand{\bbE}{{\mathbb E}}
\newcommand{\bbM}{{\mathbb M}}
\newcommand{\bbS}{{\mathbb S}}
\newcommand{\bPhi}{\mbox{\boldmath ${\rm \Phi}$}}
\newcommand{\bPsi}{\mbox{\boldmath ${\rm \Psi}$}}
\newcommand{\bphi}{\mbox{\boldmath ${\rm \phi}$}}
\newcommand{\bpsi}{\mbox{\boldmath ${\rm \psi}$}}
\newcommand{\Var}{\textnormal{Var\hspace{0.5mm}}}
\newcommand{\?}{\textbf{?}}
\newcommand{\X}{\textbf{X}}
\newcommand{\x}{\textbf{x}}
\newcommand{\Y}{\textbf{Y}}
\newcommand{\z}{\textbf{z}}
\newcommand{\W}{\textbf{W}}
\newcommand{\w}{\textbf{w}}
\newcommand{\avec}{\textbf{a}}
\newcommand{\s}{\textbf{s}}
\newcommand{\Xbar}{\overline{X}}
\newcommand{\xbar}{\overline{x}}
\newcommand{\Ybar}{\overline{Y}}
\newcommand{\ybar}{\overline{y}}
\newcommand{\MM}{\textnormal{MM}}
\newcommand{\ML}{\textnormal{ML}}
\newcommand{\UMVU}{\textnormal{UMVU}}
\newcommand{\iid}{\textnormal{iid}}
\newcommand{\Poi}{\textnormal{Poi}}
\newcommand{\E}{{\mathbf{ E}}}
\newcommand{\A}{{\mathcal A}}
\newcommand{\supp}{\textnormal{supp}}
\newcommand{\Pfam}{{\mathcal P}}
\newcommand{\Cov}{\textnormal{Cov}}
\newcommand{\U}{{\mathcal U}}
\newcommand{\uvec}{{\mathbf u}}
\newcommand{\diag}{\textnormal{diag}}
\newcommand{\imag}{{\mathbf i}}
\newcommand{\op}{\textnormal{op}}
\newcommand{\as}{\textnormal{a.s.}}

\DeclareMathOperator*{\plim}{\mathit{p}-lim}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}

\newcommand{\bW}{\mbox{\boldmath ${\rm W}$}}
\newcommand{\bU}{\mbox{\boldmath ${\rm U}$}}
\newcommand{\bV}{\mbox{\boldmath ${\rm V}$}}
\newcommand{\bA}{\mbox{\boldmath ${\rm A}$}}
\newcommand{\bC}{\mbox{\boldmath ${\rm C}$}}
\newcommand{\bI}{\mbox{\boldmath ${\rm I}$}}
\newcommand{\bB}{\mbox{\boldmath ${\rm B}$}}
\newcommand{\bF}{\mbox{\boldmath ${\rm F}$}}
\newcommand{\bX}{\mbox{\boldmath ${\rm X}$}}
\newcommand{\bY}{\mbox{\boldmath ${\rm Y}$}}
\newcommand{\bQ}{\mbox{\boldmath ${\rm Q}$}}
\newcommand{\bG}{\mbox{\boldmath ${\rm G}$}}
\newcommand{\bT}{\mbox{\boldmath ${\rm T}$}}
\newcommand{\bS}{\mbox{\boldmath ${\rm S}$}}
\newcommand{\bR}{\mbox{\boldmath ${\rm R}$}}

\newcommand{\bSigma}{\mbox{\boldmath ${\rm \Sigma}$}}
\newcommand{\bbeta}{\mbox{\boldmath ${\rm \beta}$}}

\newcommand{\ba}{\mbox{\boldmath ${\rm a}$}}
\newcommand{\bc}{\mbox{\boldmath ${\rm c}$}}
\newcommand{\bd}{\mbox{\boldmath ${\rm d}$}}
\newcommand{\bs}{\mbox{\boldmath ${\rm s}$}}
\newcommand{\bw}{\mbox{\boldmath ${\rm w}$}}
\newcommand{\bv}{\mbox{\boldmath ${\rm v}$}}
\newcommand{\bu}{\mbox{\boldmath ${\rm u}$}}
\newcommand{\bx}{\mbox{\boldmath ${\rm x}$}}
\newcommand{\by}{\mbox{\boldmath ${\rm y}$}}
\newcommand{\bxi}{\mbox{\boldmath ${\rm \xi}$}}
\newcommand{\bgee}{\mbox{\boldmath ${\rm g}$}}
\newcommand{\bepsilon}{\mbox{\boldmath ${\rm \epsilon}$}}

\newcommand{\bzero}{\mbox{\boldmath $0$}}

\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\norm}[1]{\left\|#1\right\|}
\newcommand{\GD}[1]{#1}
\newcommand{\GDcomment}[2][purple]{GD: \textcolor{#1}{#2}}

\newcommand{\y}{\mathbf{y}}
\newcommand{\QQ}{\mathbf{Q}}
\newcommand{\LLambda}{\boldsymbol{\Lambda}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\JJ}{\mathbf{J}}
\newcommand{\II}{\mathbf{I}}
\newcommand{\AAA}{\mathbf{A}}

\newcommand{\Zero}{\boldsymbol{0}}
\newcommand{\ttheta}{\boldsymbol{\theta}}
\newcommand{\mom}{\boldsymbol{\xi}}
\newcommand{\ppi}{\boldsymbol{\pi}}
\renewcommand{\P}{\mathbf{P}}
\newcommand{\p}{\mathbf{p}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\Z}{\mathbf{E}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\TT}{\mathbf{T}}
\newcommand{\RM}{\mathbf{R}}
\newcommand{\UU}{\mathbf{U}}
\newcommand{\VV}{\mathbf{V}}

\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\NN}{\mathbb{N}}

\newcommand{\alert}[1]{\textcolor{red}{#1}}

%\newcommand{\one}{\mathbbm{1}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\matSpace}{\mathcal{M}}

 \newtheorem{@definition}{\bf Definition}
 \newenvironment{definition}{\begin{@definition}\rm}{\end{@definition}}

 \newtheorem{@remark}{\bf Remark}
 \newenvironment{remark}{\begin{@remark}\rm}{\end{@remark}}

 \newtheorem{@example}{\bf Example}
 \newenvironment{example}{\begin{@example}\rm}{\end{@example}}

 \newtheorem{theorem}{\bf Theorem}
 \newtheorem{lemma}{\bf Lemma}
 \newtheorem{fact}{\bf Fact}
 \newtheorem{corollary}{\bf Corollary}
 \newtheorem{axiom}{\bf Axiom}
 \newtheorem{condition}{\bf Condition}
 \newtheorem{property}{\bf Property}
 \newtheorem{proposition}{\bf Proposition}

\allowdisplaybreaks[1]

\graphicspath{{figures/}}

\input{make-edits}

\title{On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a]{Gustavo Didier}
\author[a]{Nathan E.~Glatt-Holtz} 
\author[b,1]{Andrew J.~Holbrook}
\author[b]{Andrew F.~Magee}
\author[b,c,d]{Marc A.~Suchard}

\affil[a]{Department of Mathematics, Tulane University}
\affil[b]{Department of Biostatistics, University of California, Los Angeles}
 \affil[c]{Department of Biomathematics, University of California, Los Angeles}
 \affil[d]{Department of Human Genetics, University of California, Los Angeles}

% Please give the surname of the lead author for the running footer
\leadauthor{Didier}

% Please add a significance statement to explain the relevance of your work
\significancestatement{Recent work uses a simplistic approximation to
  the matrix exponential derivative to apply gold-standard models from
  evolutionary biology to a collection of challenging data
  analyses. Whereas one may expect the naive approach to break down
  with increasing model dimensionality, empirical results show no such
  failure.  Here, we (1) develop rigorous error bounds that
  improve---in a certain sense---as model dimension grows and (2)
  demonstrate the scalability of the naive approach to a
  higher-dimensional analysis of the global spread of the virus
  responsible for the COVID-19 pandemic.}

% Please include corresponding author, author contribution and author declaration information
% \authorcontributions{Please provide details of author contributions here.}
% \authordeclaration{Please declare any competing interests here.}
% \equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: aholbroo@g.ucla.edu}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Continuous-time Markov chains $|$  Hamiltonian Monte Carlo $|$ Matrix exponential $|$ Molecular epidemiology $|$ Random matrix theory }

\begin{abstract}
  The continuous-time Markov chain (CTMC) is the mathematical
  workhorse of evolutionary biology.  Learning CTMC model parameters
  using modern, gradient-based methods requires the derivative of the
  matrix exponential evaluated at the CTMC's infinitesimal generator
  (rate) matrix.  Motivated by the derivative's extreme computational
  complexity as a function of state space cardinality, recent work
  demonstrates the surprising effectiveness of a naive, first-order
  approximation for a host of problems in computational biology.  In
  response to this empirical success, we obtain rigorous deterministic and
  probabilistic bounds for the error accrued by the naive
  approximation and establish a ``blessing of dimensionality'' result
  that is universal for a large class of rate matrices with random
  entries.  Finally, we apply the first-order approximation within
  surrogate-trajectory Hamiltonian Monte Carlo for the analysis of the
  early spread of SARS-CoV-2 across 44 geographic regions that
  comprise a state space of unprecedented dimensionality for
  unstructured (flexible) CTMC models within evolutionary biology.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list
% environment (quote, quotation, theorem, definition, enumerate,
% itemize...), the line after the list may have some extra
% indentation. If this is the case, add \parshape=0 to the end of the
% list environment.

\dropcap{W}hen measuring the size of ``big data'', one should
consider the structures that implicitly give rise to one's data and
the computational complexities of modeling schemes that appropriately
take such structures into account. Phylogeographic methods
\cite{lemey2009bayesian, lemey2014unifying, holbrook2021massive,
  holbrook2022viral} model large-scale viral transmission between
human populations as a function of the shared evolutionary history of
the viral population of interest.  Data take the form of dates,
locations and genome sequences associated to individual viral
samples. Spatiotemporal structure interfaces with network structure
given by the phylogeny, or family tree, describing the viruses'
collective history beginning with the most recent common ancestor.
While one cannot directly observe this history, one may statistically
reconstruct the phylogenetic tree by positing that changes in the
viral genome happen randomly at regular intervals, thereby capturing
the intuition that viral samples with more differences between their
(aligned) sequences should find themselves further apart on the family
tree.

The continuous-time Markov chain (CTMC) \cite{norris1998markov}
represents the gold-standard mathematical model for such evolution of
characters (e.g., nucleotides) within a fixed span of evolutionary
time. A CTMC defined over a discrete, $d$-element state space consists
of a row vector $\ppi_0$ whose individual components describes the
probability of inhabiting each of the possible states at time $t=0$,
as well as a $d\times d$ infinitesimal generator (or rate) matrix
$\QQ$ with non-negative off-diagonal elements $q_{ij}$, $i\neq j$, and
non-positive diagonal elements $q_{ii}=-\sum_j q_{ij}$. For any lag
$t\geq 0$, the matrix exponential
\cite{moler1978nineteen,moler2003nineteen} provides the Markov chain's
transition probability matrix
\begin{align}
    \P_t := e^{t \QQ} := \sum_{n=0}^\infty \frac{t^n\QQ^n}{n!}  \, ,
\end{align}
which has elements $[\P_t]_{ij}$ that dictate the probability of the
process jumping from state $i$ to state $j$ after time $t$. It is
straightforward to verify that $\P_t$ is a valid transition matrix,
having probability vectors for rows: if $\one$ and $\Zero$ are the
column vectors of ones and zeros, respectively, then $\QQ\one=\Zero$
and, therefore, $\P_t\one=\one$. The law of total probability then
provides the marginal probability of the process at any time
$t \geq 0$ as $\ppi_t=\ppi_0 e^{t\QQ}$.

Whether frequentist \cite{felsenstein1981evolutionary} or Bayesian
\cite{sinsheimer1996bayesian, yang1997bayesian, mau1999bayesian,
  suchard2001bayesian}, likelihood-based approaches to phylogenetic
reconstruction allow phylogenetic tree branch lengths to parameterize
time lags within the CTMC framework. We present the exact statement of
the phylogenetic CTMC paradigm below (see \cref{sec:sars:cov:app}).
Here, we note that the historical importance of tree-reconstruction
from aligned sequences leads to an early emphasis on the
sparse specification of $\QQ$ based on biologically motivated
assumptions
\cite{jukes1969evolution,kimura1980simple,hasegawa1985dating}.
Classical Markov chain Monte Carlo (MCMC) procedures
\cite{metropolis1953equation,hastings1970monte} work well for such
low-dimensional models.  But the phylogenetic CTMC framework has
applications beyond simple nucleotide substitution models. Within,
e.g., Bayesian phylogeography, the work in \cite{lemey2009bayesian}
provides a phylogenetic CTMC model for the spread of avian influenza
across $d=20$ global geographic locations but, for computational
reasons, favors a low-dimensional $\order{d}$ parameterization of
$\QQ$. Similarly, \cite{lemey2014unifying} model the spread of
influenza A H1N1 and H3N2 between as many as $d=26$ geographic regions
but---again for computational reasons---fit the model with
approximation techniques that provide no inferential guarantees.

Recently, \cite{magee2023random} demonstrate the feasibility of
approximate gradient-based methods for both maximum \emph{a
  posteriori} and full Bayesian inference of flexible and
fully-parameterized rate models and apply these methods to a
gold-standard $\order{d^2}$ mixed-effect CTMC model for the spread of
A H3N2 influenza between $d=14$ geographic locations. The usual
CTMC log-likelihood gradient calculations feature the matrix
exponential derivative (see \eqref{eq:sen:eq}, \eqref{eq:var:const:2} below)
\begin{align}
\label{eq:grad:def}
  &\nabla_{\JJ} e^{t \QQ}
    := \lim_{\epsilon \to 0}
    \frac{e^{t (\QQ+ \epsilon \JJ)} - e^{t\QQ} }{\epsilon} \\
  &=  e^{t \QQ} \sum_{n = 0}^\infty \frac{t^{n+1}}{(n+1)!}  
    \left( \sum_{\ell = 0}^n
    (-1)^\ell \binom{n}{\ell} \QQ^\ell \JJ \QQ^{n -\ell}\right)
    \label{eq:grad:series}
\end{align}
computed in the direction $\JJ$ of each of the $d^2$ natural basis
elements $\JJ_{ij}$ spanning the space of real-valued, $d\times d$
matrices ${\mathcal M}(d)={\mathcal M}(d,\bbR)$, thereby requiring at least $\order{d^5}$
floating point operations \cite{najfeld1995derivatives}.

Within the phylogenetic CTMC models of Section \ref{sec:sars:cov:app}, log-likelihood derivative computations that require $\nabla_{\JJ} e^{t \QQ}$ balloon to $\order{KNd^5}$, for $N$ the number of biological specimens observed and $K$ the number of parameters parameterizing $\QQ$.
To address this overwhelming computational cost, \cite{magee2023random} leverage
the simplistic approximation obtained by setting $n=0$ within \eqref{eq:grad:series}:
\begin{align}\label{eq:firstOrder}
	\widetilde{\nabla}_{\JJ} e^{t\QQ} := t e^{t\QQ}\JJ   \, .
\end{align}
\cite{magee2023random} show that this approximation helps reduce total cost to $\order{Kd^2 + Nd^3}$ and use this speedup within
surrogate-trajectory Hamiltonian Monte Carlo (see \cref{sec:Sur:trad:HMC}) 
to obtain a
34-fold improvement in effective sample size per second (ESS/s) over
random-walk MCMC within their 14 region phylogeographic example.
When trying to explain the remarkable empirical performance of the naive
approximation, the authors derive an error upper
bound (for an arbitrary matrix norm)
\begin{align}\label{eq:initial:bnd}
  \lVert \widetilde{\nabla}_{\JJ}e^{t \QQ}  - \nabla_{\JJ} e^{t \QQ} \rVert
  \leq \frac{ \| \JJ \| \lVert\QQ \rVert }{2} (e^{2t} -2t-1) 
\end{align}
that fails to leverage the specific forms of $\QQ$ and $\JJ$.
Notably, this bound explodes as either $t$ or $\lVert \QQ \rVert$
diverges to $\infty$.  Of course, the latter quantity would be
expected to grow large with dimension $d$ without more careful
structural assumptions, e.g., that $\QQ$ is a rate matrix belonging to
the class
\begin{align}\label{eq:rate:mat}
  \mathcal{R}(d) :=
  \Big\{ \QQ \in {\mathcal M}(d) |
  \QQ_{ij} \geq 0 \text{ for } j \not= i ,\, \sum_{j =1}^d \QQ_{ij} = 0 \Big\} \, .
\end{align}

In the following, we use the finer structural properties of $\QQ$ to
obtain more precise bounds on
\begin{align}\label{eq:true:diff}
  \Z(t) := \nabla_\JJ e^{t \QQ} - t e^{t \QQ} \JJ.
\end{align}
In \cref{thm:ass:error}, we provide an affine (in $t$) correction to the 
approximation \eqref{eq:firstOrder} that yields an exponentially tight 
$t \to \infty$ asymptotic for the error \eqref{eq:true:diff}. Then, in 
\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}, we establish
precise probabilistic bounds in the high-dimensional $d \to \infty$
limit for a large class of randomly drawn rate matrices
$\QQ \in \mathcal{R}$.  Here we show, for any $\QQ \in \mathcal{R}$
whose off-diagonal elements are determined by independently and identically distributed (iid) draws from a
positive, sub-exponential distribution $F$, that all of the non-zero
singular values grow as $\sigma_j(\QQ) \sim d$ along with
asymptotically valid almost sure bounds for these rates as
$d \to \infty$.

In regards to this second result,
\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}, note that random rate (or
`Laplacian') matrices have attracted a great deal of attention from the
probability research community, especially in regard to their
high-dimensional properties; see e.g., \cite{takahashi:1969,bai:1999,
  chafai:2010, bordenave:caputo:chafai:2012, chatterjee:hazra:2022,
  nakerst:denisov:haque:2023}. In particular, seminal papers such as
\cite{bryc:dembo:jiang:2006, ding:jiang:2010,
  bordenave:caputo:chafai:2014} establish broad characterizations of
bulk behavior for the Laplacian eigenspectrum.  One contribution of
this paper, of independent interest, is a short and self-contained
construction of useful bounds for the singular values of
$\QQ \in \mathcal{R}$.

In \cref{thm:bride:of:frankinstein} and
\cref{cor:rig:det:rand:wrapup}, we show how \cref{thm:ass:error} and
\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.} combine to provide a more
refined analysis of $\E$ for suitable randomly generated
$\QQ \in \mathcal{R}(d)$. In \cref{thm:bride:of:frankinstein}, we
establish that particular terms appearing in the bound
\eqref{eq:ass:error} in \cref{thm:ass:error} decay with a rate on the
order of $1/d$ in the operator norm topology for large $d$ for certain
classes of symmetric generators $\QQ$. This class includes the random
elements considered in \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.
Here, although \cref{cor:rig:det:rand:wrapup} applies only for
symmetric matrices composed of sub-exponential draws, we provide
strong supplemental numerical evidence that our bounds remain valid
well beyond this special symmetric, sub-exponential special case; see
\cref{rmk:better:approx} and \cref{fig:app}.

One notable practical implication of \cref{thm:bride:of:frankinstein},
\cref{cor:rig:det:rand:wrapup} and \cref{rmk:better:approx} is the
identification of a further correction to
\eqref{eq:firstOrder}. Crucially this correction has the same
$\order{d^3}$ computational cost as \eqref{eq:firstOrder} while
leading to an asymptotically temporally uniformly accurate
approximation of $\nabla_{\JJ} e^{t \QQ}$.  See
\cref{rmk:better:approx} and \eqref{eq:better:help:on:sale} below.

In \cref{sec:sars:cov:app} we follow these theoretical investigations
with an application of the naive, first-order gradient approximation
\eqref{eq:firstOrder} to a challenge in phylogeography requiring
Bayesian inference of a rate-matrix of unprecedented dimensionality.
Namely, we apply the approximation to a gold-standard mixed-effects,
phylogenetic CTMC model that uses 1,897 parameters to describe the
spread of SARS-CoV-2 across a $d=44$ dimensional state space
consisting of different global geographic locations.  Such an
application complements the empirical studies of
\cite{magee2023random} in a manner that emphasizes the naive
approximation's potential for impact.

\section{Rigorous Results}
\label{sec:rig:results}

This section lays out our rigorous results \cref{thm:ass:error},
\cref{thm:bride:of:frankinstein},
\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.} and
\cref{cor:rig:det:rand:wrapup}, the proofs of which appear in the
supplemental material below.

In what follows we adopt the following notational conventions. For any
${\mathbf A} \in {\mathcal M}(d)$, we list the associated (not
necessarily distinct) eigenvalues of $\mathbf{A}$ in ascending order
according to their real part, namely,
\begin{align}\label{eq:EV:real:ass}
  \Re \lambda_1({\mathbf A}) \leq \cdots \leq \Re \lambda_d({\mathbf A}).
\end{align}
Similarly, the singular values of ${\mathbf A}$ are written in
ascending order
\begin{align}\label{eq:SV:ass}
  \sigma_1({\mathbf A}) \leq \cdots \leq \sigma_d({\mathbf A}).
\end{align}
We let ${\mathcal S}(d) = {\mathcal S}(d,\bbR)$ and ${\mathcal S}(d,\bbC)$ represent the spaces of $d \times d$ symmetric and Hermitian matrices,
respectively.

We make use of multiple matrix norms leading to
materially different bounds as $d \rightarrow \infty$ \cite{trefethen2022numerical}.  Take
\begin{align}\label{eq:frob:norm}
  \|\mathbf{A}\|_{F} := \sqrt{\sum_{i,j =1}^d \mathbf{A}_{ij}^2} =
  \sqrt{\sum_{j =1}^d \sigma_j(\mathbf{A})^2}
\end{align}
for the \emph{Frobenius norm} and
\begin{align}\label{eq:op:norm}
  \|{\mathbf A}\|_{\textnormal{op}} :=  \sqrt{\lambda_d({\mathbf
  A}^*{\mathbf A})} = \sqrt{\lambda_d({\mathbf A}{\mathbf A}^*)}
  = \sigma_d(\mathbf{A})
\end{align}
for the \emph{operator norm} of ${\mathbf A}$.  Finally note that when we
simply write $\| \cdot \|$ the statement then holds for any valid
matrix norm as in our formulation of \cref{thm:ass:error}.

\subsection*{Deterministic bounds on approximation
  error in time}

We begin by deriving a dynamical equation for the error $\Z(t)$, defined in
\eqref{eq:true:diff}. Recall that $\X(t) := e^{t \QQ}$ for any
$\QQ \in \mathcal{M}(d)$ obeys the (matrix-valued) ordinary
differential equation
\begin{align}
   \frac{d\X}{dt} = \QQ \X, \quad \X(0) = \II.
\end{align}
Setting
$\Y^\epsilon = \epsilon^{-1} ( e^{t (\QQ+ \epsilon \JJ)}- e^{t \QQ})$
and taking a limit as $\epsilon \to 0$ we find that
$\Y = \nabla_{\JJ} e^{t \QQ}$ obeys
\begin{align}
  \frac{d \Y}{dt}  = \QQ \Y + \JJ \X = \QQ \Y + \JJ e^{t \QQ},
  \quad \Y(0) = \Zero.
	\label{eq:sen:eq}
\end{align}
Thus, variation of constants yields that, for any $t \geq 0$,
\begin{align}
	\nabla_{\JJ} e^{t \QQ} &= e^{t \QQ} \int_0^t e^{-s\QQ} \JJ e^{s\QQ} ds \label{eq:var:const:1}\\
	&=  e^{t \QQ} \sum_{k,m = 0}^\infty \int_0^t \frac{(-s)^{k} \QQ^k \JJ s^{m}\QQ^m}{k!m!}ds
		\notag\\
	&=  e^{t \QQ} \sum_{n = 0}^\infty \frac{t^{n+1}}{(n+1)!}  
	\left( \sum_{\ell = 0}^n (-1)^\ell \binom{n}{\ell} \QQ^\ell \JJ \QQ^{n -\ell}\right).
	\label{eq:var:const:2}
\end{align}
Taking the first order ($n=0$) approximation in \eqref{eq:var:const:2} produces
\eqref{eq:firstOrder}.  Note that, from \eqref{eq:var:const:1}, this
approximation $\widetilde{\nabla}e^{t\QQ}\JJ$ is evidently exact in the
special case when $\JJ$ and $\QQ$ commute.

Next notice that, if we differentiate $\widetilde{\Y} := t e^{t \QQ} \JJ$ in $t$, we
find that $\widetilde{\Y}$ obeys
\begin{align}\label{eq:sen:eq:apx}
  \frac{d \widetilde{\Y}}{dt} = \QQ \widetilde{\Y} +  e^{t \QQ} \JJ,
  \quad \widetilde{\Y}(0) = \Zero.
\end{align}
Thus, taking the error $\Z(t)$ as in \eqref{eq:true:diff} and
combining \eqref{eq:sen:eq} with \eqref{eq:sen:eq:apx} yields
\begin{align}
\label{eq:Z:dym}
  \frac{d\Z}{dt} = \QQ \Z + \JJ e^{t \QQ} -  e^{t \QQ} \JJ,
  \quad \Z(0) = \Zero.
\end{align}
Hence, again integrating this expression, we find
\begin{align}
  \Z(t) &= e^{t \QQ} \int_0^t (e^{-s \QQ} \JJ e^{s \QQ} - \JJ) ds
 \label{eq:ass:procs:0}\\
	&=  e^{t \QQ} \sum_{n = 1}^\infty \frac{t^{n+1}}{(n+1)!}  
   \left( \sum_{\ell = 0}^n (-1)^\ell
   \binom{n}{\ell} \QQ^\ell \JJ \QQ^{n -\ell}\right)
   \label{eq:ass:procs}
\end{align}
as could also be directly deduced from \eqref{eq:var:const:1},
\eqref{eq:var:const:2}.

Given a rate matrix $\QQ$ in $\mathcal{R}(d)$, recall that
$\Re \lambda_{d-1}(\QQ) \leq \lambda_{d}(\QQ)= 0$ by
the Gershgorin circle theorem \cite[Theorem
6.1.1]{horn:johnson:2012}.  Imposing a further
non-degeneracy assumption (e.g., that $\Re \lambda_{d-1}(\QQ)< 0$) we
therefore have an exponential decay in $\QQ e^{t\QQ}$.  This starting
point suggests that, under fairly general conditions, we may decompose
\eqref{eq:ass:procs:0} into a component where $\QQ e^{t\QQ}$ induces
an exponential decay in time and a complementary component taking the
form of a time-affine correction term.

These observations lead to the following theorem, the proof of which appears
in the supplement as \cref{sec:Thm:det:main}.
\begin{Theorem}\label{thm:ass:error}
  Suppose that $\QQ, \JJ \in \mathcal{M}(d)$ for some $d \geq 1$.  We
  assume that we can find an element $\QQ^+ \in \mathcal{M}(d)$ such that
  $\QQ$ is a generalized inverse of $\QQ^+$, namely,
  \begin{align}\label{eq:GI:prop}
    \QQ^+ \QQ \QQ^+ = \QQ^+
  \end{align}
  and such that
  \begin{align}\label{eq:MP:proj:prop}
    e^{\tau \QQ}(\II - \QQ^+ \QQ) \!=\!  \II - \QQ^+ \QQ, \;
    (\II - \QQ \QQ^+)e^{\tau \QQ} \!=\!  \II - \QQ \QQ^+,
  \end{align}
  for any $\tau \in \bbR$. Furthermore, we suppose that $\QQ^+$ and
  $\QQ$ commute
  \begin{align}\label{eq:GI:com}
    \QQ \QQ^+ = \QQ^+ \QQ.
  \end{align}
  Finally, taking $\|\cdot\|$ be any matrix norm, we assume that
  \begin{align}\label{eq:qetq:exp:decay}
    \| \QQ e^{\tau \QQ}\| \leq C_0 e^{-\kappa \tau},
    \quad
    \text{ for all } \tau \geq 0,
  \end{align}
  where the constants $C_0 > 0, \kappa > 0$
  are independent of $\tau$.  Then, under these circumstances, 
  \begin{align}\label{eq:ass:error}
    \|\QQ^+ \JJ (\II - \QQ \QQ^{+})-& (\II -\QQ^+ \QQ) \JJ \QQ^{+}
        +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
                \notag\\
      &
        +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\| \leq C(1+ t)e^{-\kappa t}
\end{align}
for any $t \geq 0$.  Here, $C > 0$ is a $t$-independent
constant which is given explicitly as
\begin{align}
  C_0 \big(\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
  \!&+\! \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
    \!+\! \|\QQ^+ \hspace{0.5mm}\JJ\|\big)
    \notag\\
    &+ C_0^2 \| \QQ^+  \JJ  \QQ^{+} \|.
      \label{eq:exp:error:ass:er}
  \end{align}
\end{Theorem}

\begin{Remark}\label{rmk:mat:class:dcy}
  To illuminate the scope of \cref{thm:ass:error}, we have the
  following three classes of matrices maintaining the conditions
  \eqref{eq:GI:prop}--\eqref{eq:qetq:exp:decay} as follows.
  \begin{itemize}
  \item[(i)] Suppose that $\QQ \in \mathcal{M}(d)$ is such that
    \begin{align}\label{eq:ev:d:cond}
      \Re \lambda_{d-1}(\QQ) <  \Re \lambda_d({\mathbf Q}) \leq 0,
       \lambda_d({\mathbf Q})  \text{ is simple }
    \end{align}
    and such that, if $\lambda_d({\mathbf Q})$ has an imaginary component,
    then its real part is strictly negative.  Under these
    circumstances, writing $\QQ$ in its Jordan canonical form yields,
    for some $m \geq 1$,
    \begin{equation}\label{e:Q=Jordan_decomp}
      \QQ = \M \diag (J_{1},\hdots,J_{m-1},\lambda_d(\QQ))\M^{-1}.
    \end{equation}
    Here, under \eqref{eq:ev:d:cond} each of
    these blocks $J_j$ must be invertible and so
    we may take
    \begin{align}
      \QQ^+ := \M \,
      \diag (J_{1}^{-1},\hdots,J_{m-1}^{-1},\lambda_d(\QQ)^{+})
      \, \M^{-1},
    \end{align}
    where 
    \begin{align}\label{eq:eig:gen:inv}
      \lambda_d(\QQ)^{+} =
      \begin{cases}
        0& \text{ if } \lambda_d(\QQ) = 0,\\
        \lambda_d(\QQ)^{-1}& \text{ otherwise.}
       \end{cases}
    \end{align}
  \item[(ii)] We next consider the case where $\QQ \in \mathcal{M}(d)$
    is diagonalizable and its spectrum lies strictly on the left half
    plane or at the origin.  This time we can write
    \begin{align}\label{eq:spec:decomp}
    \QQ = \M \LLambda \M^{-1} \text{ where }
      \LLambda = \diag (\lambda_1(\QQ), \ldots ,\lambda_d(\QQ))
    \end{align}
    and we set 
    \begin{align}\label{eq:gen:inv:diag}
     \! \!\! \!\QQ^+ \! \! \!= \!\M \LLambda^+\M^{-1} \text{ with }
      \LLambda^{+} \! \! \!= \! \diag (\lambda_1(\QQ)^+\!\! \! \!, \ldots
      ,\lambda_d(\QQ)^+).
      \end{align}
    The complex numbers $\lambda_j(\QQ)^+$,
    $j = 1, \ldots, d$ are defined as in \eqref{eq:eig:gen:inv}.
  \item[(iii)] Finally, we specialize to the case where
    $\QQ \in \mathcal{S}(d)$ is symmetric. In this case,
    $\QQ = \UU \LLambda \UU^*$, where $\UU$ is a unitary matrix and
    $\LLambda = \diag (\lambda_1(\QQ), \ldots ,\lambda_d(\QQ))$,
    $\lambda_j(\QQ)$ are its (real) eigenvalues.  We suppose that
    these eigenvalue are all non-positive,
    \begin{align}\label{eq:ev:d:cond:sym:case}
      \lambda_d({\mathbf Q}) \leq 0,
    \end{align}
    Here we take $\QQ^+$ as the Moore-Penrose inverse, namely,
    \begin{align}\label{eq:MP:inv:sym:case}
      \QQ^+ = \UU \LLambda^+ \UU^*,
   \end{align}
    where $\LLambda^+$ is as in \eqref{eq:gen:inv:diag}.   
  \end{itemize}
\end{Remark}
In anticipation of \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.} below
and our desired application in \cref{sec:sars:cov:app} we are
preoccupied with the dimensional dependence of the constants in
\eqref{eq:qetq:exp:decay}, \eqref{eq:ass:error}
and \eqref{eq:exp:error:ass:er} in our formulation of
\cref{thm:ass:error}.  We next provide some such desirable bounds in
case (iii) of \cref{rmk:mat:class:dcy}.  Note that analogous results
for generators $\QQ$ in the classes (i) or (ii) would seemingly
require a delicate analysis of the associated eigenspaces (i.e., of the
structure of $\M$ in \eqref{e:Q=Jordan_decomp} or
\eqref{eq:spec:decomp} respectively).  However,
\cref{rmk:non:sym:num:ev} and \cref{fig:app} provide numerical evidence
of a broader scope for dimensionally improving approximations beyond
the symmetric case, at least for certain classes of randomly drawn
matrices.
\begin{Theorem}\label{thm:bride:of:frankinstein}
  Let $\QQ \in \mathcal{S}(d)$ be non-positive, i.e.,  suppose that
  \eqref{eq:ev:d:cond:sym:case} holds.  Take $\QQ^+$ as in
  \eqref{eq:MP:inv:sym:case} and define
    \begin{align}\label{eq:first:neg}
      d_{-} = \max\{ 1 \leq j \leq d | \Re \lambda_{j}(\QQ) < 0\}.
    \end{align}
    Then, for any $\JJ \in \mathcal{M}(d)$,
  \begin{align}
      \| t(\II&  -\QQ^+  \QQ) \JJ \QQ \QQ^{+}
                +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\|_{F} 
                \label{eq:bride:of:frankinstein:1}\\
    \leq&  
    \Big( \! \sqrt{d_{-}}|\lambda_1(\QQ)| \! \cdot\!  \Big[ 2 \sqrt{d-d_{-}} \| \QQ^+\|_F^2 +  \| \QQ^+\|_F
    \notag \\ 
              &\qquad \qquad \qquad \;
    + \sqrt{d_{-}}|\lambda_1(\QQ)|\| \QQ^+\|_F^2 \Big]   (1+t)e^{- t|\lambda_{d-}(\QQ)|}
   \notag \\ 
&\quad+  2\sqrt{d - d_{-}} \| \QQ^+\|_F  \Big)\|\JJ\|_{F},  \notag
  \end{align}
  with $\| \QQ^+\|_F^2 := \sum^{d_-}_{k=1}\frac{1}{\lambda^2_{k}(\QQ)}$, whereas
  \begin{align}
    \| t(&\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
         +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\|_{\op}
         \label{eq:bride:of:frankinstein}\\
    \leq &
           \Big(
           \frac{|\lambda_1(\QQ)|(2 + |\lambda_{d_-}(\QQ)|  + |\lambda_1(\QQ)|)}{\lambda^2_{d-}(\QQ)} 
           (1+t) e^{-t |\lambda_{d_-}(\QQ)|}
           \notag
    \\ 
              &\qquad+ \frac{2}{|\lambda_{d_{-}}(\QQ)|}\Big) \|\JJ\|_{\op}.
                \notag
  \end{align}
  Under the further assumption that $\lambda_{d}(\QQ) = 0$
  and
  \begin{align}
    \mu_1 d \leq |\lambda_{d-1}(\QQ)|
    \leq |\lambda_{1}(\QQ)| \leq \mu_2 d
  \end{align}
  for some $0< \mu_1\leq \mu_2$, we have
    \begin{align}\label{eq:bride:of:frankinstein:1:rm:app}
    \| t(&\II  -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
         +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\|_{F}
         \\
     \leq &    \Big(\frac{\mu_2}{\mu_1^2}  \cdot \Big[ 2 \sqrt{d} + \mu_1 d  
            +  \mu_2 d^2\Big](1+t)^{-t \mu_1 d} +  \frac{2}{\mu_1 \sqrt{d}} \Big) \|\JJ\|_{F},
            \notag
  \end{align}
  and that
    \begin{align}\label{eq:bride:of:frankinstein:0:rm:app}
    \| t(\II & -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
         +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\|_{\op}\\
      \leq &  \Big( \frac{\mu_2}{\mu^2_1}\cdot\Big[   \frac{2}{d} + \mu_1  +\mu_2\Big]
             (1+t)e^{-t \mu_1 d} + \frac{2}{\mu_1 d}\Big)
            \|\JJ\|_{\op}.\notag
  \end{align}
\end{Theorem}

\subsection*{High-dimensional asymptotics
  via random matrix theory}

We turn to our probabilistic bounds on the singular values of randomly
generated rate matrices, \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.
Although interesting in its own right, this result leads to
consequences for the bounds in \eqref{eq:qetq:exp:decay} and
\eqref{eq:ass:error} when applied in \cref{thm:bride:of:frankinstein}.
Before proceeding, we briefly introduce further mathematical
preliminaries associated with the so-called sub-exponential random
variables.
\begin{definition}
  \label{def:sub:exp:RV}
  A random variable $X$ is called \textit{sub-exponential} if there
  exists some constant $K > 0$ for which its tails satisfy
\begin{equation}\label{e:def_subexp_RV}
\bbP(|X|\geq t) \leq 2 \exp\{-t/K\}, \quad \forall t \geq 0.
\end{equation}
In this case, the \textit{sub-exponential norm} of $X$ is defined by
\begin{equation}\label{e:sub-exp_norm}
\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE \exp(|X|/s) \leq 2\big\}.
\end{equation}
The \textit{class of sub-exponential distributions} is denoted by
\begin{align*}
L_{\psi_1} = \big\{F_X(dx): \|X\|_{\psi_1} < \infty \big\}.
\end{align*}
\end{definition}

\begin{remark}\label{rmk:exp:mom}
  In fact, by \cite[Proposition 2.7.1, p.\ 33]{vershynin:2018},
  condition \eqref{e:def_subexp_RV} is equivalent to the existence of
  some $s_0 > 0$ such that $\bbE \exp(|X|/s_0) \leq 2$, namely,
  $\|X\|_{\psi_1} < \infty$ in \eqref{e:sub-exp_norm}.  As a notable
  example, if $X \sim \exp(\lambda)$, $\lambda > 0$, then it is easy
  to see that $X \in L_{\psi_1}$, indeed.
\end{remark}

We formulate our second major result as follows. 
\begin{Theorem}\label{t:lambda_n-1(Q)->-infty_non-symm_a.s.}
  Let $F_X \in L_{\psi_1}$ be a distribution such that $X \geq 0$
  a.s., $\bbE X = \mu > 0$ and $\Var X = \sigma^2 > 0$. We consider a
  sequence of random matrices
  $\QQ \equiv \QQ(d) = \{q_{ij}\}_{i,j=1,\hdots,d} \in \mathcal{R}(d)$, $d \in \bbN$,
  where either
  \begin{align}\label{e:RM_theorem_general_condition}
    \begin{aligned}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i \neq j}
    \stackrel{\textnormal{iid}}\sim F_X
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d
  \end{aligned}
  \end{align}
  or we impose that $\QQ \in \mathcal{R}(d) \cap \mathcal{S}(d, \RR)$ as 
    \begin{align}
    \begin{aligned}\label{e:RM_theorem_symmetry_condition}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i > j}
  \stackrel{\textnormal{iid}}\sim F_X,
  q_{ij} := q_{ji} \text{ for } i < j
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d.
  \end{aligned}
\end{align}
    Then, in either of these cases, for any $d \in \bbN \backslash\{1\}$, we have
    \begin{align}
 \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big) \leq& \frac{\sigma_2(\QQ)}{d} 
 \notag \\
 \leq& \frac{\sigma_d(\QQ)}{d}
 \leq \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
 \label{e:lambda_n-1(Q)->-infty_non-symm}
      \end{align}
      almost surely. In \eqref{e:lambda_n-1(Q)->-infty_non-symm},
      $O_{\textnormal{a.s.}}$ is as in
      \eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
\end{Theorem}

\begin{remark}
  The bounds constructed in Theorem
  \ref{t:lambda_n-1(Q)->-infty_non-symm_a.s.} are strongly reminiscent
  of the bounds for eigenvalues provided in Theorem 1.5 of the seminal
  paper \cite{bordenave:caputo:chafai:2014} (see also, for instance,
  Corollary 1.6 in \cite{bryc:dembo:jiang:2006} and Corollary 1.1 in
  \cite{ding:jiang:2010}).
\end{remark}

Finally, let us observe that \cref{thm:bride:of:frankinstein},
\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.} as well as the fact
that
\begin{align}
  \sigma_j(\QQ) = |\lambda_{d - j +1}(\QQ)|,
  \text{ for e.g. any }
  \QQ \in \mathcal{S}(d) \cap \mathcal{R}(d)
  \label{eq:eigen:to:SV:SYM:dummy}
\end{align}
combine to produce the following immediate corollary.
\begin{Corollary}\label{cor:rig:det:rand:wrapup}
  Consider any sequence of random matrices
  $\QQ \equiv \QQ(d) = \{q_{ij}\}_{i,j=1,\hdots,d} \in
  \mathcal{R}(d)$, $d \in \bbN$, as in
  \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.} under the second
  (symmetric) case \eqref{e:RM_theorem_symmetry_condition}.  Then,
  taking $\mu_1 = \mu_1(d)$ and $\mu_2= \mu_2(d)$ as the resulting
  lower and upper bounds defined by
  \eqref{e:lambda_n-1(Q)->-infty_non-symm}, we have that $\QQ$
  satisfies both \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app}, cf.
  \eqref{eq:eigen:to:SV:SYM:dummy} relative to this sequence of
  $\mu_1, \mu_2$ for any $\JJ \in \mathcal{M}(d)$.
\end{Corollary}

\begin{figure*}[!t]
    \centering
    \includegraphics[width=\linewidth]{appendixFig3.pdf}
    \caption{Frobenius norm errors obtained by first-order approximation $te^{t\QQ}\JJ$ and by affine-corrected first-order approximation
      $te^{t\QQ}\JJ - t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$
      (\cref{thm:bride:of:frankinstein}) under increasingly relaxed
      assumptions.  Within each assumption set, we average over 20
      independent Monte Carlo simulations of random generator matrices
      for each dimension. Plot \textbf{a} corresponds to asymmetric
      generators with off-diagonal elements having independent and
      identically distributed (iid) standard exponential random
      variables that correspond to the sub-exponential distribution
      hypothesis. Plot \textbf{b} drops the identical distribution
      assumption by allowing each row and column of the generator
      matrix to additively contribute its own mean---itself given by a
      standard exponential---to its corresponding exponentially
      distributed entries. Plot \textbf{c} features rate
      matrices with iid Cauchy entries truncated to be
      positive. Empirically, the results of
      \cref{cor:rig:det:rand:wrapup} extend beyond the symmetric,
      iid and sub-exponential hypotheses, suggesting scope of future work.}
    \label{fig:app}
\end{figure*}

\begin{Remark}\label{rmk:non:sym:num:ev}
  Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
  limited to symmetric random rate matrices whose above diagonal
  elements are independent and identically distributed draws from a
  sub-exponential distribution. However, strong numerical evidence
  suggest that the scope of the approximations
  \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond
  limitations of \cref{cor:rig:det:rand:wrapup} in several different
  ways. \cref{fig:app} explores the
  consequences of relaxing various assumptions of
  \cref{cor:rig:det:rand:wrapup}. In general, there appears to be no
  significant departure from the idea that the proxy form
  $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
  the true approximation error (\eqref{eq:true:diff}) as the dimension
  increases.
\end{Remark}

\begin{Remark}\label{rmk:better:approx}
  Calculation of
  $\widetilde{\nabla}_{\JJ} e^{t\QQ} := t e^{t\QQ} \JJ$ requires
  $\order{d^3}$ operations by, e.g., computing the spectral
  decomposition $\QQ$ as in \eqref{eq:spec:decomp}.  One may then recycle this
  decomposition to determine the
  additional term $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ for little extra cost.  In view of
  \cref{cor:rig:det:rand:wrapup} and \cref{fig:app},
  \begin{align}\label{eq:better:help:on:sale}
    \widehat{\nabla}_{\JJ} e^{t\QQ} := t e^{t\QQ} \JJ
    -t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
  \end{align}
  provides an accurate approximation of $\nabla_{\JJ} e^{t\QQ}$ for
  asymptotically for large $d$. Thus, we anticipate further computational improvements when fitting
  large, gold-standard models using this
  refined approximation.  That said, we leave the efficient and scalable application of $\widehat{\nabla}_{\JJ} e^{t\QQ}$ to future work.
\end{Remark}

\section{Application: global spread of SARS-CoV-2}
\label{sec:sars:cov:app}

\cite{magee2023random} consider phylogenetic models that involve CTMC
priors and show that the first-order approximation
\eqref{eq:firstOrder} of the matrix exponential derivative
\eqref{eq:grad:def} performs remarkably well within surrogate HMC (see
\cref{sec:Sur:trad:HMC} in the Supplement), achieving an over 30-fold
efficiency gain compared to random-walk Metropolis for a 14-state
model with over 180 model parameters.  Here, we demonstrate similar
strong performance of the first-order approximation within surrogate
HMC for a 44-state model with almost 1,900 model parameters.

\subsection*{Phylogenetic CTMC}

Start with a (possibly unknown) rooted and bifurcating phylogenetic
tree $\phylogeny$ consisting of $N$ leaf nodes that correspond to
observed biological specimens and $N-1$ internal nodes that correspond
to unobserved ancestors.  The tree also contains $2N-2$ branches of
length $t_v$ connecting each child node $v$ to its parent $u$.

Given $\phylogeny$, we model the evolution of $d$ characters along
each branch of the tree according to a CTMC model with $d\times d$
generator matrix $\QQ$ and stationary distribution
$\ppi=(\pi_1,\dots,\pi_d)=\lim_{t\rightarrow
  \infty}\widetilde{\ppi}e^{t\QQ}$, for $\widetilde{\ppi}$ for any
arbitrary probability vector.  Examples of characters are the $d=4$
nucleotides within a set of aligned genome sequences
\cite{jukes1969evolution} and the set of $d=15$ geographic regions
visited by an influenza subtype \cite{lemey2014unifying}.  We may
scale $t_v$ to be raw time (e.g., years) or the expected number of
substitutions with respect to $\ppi$ depending on the given problem.
In the former case, one may augment the model with a rate scalar
$\gamma$ that modulates the expected number of substitutions across
all branches, and the finite-time transition probability matrix along
branch $v$ becomes $\P_v:=e^{\gamma t_v \QQ}$.  In the following, we
further posit $\QQ=\QQ(\ttheta)$ for $\ttheta$ a vector of parameters.

Let data $\Y$ be the $d\times N$ matrix with columns $\y_n$, $n \in \{1,\dots,N\}$, each
having a single non-zero entry (set to 1) corresponding to the
observed state of the biological specimen.  One may use
any node $v$ to express the likelihood
 \begin{align}
p(\Y|\ttheta) = \p_v^T \q_v \, ,
 \end{align}
 where $\p_v$ and $\q_v$ are the post-order and pre-order partial
 likelihood vectors, respectively \cite{ji2020gradients}.  The former describes the
 probability of the observed states for all observed specimens (i.e.,
 leaf nodes) that descend from node $v$, conditioned on the state at
 node $v$. The latter describes analogous probabilities for all
 observed specimens not descending from node $v$. For leaf nodes,
 $\p_n:=\y_n$, $n \in \{1,\dots,N\}$, and one may specify the root node's pre-order partial likelihood to be any arbitrary probability vector \emph{a priori}. Let `$\circ$' denote the Hadamard or elementwise product between matrices or vectors of equal dimensions.  If we suppose that node $u$ gives rise to two child
 nodes, $v$ and $w$, then
\begin{align}\label{eq:recursions}
  \p_u = \P_v \p_v \circ \P_w \p_w\, ,
  \quad \q_v= \P_v^T \left( \q_u \circ  \P_w \p_w\right) \, .
\end{align}
Using the chain rule and the fact that $\p_v$ does not depend on
$\P_v$, one may write the likelihood's derivative with respect to a
single parameter $\theta_k$ thus:
\begin{align}\label{eq:phyloGrad}
	\frac{\partial}{\partial \theta_k}  p(\Y|\ttheta) 
  &\propto   \sum_{v=1}^{2N-2}
    \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v}
    \frac{\partial \P^T_v}{\partial \theta_k} \right) \\ \notag
  &= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w)
    \p_v^T\bigg( \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_k}\bigg)^{\!T}\right)  \\ \nonumber
  &=\sum_{v=1}^{2N-2} \p_v^T
    \Biggl(\sum_{i,j=1}^d  \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ij}} \frac{\partial q_{ij}}{\partial \theta_k} \Biggr)^{\!T}
    \hspace{-0.3em}(\q_u\circ \P_w\p_w)  \, ,
\end{align}
where we suppress the dependence of $u$ and $w$ on $v$.

Whereas the recursions of \eqref{eq:recursions} facilitate fast
likelihood computation, inferring $\ttheta$ using gradient-based
approaches such as HMC requires a large number of repeated evaluations
of the matrix exponential derivative.  These computations become
particularly onerous when one opts for a gold-standard mixed-effects
model \cite{magee2023random} and specifies
\begin{align}\label{eq:mixed}
\log q_{ij} = b_{ij} + \epsilon_{ij} \,, \quad i\neq j\, .
\end{align} 
Here, the fixed effects $b_{ij}$ are elements of some non-random
matrix $\B$, and the random effects $\epsilon_{ij}$ are mutually
independent \emph{a priori} and inferred as model parameters.  The
dimension $K$ of $\ttheta$ in this model is $\mathcal{O}(d^2)$, so the $\mathcal{O}(KNd^5)$ cost of the log-likelihood derivative \eqref{eq:phyloGrad}
becomes a massive $\mathcal{O}(Nd^7)$.  In this context,
\cite{magee2023random} show that an approximate log-likelihood derivative based on the first-order approximation to the matrix exponential helps achieve a considerable speedup over the exact
derivative, requiring only $\mathcal{O}(d^4 + Nd^3)$ floating-point operations yet facilitating high-quality proposals in the context
of surrogate HMC.

In the following section, we use this method to analyze the global
spread of SARS-CoV-2 and show that the first-order approximation
maintains its performance for an even higher-dimensional problem than
previously considered.

\subsection*{Bayesian analysis of SARS-CoV-2 contagion}

\begin{figure*}[!t]
	\centering
	\includegraphics[width=\linewidth]{fixedEffects.pdf}
	\vspace{-2em}
	\caption{Matrix predictors described in \eqref{eq:linear} combine in a linear manner to form the fixed-effect matrix $\B$ featured within the mixed-effects regression model \eqref{eq:mixed}.  Air traffic proximities \emph{(\textbf{a})} are proportional the number of air passengers exchanged between airports within respective regions \cite{holbrook2021massive}.  Intracontinental proximities \emph{(\textbf{b})} take values between -1 for regions on different continents to 1 for adjacent regions.  The Hubei asymmetry \emph{(\textbf{c})} roughly characterizes the Hubei quarantine of early 2020.}\label{fig:FE}
\end{figure*}

We use the phylogenetic CTMC framework to model the early spread of
SARS-CoV-2---the virus responsible for the ongoing COVID-19
pandemic---based on $N=285$ observed viral samples collected from $31$
regions worldwide between December 24, 2019 and March 19, 2020. These
regions comprise 13 provinces within China and 18 countries without.
Understanding the manner in which viruses travel between human
populations is an object of ongoing study, and phylogeographic
analyses point to the central role of travel networks including those
measured by airline passenger counts \cite{holbrook2021massive} or
google mobility data \cite{worobey2020emergence}.  Here, we include
three such predictors of travel in our CTMC model by expanding the
fixed effects in regression model \eqref{eq:mixed} to take the form
\begin{align}\label{eq:linear}
	\B = \theta_1 \X_1 + \theta_2  \X_2 + \theta_3 \X_3 
\end{align} 
for $\X_1$, $\X_2$ and $\X_3$ fixed $44\times 44$ matrix predictors and $\theta_1$, $\theta_2$ and $\theta_3$
 real-valued regression coefficients. Note we have expanded the
number of regions between which viruses may travel to $d=44$ by
including two additional Chinese provinces and 11 additional
countries.  Figure \ref{fig:FE} presents the three predictors of
interest: $\X_1$ contains air travel proximities between locations as
measured by annualized air passenger counts between airports contained
within a region \cite{holbrook2021massive}; $\X_2$ contains
intracontinental proximities arising from physical distances when two
regions inhabit the same continent and fixed at the minimum otherwise;
finally, $\X_3$ describes the Hubei asymmetry, i.e., the non-existence
of human travel out of the Hubei province in early 2020.

\begin{figure*}[!t]
	\centering
	\includegraphics[width=\linewidth]{postViz.pdf}
	\vspace{-2em}
	\caption{Posterior inference. Posterior densities
          \emph{(\textbf{a})} for the three fixed-effect regression
          coefficients corresponding to the predictor matrices of
          Figure \ref{fig:FE} reflect largely positive associations
          between predictors and infinitesimal rate matrix, although
          air traffic proximity has the only statistically significant
          coefficient with posterior mean of 0.76 and 95\% credible
          interval of (0.50, 1.03).  The posterior mean
          \emph{(\textbf{b})} infinitesimal rate matrix closely
          resembles the air traffic predictor, while reflecting the
          Hubei asymmetry to a lesser extent.}\label{fig:post}
\end{figure*}

In the context of a Bayesian analysis, we specify independent normal
priors on $\theta_1$, $\theta_2$ and $\theta_3$ with means of 0 and
variances of 2.  We also assume that the 1,892 random effects
$\epsilon_{ij}$ follow sparsity inducing Bayesian bridge priors with
global scale parameter $\tau$ and exponent $\alpha=0.25$.  Here, we
follow \cite{nishimura2022shrinkage} and specify a Gamma prior on
$\tau^{-\alpha}$ with a shape parameter of 1 and a rate parameter of
2. Finally, we place a flat prior on the rate scalar $\gamma$.
Inferring the posterior distribution of all 1,897 model parameters
requires an advanced MCMC strategy. Namely, we adopt an
HMC-within-Gibbs approach, updating the scalars $\gamma$ and $\tau$
independently but updating all 1,895 regression parameters (both fixed
and random effects) using surrogate HMC accomplished with the
first-order approximation
\begin{align*}
  \gamma t_v  e^{\gamma t_v \QQ} \JJ_{ij}
  \approx \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ij}} 
\end{align*}
within \eqref{eq:phyloGrad}.

We generate 8 million MCMC samples in this manner, saving 1 in 10,000
Markov chain states, in order to guarantee a minimum effective sample
size (ESS) greater than 100.  By far, the worst-mixing parameter is
the global scale $\tau$, which obtains an ESS of 185.  The three
fixed-effects regression coefficients $\theta_1$, $\theta_2$ and
$\theta_3$ obtain ESS's of 721, 721 and 644, respectively.  The median
and minimum ESS for the 1,892 random effects $\epsilon_{ss'}$ are 721
and 190, respectively.  We note that, after thinning and removing
burn-in, the sample only consists of 721 states, so ESS of 721 implies
negligible autocorrelation between samples.

The left plot of Figure \ref{fig:post} presents posterior densities
for the fixed-effect coefficients from regression \eqref{eq:mixed}.
The air traffic proximity, intracontinental proximity and Hubei
asymmetry coefficients have posterior means and 95\% credible
intervals of 0.76 (0.50, 1.03), 0.04 (-0.04, 0.12) and 0.27 (-0.39,
0.78), respectively. While these results suggest a positive
association between each predictor matrix and the infinitesimal
generator matrix $\QQ$, the only predictor with a statistically
significant association is air traffic proximity.  This result agrees
with a previous phylogeographic analysis of the global spread of
influenza \cite{holbrook2021massive}.  The right plot of Figure
\ref{fig:post} presents the posterior mean for generator $\QQ$.  As
one may expect, the matrix looks similar to that of the air traffic
predictor matrix in Figure \ref{fig:FE}, but one may also see the
influence of the Hubei asymmetry in, e.g., the squares corresponding
to travel between Guangdong and Hubei provinces.  Finally, we randomly
generate regions of unobserved ancestors from their posterior
predictive distributions every 100-thousandth MCMC iteration.
Figure \ref{fig:tree} projects the empirical posterior predictive mode
of these locations onto the phylogenetic tree $\phylogeny$.

\begin{figure*}[!t]
	\centering
	\includegraphics[width=0.7\textwidth]{summary_tree.pdf}
	\vspace{-1em}
	\caption{Posterior predictive modes for (unobserved) ancestral
          locations color a phylogenetic tree that describes the
          shared evolutionary history of 285 SARS-CoV-2
          samples.}\label{fig:tree}
\end{figure*}

In addition to these scientific questions of interest, we are
interested in the performance of the first-order approximation as a
surrogate gradient for HMC in such a high-dimensional setting. To
isolate this performance, we fix the Bayesian bridge global-scale
$\tau$ at 2.5$\times10^{-5}$.  We generate a Markov chain with 80,000
states, saving every tenth state and removing the first 1,000 states as
burn-in.  Despite the relatively small number of iterations, we
observe large ESS that suggest satisfactory accuracy of the
first-order approximation within the context of high-dimensional
surrogate HMC.  The ESS for the three fixed-effect regression
coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343
and 498, respectively.  The median and minimum ESS for the 1,892
random effects $\epsilon_{ij}$ are 1,514 and 1,161, respectively.

\section*{Discussion}

We develop tight probabilistic bounds on the error associated
with a simplistic approximation to the matrix exponential derivative
for a large class of CTMC infinitesimal generator matrices with random
entries.  Our ``blessing of dimensionality'' result shows that this
error improves for higher-dimensional matrices.  We apply the
numerically naive approach to the analysis of the global spread of
SARS-CoV-2 using a mixed-effects model of unprecedented dimensions.
The results obtained herein suggest the further study of CTMCs through
the lens of random matrix theory.  Furthermore, this analysis suggests a refinement of the first-order
approximation to the matrix exponential derivative that may be particularly useful within
modern, high-dimensional settings.

\acknow{GD was partially supported by the Simons Foundation
  collaboration grant $\#$714014. NEGH received support for this work
  under NSF DMS 2108790.  AJH is supported by a gift from the Karen
  Toffler Charitable Trust and by grants NIH K25 AI153816, NSF DMS
  2152774 and NSF DMS 2236854. AFM and MAS are partially supported
  through grants NIH R01AI153044 and R01AI162611.}

\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliography{refs}

\newpage

%\matmethods{

\appendix

\section{Proof of \cref{thm:ass:error}}
\label{sec:Thm:det:main}

Recall that $\Z(t)$ is defined as in \eqref{eq:true:diff}, and
consider the representation \eqref{eq:ass:procs:0}. We begin by
decomposing the first term $e^{-s\QQ} \JJ e^{s \QQ}$ inside of the
integral in \eqref{eq:ass:procs:0} as
\begin{align}
  e^{-s\QQ} &\JJ e^{s \QQ} \notag\\
  =&e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
     + e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
                  \notag\\
      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+}) e^{s\QQ}
                \notag\\
       &+e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}) e^{s\QQ}
                           \notag\\
  = &e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
      + (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
         \label{eq:big:yuck:1}                          
      \\
      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
      %\notag\\
  + (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}),
  \notag
\end{align}
where we used \eqref{eq:MP:proj:prop} for the second identity.
Regarding the middle two terms, with our commutativity
\eqref{eq:GI:com} assumption, we observe that
\begin{align}
  (\II &-\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
         + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
                \label{eq:big:yuck:2}                          
                     \\
  &=  (\II -\QQ^+ \QQ) \JJ \QQ^{+} \QQ e^{s\QQ}
    + e^{-s\QQ}\QQ \QQ^+ \JJ (\II - \QQ \QQ^{+})
    \notag\\
  &= \frac{d}{ds}
    \left(
    (\II -\QQ^+ \QQ) \JJ \QQ^{+} e^{s\QQ}
    - e^{-s\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
    \right).
 \notag
\end{align}
Hence, integrating in time, and then using \eqref{eq:GI:prop},
\eqref{eq:GI:com} for the first and last term terms on the right hand side, we have
\begin{align}
  \int_0^t& \big\{(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
                             + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})\big\}ds
                             \notag\\
   =& (\II -\QQ^+ \QQ) \JJ \QQ^{+}e^{t\QQ}
      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
      \notag\\
    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
      +  \QQ^+ \JJ (\II - \QQ \QQ^{+})
      \notag\\
   =& (\II -\QQ^+ \QQ) \JJ (\QQ^+)^2 \QQ e^{t\QQ}
      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
      \notag\\
    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
      + \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+})
             \label{eq:big:yuck:3}                          
\end{align}
Finally let us note, regarding the second term on the right hand side
of \eqref{eq:ass:procs:0}, using again 
\eqref{eq:MP:proj:prop} and \eqref{eq:GI:prop}, \eqref{eq:GI:com}
\begin{align}
  -te^{t\QQ}\JJ 
  =-te^{t\QQ}\QQ \QQ^+ \JJ  - t (\II - \QQ^+\QQ) \JJ.
               \label{eq:small:yuck:4}                          
\end{align}

By combining \eqref{eq:big:yuck:1}--\eqref{eq:small:yuck:4} and
comparing with \eqref{eq:ass:procs:0}, we now find that
\begin{align}\label{e:E(t)=integ+sum}
  \Z(t)\! =\! \! \int_0^t \! \! \! e^{(t-s)\QQ} &\QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds
           + (\II - \!\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}
           \notag\\
           &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ 
           \notag\\
           &-\QQ^+ \JJ (\II - \QQ \QQ^{+})- (\II -\QQ^+ \QQ) \JJ \QQ^{+}
             \notag\\
         &-t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}.
\end{align}
Rearranging \eqref{e:E(t)=integ+sum}, we have 
\begin{align*}
\QQ^+ \JJ (\II - \QQ \QQ^{+}) &+(\II -\QQ^+ \QQ) \JJ \QQ^{+}
   +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
                                \notag\\
  &+  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ
  = \TT_1 + \TT_2,
\end{align*}
where 
\begin{align*}
 \TT_1 :=&  \int_0^t  \! \! e^{(t-s)\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds,\\
  \TT_2 :=&  (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}\\
         &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ.
\end{align*}

We estimate each of $\TT_1$ and $\TT_2$ in turn.  Regarding $\TT_1$,
using once more \eqref{eq:GI:com}, as well as \eqref{eq:qetq:exp:decay}, we have
\begin{align*}
  \| \TT_1\| \leq&
    \int_0^t \|\QQ e^{(t-s)\QQ}\| \| \QQ^+  \JJ \QQ^{+} \|\| \QQ e^{s\QQ}\|ds
         \notag\\
   \leq& C_0^2\int_0^t  e^{-(t-s)\kappa} \| \QQ^+  \JJ \QQ^{+} \| e^{-s\kappa } ds
                         \notag\\
   =& C_0^2 t e^{-t\kappa} \| \QQ^+  \JJ  \QQ^{+} \|.
\end{align*}
Turning to $\TT_2$, by \eqref{eq:qetq:exp:decay} we have
\begin{align*}
  \| \TT_2\|
  \leq C_0 (&\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
               + \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
               \notag\\
               &+ t\|\QQ^+ \hspace{0.25mm}\JJ\|) e^{- t \kappa }.
\end{align*}
Combining these two bound completes the proof.

\section{Proof of \cref{thm:bride:of:frankinstein}}

This result is established from \cref{thm:ass:error} and some basic
properties $\QQ$ under the given assumptions. First, note that, cf.
\eqref{eq:ev:d:cond:sym:case}, (\ref{eq:EV:real:ass})
\begin{equation}\label{e:max_|lambda-j(Q)|=<|lambda-1(Q)|}
  |\lambda_d(\QQ)| \leq \max_{j =1, \ldots,d_{-}} |\lambda_{j}(\QQ)| \leq |\lambda_1(\QQ)|.
\end{equation}
Referring back to
\eqref{eq:frob:norm}, by \eqref{e:max_|lambda-j(Q)|=<|lambda-1(Q)|} we have the estimate
\begin{align}\label{eq:frob:sucks}
  \| \QQ e^{t\QQ} \|_F
  =&  \left( \sum_{j =1}^{d_{-}}
     | \lambda_j(\QQ) e^{ t  \lambda_j(\QQ)  } |^2 \right)^{1/2}
     \notag\\
  \leq& \sqrt{d_{-}} |\lambda_{1}(\QQ)| e^{-t |\lambda_{d_{-}}(\QQ)|}.
\end{align}
Similarly, from \eqref{eq:op:norm} and \eqref{e:max_|lambda-j(Q)|=<|lambda-1(Q)|}, we have that
\begin{align}\label{eq:op:better}
  \| \QQ e^{t\QQ} \|_{\op}
  \leq& \max_{j =1, \ldots,d_{-}}
        |\lambda_{j}(\QQ)| e^{-t |\lambda_{j}(\QQ)|}
        \notag\\
  \leq& |\lambda_{1}(\QQ)| e^{-t |\lambda_{d_{-}}(\QQ)|}.
\end{align}
Likewise, we observe that
\begin{align}\label{eq:thm1:extra:trms:1:F}
  \lVert \QQ\rVert^2_F = \sum_{k=1}^{d} \lambda_k(\QQ)^2 \, , \quad
  \lVert \QQ^+ \rVert^2_F = \sum_{k =1}^{d_-}\frac{1}{\lambda_k(\QQ)^2} \, ,
\end{align}
and that
\begin{align}\label{eq:thm1:extra:trms:1:F}
  \lVert \QQ\rVert^2_{\op} = \lambda_1(\QQ)^2 \, , \quad
  \lVert \QQ^+ \rVert^2_{\op} = \frac{1}{\lambda_{d_{-}}(\QQ)^2} \,.
\end{align}
Let $\UU$ be as in \eqref{eq:MP:inv:sym:case}. Noting, furthermore,
that $\II - \QQ \QQ^+ = \UU(\II - \II_{d_-})\UU^*$, where $\II_{d_-}$
is the matrix with $1$'s along the first $d_-$ diagonal elements and
zero otherwise, we have the bounds
\begin{align}\label{eq:proj:mat:bnds}
  \|\II - \QQ  \QQ^+\|_{F} = \sqrt{d - d_{-}},
  \quad
  \| \II - \QQ  \QQ^+\|_{\op} = 1.
\end{align}

Now, for any matrix norm $\|\cdot\|$, \eqref{eq:ass:error} implies
that
\begin{align}
    \| t(\II -\QQ^+ \QQ) &\JJ \QQ \QQ^{+}
                           +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\|
                           \notag\\
  &\leq C(1+ t)e^{-\kappa t}
    + 2\|   \QQ^+\| \|\JJ\| \|\II - \QQ \QQ^{+}\|
    \label{eq:ass:error_proof}
\end{align}    
for any $t \geq 0$, where $C > 0$ is given by
\eqref{eq:exp:error:ass:er}. Thus, for the choice of norm
$\|\cdot\| = \|\cdot\|_{F}$, by \eqref{eq:ass:error_proof},
\eqref{eq:frob:sucks}, \eqref{eq:thm1:extra:trms:1:F} and
\eqref{eq:proj:mat:bnds}, we obtain
\eqref{eq:bride:of:frankinstein:1}.  Similarly, for the choice of norm
$\|\cdot\| = \|\cdot\|_{\op}$, \eqref{eq:ass:error_proof}, in
conjunction with \eqref{eq:op:better}, \eqref{eq:thm1:extra:trms:1:F}
and \eqref{eq:proj:mat:bnds}, yields the bound
\eqref{eq:bride:of:frankinstein}.  Since
\eqref{eq:bride:of:frankinstein:1:rm:app} follows immediately from
\eqref{eq:bride:of:frankinstein:1}, and similarly between
\eqref{eq:bride:of:frankinstein:0:rm:app} and
\eqref{eq:bride:of:frankinstein}, the proof is now complete.

\section{Proof of \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}}
\label{sec:Thm:rand:mat:main}

For use in this section, we introduce a probabilistic version of the
typical $O$ and $o$ asymptotic notations.  Given collections of random
variable $\{X_d\}_{d \in \bbN}$, $\{Y_d\}_{d \in \bbN}$, we write
\begin{equation}\label{e:X=O_a.s.(1)_Y=o_a.s.(1)}
X_d = O_{\as}(f(d)),\quad Y_d  = o_{\as}(f(d)), 
\end{equation}
for some $f: \bbN \to \RR^+$ to mean, respectively, that there exists
a random variable $C = C(\omega)$, not dependent on $d$, such
that $|X_d|/f(d) \leq C$ for all $d \in \bbN$ a.s., and also that
$\lim_{d \rightarrow \infty}Y_d/f(d) = 0$ a.s.

In everything that follows we will make use of the so-called
\textit{Weyl's inequalities} (see e.g.
\cite[Theorem 4.3.1, p.\ 239]{horn:johnson:2012}). Namely, let
${\mathbf A},{\mathbf B} \in {\mathcal S}(d,\bbC)$, each with its own
eigenvalues listed in increasing order as in our convention
\eqref{eq:EV:real:ass}. Then, for $i = 1,\hdots,d$,
\begin{equation}\label{e:Weyl_upper}
  \lambda_i({\mathbf A}+{\mathbf B}) \leq
  \lambda_{i+j}({\mathbf A}) + \lambda_{d-j}({\mathbf B}),
  \quad j = 0,1,\hdots,d-i,
\end{equation}
and
\begin{equation}\label{e:Weyl_lower}
  \lambda_{i-j+1}({\mathbf A}) + \lambda_j({\mathbf B})
  \leq \lambda_i({\mathbf A}+{\mathbf B}), \quad j = 1,\hdots,i.
\end{equation}
In what follows, we also make use of the fact that, for ${\mathbf A} \in {\mathcal S}(d,\bbC)$,
\begin{equation}\label{e:eigenvals_of_-A}
  \lambda_\ell(-{\mathbf A}) = - \lambda_{d-\ell+1}({\mathbf A}),
  \quad \ell = 1,\hdots,d.
\end{equation}

With these preliminaries now in hand, we turn to the proof of
\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.  This result is
established with the aide of four auxiliary results, \cref{l:||Q||/n=O(1/n^(1/2-eta)},
\cref{l:max_Yn/n^eps}, \cref{l:||Q_4||=o_P(1)} and \cref{l:lambda-2(Q3)>=1+1/n}, whose
precise statements and proof are provided immediately afterward.
%in \cref{sec:aux}.
\begin{proof}
  To demonstrate our desired result,
  \eqref{e:lambda_n-1(Q)->-infty_non-symm}, it suffices to establish
  the lower bound
  \begin{equation}\label{e:sigma^2_2(Q)>=_a.s._bound}
    \sigma^2_2(\QQ) \geq
    (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
    + O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)\Big\}
  \end{equation}
  as well as the upper bound
  \begin{equation}\label{e:sigma^2_2(Q)=<_a.s._bound}
    \sigma^2_d(\QQ) \leq
    (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
    + O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)\Big\}.
  \end{equation}
  With this in mind, we now decompose $\QQ$ as follows, starting with
  the case \eqref{e:RM_theorem_general_condition}.  Fix
  $d \in \bbN \backslash\{1\}$.  Taking $F_X$ to be the distribution
  defining the elements in \eqref{e:RM_theorem_general_condition}, draw
  \begin{align}\label{eq:diag:fill:trk}
    \{\widetilde{q}_{ii}\}_{i=1,\hdots,d}
    \stackrel{\textnormal{iid}}\sim F_X
    \text{ independently of } \{q_{ij}\}_{i,j=1,\hdots,d}.
   \end{align} 
   Here note carefully that $\widetilde{q}_{ii} \stackrel{d}= q_{ij}$
   and $\widetilde{q}_{ii} \stackrel{d}\neq q_{ii}$.  Now recast
   $\QQ = \{q_{ij}\}_{i,j}$ as %{ \scriptsize
\begin{align}
  & \begin{pmatrix}
0 & q_{12}-\mu & \hdots & q_{1d}-\mu \\
q_{21}-\mu & 0 & \hdots & q_{2d}-\mu \\
\vdots & \vdots & \ddots & \vdots \\
q_{d1}-\mu & q_{d2}-\mu & \hdots & 0 \\
\end{pmatrix}
+
\begin{pmatrix}
q_{11} & \mu & \hdots & \mu \\
\mu & q_{22} & \hdots & \mu \\
\vdots & \vdots & \ddots & \vdots \\
\mu & \mu & \hdots & q_{dd} \\
\end{pmatrix} \notag \\
  &=\left\{\!
    \begin{pmatrix}
\widetilde{q}_{11}-\mu & q_{12}-\mu & \hdots & q_{1d}-\mu \\
q_{21}-\mu & \widetilde{q}_{22}-\mu & \hdots & q_{2d}-\mu \\
\vdots & \vdots & \ddots & \vdots \\
q_{d1}-\mu & q_{d2}-\mu & \hdots & \widetilde{q}_{dd}-\mu \\
\end{pmatrix}\right.
  \label{eq:En:or:Wigner}\\
& \qquad \quad \quad +
\left.\begin{pmatrix}
-\widetilde{q}_{11}+\mu & 0 & \hdots & 0 \\
0 & -\widetilde{q}_{22}+\mu & \hdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \hdots & -\widetilde{q}_{dd}+\mu \\
\end{pmatrix}\!\right\}
  \notag \\
  &\quad\;+ (1-d)
    \left\{ \!
    \left(\! \!\!\!
      \begin{array}{*{24}{c@{\hspace{2pt}}}}
\frac{-q_{11}-(d-1)\mu}{d-1} & 0 & \hdots & 0 \\
0 & \frac{-q_{22}-(d-1)\mu}{d-1} & \hdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \hdots & \frac{-q_{dd}-(d-1)\mu}{d-1}\\
    \end{array}
  \! \!  \right)
  \right.
  \notag\\ 
  &\qquad \qquad \qquad \quad+  \left. 
        \begin{pmatrix}
\mu & \frac{-\mu}{d-1} & \hdots & \frac{-\mu}{d-1} \\
\frac{-\mu}{d-1} & \mu & \hdots & \frac{-\mu}{d-1} \\
\vdots & \vdots & \ddots & \vdots \\
\frac{-\mu}{d-1} & \frac{-\mu}{d-1} & \hdots & \mu \\
\end{pmatrix}
  \!\right\}
  \label{e:Q_basic_decomp}\\
  &=: \QQ_1 + \QQ_2 + (1-d) \big\{\QQ_3 +  \QQ_4\big\}
  =: \RM + (1-d) \QQ_4  \label{e:Q1,Q2,Q3}
\end{align}

Working from \eqref{e:Q1,Q2,Q3} we start by establishing the first
bound \eqref{e:sigma^2_2(Q)>=_a.s._bound} under
\eqref{e:RM_theorem_general_condition}. Noting from
\cref{l:lambda-2(Q3)>=1+1/n} that $\QQ_4$ is symmetric and positive,
making use of \eqref{e:Weyl_lower}, first with $i =2, j =1$, then with
$i=1, j =1$, invoking \eqref{e:eigenvals_of_-A} and finally using that
$\RM\RM^*$ is symmetric and non-negative, we find
\begin{align}
  &\sigma^2_2(\QQ) = \lambda_2(\QQ\QQ^*)
  \notag\\
 &\geq (d-1)^2 \Big\{ \lambda_{2}(\QQ^2_4) + \lambda_{1}\Big(\frac{\RM\RM^*}{(d-1)^2}
   - \frac{\QQ_4\RM^*+ \RM\QQ^*_4}{d-1} \Big) \Big\}
  \notag\\
 &\geq (d-1)^2 \Big\{ \lambda_{2}(\QQ^2_4) + \lambda_{1}\Big(\frac{\RM\RM^*}{(d-1)^2}\Big)
   +\lambda_{1} \Big(\frac{\QQ_4\RM^*+ \RM\QQ^*_4}{1-d} \Big) \Big\}
  \notag\\
 &\geq (d-1)^2 \Big\{ \lambda_{2}(\QQ^2_4) 
   -\lambda_{d} \Big(\frac{\QQ_4\RM^*+ \RM\QQ^*_4}{d-1} \Big) \Big\}.
   \label{eq:sig:2:lower:setup}
\end{align}
Now observe
\begin{align}
  |\lambda_j(\AAA)| \leq \sigma_d(\AAA) = \| \AAA\|_{op} \quad
  \text{ for
  any $j =1,\ldots, d$},
  \label{eq:dumb:op:norm:ob}
\end{align}
whenever $\AAA$ is symmetric.
Furthermore,
\begin{equation}\label{e:lambda_2(Q^2_3)}
\lambda_{2}(\QQ^2_4) = \lambda^2_{2}(\QQ_4) = \frac{\mu^2d^2}{(d-1)^2},
\end{equation}
where the equality follows from
\eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n} in
\cref{l:lambda-2(Q3)>=1+1/n}.  Combining \eqref{eq:sig:2:lower:setup},
\eqref{eq:dumb:op:norm:ob} and \eqref{e:lambda_2(Q^2_3)}, we therefore
infer
\begin{equation}\label{e:sigma^2_2(Q)>=_bound}
  \sigma^2_2(\QQ) \geq (d-1)^2
  \Big\{ \Big( \frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2 
- \frac{1}{d-1} \|\QQ_4\RM^*+ \RM\QQ^*_4\|_{\op}\Big\}.
\end{equation}

However, by  \eqref{e:||Q||/n=O(1/n^(1/2-eta)} in \cref{l:||Q||/n=O(1/n^(1/2-eta)},  \eqref{e:oas:Zd:con} in \cref{l:max_Yn/n^eps}
and \eqref{e:||Q_4||=o_P(1)} from \cref{l:||Q_4||=o_P(1)}
we have
\begin{align}
  \frac{d}{d-1}  \frac{\|\RM\|_{\op}}{d}
  \leq& \frac{d}{d-1} \left(\frac{\|\QQ_1\|_{\op}}{d} +
  \frac{\|\QQ_2\|_{\op}}{d}\right) + \|\QQ_3\|_{\op}
  \notag\\
       =& O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big).
        \label{e:||Q-cal/n||=oP(1)}
\end{align}
Hence, by \eqref{e:||Q-cal/n||=oP(1)} and
\eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n} in
\cref{l:lambda-2(Q3)>=1+1/n},
\begin{align}
  \frac{1}{d-1} &\|\QQ_4\RM^*+ \RM\QQ^*_4\|_{\op}
  \leq \frac{2d}{d-1} \hspace{0.5mm} \|\QQ_4 \|_{\op} \Big\|\frac{\RM}{d} \Big\|_{\op}
  \notag\\
 &=  O(1) \cdot O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
  = O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big).
  \label{e:(1/n)||Q3_Q*-cal+Q-cal_Q3||}
\end{align}
Thus, by \eqref{e:sigma^2_2(Q)>=_bound}, \eqref{e:||Q-cal/n||=oP(1)} and \eqref{e:(1/n)||Q3_Q*-cal+Q-cal_Q3||}, the lower bound \eqref{e:sigma^2_2(Q)>=_a.s._bound} holds.

We now turn to verify \eqref{e:sigma^2_2(Q)=<_a.s._bound}. Here, by
\eqref{e:Q1,Q2,Q3}, \eqref{e:Weyl_upper} with $j = 0, i = d$, 
\eqref{e:lambda_2(Q^2_3)} and finally \eqref{eq:dumb:op:norm:ob},
\begin{align*}
  &\sigma^2_d(\QQ)
  = \lambda_d(\QQ \QQ^*)
  \notag\\
   &\leq (d-1)^2 \Big\{ \lambda_{d}(\QQ^2_4)
     + \lambda_{d}\Big(\frac{{\mathbf R}{\mathbf R}^*}{(d-1)^2}\Big) +
     \lambda_d\Big(\frac{\QQ_4{\mathbf R}^*+ {\mathbf R}\QQ^*_4}{1-d} \Big) \Big\}
     \notag\\
   % &\leq (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
   %   + \lambda_{d}\Big(\frac{{\mathbf R}{\mathbf R}^*}{(d-1)^2}\Big) +
   %    \lambda_d\Big(\frac{\QQ_4{\mathbf R}^*+ {\mathbf R}\QQ^*_4}{1-d} \Big) \Big\}     
   %   \notag\\
  &\leq (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
    + \Big\|  \frac{{\mathbf R}}{d-1}\Big\|^2_{\op}
    + \frac{1}{d-1} \|\QQ_4{\mathbf R}^*+ {\mathbf R}\QQ^*_4\|_{\op}\Big\}.
\end{align*}
The upper bound \eqref{e:sigma^2_2(Q)=<_a.s._bound} now follows from
\eqref{e:||Q-cal/n||=oP(1)} and \eqref{e:(1/n)||Q3_Q*-cal+Q-cal_Q3||}.
Thus, as a consequence of \eqref{e:sigma^2_2(Q)>=_a.s._bound} and
\eqref{e:sigma^2_2(Q)=<_a.s._bound},
\eqref{e:lambda_n-1(Q)->-infty_non-symm} holds under condition
\eqref{e:RM_theorem_general_condition}, as claimed.  We have
established the desired result in the first case \eqref{e:RM_theorem_general_condition}.

Now suppose that condition \eqref{e:RM_theorem_symmetry_condition}
holds. In order to establish \eqref{e:lambda_n-1(Q)->-infty_non-symm},
it suffices to replace $q_{ij}$, $i < j$, with $q_{ji}$ in expression
\eqref{eq:En:or:Wigner} and in the definition of $\QQ_3$.  We then
follow the rest of the argument for proving
\eqref{e:lambda_n-1(Q)->-infty_non-symm} under condition
\eqref{e:RM_theorem_general_condition} noting that
\cref{l:||Q||/n=O(1/n^(1/2-eta)}, \cref{l:||Q_4||=o_P(1)} also hold in
the symmetric case.  This concludes the proof.
\end{proof}

% \section{Auxiliary results}
% \label{sec:aux}

We turn now to the Lemmata \ref{l:||Q||/n=O(1/n^(1/2-eta)},
\ref{l:max_Yn/n^eps}, \ref{l:||Q_4||=o_P(1)} and
\ref{l:lambda-2(Q3)>=1+1/n}, which are used in the proof of
\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.  These results
correspond to the bounds we use for each of elements in the
decomposition \eqref{e:Q1,Q2,Q3}. We start off with
\cref{l:||Q||/n=O(1/n^(1/2-eta)} which essentially packages results
found in \cite{yin:bai:krishnaiah:1988}, \cite{bai:1999}.

\begin{lemma}\label{l:||Q||/n=O(1/n^(1/2-eta)}
  Let $\QQ_1 = \QQ_1(d)$, $d \in \bbN \setminus \{1\}$, be the
  sequence of random matrices defined as in \eqref{eq:En:or:Wigner}.
  Here we suppose that the off diagonal elements $q_{ij}$ defining
  $\QQ_1$ are either as in \eqref{e:RM_theorem_general_condition} or
  as in \eqref{e:RM_theorem_symmetry_condition} and that the diagonal
  elements $\widetilde{q}_{ii}$ are drawn as in
  \eqref{eq:diag:fill:trk}.  Then, in either case, we have
  \begin{equation}\label{e:||Q||/n=O(1/n^(1/2-eta)}
    \frac{\|\QQ_1\|_{\op}}{d} = O_{\as}\Big( \frac{1}{\sqrt{d}}\Big).
  \end{equation}
\end{lemma}
\begin{proof}
  To establish \eqref{e:||Q||/n=O(1/n^(1/2-eta)} in the first case,
  \eqref{e:RM_theorem_general_condition}, we note that
  $d^{-1}\QQ_1 \QQ^*_1$ forms a sample covariance matrix, where each
  entry of $\QQ_1$ is centered and has infinitely many moments; cf.
  \cref{rmk:exp:mom}. Thus, by \cite[Theorem 3.1, p.\
  517]{yin:bai:krishnaiah:1988},
\begin{equation}\label{e:|Q-tilde_1|=sqrt(n)OP(1)}
  \|\QQ_1\|_{\op} = \sqrt{d}\cdot \sqrt{\lambda_{d}\Big(\frac{\QQ_1 \QQ^*_1}{d}\Big)}
  = \sqrt{d}\cdot \sqrt{4 \sigma^2 + o_{\as}(1)},
\end{equation}
which immediately yields \eqref{e:||Q||/n=O(1/n^(1/2-eta)}.

In the second case, where all the off-diagonal elements are determined
starting from \eqref{e:RM_theorem_symmetry_condition}, $\QQ_1$ is a
centered (mean-zero) Wigner matrix in the typical nomenclature
followed by \cite{bai:1999}.  Thus, noting \eqref{eq:op:norm} and
that, as in the previous case, $\QQ_1$ has infinitely many moments
\eqref{e:||Q||/n=O(1/n^(1/2-eta)} is thus a direct consequence of
\cite[Theorem 2.12, p.\ 630]{bai:1999}.
\end{proof}

We next state and establish \cref{l:max_Yn/n^eps}. Note that the
assumption of independence among the random variables is not needed
for this result.  Here and below in what follows we adopt the useful
notational convention that $o_{\textnormal{a.s.}}(1)$ denotes a random
variable that vanishes almost surely, as $d\to \infty$ (cf.\
\eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}).
\begin{lemma}\label{l:max_Yn/n^eps}
  Let $\{Y_{d}\}_{d \in \bbN}$ be an identically distributed sequence
  of sub-exponential variables with mean $\mu$;
  cf. \cref{def:sub:exp:RV}. Let
  \begin{align}
    Z_{d} = \max_{k=1,\hdots,d} |Y_k - \mu|, d \in \bbN.
    \label{eq:max:sub:exp}
  \end{align} 
  Then, for any fixed $\varepsilon_0 > 0$, 
  \begin{align}\label{e:oas:Zd:con}
   \frac{Z_d}{d^{\varepsilon_0}} = o_{\as}(1).
  \end{align}
   In particular, for the sequence of diagonal matrices
   $\QQ_2 \equiv \QQ_2(d)$ given as in \eqref{e:Q1,Q2,Q3}, and any
   $\varepsilon_0 > 0$,
\begin{align}
  \frac{\|\QQ_2\|_{op}}{d^{\varepsilon_0}} = o_{\as}(1).
  \label{e:QQ2=oP(1)}
\end{align}
\end{lemma}
\begin{proof}
For any $M > 0$,
\begin{align*}
  \bbP\Big(\frac{Z_d}{d^{\varepsilon_0}} \geq \frac{M}{d^{\varepsilon_0/2}}\Big)
  &= \bbP\Big(\bigcup^{d}_{k=1}\Big\{\frac{|Y_k - \mu|}{d^{\varepsilon_0}}
  \geq \frac{M}{d^{\varepsilon_0/2}}\Big\}\Big) \\
  &\leq d \cdot\bbP\big(|Y_1 - \mu| \geq d^{\varepsilon_0/2}M\big) 
 \leq  2d\cdot e^{- \frac{d^{\varepsilon_0/2}M}{K}}.
\end{align*}
In the first and second inequalities, respectively, we use the
assumption that the random variables $\{Y_{d}\}_{d \in \bbN}$ are
identically distributed as well as relation \eqref{e:def_subexp_RV}
for some suitable value $K > 0$. Therefore,
$\sum^{\infty}_{d=1}\bbP(Z_d/d^{\varepsilon_0/2} \geq
M/d^{\varepsilon_0/2})< \infty.$ Since
$\lim_{d \rightarrow \infty}M/d^{\varepsilon_0/2} = 0$,
\eqref{e:oas:Zd:con} now follows as a consequence of the
Borel-Cantelli lemma.

Regarding the second claim observe that, cf. \eqref{eq:diag:fill:trk},
$\QQ_2$ is a diagonal matrix containing identically distributed
sub-exponential random variables.  This mean that, taking
$Y_k :=\tilde{q}_{kk}$, $\|\QQ_2\|$ is of the form
\eqref{eq:max:sub:exp}. As such the second claim \eqref{e:QQ2=oP(1)}
now follows from the first \eqref{e:oas:Zd:con}, completing the proof.
\end{proof}

Turning to our bounds on $\|\QQ_3\|_{op}$, we have the following lemma.
\begin{lemma}\label{l:||Q_4||=o_P(1)} Let $\QQ_3 \equiv \QQ_3(d)$,
  $d \in \bbN$, be a sequence of random matrices as in
  \eqref{e:Q1,Q2,Q3}. Here we assume that the $q_{ii}$ are determined
  either according to \eqref{e:RM_theorem_general_condition} or
  according to \eqref{e:RM_theorem_symmetry_condition}. Then, in
  either of these cases,
    \begin{equation}\label{e:||Q_4||=o_P(1)}
      \|\QQ_3\|_{\op} = O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big).
    \end{equation}
\end{lemma}
\begin{proof}
  Under either \eqref{e:RM_theorem_general_condition} or
  \eqref{e:RM_theorem_symmetry_condition}, each of the entries along
  the main diagonal is a renormalized sums of iid sub-exponential
  random variables i.e.
  \begin{align}
    \frac{-q_{ii} - \mu(d-1)}{d-1}
    = \frac{1}{d-1} \sum_{j \not =i}^d (q_{ij} - \mu)
  \end{align}
  While these diagonal elements are not independent under
  \eqref{e:RM_theorem_symmetry_condition} note carefully
  that we do not use the independence of the rows of $\QQ_3$
  in the arguments that follow.

  Now, by Bernstein's inequality,
  \cite[p.\ 29, Theorem 2.8.1]{vershynin:2018}, there exists a
  constant $C > 0$ such that, for any $\varepsilon > 0$,
  \GD{\begin{align}
        \bbP&\big(\big|-q_{ii}-\mu(d-1)\big| > \varepsilon\big)\notag\\
            &\leq 2 \cdot \exp\Big\{ - C \min\Big\{
              \frac{\varepsilon^2}{(d-1)\|X_\mu\|^{2}_{\psi_1}},
              \frac{\varepsilon}{\|X_\mu\|_{\psi_1}} \Big\}\Big\},
    \label{e:P(max|-qii-n|/n>eps)}
\end{align}
for $i = 1, \ldots, d$ where $X_{\mu}:= X - \mu$ for $X \sim F_X$ and
the right-hand side of \eqref{e:P(max|-qii-n|/n>eps)} involves the
sub-exponential norm \eqref{e:sub-exp_norm}. Fix $\delta > 0$, and let
$\eta:= (2 + \delta)/C$ where $C$ is the constant in the upper
bound in \eqref{e:P(max|-qii-n|/n>eps)}. Hence, for all
$d \in \bbN \backslash \{1\}$,
\begin{align}\label{e:P(|Q3|>eps)}
  \bbP&\Big( \sqrt{\frac{d-1}{\eta \log d}}
        \hspace{0.5mm}\|\QQ_3(d)\|_{\op}
        > \|X_\mu\|_{\psi_1}\Big)\notag\\
      &= \bbP\Big(
        \max_{i=1,\hdots,d}\Big|\frac{-q_{ii}-\mu(d-1)}{d-1}\Big|
        > \|X_\mu\|_{\psi_1}\sqrt{\frac{\eta \log d}{d-1}}\Big) \notag\\
      &= \bbP\Big(\bigcup^{d}_{i=1}\big\{\big|-q_{ii}-\mu(d-1)\big|
        > \|X_\mu\|_{\psi_1}\sqrt{\eta (d-1)\log d}\big\}\Big) \notag \\ 
      &\leq \sum^{d}_{i=1}\bbP\Big(\big|-q_{ii}-\mu(d-1)\big|
        > \|X_\mu\|_{\psi_1}\sqrt{\eta (d-1)\log d}\Big) \notag \\
      & \leq  2d \cdot \exp\big\{ -  C \min\big\{ \eta  \log d ,
        \sqrt{ \eta (d-1) \log d } \big\}\big\}.
\end{align}}
In the second inequality in \eqref{e:P(|Q3|>eps)},
we use \eqref{e:P(max|-qii-n|/n>eps)} with
$\varepsilon := \|X_\mu\|_{\psi_1} \sqrt{\eta (d-1)\log d}$
and the fact that $\{q_{ii}\}_{i=1,\hdots,d}$ are identically
(but not necessarily independently) distributed. Therefore,
for every $d$ sufficiently large,
\begin{align*}
\bbP\Big( \sqrt{\frac{d-1}{\eta \log d}}\|\QQ_3(d)\|_{\op}
> \|X_\mu\|_{\psi_1}\Big)\leq  \frac{2}{d^{1 + \delta}},
\end{align*}
and as a consequence,
\begin{align*}
\sum^{\infty}_{d=1}\bbP\big(\|\QQ_3(d)\|_{\op} >
  \|X_\mu\|_{\psi_1}\sqrt{\frac{\eta \log d}{d-1}}\big)< \infty.
\end{align*}
Hence, by the Borel-Cantelli lemma,
\begin{align*}
  \bbP\Big(\|\QQ_3(d)\|_{\op} >
  \|X_\mu\|_{\psi_1}\sqrt{\frac{\eta \log d}{d-1}}
  \textnormal{ i.o.}\Big)  = 0,
\end{align*}
so that the relation \eqref{e:||Q_4||=o_P(1)} holds. The
proof is complete.
\end{proof}

Finally we conclude with \cref{l:lambda-2(Q3)>=1+1/n} as follows.
\begin{lemma}\label{l:lambda-2(Q3)>=1+1/n}
  Let $\QQ_4 \equiv \QQ_4(d)$, $d \in \bbN$, be the sequence of
  symmetric matrices defined in \eqref{e:Q1,Q2,Q3}. Then, for any
  $d \in \bbN \backslash\{1\}$,
\begin{equation}\label{e:lambda-1(Q3)>=1/n}
  \lambda_1(\QQ_4) = 0,
\end{equation}
and
\begin{equation}\label{e:lambda-ell(Q3)=1+1/n,ell=2,...,n}
  \lambda_\ell(\QQ_4) = \GD{\frac{\mu \hspace{0.5mm} d}{d-1} }, \quad
  \text{ for } \ell = 2,\hdots,d.
\end{equation}
\end{lemma}
\begin{proof}
  The statement \eqref{e:lambda-1(Q3)>=1/n} is a consequence of the
  fact that
  $\QQ_4 \hspace{0.5mm}(1,\hdots,1)^\top = {\mathbf 0} \in \bbR^n$. To
  establish \eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n}, it suffices to
  prove that
\begin{equation}\label{e:lambda-2(Q3)>=1+1/n}
  \frac{d}{d-1} \leq \GD{\frac{\lambda_2(\QQ_4)}{\mu}
    \leq \frac{\lambda_d(\QQ_4)}{\mu}} \leq \frac{d}{d-1},
  \quad d \in \bbN \backslash\{1\}.
\end{equation}
Let ${\mathbf 1} \in {\mathcal S}(d,\bbR)$ be a matrix of
ones and recast
\begin{equation}\label{e:Q3/mu}
\frac{1}{\mu}\QQ_4 = \II + \frac{1}{d-1}(\II-{\mathbf 1}). 
\end{equation}
By \eqref{e:Q3/mu}, \eqref{e:Weyl_lower} with $i = 2, j=1$
and \eqref{e:eigenvals_of_-A},
\begin{align}
  \frac{\lambda_2(\QQ_4)}{\mu}
  &\geq \lambda_{1}(\II) + \lambda_{2}\Big(\frac{1}{d-1}(\II-{\mathbf 1})\Big)
    \notag\\
  &= \lambda_{1}(\II) - \frac{1}{d-1}\lambda_{d-1}({\mathbf 1}-\II).
    \label{e:lambda-2(Q3)>=_Weyl_bound}
\end{align}
However, by \eqref{e:Weyl_upper}, in the case $i = d-1, j = 1$,
\begin{equation}\label{e:lambda_n-1(1-I)=-1}
  \lambda_{d-1}({\mathbf 1}-\II)
  \leq \lambda_{d-1}({\mathbf 1}) + \lambda_d(-\II) = -1,
\end{equation}
where we used that
\begin{equation}\label{e:rank(1)=1}
  \lambda_{1}(\mathbf{1}) = \cdots = \lambda_{d-1}({\mathbf 1}) = 0.
\end{equation}
This latter claim follows immediately from the fact that
$\textnormal{rank}({\mathbf 1}) = 1$ and
$\mathbf{1} \hspace{0.5mm}(1,\hdots,1)^\top= d(1,\hdots,1)^\top$.  The
first inequality in \eqref{e:lambda-2(Q3)>=1+1/n} is now a consequence
of \eqref{e:lambda-2(Q3)>=_Weyl_bound} and
\eqref{e:lambda_n-1(1-I)=-1}.

On the other hand, again by \eqref{e:Q3/mu}, \eqref{e:Weyl_upper}, this
time with $i = d, j = 0$, and \eqref{e:eigenvals_of_-A},
\begin{align}
  \GD{\frac{\lambda_d(\QQ_4)}{\mu}}
  &\leq \lambda_d(\II)
         + \lambda_d\Big(\frac{1}{d-1}(\II - {\mathbf 1})\Big)
    \notag\\
  &= \lambda_d(\II) - \lambda_1\Big(\frac{1}{d-1}({\mathbf 1}-\II)\Big).
        \label{e:lambda_n(Q3)=<_upper}
\end{align}
However, by \eqref{e:Weyl_lower} and \eqref{e:rank(1)=1},
\begin{equation}\label{e:lambda_1(1-I)>=_lower}
\lambda_1({\mathbf 1}-\II) \geq \lambda_1({\mathbf 1}) + \lambda_1(-\II)= -1.
\end{equation}
The third inequality in \eqref{e:lambda-2(Q3)>=1+1/n} is now a
consequence of \eqref{e:lambda_n(Q3)=<_upper} and
\eqref{e:lambda_1(1-I)>=_lower}. This establishes
\eqref{e:lambda-2(Q3)>=1+1/n} and, hence,
\eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n}, completing the proof.
\end{proof}

\section{Surrogate-trajectory Hamiltonian Monte Carlo}
\label{sec:Sur:trad:HMC}

\newcommand{\G}{\mathbf{G}}

Hamiltonian Monte Carlo (HMC) \cite{duane1987hybrid,neal2011mcmc} is
an advanced MCMC procedure that uses numerical approximations to
Hamiltonian trajectories to generate Metropolis-type
\cite{metropolis1953equation} proposals far away from the current
Markov chain state.  On the one hand, this approach to proposal
generation helps reduce autocorrelation between chain states and is
particularly helpful within higher-dimensional state spaces.  On the
other hand, numerical integration of Hamilton's equations requires
repeated evaluation of the Hamiltonian potential energy's gradient,
and these repeated floating-point operations may become
computationally burdensome.

Lets briefly recall this HMC approach to resolve a given density
$\pi(\cdot)$ of a `target' probability measure of
interest.\footnote{In the Bayesian inference context, we are typically
  considering the posterior density function
  $\pi(\ttheta):=p(\ttheta|\Y)$ for the parameter
  $\ttheta\in\mathbb{R}^K$ given observed data $\Y$.}  We proceed by
considering a `potential energy' of the form
$U(\ttheta):=-\log \pi(\ttheta)$.  We then select an associated
potential energy $V(\mom) := | \G^{-1/2} \mom|^2$ for an appropriately
chosen symmetric-positive-definite mass matrix $\G$ (which is often
taken as the identity for simplicity).  One then observes that the
Gibbs measure, proportional to $e^{-H(\ttheta, \mom)}$ where
$H = U +V$, is invariant under the associated Hamiltonian dynamic.
Here note that the $\ttheta$ marginal of $e^{-H(\ttheta, \mom)}$
coincides with $\pi$ while the $\mom$ marginal is normally distributed
as $\mathcal{N}(\Zero,\G)$.

One operationalizes these observations as an algorithmic sampling
procedure as follows. At each step, given a current sample
$\ttheta^{(n)}$, one draws $\mom^{(n)} \sim \mathcal{N}(\Zero,\G)$.
From this initial state
$(\ttheta(0), \mom(0)) := (\ttheta^{(n)},\mom^{(n)})$ one then
numerically approximates the associated Hamiltonian dynamics using a
St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator up to a
total integration time $\tau >0$ and using integration step size
$\epsilon > 0$.  In this context, note that a single iteration of this
integrator takes the form, \cite{leimkuhler2004simulating},
\begin{align}
  \mom\left(s + \frac{\epsilon}{2}\right)
  &:= \mom(s) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(s)),  
	 \nonumber \\
  \ttheta(s+\epsilon) &:= \ttheta(s) + \epsilon \, \G^{-1}\mom(s+\frac{\epsilon}{2}), 
		 \label{eq:leap}\\ 
  \mom(s + \epsilon)
  &:= \mom\left(s+\frac{\epsilon}{2}\right)
    + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(s+\epsilon))  \, .
	\nonumber
\end{align}
In this fashion the proposed new state is given through
\eqref{eq:leap} by $\ttheta(\tau)$.  To remove bias this procedure can
be augmented with an acceptance probability of the form
$\alpha^{(n)} := \exp( H( \ttheta(0), \mom(0)) - H(\ttheta(\tau),
\mom(\tau)) \wedge 1$.

Different strategies aim to speed up the leapfrog integrator's many
log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$, as
these numerical routines often represent the algorithm's computational
bottleneck.  In model-specific contexts,
\cite{holbrook2021massive,holbrook2022bayesian,holbrook2022viral}
yield parallelization strategies, and \cite{ji2020gradients} develops
dynamic programming techniques, to accelerate the evaluation of
$\nabla \log \pi (\ttheta)$.

A small body of work considers another approach by replacing
$\nabla \log \pi (\ttheta)$ with a suitable approximation
$\widetilde{\nabla} \log \pi (\ttheta)$ and recognizing that the
modified \eqref{eq:leap} continues to satisfy path reversibility and
volume preservation, two essential ingredients for well-specified
HMC. It follows immediately that the resulting `surrogate trajectory
HMC' continues to sample the correct target distribution $\pi(\cdot)$;
see further details \cite{neal2011mcmc, glatt2020accept}.
Nonetheless, the acceptance rates and overall efficiency of such
samplers may suffer when approximations are poor.

The majority of surrogate HMC methods first obtain a small sample of
exact gradient evaluations and then use some model to interpolate:
\cite{zhang2017precomputing} assumes the approximate gradient follows
a piecewise constant form across a grid;
\cite{rasmussen2003gaussian,lan2016emulation} construct approximations
using Gaussian processes; and \cite{zhang2017hamiltonian,li2019neural}
do the same using neural networks.  But an even simpler approach to
surrogate HMC may be appropriate when gradients have series
representations as we can leverage here in \eqref{eq:grad:series}.

%}
%\showmatmethods{} % Display the Materials and Methods section

\end{document}

### 2023-09-12 manuscript/revision_letter.tex
revisions commit 1

%\documentclass[standard,
%               backaddress=off,
%               foldmarks=false,
%               enlargefirstpage,
%               pagenumber=off,
%               parskip=half
%               ]{scrlttr2}
%
%\setkomavar{fromname}{Gabriel Hassler}
%\setkomavar{fromaddress}{Department of Computational Medicine / UCLA \\ 5303 Life Sciences \\ Box 951766 \\ Los Angeles, CA 90095}
%\setkomavar{fromphone}{(202) 557-5140}
%\setkomavar{fromemail}{ghassler@g.ucla.edu}
%%\setkomavar{department}{Department of Computational Medicine}

\documentclass[12pt]{article}

\usepackage{amssymb}
\usepackage{geometry}
\usepackage{url}
\geometry{letterpaper} % or letterpaper (US) or a5paper or....
\geometry{margin=1in} % for example, change the margins to 2 inches all round
\usepackage[table]{xcolor}

\newenvironment{reply}{$\triangleright$\bfseries}{$\triangleleft$}
\renewenvironment{quote}
               {\list{}{\rightmargin\leftmargin}%
                \item\relax\normalfont}
               {\endlist}

%
%\newcommand{\bibsection}[1]{}
%\newcommand{\section}[1]{}
%\newcommand{\newblock}{}
%
%   \newenvironment{thebibliography}[1]%
%      {References\begin{description}}{\end{description}}
%   \newcommand{\htmlbibitem}[2]{\label{#2}\item[{[#1]}]}
\usepackage{natbib}
\usepackage[colorlinks, citecolor={blue}]{hyperref}
\usepackage{amsmath}
\usepackage{float}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{setspace}
\usepackage{xr}
\usepackage[textsize=tiny]{todonotes}
\usepackage{algorithm,algorithmicx,algpseudocode}
%\usepackage{subfigure}
\setlength{\parindent}{0pt}
\setlength{\parskip}{12pt}

\usepackage{braket}

\usepackage{amsthm}

\definecolor{trevorblue}{rgb}{0.330, 0.484, 0.828}
\definecolor{trevoryellow}{rgb}{0.829, 0.680, 0.306}

\def\suppDoc{pnas_revision}
\externaldocument{\suppDoc}
%\input{missing_traits_supplement_revision.xtr}
%\input{revision_values.sty}
%\input{new_floats.tex}

\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{bm}

\usepackage{todonotes}

\usepackage[format=plain,
labelfont={it,bf},
textfont=it]{caption}

\newcommand{\mb}{\mathbf}
\newcommand{\mc}{\mathcal}
\newcommand{\dx}{\mbox{d}}

\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand*{\fplus}{\genfrac{}{}{0pt}{}{}{+}}
\newcommand*{\fdots}{\genfrac{}{}{0pt}{}{}{\cdots}}

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\numTaxa}{N}
\newcommand{\numTraits}{D}
\newcommand{\numDatasets}{M}
%\newcommand{\numLatent}{D}
\newcommand{\taxonIndex}{i}
\newcommand{\traitIndex}{j}
\newcommand{\traitData}{\vec{Y}}
\newcommand{\traitDatum}{y}
\newcommand{\datasetIndex}{m}
\newcommand{\exemplar}{\text{e}}

\newcommand{\sequences}{\vec{S}}
\newcommand{\latentData}{\vec{X}}
\newcommand{\latentdata}{\vec{z}}
\newcommand{\latentDatum}{x}
\newcommand{\phylogeneticParameters}{\boldsymbol{\phi}}
\newcommand{\phylogeny}{{\cal G}}
\newcommand{\tree}{\phylogeny}
%\newcommand{\otherParameters}{\boldsymbol{\
\newcommand{\transpose}{^{t}}

\newcommand{\distanceMatrix}{\mathbf{Y}}
\newcommand{\distance}{y}
\newcommand{\summant}{r}

\newcommand{\cdensity}[2]{\ensuremath{p(#1 \,|\,#2)}}
\newcommand{\density}[1]{\ensuremath{p(#1 )}}

\newcommand{\treeNode}{\nu}

\newcommand{\traitVariance}{\mathbf{\Sigma}}
\newcommand{\nodeIndex}{c}
%\newcommand{\parent}[1]{\mbox{\tiny pa}(#1)}
\newcommand{\parentBig}[1]{\mbox{pa}(#1)}

\newcommand{\sibling}[1]{\mbox{\tiny sib}(#1)}
\newcommand{\siblingBig}[1]{\mbox{sib}(#1)}

\newcommand{\rootMean}{\boldsymbol{\mu}_0}
\newcommand{\rootVarianceScalar}{\tau_0}
\newcommand{\unsequencedVarianceScalar}{\tau_{\exemplar}}
\newcommand{\treeVariance}{\vec{V}_{\tree}}
\newcommand{\hatTreeVariance}{\hat{\vec{V}}_{\tree}}
\newcommand{\mdsSD}{\sigma}
\newcommand{\mdsVariance}{\mdsSD^2}
\newcommand{\residual}{\hat{\traitDatum}}
\newcommand{\modelDistance}{\delta}
\newcommand{\cdf}{\phi}
\newcommand{\normalCDF}[1]{\Phi \left( #1 \right)}

\newcommand{\order}[1]{{\cal O}\hspace{-0.2em}\left( #1 \right)}

\newcommand{\rootNode}{\nu^{\datasetIndex}_{2 \numTaxa_{\datasetIndex} -1 }}
\newcommand{\pathLength}[1]{d(F, #1 )}
\newcommand{\pathLengthNew}[2]{
	d_{F}
	(
	{#1}, {#2}
	)
}
\newcommand{\J}{\vec{J}}
\newcommand{\pprime}{^{\prime}}
\newcommand{\otherIndex}{i \pprime}
\def\kronecker{\raisebox{1pt}{\ensuremath{\:\otimes\:}}}

\newcommand{\x}{\mathbf{x}}

%\graphicspath{{../../Graphs/}}

\def\journalName{Journal of Multivariate Analysis}
\def\journalAbbr{JMVA}
\def\editor{Hongyu Zhao}
\def\editorTitle{Theory and Methods Co-Editor}
\def\editorLast{Zhao}
\def\pubOrg{American Statistical Association}
\def\pubAddressOne{732 North Washington Street}
\def\pubAddressTwo{Alexandria, VA 22314-1943}

\def\paperTitle{A quantum parallel Markov chain Monte Carlo}
\def\paperID{JCGS-22-275}

%\newcommand{\bigo}[1]{\cal{O}\hspace{-.5mm}\left(#1\right)}
\onehalfspacing

\renewcommand{\x}{\mathbf{x}}
\renewcommand{\v}{\mathbf{v}}
\newcommand{\TTheta}{\boldsymbol{\theta}}
\newcommand{\TTTheta}{\boldsymbol{\Theta}}
\newcommand{\tr}{\mbox{tr}}
\newcommand{\X}{\mathbf{X}}

\usepackage{tikz}
\usetikzlibrary{matrix}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{arrows,shapes}
\usetikzlibrary{decorations}
\usetikzlibrary{decorations.pathmorphing}
\usetikzlibrary{fit}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.geometric}

\newtheorem{thm}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{prop}{Proposition}
\newtheorem{cor}{Corollary}
\newtheorem{remark}{Remark}
\newtheorem{example}{Example}
\newtheorem{mydef}{Definition}
\newtheorem*{assumption}{Assumption}
\newtheorem{clm}{Claim}

\newcommand{\ttheta}{\boldsymbol{\theta}}
\newcommand{\Ttheta}{\boldsymbol{\Theta}}
\newcommand{\dd}{\mbox{d}}
\newcommand{\ppsi}{\boldsymbol{\psi}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\I}{\mathbf{I}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\G}{\mathbf{G}}
\newcommand{\ppi}{\boldsymbol{\pi}}
\newcommand{\llambda}{\boldsymbol{\lambda}}

\begin{document}

\input{pnas_revision.xtr}

\centerline{\large \bf Reponse to Editor and Reviewer Comments}

To the Editor and Reviewers:

I appreciate the thoughtful comments and recommendations made by the AE and Reviewers and the opportunity to revise my manuscript ``\paperTitle" (\paperID).
I have carefully considered these extremely helpful suggestions and have addressed them to the best of my abilities.  I believe that the manuscript is now significantly improved and am thankful for the time and energy put forth by the Editors and Reviewers.

In particular, I have:
\begin{itemize}
	\item clarified how the present work compares to other quantum Monte Carlo methods;

	\item expanded the scope of the methods to include discrete target distributions (Section \ref{sec:discrete});

	\item included an application to Bayesian image analysis (Section \ref{sec:bayesIm});

	\item included a short discussion around challenges reading classical data into quantum machines; and

	\item responded to all other Reviewer/AE questions and comments.
\end{itemize}

The document below contains my replies (in \textbf{bold}) alongside the AE and Reviewers' comments (normal text).  I list changes to the text with indented normal text set off by quotation marks.

With sincere gratitude,

Andrew Holbrook
\clearpage

{\Large \bf Editor and Reviewer Comments:}

\section*{Editor}

Dear Andrew,

Two reviews of your manuscript are somewhat split, but they also focus on different aspects of your paper. So the paper cannot be accepted. Rev 2 addresses the mathematical dimension of your study (as well as the awkward juxtaposition of the mathematical presentation and the data example); Rev 1 seems to focus on details of the covid data but also has some queries about the methods. The paper is well outside anything to do with my expertise, but I can certainly identify with the general points of Rev 2, especially the absence of demonstrated method performance on data where you know the answer.

If you think you can revise to overcome reviewer concerns, you are welcome to do so. Any revision will be subjected to re-review with no prejudice toward acceptance. I think that both reviewers are highly appropriate, and given that their expertise far exceeds mine, I cannot apprise you of whether a successful revision is feasible.

Jim

\section*{Reviewer 1}

Phylogeography inference has been a very difficult problem in the sense of the tree size and the dimen- sionality of states. However, in this manuscript, the authors have shown that, based on the CTMC model, the first-order approximation of the gradient towards the rate matrix can improve the computational efficiency with acceptably bounded errors, which can be hugely confined with correction. The method is applied to SARS-CoV-2 transmission to infer the traveling factors across 33 regions, framed in a mixed-effects model. I am not an expert in Graph Theory and Random Matrix Theory but the manuscript is clear enough for me to understand. I, therefore, am strongly recommending for publication in PNAS, while there are still a few issues that the authors need to be addressed properly:

\begin{enumerate}
	\item The authors spend 3 out of 7 pages in the main text and three corresponding sections in the Appendix discussing the refined error and the proof, while postpone the application with the affine-corrected error to future work. I would expect more than a relatively abstract error comparison figure for this issue. For example, a simple numerical simulation similar to the last paragraph in Application to compute the posterior likelihood ratio or posterior density ratio of the corrected and the uncorrected methods, because the authors eventually applied the method within a Bayesian framework.

	\item I am curious about the selection of the extra two additional Chinese provinces and 11 additional countries when fitting the mixed-effects model and how the inclusion of the extra regions improve the fitting. Is there any criteria to determine the inclusion?

	\item For the mixed-effects model, do the authors assume independent predictors for the fixed effects? It seems to me that, in Fig. 2, the air traffic proximity and the intracontinental proximity are highly correlated, which can be the reason why θ2 is not statistically significant. Then what’s the justification for the predictor selection?

	\item Am I correct that the total number of particles (i.e., samples) in the MCMC framework is 2,000? The ESS for τ is extremely small, do the authors have any explanations?

	\item What does the random effect matrix look like? The authors indicate that air traffic proximity is the only significant fixed-effect predict, then I would wonder how much does the random effect contribute to the model?

	\item The ranges for predictor matrices and posterior mean rate matrix are different. However, the color palettes are identical, which makes it extremely hard to compare the impact of each predictor. For example, I cannot see how the sentence “... but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.” makes sense visually from Fig. 2 and 3. It also doesn’t make sense to me conceptually, because air traffic proximity is the only significant predictor.

	\item 
	The color palette in Fig. 4 makes it really hard to distinguish Hubei, other provinces in China, and other countries. It will be great if the authors can highlight Hubei in one color, other provinces in another color with gradients, and other countries in a third color with gradients. And why there are only 29 regions?

	\item 
	The proposed method is able to include higher-dimension rate matrix compared with \citet{lemey2009bayesian,lemey2014unifying} [1, 2]. Can the authors then have an external comparison to the results in \citet{lemey2020accommodating} just visually to see whether the color pattern in the phylogenetic tree is matched?
\end{enumerate}

\section*{Reviewer 2}

This is a review for the manuscript "On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2". 
For the benefit of the authors on how they interpret my review, I am a theoretical/computational statistician with background on theoretical population genetics, interested in Markov chains and its implementations in Monte Carlo methods.

Before content-specific comments let me first clarify that there are issues with the numbering of theorems and remarks in the manuscript that I have received (e.g., theorem 2 is missing, and there are two remarks labeled as Remark 2), and below I stick to the numbering used in the manuscript that I have received for review to avoid any confusion.

The main idea in the manuscript revolves around the three theorems presented and how the approximations provided by these theorems improve computational efficiency in large matrices involved in the continuous time markov chains. The positives are: 1) The theorems are stated well as in solid applied mathematical work, 2) Their proofs seem to hold as far as I can tell. 
To me it is clear that if there are any gaps or minor incorrect statements they can be remedied with little effort and the theorems and corollaries will hold. It is also clear that a lot of effort and thought was put into these theorems and corollaries and I appreciate that, along with the clear exposition which made following the mathematical reasoning easy.

Unfortunately, the manuscript did not sit well with me, so to the negatives. There are two major negatives for me. The second item is the more important one, but the first motivates the second, and it deeply troubled me.

1) I think the way the manuscript is positioned with respect to the interpretation of the scope of the work is inaccurate. After a first read of the manuscript, I was left with the impression that the work was originally conceived and intended as an applied mathematics or a stochastic process paper, which --for example-- with the omission of opening paragraph of the manuscript would stop before section 2. Presented this way, I think its scope is ideal for a specific technical subject-matter journal (e.g., applied math, stochastic process, statistics). With the inclusion of the first paragraph (which I do not find clear, see below specific comments on this), and section 2, I felt that there was an effort to make the paper appealing to a broader audience. I think this effort falls short of its aim and the integration of theoretical results with data looks ad-hoc. In particular, the example data set chosen to demonstrate the results does not have much information about the parameters, and it is as if it was included just because its dimensionality is large. This tells us nothing about the quality of inference (see 2 below) and also misleads us to believe that the approximation method presented is successful as a statistical computational --in the sense of methods that facilitate inference but they are not inferential methods themselves-- tool. I note that the example consists of a real data set in which we do not know whether the assumed probability model generating the data is true, let alone the true parameter values of that model. Even if there was a signal in the data, I would still be hesitant to accept the approximation method presented as a valid statistical computational tool in the context of MCMC, before it was well established with some simulation studies (again, see 2 below). Therefore, there is no way to assess the quality of inference about parameters with this example. The only use of this data set can be in supporting the claim that the approximation presented through the theorems provides a computational advantage at higher dimensions. But then, why use SARS-Cov-2 data for this goal, whereas it can be better established by a simulated data set?

2) To me, a huge barrier to accept that the approximations presented can be successful in aiding statistical inference is the lack of simulation studies to assess the quality of inference. (This is not about the type of simulations shown in Figure 1 as detailed below). For a work like this, before applying to real data, the first test would be to design a large simulation study, where the model and the parameters are known and to test whether the application of the approximation method when implemented within the Hamiltonian Monte Carlo has good statistical inference performance. MCMC makes its own approximations (mixing, convergence to stationarity, etc..) and errors introduced by an approximative method injected to an MCMC sampling might be exacerbated. Even the statistical performance of well-known MCMC algorithms can vary with different signal strengths (e.g., magnitude of the parameters relative to each other in a model), as well as the model size. It would be premature to think that a method is valid based solely on its computational performance (as it is presented in the manuscript) without first assessing its statistical performance of recovering an approximate joint posterior distribution of parameters and it passes rigorous tests. A way to check this is by simulating data from target distributions of interest under a variety of parameter value combinations and then apply the method to see whether the posterior distribution of parameters are recovered successfully (in the case of Bayesian inference for this manuscript). I am surprised that the authors did not see this as necessary especially because they are working with large mixed effect models where things can go awry pretty quickly. As an example, I would have liked to see a comparison of the statistical performance in say, a 3 x 3 simulation experiment design. Variable 1, The method with 3 levels: Standard Hamiltonian Monte Carlo, the naive method of reference (18), and the method of the manuscript; Variable 2, The model size with 3 levels: Small, Intermediate, Large; Variable 3, The number of parameters with considerable signal with 3 levels: 1 parameter, half of the parameters, all parameters in the model.

Specific Comments:

* Please check numbering of theorems and remarks.

* Regarding the opening paragraph. The first sentence did not make sense to me. What is big data (why is it in double quotes, anyways?), what is modeling schemes, what is structures. The whole paragraph is written vaguely and I felt dismissive of the applications of the work being presented.

* Figure 1 is designed to show how the approximation behaves beyond the assumptions of theorems and corollaries, and support the claim that the results apply more generally by numerical evidence. I am not convinced that this constitutes strong evidence (as it is claimed in the manuscript), however. My feeling is that the claim holds, but it needs to be demonstrated with distributions that are not so docile. I think that deviations from the assumptions of theorems and corollary are not so large for cases a, b, and c. For Figure 1c, I have a specific comment: I am not sure about what the authors mean by "Cauchy entries truncated to be positive", but my understanding is that this is a folded Cauchy distribution at zero with the support of the probability density function defined on the non-negative real line (if this is not correct please add a sentence to describe). Cauchy distribution is a member of the subexponential family and I wonder whether the tail properties of the folded Cauchy are distinct from those of the subexponential family (I do not have an answer to this, so I suggest that the authors check and provide a brief explanation).

* Equations 12-19 seems to follow straightforwardly from the section where equation 2-7 are presented. I agree that writing them explicitly is better, but I felt that they disrupt the flow of the section "Deterministic bounds on approximation error in time". Please consider emulating them to somewhere in 2-7 or following, intermediate steps in 12-19 maybe put in an appendix.

* In presenting the subexponential families in formulae 40-41, notation "exp" is used instead of "e". Please change to "e" to be consistent with the rest of the manuscript. Also, I think this definition comes from Vershynin. There might be some regularity conditions attached to this specific definition, although I cannot remember exactly. Please check. There is a more fundamental definition of subexponential families using the cumulative distribution functions, maybe that serves better here. Your choice of course.

* Remark 6. The sentence starting with "In general..." is not clear in which sense it is general. "Proxy" is also not clear.

* The paragraph starting with "In addition..." on page 7 is too short for the reader to understand the implications of the analysis. Please consider expanding it.

* Appendix D. Surrogate-trajectory Hamiltonian Monte Carlo is described quite detached from the work presented in the manuscript. Please consider describing it in the context of the work.

* For completeness, it is worth restating that Q is symmetric in Theorem 3. (Or is it not?)

* Remark 2 (iii). "eigenvalue" should be "eigenvalues".

* References (5) is a book, but there is no section or page.

%This paper presented an interesting framework for large-scale MCMC. It leverages the quantum speedup of Grover's searching algorithm to accelerate the treatment of multi-proposals in MCMC. However, an implementation of a quantum algorithm involves nontrivial steps, which need to be explained before this paper can be accepted.
%
%\begin{reply}
%	I thank the Reviewer for taking the time to review the manuscript and for their positive reception.
%\end{reply}
%
%The author did not clarify whether this is a fully quantum algorithm, or a hybrid one. It is important to explain what is the input of the quantum computer and what to extract. For the input, one has to encode it coherently into a quantum state $\ket{\psi}$, and estimate the complexity associated with such preparation.
%
%
%
%To extract quantities from a quantum computer, one has to express it as an expectation (observable), and estimate the probability of obtaining the correct output. 
%
%\begin{reply}
%	I have clarified the hybrid nature of the QPMCMC computing scheme.  The first two sentences of the abstract now read:
%	\begin{quote}
%		We propose a novel hybrid quantum computing strategy for parallel MCMC algorithms that generate multiple proposals at each step. This strategy makes the rate-limiting step within parallel MCMC amenable to quantum parallelization by using the Gumbel-max trick to turn the generalized accept-reject step into a discrete optimization problem.
%	\end{quote}
%
%	In response to the Reviewer's important points regarding on-loading and off-loading data to and from the quantum machine, I have added the step $\ket{\theta_p} \leftarrow \ttheta_p$ within Algorithm \ref{alg:qpMCMC}, along with the comment ``\verb|Load proposal onto quantum computer.|''. Algorithm 4, which is nested within Algorithm 6, already addresses extraction.  Perhaps more importantly, I have added the following paragraph to the Discussion (Section \ref{sec:disc}):
%	
%	\begin{quote}
%There are major technical barriers to the practical implementation of QPMCMC.  The framework, like other quantum machine learning (QML) schemes, requires on-loading and off-loading classical data to and from a quantum machine.  In the seminal review of QML, \citet{biamonte2017quantum} discuss what they call the `input problem'.  \citet{cortese2018loading} present a general purpose quantum circuit for loading $B$ classical bits into a quantum data structure comprising $\log_2 (B)$ qubits with circuit depth $\order{\log (B)}$.  For continuous targets, QPMCMC requires reading $\order{P}$ real vectors $\ttheta_p\in \mathbb{R}^D$ onto the quantum machine at each MCMC iteration.  If $b$ is the number of bits used to represent a single real value, then reading $B=\order{PDb}$ classical bits into the quantum machine requires time $\order{\log (PDb)}$.  This is true whether one opts for fixed-point \citep{jordan2005fast} or floating-point \citep{haener2018quantum} representations for real values.  The outlook can be better for discrete distributions.  For example, the QPMCMC scheme presented in Section \ref{sec:disc} only requires loading the entire configuration state once prior to sampling.  A $D$-spin Ising model requires $D$ classical bits to encode, and these bits load onto a quantum machine in time $\order{\log(D)}$.  Once one has encoded the entire system state, each QPMCMC iteration only requires loading the addresses of the $\order{P}$ proposal states. If one uses $b$ bits to encode each address, then the total time required to load data onto the quantum machine is $\order{\log(Pb)}$ for each QPMCMC iteration.  On the other hand, the speedup for discrete targets assumes the ability to hold an entire configuration in QRAM.
%Conveniently, the `output problem' is less of an issue for QPMCMC, as only a single integer $\hat{p} \in \{0,\dots,P\}$ need be extracted within Algorithm \ref{alg:min}.
%	\end{quote}
%\end{reply}
%
%There have already been numerous quantum algorithms for MCMC, e.g.,
%\begin{enumerate}
%	\item Montanaro, A. Quantum Speedup of Monte Carlo Methods. Proc. R. Soc. A. 2015, 471 (2181), 20150301. \citep{montanaro}
%	
%	\item Wocjan, P.; Abeyesinghe, A. Speedup via Quantum Sampling. Phys. Rev. A 2008, 78 (4), 042336. \citep{wocjan2008speedup}
%\end{enumerate}
%The author should comment on how the present approach would compare to these existing methods.
%
%\begin{reply}
%I thank the Reviewer for the helpful suggestion.  To clarify the relationship between the proposed method and other quantum Monte Carlo algorithm, I have included the following paragraph in Section \ref{sec:disc}:
%\begin{quote}
%Second, can QPMCMC be combined with established quantum algorithms that make use of `quantum walks' on graphs to sample from discrete target distributions?  \citet{szegedy2004quantum} presents a quantum analog to classical ergodic reversible Markov chains and shows that such quantum walks sometimes provide quadratic speedups over classical Markov chains.  \citet{szegedy2004quantum} also points out that Grover's search, a key component within QPMCMC, may be interpreted as performing just such a quantum walk on a complete graph.  \citet{wocjan2008speedup} accelerate the quantum walk by using ancillary Markov chains to improve mixing and apply their method to simulated annealing.  Given a quantum algorithm $\mathbb{A}$ for producing a discrete random sample with variance $\sigma^2$, \citet{montanaro} develops a quantum algorithm for estimating the mean of algorithm $\mathbb{A}$'s output with error $\epsilon$ by running algorithm $\mathbb{A}$ a mere $\widetilde{\mathcal{O}}(\sigma/\epsilon)$ times, where $\widetilde{\mathcal{O}}$ hides polylogarithmic terms. Importantly, this quadratic quantum speedup over classical Monte Carlo applies for quantum algorithms $\mathbb{A}$ that only feature a single measurement such as certain simple quantum walk algorithms. Unfortunately, this assumption fails for quantizations of Metropolis-Hastings, in general, and QPMCMC, in particular. More promising for QPMCMC, \citet{lemieux2020efficient} develop a quantum circuit that applies Metropolis-Hastings to the Ising model without the need for oracle calls.  An interesting question is whether similar non-oracular quantum circuits exist for the basic parallel MCMC backbone to QPMCMC.  In general, however, comparison between QPMCMC and other quantum Monte Carlo techniques is challenging because the foregoing literature (\textcolor{red}{a}) largely focuses on MCMC as a tool for discrete optimization, with algorithms that only ever return a single Monte Carlo sample or function thereof, and (\textcolor{red}{b}) restricts itself to a handfull of simple, stylized and discrete target distributions.  On the other hand, QPMCMC is a general inferential framework for sampling from general discrete and continuous distributions alike.
%\end{quote}
%\end{reply}
%
%\begin{reply}
%	Again, I'd like to thank the Reviewer for suggesting the above additions/edits, and I believe the manuscript is now much improved as a result.
%\end{reply}
%
%
%\section*{Reviewer 2}
%
%
%This paper seeks to further develop parallel Markov chain Monte Carlo in the quantum computing hardware. The topic is very interesting. The author has a nice comprehensive review of multi-proposal parallel MCMC, and is very current in literature. The review of established quantum computing algorithms is also very helpful for statistician who might have limited knowledge in the area. 
%
%The key contribution is connecting parallel Markov chain Monte Carlo to the quantum computing. The key step is to employ Gumbel-max trick to transform the generalized accept-reject step (i.e., Dirichlet sampling) into a discrete optimization procedure, so that quantum optimization algorithm can be used for sampling. The idea is simple and elegant. It just connects the dots.
%
%\begin{reply}
%	I thank the Reviewer for their kind words and for taking the time to read the manuscript.
%\end{reply}
%
%I still have a few concerns, more generally related to the “bedrock” parallel Markov chain Monte Carlo:
%
%\begin{enumerate}
%	\item Although named as parallel MCMC, the algorithm is still MCMC, which relies on the back-bone Markov chain to achieve equilibrium. The algorithm still heavily relies on the long sequence for the Markov chain to reach equilibrium. The quantum part doesn’t seem to be able to solve this major drawback of parallel MCMC. 
%\end{enumerate}	
%
%\begin{reply}
%	I absolutely agree with the Reviewer's judgment.   To clarify that my method does not overcome this basic shortcoming of MCMC, I've included the following text in Section \ref{sec:disc}:
%	\begin{quote}
%		While the algorithm still must construct long Markov chains to reach equilibrium, generating each individual Markov chain state requires significantly fewer target evaluations.
%	\end{quote} 
%\end{reply}
%	
%	
%\begin{enumerate}
%	  \setcounter{enumi}{1}
%	\item I see more potential of the Gumbel-max trick $+$ quantum computing in sequential Monte Carlo. Unlike MCMC, SMC only requires a handful of sequential chains, and mostly need to have a very large number of proposals for re-sampling, where your trick could have more impact.
%\end{enumerate}
%
%\begin{reply}
%	This is a good point.  I've added the following to the second-to-last paragraph of Section \ref{sec:disc}:
%	\begin{quote}
%		The trick may also find use within sequential Monte Carlo \citep{doucet2001sequential}.  For example, \citet{berzuini2001resample} present a sequential importance resampling algorithm that uses MCMC-type moves to encourage particle diversity and avoid the need for bootstrap resampling.  Multiproposals accelerated by the quantum Gumbel-max trick could add speed and robustness to such MCMC-resampling.
%	\end{quote}	
%\end{reply}
%
%\begin{enumerate}
%	  \setcounter{enumi}{2}
%	\item A last comment is on computer hardware structure. In the QPMCMC, the Markov chain for loop requires CPU, the evaluation of log-likelihood requires GPU, and the sample from Dirichlet requires quantum processor. Will there be I/O communication overhead for these steps on distinct processors?
%\end{enumerate}
%
%\begin{reply}
%	I thank the Reviewer for raising this possible point of confusion.  The multiproposal is generated on a CPU or GPU.  The quantum computer both (a) evaluates the log-posterior and (b) samples from the Dirichlet.  This is the nice thing about combining the Gumbel-max trick with quantum optimization.  The question about I/O is important, and I have added the following paragraph to Section \ref{sec:disc}:
%	\begin{quote}
%There are major technical barriers to the practical implementation of QPMCMC.  The framework, like other quantum machine learning (QML) schemes, requires on-loading and off-loading classical data to and from a quantum machine.  In the seminal review of QML, \citet{biamonte2017quantum} discuss what they call the `input problem'.  \citet{cortese2018loading} present a general purpose quantum circuit for loading $B$ classical bits into a quantum data structure comprising $\log_2 (B)$ qubits with circuit depth $\order{\log (B)}$.  For continuous targets, QPMCMC requires reading $\order{P}$ real vectors $\ttheta_p\in \mathbb{R}^D$ onto the quantum machine at each MCMC iteration.  If $b$ is the number of bits used to represent a single real value, then reading $B=\order{PDb}$ classical bits into the quantum machine requires time $\order{\log (PDb)}$.  This is true whether one opts for fixed-point \citep{jordan2005fast} or floating-point \citep{haener2018quantum} representations for real values.  The outlook can be better for discrete distributions.  For example, the QPMCMC scheme presented in Section \ref{sec:disc} only requires loading the entire configuration state once prior to sampling.  A $D$-spin Ising model requires $D$ classical bits to encode, and these bits load onto a quantum machine in time $\order{\log(D)}$.  Once one has encoded the entire system state, each QPMCMC iteration only requires loading the addresses of the $\order{P}$ proposal states. If one uses $b$ bits to encode each address, then the total time required to load data onto the quantum machine is $\order{\log(Pb)}$ for each QPMCMC iteration.  On the other hand, the speedup for discrete targets assumes the ability to hold an entire configuration in QRAM.
%Conveniently, the `output problem' is less of an issue for QPMCMC, as only a single integer $\hat{p} \in \{0,\dots,P\}$ need be extracted within Algorithm \ref{alg:min}.
%	\end{quote}
%\end{reply}
%
%
%\begin{reply}
%	Again, I'd like to thank the Reviewer for taking the time to review this manuscript and for their positive reception.  I hope that my edits have addressed their concerns.
%\end{reply}
%
%
%
%\section*{Associate Editor}
%
%
%
%The author proposes a quantum parallel MCMC algorithm that uses a Gumbel-max trick to quantum parallelize a multiple-proposal MCMC algorithm. The resulting algorithm drastically reduces the number evaluations of the target density. The manuscript is quite well written and motivated.
%
%\begin{reply}
%	I thank the AE for their generally positive reception!
%\end{reply}
%
%
%The referees are relatively positive about the work, however, certain improvements are deemed necessary and the points raised by them are important. I highlight a few points here, and have some additional comments.
%
%
%The first referee highlights that more specific details about the quantum algorithm are required. The second referee asks an important question regarding the cost of the I/O communication. I would further add that in the displayed Algorithms, it would be useful to indicate which components are CPU parallelized, GPU parallelized, and quantum parallelized.
%
%\begin{reply}
%I completely agree with the Reviewers' points and have added the following paragraph describing the burden associated with loading classical data onto a quantum machine:
%\begin{quote}
%There are major technical barriers to the practical implementation of QPMCMC.  The framework, like other quantum machine learning (QML) schemes, requires on-loading and off-loading classical data to and from a quantum machine.  In the seminal review of QML, \citet{biamonte2017quantum} discuss what they call the `input problem'.  \citet{cortese2018loading} present a general purpose quantum circuit for loading $B$ classical bits into a quantum data structure comprising $\log_2 (B)$ qubits with circuit depth $\order{\log (B)}$.  For continuous targets, QPMCMC requires reading $\order{P}$ real vectors $\ttheta_p\in \mathbb{R}^D$ onto the quantum machine at each MCMC iteration.  If $b$ is the number of bits used to represent a single real value, then reading $B=\order{PDb}$ classical bits into the quantum machine requires time $\order{\log (PDb)}$.  This is true whether one opts for fixed-point \citep{jordan2005fast} or floating-point \citep{haener2018quantum} representations for real values.  The outlook can be better for discrete distributions.  For example, the QPMCMC scheme presented in Section \ref{sec:disc} only requires loading the entire configuration state once prior to sampling.  A $D$-spin Ising model requires $D$ classical bits to encode, and these bits load onto a quantum machine in time $\order{\log(D)}$.  Once one has encoded the entire system state, each QPMCMC iteration only requires loading the addresses of the $\order{P}$ proposal states. If one uses $b$ bits to encode each address, then the total time required to load data onto the quantum machine is $\order{\log(Pb)}$ for each QPMCMC iteration.  On the other hand, the speedup for discrete targets assumes the ability to hold an entire configuration in QRAM.
%Conveniently, the `output problem' is less of an issue for QPMCMC, as only a single integer $\hat{p} \in \{0,\dots,P\}$ need be extracted within Algorithm \ref{alg:min}.
%\end{quote}
%I have also expanded the third paragraph of Section \ref{sec:intro} to help clarify the nature of parallelization within the proposed framework:
%\begin{quote}
%	One need not use parallel computing to implement Algorithm \ref{alg:pMCMC}, but the real promise and power of parallel MCMC comes from its natural parallelizability \citep{calderhead2014general}. 
%	Contemporary hardware design emphasizes architectures that enable execution of multiple mathematical operations simultaneously. Parallel MCMC techniques stand to leverage technological developments that keep modern computation on track with Moore's Law, which predicts that processing power doubles every two years.  For example, the algorithm of \citet{tjelmeland2004using} generates $P$ conditionally independent proposals and then evaluates the probabilities of \eqref{eq:probs}.  One may parallelize the proposal generation step using parallel pseudorandom number generators (PRNG) such as those advanced in \citet{salmon2011parallel}. The computational complexity of the target evaluations $\pi(\ttheta_p)$ is linear in the number of proposals. This presents a significant burden when $\pi(\cdot)$ is computationally expensive, e.g., in big data settings or for Bayesian inversion, but evaluation of the target density over the $P$ proposals is again a naturally parallelizable task.  Moreover, widely available machine learning software such as \textsc{TensorFlow} allows users to easily parallelize both random number generation and target evaluations on general purpose graphics processing units (GPU) \citep{lao2020tfp}. Finally, when generating independent proposals using a proposal distribution of the form $q(\ttheta_0,\Ttheta_{-0})=\prod_{p=1}^Pq(\ttheta_0,\ttheta_{p})$, the acceptance probabilities \eqref{eq:probs} require the $\order{P^2}$ evaluation of the $P+1\choose 2$ terms $q(\ttheta_{p},\ttheta_{p'})$, but \citet{massive,holbrook2021scalable} demonstrate the natural parallelizability of such pairwise operations, obtaining orders-of-magnitude speedups with contemporary GPUs.  The proposed method directly addresses the acceptance step of Algorithm \ref{alg:pMCMC}, while leaving the choice of parallelizing (or not parallelizing) the proposal step to the practitioner.  
%\end{quote}
%\end{reply}
%
%
%The author has motivated the need for quantum algorithms for big-data problems where evaluation of the target is expensive. However, only data-less examples have been implemented. Implementation on a real example is warranted. Further, comparisons with other quantum methods (as suggested by referee) would be useful to understand the effectiveness of the proposed methodology. Finally, comparison with the non-quantum version of the multiple-proposal would be helpful to make the point that the mixing of the algorithm remains unaffected.
%
%\setcounter{figure}{7}
% \begin{figure}[!t]
%	\centering
%	\includegraphics[width=0.9\linewidth]{figures/blackHole.png}
%	\caption{On the left is a 4,076-by-4,076 intensity map of the shadow of supermassive black hole Sagittarius A*  \citep{akiyama2022first}.  On the right is the pixelwise posterior mode of a Bayesian image classification model fit to intensity data.  Within a Metropolis-in-Gibbs routine, quantum parallel MCMC using 1,024 proposals requires less that one-tenth the posterior evaluations required by conventional parallel MCMC.}\label{fig:blackHole}
%\end{figure}
%
%\begin{reply}
%	I thank the AE for these helpful suggestions. In response, I have added Section \ref{sec:bayesIm}, in which I apply a Bayesian image classification model to a recently released 4000$\times$4000 image of the supermassive black hole Sag A*.  For this target, QPMCMC scores a 10-fold speedup over parallel MCMC:
%	\begin{quote}
%		We apply a Bayesian image classification model to an intensity map (Figure \ref{fig:blackHole}) of the newly imaged supermassive black hole, Sagittarius A*, located at the Galactic Center of the Milky Way \citep{akiyama2022first}.  Whereas one cannot see the black hole itself, one may see the shadow of the black hole cast by the hot, swirling cloud of gas surrounding it.  We extract the black hole from the surrounding cloud by modeling the intensity at each of the $D=$4,076$^2=$16,613,776 pixels as belonging to a mixture of two truncated normals with values $y_{d}$ restricted to the intensity range $[0,255]$.  Namely, we follow \citet{hurn1997difficulties} and specify the latent variable model
%		\begin{align*}
%			y_d |(\mu_\ell, \sigma^2, \theta_d) &\stackrel{ind}{\sim} \mbox{Normal}(\mu_\ell,\sigma^2 ) \, ,\quad y_d \in [0,255] \, , \quad \theta_d=\ell \,,\quad d\in \{1,\dots,D\}\, ,\\
%			\mu_\ell &\stackrel{iid}{\sim} \mbox{Uniform} (0,255) \,  , \quad \ell \in \{-1,1\} \, , \\
%			\frac{1}{\sigma^2} &\sim  \mbox{Gamma}\left(\frac{1}{2}, \frac{1}{2} \right)  \, ,
%		\end{align*}
%		where $\ttheta=(\theta_1,\dots,\theta_D)$ share for a prior the Ising model \eqref{eq:ising} with edges joining neighboring pixels and interaction $\rho=1.2$. 
%		
%		We use a QPMCMC-within-Gibbs scheme to infer the join posterior of $\ttheta$ and the three mixture model parameters.  For the former, we use the same QPMCMC scheme as in Section \ref{sec:ising} with 1,024 proposals at each iteration. For the latter, we use the closed-form updates presented in \citet{hurn1997difficulties}.  We run this scheme for 20 million iterations, discarding the first 10 million as burnin.  We thin the remaining sample at a ratio of 1 to 40,000 for the latent memberships $\ttheta$ and 1 to 4,000 for the three parameters $\mu_{-1}$, $\mu_1$ and $\sigma^2$.  Using the \textsc{R} package \textsc{coda} \citep{coda}, we calculate effective sample sizes of the log-posterior (257.1), $\mu_{-1}$ (1,578.7), $\mu_1$ (257.6) and $\sigma^2$ (2,500.0), suggesting adequate convergence.   Figure \ref{fig:blackHole} shows both the intensity data and the pixelwise posterior mode of the latent membership vector $\ttheta$.  The QPMCMC requires only 1,977,553,608 target evaluations compared to the 1,024 $\times$ 20,000,000 $=$ 2.048$\times10^{10}$ evaluations required for the analogous parallel MCMC scheme implemented on a conventional computer, representing a 10.36-fold speedup.
%	\end{quote}
%	I have also added the following paragraph to Section \ref{sec:disc} comparing QPMCMC with other quantum methodologies:
%	\begin{quote}
%Second, can QPMCMC be combined with established quantum algorithms that make use of `quantum walks' on graphs to sample from discrete target distributions?  \citet{szegedy2004quantum} presents a quantum analog to classical ergodic reversible Markov chains and shows that such quantum walks sometimes provide quadratic speedups over classical Markov chains.  \citet{szegedy2004quantum} also points out that Grover's search, a key component within QPMCMC, may be interpreted as performing just such a quantum walk on a complete graph.  \citet{wocjan2008speedup} accelerate the quantum walk by using ancillary Markov chains to improve mixing and apply their method to simulated annealing.  Given a quantum algorithm $\mathbb{A}$ for producing a discrete random sample with variance $\sigma^2$, \citet{montanaro} develops a quantum algorithm for estimating the mean of algorithm $\mathbb{A}$'s output with error $\epsilon$ by running algorithm $\mathbb{A}$ a mere $\widetilde{\mathcal{O}}(\sigma/\epsilon)$ times, where $\widetilde{\mathcal{O}}$ hides polylogarithmic terms. Importantly, this quadratic quantum speedup over classical Monte Carlo applies for quantum algorithms $\mathbb{A}$ that only feature a single measurement such as certain simple quantum walk algorithms. Unfortunately, this assumption fails for quantizations of Metropolis-Hastings, in general, and QPMCMC, in particular. More promising for QPMCMC, \citet{lemieux2020efficient} develop a quantum circuit that applies Metropolis-Hastings to the Ising model without the need for oracle calls.  An interesting question is whether similar non-oracular quantum circuits exist for the basic parallel MCMC backbone to QPMCMC.  In general, however, comparison between QPMCMC and other quantum Monte Carlo techniques is challenging because the foregoing literature (\textcolor{red}{a}) largely focuses on MCMC as a tool for discrete optimization, with algorithms that only ever return a single Monte Carlo sample or function thereof, and (\textcolor{red}{b}) restricts itself to a handfull of simple, stylized and discrete target distributions.  On the other hand, QPMCMC is a general inferential framework for sampling from general discrete and continuous distributions alike.
%	\end{quote}
%Finally, I've added Section \ref{sec:mixing}, which compares mixing between conventional parallel MCMC and QPMCMC:
%\begin{quote}
%	To ascertain whether QPMCMC mixes differently compared to conventional parallel MCMC, we run both algorithms for a range of proposal counts to sample from a 10-dimensional standard normal distribution. For each algorithm and proposal setting, we run 100 independent chains for 10,000 iterations and obtain effective sample sizes ESS$_d$ for $d \in \{1,\dots,10\}$.   We then calculate the relative differences between the means and minima of one chain generated using parallel MCMC and one chain generated using QPMCMC; for example:
%	\begin{align*}
%		\mbox{Relative difference between means} \: \overline{\mbox{ESS}}_{(\cdot)} = \frac{\left|\overline{\mbox{ESS}}_{pMCMC} - \overline{\mbox{ESS}}_{QPMCMC}\right| }{\overline{\mbox{ESS}}_{pMCMC} }
%	\end{align*}
%	Figure \ref{fig:mixing} shows results.  In general, the majority relative differences are small.  For both statistics, mean relative differences are less than 0.05, regardless of proposal count.  Again for both statistics, more than 75\% of the independent runs result in relative differences below 0.1.  We note that relative differences between means (blue) appear to decrease with the number of proposals, but the same does not hold for relative differences between minima (red).
%\end{quote}
%\end{reply}
%
% \begin{figure}[!t]
%	\centering
%	\includegraphics[width=0.7\linewidth]{figures/pMCMCvsQPMCMC.pdf}
%	\caption{Relative differences between effective sample sizes (ESS) for parallel MCMC (pMCMC) and quantum parallel MCMC (QPMCMC) across a range of proposal counts. We target a 10-dimensional standard normal distribution.  For each algorithm and each proposal setting, we generate 100 independent chains of length 10,000 and calculate the mean and minimum ESS across dimensions.}\label{fig:mixing} 
%\end{figure}
%
%
%\subsection*{Additional Comments}
%
%
%\begin{itemize}
%	\item (I am assuming the following is true.) It would be helpful to state clearly at some point that the resulting QPMCMC chain has the same transition kernel as the non-quantum version. That is, it remains exact.
%\end{itemize}	
%
%\begin{reply}
%	Sometimes Algorithm \ref{alg:min} does not find the minimum. For example, in the experiments described in Figures \ref{fig:mcmcIterations} and \ref{fig:qq}, the correct kernel is used 99.4\% of the time.  Because of this the QPMCMC chain does not have the same transition kernel as the non-quantum version once every 200 QPMCMC iterations.  Nonetheless, results like those in Figure \ref{fig:qq} show that bias remains small.  The second paragraph of Section \ref{sec:disc} reads:
%	\begin{quote}
%This work leads to three interesting questions.   First, what is the status of inference obtained by QPMCMC?  The QPMCMC selection step relies on quantum minimization, an algorithm that only achieves success with probability $1-\epsilon$.  While empirical studies suggest that this error induces little bias, it would be helpful to use this $\epsilon$ to bound the distance between the target distribution and the distribution generated by QPMCMC.  Such theoretical efforts would need to extend recent formalizations of the multiproposal based parallel MCMC paradigm \citep{glatt} to account for biased MCMC kernels.
%	\end{quote}
%I am currently working with collaborators to bound the Wasserstein distance between the target and the distribution actually sampled.  At the moment, I must leave such results for future work. 
%\end{reply}
%	
%\begin{itemize}	
%	\item Equation (1): It is not clear to me how the ``a.s.'' applies to this definition of 1
%	stationarity.
%\end{itemize}
%
%\begin{reply}
%	I thank the AE for catching this silly mistake, which  I have fixed.
%\end{reply}
%
%\begin{itemize}
%	
%	\item Algorithm 1: The algorithm of Tjelmeland (2004) is described here as ``Parallel MCMC'', but it is not immediately clear which part of the algorithm will be implemented in a parallelizable way. I would recommend changing the caption.
%	
%\end{itemize}
%
%\begin{reply}
%	This caption uses the term `parallel MCMC' in the same way that it is used in \citet{calderhead2014general}, \citet{schwedes2019parallel}, \citet{schwedes2021rao} and \citet{glatt}, i.e., to indicate the use of multiple proposals.  To help resolve the potential confusion, I have changed the caption to `Parallel (multiproposal) MCMC' and added the following text to Section \ref{sec:intro}:  
%	\begin{quote}
%		One need not use parallel computing to implement Algorithm \ref{alg:pMCMC}, but the real promise and power of parallel MCMC comes from its natural parallelizability \citep{calderhead2014general}. 
%	\end{quote}
%\end{reply}
%
%\begin{itemize}
%
%	\item page 4, line 10: I believe the citation for \citet{gelman1992inference} is incorrect. Did the author mean \citet{gelman1992single}?
%
%\end{itemize}
%
%\begin{reply}
%	I really did mean \citet{gelman1992inference}, but I have now changed the text to cite both papers following the AE's helpful recommendation.  All I'm saying here is that there are different kinds of parallelization available, and that computational statisticians should engage the entire spectrum.  
%\end{reply}
%
%\begin{itemize}
%	
%	\item Algorithm 5: Should the last step be ``return $\hat{p}$''.
%	
%\end{itemize}
%
%\begin{reply}
%	I thank the AE for catching this, and I have changed the text accordingly.
%\end{reply}
%
%\begin{itemize}
%	
%	\item Section 3.1: The proof of the Gumbel-max trick can likely be either removed or
%	moved to the Appendix, since this is a fairly well-known result.
%
%\end{itemize}
%
%\begin{reply}
%	I have moved the proof to Section \ref{sec:gm} of the Appendix.
%\end{reply}
%
%\begin{itemize}	
%	
%	\item Algorithm 6: I would recommend explicitly highlighting which parts of this algorithm are quantum parallelized.
%
%\end{itemize}
%
%\begin{reply}
%	I have highlighted the lines accordingly.  Also, I have edited the relevant text to read:
%	\begin{quote}
%		Algorithm \ref{alg:qpMCMC} presents the details of QPMCMC for a continuously-valued target.  The algorithm uses conventional (perhaps parallel) computing almost entirely, excluding two lines that are highlighted.  The first of these loads proposals onto the quantum machine, and the second line that calls the quantum minimization algorithm presented in Algorithm \ref{alg:min}. 
%	\end{quote}
%\end{reply}
%
%\begin{itemize}
%	
%	\item Figure 5: The caption here is confusing. It reads ``Total number of oracle evaluations required for each of 2000 parallel MCMC iterations''. Traditionally, this would imply there were 2000 independent chains runs. However, the text describing this Figure implies that there were 2000 Gaussian proposals.
%	
%\end{itemize}
%
%\begin{reply}
%	I thank the AE for pointing out this error on my part.  In fact, the first line should have included `quantum parallel MCMC' instead of just `parallel MCMC'.
%	
%	\begin{quote}
%		Total number of oracle evaluations required for each of 2,000 quantum parallel MCMC (QPMCMC) iterations for standard multivariate normal targets of five different dimensionalities. Regardless of target dimension, the individual QPMCMC runs require roughly $7$\% of the usual 4 million target evaluations. Over  99.4\% of the 10,000 MCMC iterations across all dimensions successfully sample from the discrete distribution with probabilities of \eqref{eq:probs}. Burn-in iterations require moderately more evaluations because the current state occupies a lower density region and represents a `less good' warm-start.
%	\end{quote}
%\end{reply}
%
%\begin{itemize}
%	
%	\item page 22: The authors tune to 50\% acceptance rate, but I believe the optimal in high- dimensions from the reference \citet{rosenthal2011optimal} is 23.4\%. Further, the tuning to 23.4\% acceptance rate is strictly for the Metropolis-Hastings acceptance function. In the multiple-proposal world, the acceptance function employed is completely different, implying different optimality results. See \citet{agrawal2021optimal}.
%\end{itemize}
%
%\begin{reply}
%	I completely agree with the AE on this one.  The reference to \citet{rosenthal2011optimal} was for the `standard adaptation procedure', and not the 50\% tuning target.  I use this target because it is close to the roughly 60\% target I found optimal in my paper about another multiproposal technique \citep{holbrook2021generating} and because there are no theoretical optimality results in the multiproposal setting that I am aware of.  I've edited the text to make this clear:
%	\begin{quote}
%		We fix the number of Gaussian proposals to be 2,000 and use standard adaptation procedures \citep{rosenthal2011optimal} to target a 50\% acceptance rate while guaranteeing convergence. Although no theoretical results validate this target acceptance rate, the rate is close to empirically optimal rates from \citet{holbrook2021generating}.
%	\end{quote}
%\end{reply}
%
%\begin{itemize}	
%	
%	\item Figure 5: ``Over 99.4\% of the 10,000 MCMC iterations successfully sampled from the discrete distribution with probabilities of Equation (2)''. It’s unclear exactly how this number comes about and what it means. Once the quantum minimization returns a $\hat{p}$, the authors have shown that it follows the right distribution. How can then (1) .6\% not be from the right distribution, and (2) how does one even quantify this.
%\end{itemize}
%
%\begin{reply}
%For that and other related examples, I found that roughly 1 in 200 implementations of Algorithm 4 failed to find the minimum, i.e., failed to use the theoretically justified kernel.  We can make this failure happen less often by running Algorithm 4 longer, but there is a tradeoff between accuracy and target evaluations. Empirical results like those of Figure \ref{fig:qq} show that bias is relatively small, but I am currently working with collaborators to better understand how intermittent failure leads to biased sampling.  
%
%I quantify these things by running Monte Carlo simulations of all nested quantum algorithms beginning with the Grover iteration as basic building block.  This is implemented in R as described in Section \ref{sec:sim}.
%\end{reply}
%
%\begin{itemize}	
%	\item Figure 6: Considering the target is known, burn-in seems incredibly wasteful. Once can just start from stationarity and remove the need to throw away samples.
%\end{itemize}
%
%\begin{reply}
%	I include burnin here in order to demonstrate that warm-starting Algorithm 4 with the current Markov chain state becomes more beneficial as one converges the the target's high density regions. This explains why more target evaluations are required earlier on in the trajectory.
%\end{reply}
%
%\begin{itemize}	
%	\item Table 1: In producing this table, what effective sample size is being used for this two-dimensional problem?
%\end{itemize}
%
%\begin{reply}
%	I thank the AE for pointing out this omission. Here, I am using the minimum ESS between the two dimensions.  I  have edited the first sentence of the caption to say:
%	\begin{quote}
%		Racing to a minimum (between the two dimensions) of 100 effective samples for a target with 1,000 disjoint modes. 
%	\end{quote}
% In case this was a software question, I have also added information on ESS to the first paragraph of Section \ref{sec:sim}:
% \begin{quote}
%We use \textsc{R} \citep{rlang}, \textsc{Python} \citep{vanrossum1995python}, \textsc{TensorFlow} \citep{abadi2016tensorflow} and  \textsc{TensorFlow Probability MCMC} \citep{lao2020tfp} in our simulations and the \textsc{ggplot2} package to generate all figures \citep{ggplot}.   In \textsc{R}, we use the package \textsc{coda} \citep{coda} to calculate effective sample sizes.  In \textsc{Python}, we use a built-in function from \textsc{TensorFlow Probability MCMC}.  The primary color palette comes from \citet{wes}.
% \end{quote}
%\end{reply}
%
%\begin{itemize}	
%	\item Could the authors comment on the potential applicability of the algorithm on discrete state spaces? This may be useful for Bayesian record linkage, and variable selection problems \citep{zanella2020informed}.
%\end{itemize}
%
%\begin{reply}
%	This is a great question. I thank the AE for inspiring some additions to this manuscript.  I have added experiments in which I used QPMCMC to sample from an Ising model and a Bayesian classification model.  Figure \ref{fig:ising2d} shows benefits for using many proposals within QPMCMC and that speedups increase with the number of proposals!
%\end{reply}
%
%\setcounter{figure}{6}
% \begin{figure}[!t]
%	\centering
%	\includegraphics[width=0.7\linewidth]{figures/Ising2dFig.pdf}
%	\caption{Convergence of log-posterior for a parallel MCMC sampler targeting an Ising model with uniform interaction $\rho=1$ and no external field. All chains begin at minimal probability `checkerboard' state. Larger proposal counts allow discovery of higher probability configurations.}\label{fig:ising2d}
%\end{figure}
%
%\begin{reply}
%	Again, I thank the AE for all their helpful suggestions, and I am confident that the manuscript is now significantly improved as a result.
%\end{reply}

%\clearpage
\bibliographystyle{sysbio}
\bibliography{refs}

\end{document}

### 2023-09-12 manuscript/revision_letter.tex
revision letter started

@@
 
-\def\paperTitle{A quantum parallel Markov chain Monte Carlo}
-\def\paperID{JCGS-22-275}
+\def\paperTitle{On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2}
+\def\paperID{MS\# 2023-10809}
 
@@
 
-I appreciate the thoughtful comments and recommendations made by the AE and Reviewers and the opportunity to revise my manuscript ``\paperTitle" (\paperID).
-I have carefully considered these extremely helpful suggestions and have addressed them to the best of my abilities.
-I believe that the manuscript is now significantly improved and am thankful for the time and energy put forth by the Editors and Reviewers.
+On behalf of my coauthors, I thank you for your thoughtful recommendations and the opportunity to revise our manuscript ``\paperTitle" (\paperID).
+We have carefully considered your helpful suggestions and have addressed them to the best of our abilities.
+We believe that the manuscript is now significantly improved and are thankful for the time and energy you have put forth.
 
-In particular, I have:
+In particular, we have:
 \begin{itemize}
-	\item clarified how the present work compares to other quantum Monte Carlo methods;
+	\item clarified how the present work complements its companion paper \citet{magee2023random}, which presents scalability and simulation studies;
 
-	\item expanded the scope of the methods to include discrete target distributions (Section \ref{sec:discrete});
+	\item implemented an additional simulation study which considers accuracy for multiple algorithms, different model sizes and different signal strength regimes;
 
-	\item included an application to Bayesian image analysis (Section \ref{sec:bayesIm});
+    \item compared scalability of surrogate-trajectory HMC with the naive and affine-corrected approximations;
 
-	\item included a short discussion around challenges reading classical data into quantum machines; and
+    \item improved upon the SARS-CoV-2 application, relating it to previously established results;
 
-	\item responded to all other Reviewer/AE questions and comments.
+	\item responded to all other Editor/Reviewer questions and comments.
 \end{itemize}
 
-The document below contains my replies (in \textbf{bold}) alongside the AE and Reviewers' comments (normal text).  I list changes to the text with indented normal text set off by quotation marks.
+The document below contains my replies (in \textbf{bold}) alongside the Editor's and Reviewers' comments (normal text).
+I list changes to the text with indented normal text set off by quotation marks.
 
@@
 
+\begin{reply}
+	Thank you for the opportunity to submit a revised manuscript.
+We have made ample additions to the manuscript in response to the suggestions made by both reviewers.
+Among the many improvements we have made to this manuscript, the simulation study suggested by Reviewer 2 helps the paper smoothly transition between theory and application.
+\end{reply}
+
 \section*{Reviewer 1}
 
-Phylogeography inference has been a very difficult problem in the sense of the tree size and the dimen- sionality of states.
+Phylogeography inference has been a very difficult problem in the sense of the tree size and the dimensionality of states.
 However, in this manuscript, the authors have shown that, based on the CTMC model, the first-order approximation of the gradient towards the rate matrix can improve the computational efficiency with acceptably bounded errors, which can be hugely confined with correction.
@@
 I am not an expert in Graph Theory and Random Matrix Theory but the manuscript is clear enough for me to understand.
-I, therefore, am strongly recommending for publication in PNAS, while there are still a few issues that the authors need to be addressed properly:
+I, therefore, am strongly recommending for publication in PNAS, while there are still a few issues that the authors need to be addressed properly.
+
+\begin{reply}
+	Thank you for taking the time to review our manuscript and for your positive reception of our work.
+We have carefully addressed each of your helpful recommendations in our revised manuscript, which is now greatly improved as a result.
+Please find our responses listed below.
+\end{reply}
 
@@
 For example, a simple numerical simulation similar to the last paragraph in Application to compute the posterior likelihood ratio or posterior density ratio of the corrected and the uncorrected methods, because the authors eventually applied the method within a Bayesian framework.
+\end{enumerate}
 
+\begin{reply}
+	content...
+\end{reply}
+
+\begin{enumerate}	
+	  \setcounter{enumi}{1}
 	\item I am curious about the selection of the extra two additional Chinese provinces and 11 additional countries when fitting the mixed-effects model and how the inclusion of the extra regions improve the fitting.
 Is there any criteria to determine the inclusion?
+\end{enumerate}
+
+\begin{reply}
+content...
+\end{reply}
+
+\begin{enumerate}	
+\setcounter{enumi}{2}
 
@@
 Then what’s the justification for the predictor selection?
+\end{enumerate}
 
+\begin{reply}
+content...
+\end{reply}
+
+\begin{enumerate}	
+\setcounter{enumi}{3}	
 	\item Am I correct that the total number of particles (i.e., samples) in the MCMC framework is 2,000? The ESS for τ is extremely small, do the authors have any explanations?
+\end{enumerate}
 
+\begin{reply}
+content...
+\end{reply}
+
+\begin{enumerate}	
+\setcounter{enumi}{4}	
 	\item What does the random effect matrix look like?
 The authors indicate that air traffic proximity is the only significant fixed-effect predict, then I would wonder how much does the random effect contribute to the model?
+\end{enumerate}
 
+\begin{reply}
+content...
+\end{reply}
+
+\begin{enumerate}	
+\setcounter{enumi}{5}	
 	\item The ranges for predictor matrices and posterior mean rate matrix are different.
@@
 It also doesn’t make sense to me conceptually, because air traffic proximity is the only significant predictor.
+\end{enumerate}
 
+\begin{reply}
+content...
+\end{reply}
+
+\begin{enumerate}	
+\setcounter{enumi}{6}	
 	\item 
@@
 And why there are only 29 regions?
+\end{enumerate}
 
+\begin{reply}
+content...
+\end{reply}
+
+\begin{enumerate}	
+\setcounter{enumi}{7}	
 	\item 

### 2023-09-12 manuscript/revision_letter.tex
more revisions

@@
 
+\newcommand{\abs}[1]{\left|#1\right|}
+\newcommand{\norm}[1]{\left\|#1\right\|}
+\newcommand{\GD}[1]{#1}
+\newcommand{\GDcomment}[2][purple]{GD: \textcolor{#1}{#2}}
+
+\newcommand{\y}{\mathbf{y}}
+\newcommand{\QQ}{\mathbf{Q}}
+\newcommand{\LLambda}{\boldsymbol{\Lambda}}
+\newcommand{\M}{\mathbf{M}}
+\newcommand{\JJ}{\mathbf{J}}
+\newcommand{\II}{\mathbf{I}}
+\newcommand{\AAA}{\mathbf{A}}
+
+\newcommand{\Zero}{\boldsymbol{0}}
 \newcommand{\ttheta}{\boldsymbol{\theta}}
-\newcommand{\Ttheta}{\boldsymbol{\Theta}}
-\newcommand{\dd}{\mbox{d}}
-\newcommand{\ppsi}{\boldsymbol{\psi}}
-\newcommand{\U}{\mathbf{U}}
-\newcommand{\I}{\mathbf{I}}
-\renewcommand{\H}{\mathbf{H}}
-\newcommand{\A}{\mathbf{A}}
+\newcommand{\mom}{\boldsymbol{\xi}}
+\newcommand{\ppi}{\boldsymbol{\pi}}
+\renewcommand{\P}{\mathbf{P}}
+\newcommand{\p}{\mathbf{p}}
+\newcommand{\q}{\mathbf{q}}
 \newcommand{\B}{\mathbf{B}}
-\newcommand{\G}{\mathbf{G}}
-\newcommand{\ppi}{\boldsymbol{\pi}}
-\newcommand{\llambda}{\boldsymbol{\lambda}}
+\newcommand{\Z}{\mathbf{E}}
+\newcommand{\T}{\mathbf{T}}
+\newcommand{\TT}{\mathbf{T}}
+\newcommand{\RM}{\mathbf{R}}
+\newcommand{\UU}{\mathbf{U}}
+\newcommand{\VV}{\mathbf{V}}
+
+\newcommand{\RR}{\mathbb{R}}
+\newcommand{\CC}{\mathbb{C}}
+\newcommand{\NN}{\mathbb{N}}
 
@@
 \begin{itemize}
-	\item clarified how the present work complements its companion paper \citet{magee2023random}, which presents scalability and simulation studies;
+	\item clarified how the present work complements its companion paper \citet{magee2023random}, which presents scaling and simulation studies;
 
-	\item implemented an additional simulation study which considers accuracy for multiple algorithms, different model sizes and different signal strength regimes;
+	\item implemented an additional simulation study which compares accuracy for multiple algorithms, different model sizes and different signal strength regimes;
 
@@
 \begin{reply}
-	content...
+TBD (Simulation related)
 \end{reply}
@@
 \begin{reply}
-content...
+TBD (Simulation related)
+
+Recent work \citep{gotovos2021scaling} suggests that inclusion of auxiliary states can lead to improved model fit even when these additional states do not correspond to observed data.
+In future analyses, we aim to include hundreds of global regions.
+For the time being, we have incorporated this question into the simulation study...
 \end{reply}
@@
 \setcounter{enumi}{2}
-
 	\item For the mixed-effects model, do the authors assume independent predictors for the fixed effects?
 It seems to me that, in Fig.
-2, the air traffic proximity and the intracontinental proximity are highly correlated, which can be the reason why θ2 is not statistically significant.
+2, the air traffic proximity and the intracontinental proximity are highly correlated, which can be the reason why $\theta_2$ is not statistically significant.
 Then what’s the justification for the predictor selection?
@@
 \begin{reply}
-content...
+Multiple regression models (and extensions) do not assume independence between fixed-effect predictors, but the interpretations of the individual regression coefficients do need to be adjusted accordingly.
+We must interpret each association holding all other predictors fixed.
+Yes, intracontinental proximity is no longer statistically significant, but the positive association it does maintain has a stronger interpretation because we are accounting for the role played by air traffic proximity.
+If anything, we would like to include larger numbers of fixed-effect predictors---such as google mobility flows \citep{lemey2021untangling}---in future analyses.
 \end{reply}
@@
 \setcounter{enumi}{3}	
-	\item Am I correct that the total number of particles (i.e., samples) in the MCMC framework is 2,000? The ESS for τ is extremely small, do the authors have any explanations?
+	\item Am I correct that the total number of particles (i.e., samples) in the MCMC framework is 2,000? The ESS for $\tau$ is extremely small, do the authors have any explanations?
 \end{enumerate}
@@
 \begin{reply}
-content...
+We generate 8 million samples but throw away the first 10\% (80,000) as burnin and save only 1 sample out of every 10,000.
+Such thinning does not increase ESS, but it helps with memory constraints.
+After removing burnin and thinning, we have 721 samples.
+The ESS of 185 for $\tau$ is not wonderful but not bad.
+Having fit these Bayesian bridge models in a number of different contexts, we have come to expect $\tau$ to be difficult.
+Hierarchical hyperparameters mix more poorly than those that are close to the data, and $\tau$ is a global hyperparameter that dictates the distribution of a large number of random-effects that stand between it and the data.
 \end{reply}
@@
 \begin{reply}
-content...
+This is what 
 \end{reply}

### 2023-09-12 code/analyzeSARS2.R
random effects imagwe

@@ -266,3 +266,72 @@ system2(command = "pdfcrop",
         args    = c("~/expmDerivative/figures/postViz.pdf",
                     "~/expmDerivative/figures/postViz.pdf")
 )
+
+
+#
+####
+######### visualize random effects for the supplement
+####
+#
+df2 <- df[,12:1903]
+nSamps <- dim(df2)[1]
+randCoeffs <- list()
+
+for(n in 1:nSamps) {
+  k <- 1
+  rcMat <- matrix(0,44,44)
+  for(i in 1:43) {
+    for(j in (i+1):44) {
+      rcMat[i,j] <- unlist(df2[n,k])
+      k <- k+1
+    }
+  }
+  for(i in 1:43) {
+    for(j in (i+1):44) {
+      rcMat[j,i] <- unlist(df2[n,k])
+      k <- k+1
+    }
+  }
+  randCoeffs[[n]] <- rcMat
+}
+
+totalCoeffs <- list()
+for (n in 1:nSamps) {
+  totCoeffMat <- randCoeffs[[n]] +
+    atMat * 0 +
+    hubMat * 0 +
+    contMat * 0
+  
+  totalCoeffs[[n]] <- exp(totCoeffMat)
+  diag(totalCoeffs[[n]]) <- 0
+}
+
+meanMat <- matrix(0,44,44)
+for (n in ceiling(nSamps/10):nSamps) {
+  meanMat <- meanMat + totalCoeffs[[n]]
+}
+meanMat <- meanMat/length(ceiling(nSamps/10):nSamps)
+
+tC <- reshape2::melt(meanMat)
+colnames(tC) <- c("Var1","Var2","Rate")
+gg3 <- ggplot(data = tC, aes(x=Var1, y=Var2, fill=Rate)) + 
+  geom_tile() +
+  guides(x =  guide_axis(angle = 45)) +
+  xlab(NULL) + ylab(NULL) +
+  ggtitle("Posterior mean of random effects") +
+  scale_fill_gradientn(#breaks=c(-1,0,1),
+    colours = tableau_div_gradient_pal("Temperature Diverging")(seq(0, 1, length = 25))) +
+  theme(plot.title = element_text(vjust = -1),
+        axis.text = element_text(size = 5),
+        legend.text = element_text(),
+        legend.margin=margin(0,0,0,0),
+        legend.box.spacing = unit(5, "pt"))
+gg3
+
+ggsave(gg3,
+       width = 5,height = 4,
[…]

### 2023-09-13 manuscript/pnas_revision.tex
more response

@@
 
+\section{Visualizing the posterior mean random effects}
+
+Figure
+
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=0.8\linewidth]{figures/randEffects.pdf}
+	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.
+Notably, we infer a roughly 1.8-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.
+Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}\label{fig:randEffs}
+\end{figure}
+
 %}

### 2023-09-13 manuscript/revision_letter.tex
random effects imagwe
more response

@@
 \end{enumerate}
+
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=0.7\linewidth]{figures/randEffects.pdf}
+	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.
+Notably, we infer a roughly 1.8-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.
+Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}
+\end{figure}
 

### 2023-09-13 code/analyzeSARS2.R

@@ -318,7 +318,7 @@ gg3 <- ggplot(data = tC, aes(x=Var1, y=Var2, fill=Rate)) +
   geom_tile() +
   guides(x =  guide_axis(angle = 45)) +
   xlab(NULL) + ylab(NULL) +
-  ggtitle("Posterior mean of random effects") +
+  ggtitle("Posterior mean of exponentiated random effects") +
   scale_fill_gradientn(#breaks=c(-1,0,1),
     colours = tableau_div_gradient_pal("Temperature Diverging")(seq(0, 1, length = 25))) +
   theme(plot.title = element_text(vjust = -1),
@@ -329,7 +329,7 @@ gg3 <- ggplot(data = tC, aes(x=Var1, y=Var2, fill=Rate)) +
 gg3
 
 ggsave(gg3,
-       width = 5,height = 4,
+       width = 6,height = 4,
        file="figures/randEffects.pdf")
 system2(command = "pdfcrop",
         args    = c("~/expmDerivative/figures/randEffects.pdf",

### 2023-09-13 manuscript/pnas_revision.tex
more revisions

@@
 
-\section{Visualizing the posterior mean random effects}
-
-Figure
+\section{Visualizing the posterior mean random effects}\label{sec:randomEffs}
+
+Figure \ref{fig:randEffs} displays posterior means of the exponentiated random effects, which one may interpret as multiplicative deviations from the fixed effects' contributions to the generator matrix.
+Whereas the vast majority of exponentiated random effects have posterior means close to 1, indicating no deviation from the fixed-effect model, a few exhibit posterior means that are significantly greater than 1.
+In particular, the rate element corresponding to transfer from the US to Hubei, CN, exhibits a 1.81-fold random-effect derived increase to the fixed-effect component.
+This large multiplicative increase agrees with, but goes beyond, the influence that the Hubei asymmetry holds for the entire generator matrix model.
 
@@
 	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.
-Notably, we infer a roughly 1.8-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.
+Notably, we infer a roughly 1.81-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.
 Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}\label{fig:randEffs}

### 2023-09-13 manuscript/revision_letter.tex
more revisions

--- 2023-09-12 manuscript/revision_letter.tex
@@
 
+\setcounter{figure}{4}
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=0.8\linewidth]{figures/randEffects.pdf}
+	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.
+Notably, we infer a roughly 1.81-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.
+Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}%\label{fig:randEffs}
+\end{figure}
+
 \begin{reply}
-This is what 
+This is a good question.
+Figure \ref{fig:randEffs} show posterior means of the exponentiated random effects, which amount to multiplicative deviations from the fixed-effect model.
+As one might expect, only a small number of random effects deviate from 0, but those that do deviate are highly suggestive.
+We stop short of interpreting exactly why the transition from the US to Hubei, CN, exhibits largest deviation (a 1.81-fold increase).
+We have added this figure and the following short discussion to the appendix:
+	\begin{quote}
+		Figure \ref{fig:randEffs} displays posterior means of the exponentiated random effects, which one may interpret as multiplicative deviations from the fixed effects' contributions to the generator matrix.
+Whereas the vast majority of exponentiated random effects have posterior means close to 1, indicating no deviation from the fixed-effect model, a few exhibit posterior means that are significantly greater than 1.
+In particular, the rate element corresponding to transfer from the US to Hubei, CN, exhibits a 1.81-fold random-effect derived increase to the fixed-effect component.
+This large multiplicative increase agrees with, but goes beyond, the influence that the Hubei asymmetry holds for the entire generator matrix model.
+	\end{quote}	
 \end{reply}
@@
 However, the color palettes are identical, which makes it extremely hard to compare the impact of each predictor.
-For example, I cannot see how the sentence “...
-but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.” makes sense visually from Fig.
+For example, I cannot see how the sentence ``...
+but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.'' makes sense visually from Fig.
 2 and 3.
@@
 \begin{reply}
-content...
+The ranges are different because the predictors are defined on the log scale, whereas the generator matrix is obtained by the (element-wise) exponentiation of the predictors times their respective regression coefficients plus the random-effects matrix.
+Even though the Hubei asymmetry is not statistically significant (i.e., its 95\% credible interval contains 0), it still has a non-negligible effect size ($\approx 0.25$ posterior mean per Fig.
+3a) that helps explain its appearance in Figure 3b.
 \end{reply}

### 2023-09-13 manuscript/revision_letter.tex
updated phylogenetic tree

@@
 
+\setcounter{figure}{3}
+\begin{figure*}[!t]
+	\centering
+	\includegraphics[width=0.7\textwidth]{figures/summary2.tree.pdf}
+	\vspace{-0.5em}
+	\caption{Posterior predictive modes for (unobserved) ancestral
+		locations color a phylogenetic tree that describes the
+		shared evolutionary history of 285 SARS-CoV-2
+		samples.}\label{fig:tree}
+\end{figure*}
+
 \begin{reply}
-content...
+We have updated Figure \ref{fig:tree} to the best of our ability to make the main geographic blocks visible.
+The previous limitation to 29 regions was caused by software limitation that our updated figure also avoids.
 \end{reply}
@@
 	\item 
-	The proposed method is able to include higher-dimension rate matrix compared with \citet{lemey2009bayesian,lemey2014unifying} [1, 2].
+	The proposed method is able to include higher-dimension rate matrix compared with \citet{lemey2009bayesian,lemey2014unifying}.
 Can the authors then have an external comparison to the results in \citet{lemey2020accommodating} just visually to see whether the color pattern in the phylogenetic tree is matched?

### 2023-09-13 simulations/src/condenseStates.R
Groundwork for simulations

@@ -1,0 +1,111 @@
+# All 44 locations
+locations_44 <- c("Australia","Belgium","Brazil","Cambodia","Canada","ChinaAnhui","ChinaBeijing","ChinaChongqing","ChinaFujian","ChinaGuangdong","ChinaHenan","ChinaHongKong","ChinaHubei","ChinaHunan","ChinaJiangsu","ChinaJiangxi","ChinaShandong","ChinaSichuan","ChinaYunnan","ChinaZhejiang","Finland","France","Germany","India","Iran","Italy","Japan","Luxembourg","Mexico","Nepal","Netherlands","NewZealand","Nigeria","Portugal","Singapore","SouthKorea","Spain","Sweden","Switzerland","Taiwan","Thailand","UK","USA","Vietnam")
+
+### If making new smaller regions sets, leave Hubei in as its own area, this script is not designed to remove any of the covariates
+
+# A smaller 31 location set with only one for China
+locations_31 <- c(locations_44[1:5],locations_44[13],"ChinaOther",locations_44[21:44])
+map_44_to_31 <- as.list(match(locations_31,locations_44))
+map_44_to_31[[7]] <- which(grepl("China",locations_44) & (!grepl("Hubei",locations_44)))
+
+# China only
+locations_15 <- locations_44[grepl("China",locations_44)]
+map_44_to_15 <- match(locations_15,locations_44)
+
+# The covariates 
+covariates <- c("intraContinent_distances","airTravel","Hubei_assymetry")
+
+# Assemble
+location_list <- list(locations_31,locations_15)
+map_list <- list(map_44_to_31,map_44_to_15)
+xml_names <- c("real_data_31_state_analysis.xml","real_data_15_state_analysis.xml")
+
+full_xml <- scan("xml/original/real_data_44_state_analysis.xml",sep="\n",what=character(),blank.lines.skip=FALSE)
+
+for (i in 1:length(location_list)) {
+  this_xml <- full_xml
+  
+  nstates <- length(location_list[[i]])
+  
+  # Change DataType
+  dt_open <- grep('<generalDataType id="sampleLoc.dataType">',this_xml)
+  dt_close <- grep('</generalDataType>',this_xml)
+  before <- this_xml[1:dt_open]
+  after <- this_xml[dt_close:length(this_xml)]
+  new_dt <- paste0('\t\t<state code="',location_list[[i]],'"/>')
+  if ( any(lengths(map_list[[i]]) > 1) ) {
[…]

### 2023-09-13 manuscript/pnas_revision.tex
finished rev 1 other than sim

@@
 predictive distributions every 100-thousandth MCMC iteration.
-Figure \ref{fig:tree} projects the empirical posterior predictive mode
-of these locations onto the phylogenetic tree $\phylogeny$.
+After collapsing regions into 8 major blocks, Figure \ref{fig:tree} projects the empirical posterior predictive mode of these blocks onto the phylogenetic tree $\phylogeny$.
+The general pattern looks similar to that of Figure 1 from \cite{lemey2020accommodating}, although the geographic blocking scheme differs slightly.
 
@@
 	\centering
-	\includegraphics[width=0.7\textwidth]{summary_tree.pdf}
-	\vspace{-1em}
+	\includegraphics[width=0.7\textwidth]{summary2.tree.pdf}
+	\vspace{-0.5em}
 	\caption{Posterior predictive modes for (unobserved) ancestral

### 2023-09-13 simulations/src/condenseStates.R
Draft of piBUSS-script-generator

@@ -15,6 +15,8 @@ map_44_to_15 <- match(locations_15,locations_44)
 # The covariates 
 covariates <- c("intraContinent_distances","airTravel","Hubei_assymetry")
 
+### If adding more, make sure that the state space size is unique and the name matches this format exactly
+
 # Assemble
 location_list <- list(locations_31,locations_15)
 map_list <- list(map_44_to_31,map_44_to_15)

### 2023-09-13 simulations/src/makePiBUSS.R

@@ -1,0 +1,94 @@
+filepaths <- c("xml/original/real_data_44_state_analysis.xml",list.files("xml/condensed_original",full.names=TRUE))
+xmls <- lapply(filepaths,scan,sep="\n",what=character(),blank.lines.skip=FALSE)
+
+# piBUSS block
+pibuss_text <- 
+'	<beagleSequenceSimulator id="simulator" parallel="false" output="XML">
+		<partition from="1" to="1">
+			<treeModel idref="treeModel"/>
+			<glmSubstitutionModel idref="sampleLoc.model"/>
+			<siteModel idref="sampleLoc.siteModel"/>
+			<strictClockBranchRates idref="sampleLoc.branchRates"/>
+			<frequencyModel idref="frequencyModel"/>
+		</partition>
+	</beagleSequenceSimulator>
+
+	<report fileName="sequences.xml">
+		<beagleSequenceSimulator idref="simulator"/>
+	</report>
+'
+
+# Random-effects stuff
+weak_mean <- 0.0
+weak_sd <- 0.1
+
+strong_abs_mean <- log(2.0)
+strong_sd <- 0.1
+
+percent_strong <- c(0.01,0.5,1.0)
+
+cols <- viridis::rocket(2,begin=0.3,end=0.8)
+
+pdf(file="simulating_distributions.pdf",width=4,height=3)
+  par(mai=c(0.8,0.8,0.01,0.01))
+  curve(dlnorm(x,-strong_abs_mean,weak_sd),0,3,1001,add=FALSE,col=cols[2],lwd=2,
+        xlab="Rate-scale random-effect",ylab="Density")
+  curve(dlnorm(x,weak_mean,weak_sd),0,3,1001,add=TRUE,col=cols[1],lwd=2)
+  curve(dlnorm(x,strong_abs_mean,weak_sd),0,3,1001,add=TRUE,col=cols[2],lwd=2)
+dev.off()
+  
+set.seed(42)
+for (i in 1:3) {
+  nstates <- basename(filepaths[i])
+  nstates <- gsub("real_data_","",nstates)
+  nstates <- gsub("_state_analysis.xml","",nstates)
+  nstates <- as.numeric(nstates)
+  refxdim <- 2 * choose(nstates,2)
+  for (j in 1:3) {
+    sim_name <- paste0("nstates_",nstates,"_frac_",percent_strong[j])
+    
+    mean_vec <- rep(weak_mean,refxdim)
[…]

### 2023-09-13 manuscript/revision_letter.tex

@@
 
-\begin{enumerate}
-	\item The authors spend 3 out of 7 pages in the main text and three corresponding sections in the Appendix discussing the refined error and the proof, while postpone the application with the affine-corrected error to future work.
+1.
+The authors spend 3 out of 7 pages in the main text and three corresponding sections in the Appendix discussing the refined error and the proof, while postpone the application with the affine-corrected error to future work.
 I would expect more than a relatively abstract error comparison figure for this issue.
 For example, a simple numerical simulation similar to the last paragraph in Application to compute the posterior likelihood ratio or posterior density ratio of the corrected and the uncorrected methods, because the authors eventually applied the method within a Bayesian framework.
-\end{enumerate}
 
@@
 
-\begin{enumerate}	
-	  \setcounter{enumi}{1}
-	\item I am curious about the selection of the extra two additional Chinese provinces and 11 additional countries when fitting the mixed-effects model and how the inclusion of the extra regions improve the fitting.
+2.
+I am curious about the selection of the extra two additional Chinese provinces and 11 additional countries when fitting the mixed-effects model and how the inclusion of the extra regions improve the fitting.
 Is there any criteria to determine the inclusion?
-\end{enumerate}
 
@@
 
-\begin{enumerate}	
-\setcounter{enumi}{2}
-	\item For the mixed-effects model, do the authors assume independent predictors for the fixed effects?
+3.
+For the mixed-effects model, do the authors assume independent predictors for the fixed effects?
 It seems to me that, in Fig.
@@
 Then what’s the justification for the predictor selection?
-\end{enumerate}
 
@@
 
-\begin{enumerate}	
-\setcounter{enumi}{3}	
-	\item Am I correct that the total number of particles (i.e., samples) in the MCMC framework is 2,000? The ESS for $\tau$ is extremely small, do the authors have any explanations?
-\end{enumerate}
+4. Am I correct that the total number of particles (i.e., samples) in the MCMC framework is 2,000? The ESS for $\tau$ is extremely small, do the authors have any explanations?
 
@@
 
-\begin{enumerate}	
-\setcounter{enumi}{4}	
-	\item What does the random effect matrix look like?
+5.
+What does the random effect matrix look like?
 The authors indicate that air traffic proximity is the only significant fixed-effect predict, then I would wonder how much does the random effect contribute to the model?
-\end{enumerate}
 
@@
 	\centering
-	\includegraphics[width=0.8\linewidth]{figures/randEffects.pdf}
+	\includegraphics[width=0.7\linewidth]{figures/randEffects.pdf}
 	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.
@@
 
-\begin{enumerate}	
-\setcounter{enumi}{5}	
-	\item The ranges for predictor matrices and posterior mean rate matrix are different.
+6.
+The ranges for predictor matrices and posterior mean rate matrix are different.
 However, the color palettes are identical, which makes it extremely hard to compare the impact of each predictor.
@@
 It also doesn’t make sense to me conceptually, because air traffic proximity is the only significant predictor.
-\end{enumerate}
 
@@
 
-\begin{enumerate}	
-\setcounter{enumi}{6}	
-	\item 
-	The color palette in Fig.
+7.	The color palette in Fig.
 4 makes it really hard to distinguish Hubei, other provinces in China, and other countries.
@@
 And why there are only 29 regions?
-\end{enumerate}
 
@@
 
-\begin{enumerate}	
-\setcounter{enumi}{7}	
-	\item 
-	The proposed method is able to include higher-dimension rate matrix compared with \citet{lemey2009bayesian,lemey2014unifying}.
+8.	The proposed method is able to include higher-dimension rate matrix compared with \citet{lemey2009bayesian,lemey2014unifying}.
 Can the authors then have an external comparison to the results in \citet{lemey2020accommodating} just visually to see whether the color pattern in the phylogenetic tree is matched?
-\end{enumerate}
+
+\begin{reply}
+	Thank you for the suggestion.
+We now notice a strong correspondence between our updated Figure \ref{fig:tree} and Fig1 of \citet{lemey2020accommodating}.
+We have accordingly added the following text to the manuscript:
+	\begin{quote}
+		After collapsing regions into 8 major blocks, Figure \ref{fig:tree} projects the empirical posterior predictive mode of these blocks onto the phylogenetic tree $\phylogeny$.
+The general pattern looks similar to that of Figure 1 from \cite{lemey2020accommodating}, although the geographic blocking scheme differs slightly.
+	\end{quote}
+\end{reply}
+
+\begin{reply}
+	Again, we wish to thank you for taking the time to review our manuscript.  We are confident that the revised manuscript is much improved as a result of your efforts.
+\end{reply}
 

### 2023-09-13 simulations/src/makePiBUSS.R
Try to remove some headaches

@@ -64,6 +64,18 @@ for (i in 1:3) {
     # Put things in the XML
     this_xml <- xmls[[i]]
     
+    # We want the taxa in the XML, but having the old data causes headaches
+    attribute_ends <- grep('</attr>',this_xml)
+    sample_loc_starts <- grep('<attr name="sampleLoc">',this_xml)
+    sample_ori_starts <- grep('<attr name="oriLoc">',this_xml)
+    attribute_starts <- c(sample_loc_starts,sample_ori_starts)
+    to_remove <- lapply(attribute_starts,function(as){
+      ae <- min(attribute_ends[attribute_ends > as])
+      return(as:ae)
+    })
+    to_remove <- unlist(to_remove)
+    this_xml <- this_xml[-to_remove]
+    
     # Replace fixed-effect values with posterior means
     from <- '<parameter id="glmCoefficients" value="0.1 0.8 2.0"/>'
     to <- '<parameter id="glmCoefficients" value="0.04 0.76 0.27"/>'
@@ -88,6 +100,9 @@ for (i in 1:3) {
                              paste0("sequences_",sim_name,".xml"),
                              pibuss_text,fixed=TRUE)
     this_xml <- c(this_xml,this_pibuss_text)
+    
+    # close the xml
+    this_xml <- c(this_xml,'\n</beast>')
     
     cat(this_xml,file=paste0("xml/piBUSS/",sim_name,".xml"),sep="\n")
   }

### 2023-09-13 simulations/src/makePiBUSS.R
We can now make working piBUSS scripts

@@ -76,19 +76,28 @@ for (i in 1:3) {
     to_remove <- unlist(to_remove)
     this_xml <- this_xml[-to_remove]
     
-    # Replace fixed-effect values with posterior means
-    from <- '<parameter id="glmCoefficients" value="0.1 0.8 2.0"/>'
-    to <- '<parameter id="glmCoefficients" value="0.04 0.76 0.27"/>'
-    this_xml[grep(from,this_xml)] <- to
+    # We don't want the real patterns either, and don't need the coalescent
+    to_remove <- grep('<!-- Data pattern for discrete trait',this_xml):grep('</exponentialGrowth>',this_xml)
+    this_xml <- this_xml[-to_remove]
     
     # Replace clock rate with posterior mean
     from <- '<parameter id="sampleLoc.clock.rate" value="3.5" lower="0.0"/>'
-    to <- '<parameter id="sampleLoc.clock.rate" value="4.02" lower="0.0"/>'
+    to <- '\t\t<parameter id="sampleLoc.clock.rate" value="4.02" lower="0.0"/>'
+    this_xml[grep(from,this_xml)] <- to
+    
+    # Name the frequency model so we can reference it in the piBUSS block
+    from <- '<frequencyModel normalize="true">'
+    to <- '\t\t\t<frequencyModel id="frequencyModel" normalize="true">'
+    this_xml[grep(from,this_xml)] <- to
+
+    # Replace fixed-effect values with posterior means
+    from <- '<parameter id="glmCoefficients" value="0.1 0.8 2.0"/>'
+    to <- '\t\t\t<parameter id="glmCoefficients" value="0.04 0.76 0.27"/>'
     this_xml[grep(from,this_xml)] <- to
     
     # Replace random-effects
     from <- paste0('<parameter id="glmRandCoefficients" dimension="',refxdim,'" value="0.0"/>')
-    to <- paste0('<parameter id="glmRandCoefficients" dimension="',paste0(refx_drawn,collapse=""),'" value="0.0"/>')
+    to <- paste0('\t\t\t<parameter id="glmRandCoefficients" dimension="',refxdim,'" value="',paste0(refx_drawn,collapse=" "),'"/>')
     this_xml[grep(from,this_xml)] <- to
     
     # Remove detritus
@@ -99,7 +108,7 @@ for (i in 1:3) {
     this_pibuss_text <- gsub("sequences.xml",
                              paste0("sequences_",sim_name,".xml"),
[…]

### 2023-09-13 simulations/src/makeAnalysisXMLs.R
Draft script to make analysis files

@@ -1,0 +1,51 @@
+pibuss_xml <- list.files("xml/piBUSS",full.names=TRUE)
+
+beastjar <- "~/git_repos/beast-mcmc/build/dist/beast.jar"
+beaglepath <- "/usr/local/lib"
+
+for (i in 1:length(pibuss_xml)) {
+  # Simulate
+  system(paste0("java  -Djava.library.path=",beaglepath," -jar ",beastjar," -working -overwrite ",pibuss_xml[i]))
+  
+  # Get simulated data
+  seq_xml <- list.files("xml/piBUSS",full.names=TRUE)
+  seq_xml <- seq_xml[grepl("sequence",seq_xml)]
+  simulated_data <- scan(seq_xml,sep="\n",what=character(),blank.lines.skip=TRUE)
+  
+  # Get XML template for analysis
+  id <- basename(pibuss_xml[i])
+  id <- strsplit(id,"_frac")[[1]][1]
+  if ( id == "nstates_44" ) {
+    template_file <- "xml/original/real_data_44_state_analysis.xml"
+  } else {
+    template_file <- list.files("xml/condensed_original/",full.names=TRUE)
+    template_file <- template_file[grepl(gsub("nstates_","",id),template_file)]
+  }
+  template <- scan(template_file,sep="\n",what=character(),blank.lines.skip=FALSE)
+  
+  attribute_ends <- grep('</attr>',template)
+  sample_loc_starts <- grep('<attr name="sampleLoc">',template)
+  sample_ori_starts <- grep('<attr name="oriLoc">',template)
+  
+  taxa <- simulated_data[grepl('<taxon idref=',simulated_data)]
+  taxa <- gsub('\"/>','',taxa)
+  taxa <- do.call(rbind,strsplit(taxa,'idref=\"'))[,2]
+  all_states <- c()
+  for (taxon in taxa) {
+    charstate <- simulated_data[grep(taxon,simulated_data,fixed=TRUE)+1]
+    charstate <- trimws(charstate)
+    all_states <- c(all_states,charstate)
+    charstate <- paste0("\t\t\t\t",charstate)
+    
+    taxon_line <- grep(paste0('taxon id="',taxon),template,fixed=TRUE)
+    
+    taxon_loc <- min(sample_loc_starts[sample_loc_starts > taxon_line])
+    template[taxon_loc+1] <- charstate
+    
[…]

### 2023-09-14 simulations/src/makeAnalysisXMLs.R
Clean up

@@ -23,6 +23,24 @@ for (i in 1:length(pibuss_xml)) {
   }
   template <- scan(template_file,sep="\n",what=character(),blank.lines.skip=FALSE)
   
+  # Make datatype codes upper-case because they come out of piBUSS that way
+  dt_open <- grep('<generalDataType id="sampleLoc.dataType">',template)
+  dt_close <- grep('</generalDataType>',template)
+  for (l in (dt_open+1):(dt_close-1)) {
+    line <- template[l]
+    has_code <- grepl("code",line)
+    has_alias <- grepl("alias",line)
+    line <- strsplit(line,'"')[[1]]
+    if (has_code) {
+      line[grep("code",line)+1] <- toupper(line[grep("code",line)+1])
+    }
+    if (has_alias) {
+      line[grep("alias",line)+1] <- toupper(line[grep("code",line)+1])
+    }
+    template[l] <- paste0(line,collapse='"')
+  }
+  
+  # Locate where we need to put simulated data and add it
   attribute_ends <- grep('</attr>',template)
   sample_loc_starts <- grep('<attr name="sampleLoc">',template)
   sample_ori_starts <- grep('<attr name="oriLoc">',template)
@@ -47,5 +65,5 @@ for (i in 1:length(pibuss_xml)) {
   }
   
   cat(template,sep="\n",file=paste0("xml/simulated_datasets_to_analyze/",basename(pibuss_xml[i])))
-  
+  system(paste0("rm ",seq_xml))
 }

### 2023-09-14 simulations/src/makeAnalysisXMLs.R
Change MCMC settings a bit

@@ -7,11 +7,6 @@ for (i in 1:length(pibuss_xml)) {
   # Simulate
   system(paste0("java  -Djava.library.path=",beaglepath," -jar ",beastjar," -working -overwrite ",pibuss_xml[i]))
   
-  # Get simulated data
-  seq_xml <- list.files("xml/piBUSS",full.names=TRUE)
-  seq_xml <- seq_xml[grepl("sequence",seq_xml)]
-  simulated_data <- scan(seq_xml,sep="\n",what=character(),blank.lines.skip=TRUE)
-  
   # Get XML template for analysis
   id <- basename(pibuss_xml[i])
   id <- strsplit(id,"_frac")[[1]][1]
@@ -22,6 +17,11 @@ for (i in 1:length(pibuss_xml)) {
     template_file <- template_file[grepl(gsub("nstates_","",id),template_file)]
   }
   template <- scan(template_file,sep="\n",what=character(),blank.lines.skip=FALSE)
+  
+  # Get simulated data
+  seq_xml <- list.files("xml/piBUSS",full.names=TRUE)
+  seq_xml <- seq_xml[grepl("sequence",seq_xml) & grepl(gsub(".xml","",basename(pibuss_xml[i])),seq_xml)]
+  simulated_data <- scan(seq_xml,sep="\n",what=character(),blank.lines.skip=TRUE)
   
   # Make datatype codes upper-case because they come out of piBUSS that way
   dt_open <- grep('<generalDataType id="sampleLoc.dataType">',template)
@@ -65,5 +65,6 @@ for (i in 1:length(pibuss_xml)) {
   }
   
   cat(template,sep="\n",file=paste0("xml/simulated_datasets_to_analyze/",basename(pibuss_xml[i])))
-  system(paste0("rm ",seq_xml))
 }
+
+system("rm xml/piBUSS/sequences*")

### 2023-09-14 simulations/src/makeAnalysisXMLs.R
Stupid mistake

@@ -35,7 +35,7 @@ for (i in 1:length(pibuss_xml)) {
       line[grep("code",line)+1] <- toupper(line[grep("code",line)+1])
     }
     if (has_alias) {
-      line[grep("alias",line)+1] <- toupper(line[grep("code",line)+1])
+      line[grep("state",line)+1] <- toupper(line[grep("state",line)+1])
     }
     template[l] <- paste0(line,collapse='"')
   }

### 2023-09-14 manuscript/pnas_revision.tex
fixed numbering issues

@@
 
+% \newtheorem{remark}{\bf Remark}
  \newtheorem{theorem}{\bf Theorem}
@@
 in the supplement as \cref{sec:Thm:det:main}.
-\begin{Theorem}\label{thm:ass:error}
+\begin{theorem}\label{thm:ass:error}
   Suppose that $\QQ, \JJ \in \mathcal{M}(d)$ for some $d \geq 1$.  We
@@
   \end{align}
-\end{Theorem}
-
-\begin{Remark}\label{rmk:mat:class:dcy}
+\end{theorem}
+
+\begin{remark}\label{rmk:mat:class:dcy}
   To illuminate the scope of \cref{thm:ass:error}, we have the
@@
   \end{itemize}
-\end{Remark}
+\end{remark}
 In anticipation of \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.} below
@@
 matrices.
-\begin{Theorem}\label{thm:bride:of:frankinstein}
+\begin{theorem}\label{thm:bride:of:frankinstein}
   Let $\QQ \in \mathcal{S}(d)$ be non-positive, i.e.,  suppose that
@@
   \end{align}
-\end{Theorem}
+\end{theorem}
 
@@
 We formulate our second major result as follows. 
-\begin{Theorem}\label{t:lambda_n-1(Q)->-infty_non-symm_a.s.}
+\begin{theorem}\label{t:lambda_n-1(Q)->-infty_non-symm_a.s.}
   Let $F_X \in L_{\psi_1}$ be a distribution such that $X \geq 0$
@@
       \eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
-\end{Theorem}
+\end{theorem}
 
@@
 
-\begin{Remark}\label{rmk:non:sym:num:ev}
+\begin{remark}\label{rmk:non:sym:num:ev}
   Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
@@
   increases.
-\end{Remark}
-
-\begin{Remark}\label{rmk:better:approx}
+\end{remark}
+
+\begin{remark}\label{rmk:better:approx}
   Calculation of
@@
   refined approximation.  That said, we leave the efficient and scalable application of $\widehat{\nabla}_{\JJ} e^{t\QQ}$ to future work.
-\end{Remark}
+\end{remark}
 

### 2023-09-14 manuscript/revision_letter.tex
fixed numbering issues

@@
 
-This is a review for the manuscript "On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2". 
+This is a review for the manuscript ``On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2''. 
 For the benefit of the authors on how they interpret my review, I am a theoretical/computational statistician with background on theoretical population genetics, interested in Markov chains and its implementations in Monte Carlo methods.
 
+\begin{reply}
+	We thank you for lending your effort and expertise in reviewing our manuscript and for your helpful recommendations.
+We have thoroughly implemented your suggestions and believe that the manuscript is greatly improved.
+\end{reply}
+
 Before content-specific comments let me first clarify that there are issues with the numbering of theorems and remarks in the manuscript that I have received (e.g., theorem 2 is missing, and there are two remarks labeled as Remark 2), and below I stick to the numbering used in the manuscript that I have received for review to avoid any confusion.
+
+\begin{reply}
+	We thank you for catching this numbering error, which we have addressed.  Now there are 3 theorems (1,2,3) and 5 remarks (1,2,3,4,5).
+\end{reply}
 
@@
 It is also clear that a lot of effort and thought was put into these theorems and corollaries and I appreciate that, along with the clear exposition which made following the mathematical reasoning easy.
+
+\begin{reply}
+	We thank you for your positive reception of our mathematical work.
+\end{reply}
 
@@
 But then, why use SARS-Cov-2 data for this goal, whereas it can be better established by a simulated data set?
+
+\begin{reply}
+	We thank you for sharing these concerns.
+Before responding, let us point out that we have now added a simulation section (similar to that which you specify below) to the manuscript.
+Now, we would like to argue that the decisions we made in structuring our manuscript are well-motivated based on our work in two other papers.
+	\begin{enumerate}
+		\item First,
+	\end{enumerate}
+\end{reply}
 

### 2023-09-14 manuscript/revision_letter.tex
more response

@@
 \begin{reply}
-	We thank you for sharing these concerns.
+	We thank you for sharing your concerns, which we take seriously.
 Before responding, let us point out that we have now added a simulation section (similar to that which you specify below) to the manuscript.
-Now, we would like to argue that the decisions we made in structuring our manuscript are well-motivated based on our work in two other papers.
+Now, we would like to argue that the decisions we made in structuring our manuscript are well-motivated based on our other work.
 	\begin{enumerate}
-		\item First,
+		\item First, we have written a companion paper \citet{magee2023random} (minor revisions at Systematic Biology), which presents the basic approach, implements a thorough simulation study and applies the methodology to 3 real-world data analyses of varying dimensions and differing fixed-effect structures.
+
+		\item Second, when writing this paper, we had a clear understanding of the theoretical guarantees provided by surrogate-trajectory HMC.
+Our paper \citet{glatt2020accept} (to appear at Annals of Applied Probability) proves that the surrogate-trajectory HMC transition kernel maintains detailed balance and that the algorithm samples from the target distribution with the same guarantees as vanilla HMC.
+The results of \citet{glatt2020accept} hold for general state spaces, regardless of dimensionality.
 	\end{enumerate}
+
+Despite these theoretical guarantees, the method of \citet{magee2023random} performed even better than expected in terms of relatively high acceptance rates and relatively low autocorrelation between samples.
+From our other previous work with surrogate-trajectory HMC \citep{li2019neural}, we knew this must suggest that the first-order matrix exponential gradient approximation must be pretty good.
+As a result, our goal in this paper was to find out what makes the first-order approximation so good and provide a salient example that illustrates the efficacy of the approximation in higher dimensions.
+
+In short, our original goal was not to prove validity of the general methodology, which we saw as already established through simulation studies, real-world analyses and measure theoretic probability.
+
+Again, we thank you for sharing your concerns and hope that the broader context of our work helps explain the positioning of our manuscript.
+Probably more importantly, we have updated our manuscript with a thorough simulation study (see below).
 \end{reply}
@@
 Variable 1, The method with 3 levels: Standard Hamiltonian Monte Carlo, the naive method of reference (18), and the method of the manuscript; Variable 2, The model size with 3 levels: Small, Intermediate, Large; Variable 3, The number of parameters with considerable signal with 3 levels: 1 parameter, half of the parameters, all parameters in the model.
+
+\begin{reply}
+Simulaiton. TBD.
+\end{reply}
 

### 2023-09-14 manuscript/revision_letter.tex
first paragraph

@@
 \begin{itemize}
-	\item clarified how the present work complements its companion paper \citet{magee2023random}, which presents scaling and simulation studies;
+	\item clarified how the present work relates to our other work that includes thorough simulation studies \citet{magee2023random} and provides strong theoretical guarantees for surrogate-trajectory HMC \citet{glatt2020accept};
 
@@
 
-    \item compared scalability of surrogate-trajectory HMC with the naive and affine-corrected approximations;
-
-    \item improved upon the SARS-CoV-2 application, relating it to previously established results;
+    \item expanded upon the SARS-CoV-2 application, relating it to previously established results and improving a key figure;
 
@@
 
-In short, our original goal was not to prove validity of the general methodology, which we saw as already established through simulation studies, real-world analyses and measure theoretic probability.
+In short, our original goal was not to prove validity of the general methodology, which we saw as already established through simulation studies, real-world analyses and measure theoretic arguments.
 
@@
 
+\begin{reply}
+	Again, we thank you for catching this error and have fixed the numbering accordingly. 
+\end{reply}
+
 * Regarding the opening paragraph.
@@
 The whole paragraph is written vaguely and I felt dismissive of the applications of the work being presented.
+
+\begin{reply}
+	It is clear that the opening sentence missed its mark.  We have edited the first paragraph to read:
+	\begin{quote}
+		Phylogeographic methods
+	\cite{lemey2009bayesian, lemey2014unifying, holbrook2021massive,
+	holbrook2022viral} model large-scale viral transmission between
+human populations as a function of the shared evolutionary history of
+the viral population of interest.  Data take the form of dates,
+locations and genome sequences associated to individual viral
+samples. Spatiotemporal structure interfaces with network structure
+given by the phylogeny, or family tree, describing the viruses'
+collective history beginning with the most recent common ancestor.
+While one cannot directly observe this history, one may statistically
+reconstruct the phylogenetic tree by positing that changes in the
+viral genome happen randomly at regular intervals, thereby capturing
+the intuition that viral samples with more differences between their
+(aligned) sequences should find themselves further apart on the family
+tree.
+	\end{quote}
+\end{reply}
 

### 2023-09-14 manuscript/pnas_revision.tex
first paragraph
more revisions

@@
 
-\dropcap{W}hen measuring the size of ``big data'', one should
-consider the structures that implicitly give rise to one's data and
-the computational complexities of modeling schemes that appropriately
-take such structures into account. Phylogeographic methods
+\dropcap{P}hylogeographic methods
 \cite{lemey2009bayesian, lemey2014unifying, holbrook2021massive,
@@
 variables.
+To avoid confusion, note that the following definition uses the term in the same way as, e.g., \cite{vershynin:2018}, but that other definitions that mean quite the opposite (i.e., heavier than exponential tails) appear in the literature \cite{goldie1998subexponential}.
 \begin{definition}
@@
   consequences of relaxing various assumptions of
-  \cref{cor:rig:det:rand:wrapup}. In general, there appears to be no
+  \cref{cor:rig:det:rand:wrapup}.
+The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (\cref{def:sub:exp:RV}).
+In general, there appears to be no
   significant departure from the idea that the proxy form

### 2023-09-14 manuscript/revision_letter.tex
more revisions

@@
 \newcommand{\NN}{\mathbb{N}}
+
+ \newtheorem{@definition}{\bf Definition}
+\newenvironment{definition}{\begin{@definition}\rm}{\end{@definition}}
 
@@
 Cauchy distribution is a member of the subexponential family and I wonder whether the tail properties of the folded Cauchy are distinct from those of the subexponential family (I do not have an answer to this, so I suggest that the authors check and provide a brief explanation).
+
+\newcommand{\bbP}{{\mathbb P}}
+\newcommand{\bbE}{{\mathbb E}}
+\begin{reply}
+	You are correct that the truncated Cauchy (at 0) and the folded Cauchy are the same, but this distribution is not sub-exponential as we have defined the term in Definition \ref{def:sub:exp:RV}:
+	\begin{quote}
+\begin{definition}
+	\label{def:sub:exp:RV}
+	A random variable $X$ is called \textit{sub-exponential} if there
+	exists some constant $K > 0$ for which its tails satisfy
+	\begin{equation}\label{e:def_subexp_RV}
+		\bbP(|X|\geq t) \leq 2 \exp\{-t/K\}, \quad \forall t \geq 0.
+	\end{equation}
+	In this case, the \textit{sub-exponential norm} of $X$ is defined by
+	\begin{equation}\label{e:sub-exp_norm}
+		\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE \exp(|X|/s) \leq 2\big\}.
+	\end{equation}
+	The \textit{class of sub-exponential distributions} is denoted by
+	\begin{align*}
+		L_{\psi_1} = \big\{F_X(dx): \|X\|_{\psi_1} < \infty \big\}.
+	\end{align*}
+\end{definition}
+	\end{quote}
+Here, sub-exponential means the opposite of heavy-tailed.
+The folded Cauchy distribution represents a distinct failure of the sub-exponential assumption.
+In order to avoid further confusion, we have added the following sentence prior to Definition 1:
+\begin{quote}
+	To avoid confusion, note that the following definition uses the term in the same way as, e.g., \cite{vershynin:2018}, but that other definitions that mean quite the opposite (i.e., heavier than exponential tails) appear in the literature \citep{goldie1998subexponential}.
+\end{quote}
+Additionally, we have added the following sentence to Remark 4:
+\begin{quote}
+	The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (Definition \ref{def:sub:exp:RV}).
+\end{quote}
+\end{reply}
 

### 2023-09-14 manuscript/pnas_revision.tex
more revisions

@@
 \begin{equation}\label{e:def_subexp_RV}
-\bbP(|X|\geq t) \leq 2 \exp\{-t/K\}, \quad \forall t \geq 0.
+\bbP(|X|\geq t) \leq 2 e^{-t/K}, \quad \forall t \geq 0.
 \end{equation}
@@
 \begin{equation}\label{e:sub-exp_norm}
-\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE \exp(|X|/s) \leq 2\big\}.
+\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE e^{|X|/s} \leq 2\big\}.
 \end{equation}
@@
   condition \eqref{e:def_subexp_RV} is equivalent to the existence of
-  some $s_0 > 0$ such that $\bbE \exp(|X|/s_0) \leq 2$, namely,
+  some $s_0 > 0$ such that $\bbE e^{|X|/s_0} \leq 2$, namely,
   $\|X\|_{\psi_1} < \infty$ in \eqref{e:sub-exp_norm}.  As a notable
@@
 The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (\cref{def:sub:exp:RV}).
-In general, there appears to be no
-  significant departure from the idea that the proxy form
+There appears to be no
+  significant departure from the idea that the form
   $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
@@
 interested in the performance of the first-order approximation as a
-surrogate gradient for HMC in such a high-dimensional setting. To
-isolate this performance, we fix the Bayesian bridge global-scale
+surrogate gradient for HMC in such a high-dimensional setting.
+Whereas we know that the surrogate-trajectory HMC transition kernel leaves its target distribution invariant regardless of the approximation quality \cite{glatt2020accept}, transitions that rely on poor gradient approximations result in small acceptance rates, more random walk behavior and high autocorrelation between samples.
+Since ESS is inversely proportional to a Markov chain's asymptotic autocorrelation, larger ESS suggest a useful gradient approximation.
+To
+isolate the approximation's performance, we fix the Bayesian bridge global-scale
 $\tau$ at 2.5$\times10^{-5}$.  We generate a Markov chain with 80,000

### 2023-09-14 manuscript/revision_letter.tex
more revisions

@@
 	\begin{quote}
+
+\setcounter{equation}{39}
 \begin{definition}
@@
 	\begin{equation}\label{e:def_subexp_RV}
-		\bbP(|X|\geq t) \leq 2 \exp\{-t/K\}, \quad \forall t \geq 0.
+		\bbP(|X|\geq t) \leq 2 e^{-t/K}, \quad \forall t \geq 0.
 	\end{equation}
@@
 	\begin{equation}\label{e:sub-exp_norm}
-		\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE \exp(|X|/s) \leq 2\big\}.
+		\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE e^{|X|/s} \leq 2\big\}.
 	\end{equation}
@@
 * Equations 12-19 seems to follow straightforwardly from the section where equation 2-7 are presented.
-I agree that writing them explicitly is better, but I felt that they disrupt the flow of the section "Deterministic bounds on approximation error in time".
+I agree that writing them explicitly is better, but I felt that they disrupt the flow of the section ``Deterministic bounds on approximation error in time''.
 Please consider emulating them to somewhere in 2-7 or following, intermediate steps in 12-19 maybe put in an appendix.
+
+\begin{reply}
+We thank you for the suggestion, but we would really like to keep 12-19 where they are if it's alright.
+We do not want to place them directly following 2-7 because we would prefer to keep the Introduction at a high level, and we do not want to put them in the Appendix because we think they communicate valuable insight into the dynamical systems view of our formal argument.
+We hope you understand.
+\end{reply}
 
@@
 
+\begin{reply}
+	Thank you for catching our inconsistent use of $\exp$.
+We have updated the manuscript accordingly.
+You are also correct that Vershynin provides this definition.
+We have re-checked and there are no additional regularity conditions attached.
+We wish to proceed with this definition, but please note that there exists a definition of sub-exponential based on CDFs that actually means heavier-tailed than exponential \citep{goldie1998subexponential}.
+We point this out again to help avoid confusion.
+\end{reply}
+
 * Remark 6. The sentence starting with "In general..." is not clear in which sense it is general. "Proxy" is also not clear.
 
+\begin{reply}
+	We thank you for pointing this out. We have changed the sentence to the following:
+	\begin{quote}
+		There appears to be no
+		significant departure from the idea that the form
+		$t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
+		the true approximation error \eqref{eq:true:diff} as the dimension
+		increases.
+	\end{quote}
+\end{reply}
+
 * The paragraph starting with "In addition..." on page 7 is too short for the reader to understand the implications of the analysis. Please consider expanding it.
+
+\begin{reply}
+	Thank you for pointing this out.  We have added the following sentences to the paragraph:
+	\begin{quote}
+		Whereas we know that the surrogate-trajectory HMC transition kernel leaves its target distribution invariant regardless of the approximation quality \citep{glatt2020accept}, transitions that rely on poor gradient approximations result in small acceptance rates, more random walk behavior and high autocorrelation between samples.
+Since ESS is inversely proportional to a Markov chain's asymptotic autocorrelation, larger ESS suggest a useful gradient approximation.
+	\end{quote}
+\end{reply}
 

### 2023-09-14 manuscript/pnas_revision.tex
first pass done

@@
 
-The continuous-time Markov chain (CTMC) \cite{norris1998markov}
+The continuous-time Markov chain (CTMC) \cite{norris_1997}
 represents the gold-standard mathematical model for such evolution of
@@
     $\lambda_j(\QQ)$ are its (real) eigenvalues.  We suppose that
-    these eigenvalue are all non-positive,
+    these eigenvalues are all non-positive,
     \begin{align}\label{eq:ev:d:cond:sym:case}

### 2023-09-14 manuscript/revision_letter.tex
first pass done

@@
 
+\newcommand{\Y}{\mathbf{Y}}
+\begin{reply}
+	We would prefer to keep this section the same, but we do relate the general statement to Bayesian inference in a footnote:
+	\begin{quote}
+		In the Bayesian inference context, we are typically
+		considering the posterior density function
+		$\pi(\ttheta):=p(\ttheta|\Y)$ for the parameter
+		$\ttheta\in\mathbb{R}^K$ given observed data $\Y$.
+	\end{quote}
+\end{reply}
+
 * For completeness, it is worth restating that Q is symmetric in Theorem 3. (Or is it not?)
 
+\begin{reply}
+TBD	
+\end{reply}
+
 * Remark 2 (iii). "eigenvalue" should be "eigenvalues".
 
+\begin{reply}
+Thanks for catching this!  We have amended accordingly. 
+\end{reply}
+
 * References (5) is a book, but there is no section or page.
+
+\begin{reply}
+	Again, thanks for catching this.  We have updated the citation to reference the specific chapter \cite{norris_1997}.
+\end{reply}
+
+\begin{reply}
+	Again, we wish to thank you for putting so much time and effort in to your review of our manuscript.
+It is greatly appreciated.
+We have addressed and implemented your suggestions to the best of our abilities and are confident that the manuscript is now much improved.
+\end{reply}
 

### 2023-10-03 code/simulations.R
sim and likelihood

@@ -1,0 +1,69 @@
+setwd("~/expmDerivative/")
+
+##### simulation design (3x3x3)
+# 1) 3 significance proportions: 5 percent, 50 percent, 95 percent
+# 2) 3 gradients: exact, affine-corrected approx, naive approx
+# 3) 3 state space sizes: 15, 30, 45
+
+library(expm)
+
+################################################################################
+# code for simulating data
+
+simCTMC <- function(Q, initialStates, time) {
+  S <- dim(Q)[1]
+  N <- length(initialStates)
+  if(any(initialStates>S)) stop("Initial states cannot be larger than state space.")
+  
+  transMat <- expm(time*Q)
+  
+  output <- rep(0,N)
+  for(n in 1:N) {
+    ntlStt <- rep(0,S)
+    ntlStt[initialStates[n]] <- 1
+    outputProbs <- t(transMat) %*% ntlStt
+    output[n] <- sample(x=1:S,size=1,prob=outputProbs)
+  }
+  return(output)
+}
+
+# # test
+# S <- 15
+# Q       <- matrix(rexp(S^2),S,S)
+# diag(Q) <- 0
+# diag(Q) <- - rowSums(Q)
+# initialStates <- sample(x=1:S,size=100,replace=TRUE)
+# simCTMC(Q,initialStates,1)
+
+lg_lklhd <- function(Q,input,output,time) {
+  if(! length(input)==length(output)) stop("Input and output must have same length.")
+  S <- dim(Q)[1]
+  if(any(input>S)) stop("Input states cannot be larger than state space.")
+  if(any(output>S)) stop("Output states cannot be larger than state space.")
+  N <- length(input)
+  transMat <- expm(time*Q)
+  
+  lklhd <- 0
+  for(n in 1:N) {
+    ntlStt <- rep(0,S)
+    tptStt <- rep(0,S)
+    ntlStt[input[n]] <- 1
+    tptStt[output[n]] <- 1
+    lklhd <- lklhd + log(t(tptStt) %*% t(transMat) %*% ntlStt)
+  }
+  return(lklhd)
+}
+
+# # test
+# S <- 15
+# Q       <- matrix(rexp(S^2),S,S)
+# diag(Q) <- 0
+# diag(Q) <- - rowSums(Q)
+# initialStates <- sample(x=1:S,size=100,replace=TRUE)
+# output <- simCTMC(Q,initialStates,1)
+# 
+# ll <- lg_lklhd(Q,initialStates,output,1)
+# ll
+
+
+

### 2023-10-03 code/simulations.R
derivs appear correct

@@ -51,7 +51,7 @@ lg_lklhd <- function(Q,input,output,time) {
     tptStt[output[n]] <- 1
     lklhd <- lklhd + log(t(tptStt) %*% t(transMat) %*% ntlStt)
   }
-  return(lklhd)
+  return(as.numeric(lklhd))
 }
 
 # # test
@@ -65,5 +65,121 @@ lg_lklhd <- function(Q,input,output,time) {
 # ll <- lg_lklhd(Q,initialStates,output,1)
 # ll
 
+blockwise <- function(t,Q,E) {
+  B1  <- cbind(Q,E)
+  B2  <- cbind(Q*0,Q)
+  B   <- rbind(B1,B2) 
+  etB <- expm::expm(t*B)
+  return(etB[1:(dim(etB)[1]/2),(dim(etB)[1]/2+1):dim(etB)[1]])
+}
 
+approx <- function(t, Q, Jij) {
+  n <- dim(Q)[1]
+  eig.obj <- eigen(Q)
+  inv.spctrm <- 1/eig.obj$values
+  inv.spctrm[length(inv.spctrm)] <- 0
+  pinv    <- Re(eig.obj$vectors %*% diag(inv.spctrm) %*% solve(eig.obj$vectors))
+  mat <- t*(diag(n) - pinv%*%Q) %*% Jij %*% Q %*% pinv
+  return(mat)
+}
+
+frstRdr <- function(t,Q,E){
+  return(expm::expm(t*Q) %*% E * t)
+}
+
+frstRdrCrrctd <- function(t,Q,E) {
+  out <- frstRdr(t,Q,E) - approx(t,Q,E)
+  return(out)
+}
+
+lg_prr <- function(Q) {
+  return(0) # TODO: add bayesian bridge log prior
+}
+
+lg_prr_grd <- function(Q) {
+  return(0) # TODO: add bayesian bridge log prior gradient
+}
+
+lg_lklhd_grd <- function(t,Q,input,output,method) {
+  if(! length(input)==length(output)) stop("Input and output must have same length.")
+  S <- dim(Q)[1]
+  if(any(input>S)) stop("Input states cannot be larger than state space.")
+  if(any(output>S)) stop("Output states cannot be larger than state space.")
+  N <- length(input)
+  if(! method %in% c("exact","approx","corrected")) stop("Use prescribed gradient method!")
[…]

### 2023-10-03 code/simulations.R
hmc not quite working

@@ -41,7 +41,7 @@ lg_lklhd <- function(Q,input,output,time) {
   if(any(input>S)) stop("Input states cannot be larger than state space.")
   if(any(output>S)) stop("Output states cannot be larger than state space.")
   N <- length(input)
-  transMat <- expm(time*Q)
+  transMat <- expm::expm(time*Q)
   
   lklhd <- 0
   for(n in 1:N) {
@@ -135,51 +135,158 @@ lg_lklhd_grd <- function(t,Q,input,output,method) {
     }
   }
   
-  # don't forget to subtract log-posterior
-  gradient <- gradient - lg_lklhd(Q,input,output,t) 
+  # don't forget to divide by log-likelihood
+  gradient <- gradient / lg_lklhd(Q,input,output,t) 
   diag(gradient) <- 0
   
   return(gradient)
  }
 
 # test
+# S <- 5
+# Q       <- matrix(rexp(S^2),S,S)
+# diag(Q) <- 0
+# diag(Q) <- - rowSums(Q)
+# initialStates <- sample(x=1:S,size=100,replace=TRUE)
+# output <- simCTMC(Q,initialStates,1)
+# 
+# ll_grd <- lg_lklhd_grd(1,Q,initialStates,output,method="exact")
+# ll_grd
+# 
+# ll_grd_prx <- lg_lklhd_grd(1,Q,initialStates,output,method="approx")
+# ll_grd_prx
+# 
+# ll_grd_crctd <- lg_lklhd_grd(1,Q,initialStates,output,method="corrected")
+# ll_grd_crctd
+# 
+# norm(ll_grd - ll_grd_prx)
+# 
+# norm(ll_grd - ll_grd_crctd)
+# 
+# S <- 45
+# Q       <- matrix(rexp(S^2),S,S)
+# diag(Q) <- 0
+# diag(Q) <- - rowSums(Q)
+# initialStates <- sample(x=1:S,size=100,replace=TRUE)
+# output <- simCTMC(Q,initialStates,1)
+# 
+# ll_grd <- lg_lklhd_grd(1,Q,initialStates,output,method="exact")
+# #ll_grd
+# 
+# ll_grd_prx <- lg_lklhd_grd(1,Q,initialStates,output,method="approx")
+# #ll_grd_prx
+# 
+# ll_grd_crctd <- lg_lklhd_grd(1,Q,initialStates,output,method="corrected")
[…]

### 2023-10-04 code/simulations.R
things seems to be working but slowly

@@ -93,11 +93,17 @@ frstRdrCrrctd <- function(t,Q,E) {
 }
 
 lg_prr <- function(Q) {
-  return(0) # TODO: add bayesian bridge log prior
+  priors <- dnorm(log(Q),log = TRUE)
+  diag(priors) <- 0
+  return(sum(priors)) # TODO: add bayesian bridge log prior
 }
 
 lg_prr_grd <- function(Q) {
-  return(0) # TODO: add bayesian bridge log prior gradient
+  # start with gaussian prior
+  gradient <- - log(Q)
+  diag(gradient) <- 0
+  
+  return(gradient) # TODO: add bayesian bridge log prior gradient
 }
 
 lg_lklhd_grd <- function(t,Q,input,output,method) {
@@ -118,11 +124,11 @@ lg_lklhd_grd <- function(t,Q,input,output,method) {
         E <- matrix(0,S,S)
         E[i,j] <- 1
         if (method=="exact") {
-          patty <- patty + blockwise(t,Q,E) * Q[i,j] + blockwise(t,Q,Eii) * Q[i,i]
+          patty <- patty + blockwise(t,Q,E) * Q[i,j] - blockwise(t,Q,Eii) * Q[i,i]
         } else if (method=="approx") {
-          patty <- patty + frstRdr(t,Q,E) * Q[i,j] + frstRdr(t,Q,Eii) * Q[i,i]
+          patty <- patty + frstRdr(t,Q,E) * Q[i,j] - frstRdr(t,Q,Eii) * Q[i,i]
         } else {
-          patty <- patty + frstRdrCrrctd(t,Q,E) * Q[i,j] + frstRdrCrrctd(t,Q,Eii) * Q[i,i]
+          patty <- patty + frstRdrCrrctd(t,Q,E) * Q[i,j] - frstRdrCrrctd(t,Q,Eii) * Q[i,i]
         }
         for(n in 1:N) {
           ntlStt <- rep(0,S)
@@ -197,54 +203,65 @@ grad <- function(t,Q,input,output,method) {
 
 delta <- function(n) {
   #return( min(0.01,n^(-0.5)) )
-  return(n^(-0.5))
+  return(n^(-0.1))
+}
+
+lg_Q_to_Q <- function(lg_Q) {
+  Q <- exp(lg_Q)
+  diag(Q) <- 0
+  diag(Q) <- - rowSums(Q)
+  return(Q)
 }
 
[…]

### 2023-10-04 code/simulations.R
correct but slow

@@ -51,6 +51,12 @@ lg_lklhd <- function(Q,input,output,time) {
     tptStt[output[n]] <- 1
     lklhd <- lklhd + log(t(tptStt) %*% t(transMat) %*% ntlStt)
   }
+  # ntlStts <- matrix(0,N,S)
+  # tptStts <- matrix(0,N,S)
+  # ntlStts[cbind(1:N,input)] <- 1
+  # tptStts[cbind(1:N,output)] <- 1
+  # lklhd <- log(sum(diag(tptStts %*% t(transMat) %*% t(ntlStts))))
+  
   return(as.numeric(lklhd))
 }
 
@@ -118,17 +124,26 @@ lg_lklhd_grd <- function(t,Q,input,output,method) {
   for(i in 1:S) {
     Eii <- matrix(0,S,S)
     Eii[i,i] <- 1
+    # get diagonal contributions
+    if (method=="exact") {
+      diagContrib <- - blockwise(t,Q,Eii) * Q[i,i]
+    } else if (method=="approx") {
+      diagContrib <- - frstRdr(t,Q,Eii) * Q[i,i]
+    } else {
+      diagContrib <- - frstRdrCrrctd(t,Q,Eii) * Q[i,i]
+    }
+    
     for(j in 1:S) {
       if(i != j) {
         patty <- matrix(0,S,S)
         E <- matrix(0,S,S)
         E[i,j] <- 1
         if (method=="exact") {
-          patty <- patty + blockwise(t,Q,E) * Q[i,j] - blockwise(t,Q,Eii) * Q[i,i]
+          patty <- patty + blockwise(t,Q,E) * Q[i,j] + diagContrib
         } else if (method=="approx") {
-          patty <- patty + frstRdr(t,Q,E) * Q[i,j] - frstRdr(t,Q,Eii) * Q[i,i]
+          patty <- patty + frstRdr(t,Q,E) * Q[i,j] + diagContrib
         } else {
-          patty <- patty + frstRdrCrrctd(t,Q,E) * Q[i,j] - frstRdrCrrctd(t,Q,Eii) * Q[i,i]
+          patty <- patty + frstRdrCrrctd(t,Q,E) * Q[i,j] + diagContrib
         }
         for(n in 1:N) {
           ntlStt <- rep(0,S)
@@ -307,3 +322,41 @@ for(i in 1:maxIts) {
 plot(variable,type="l")
[…]

### 2023-10-04 code/lowDimSim1.R
ready to run scripts

@@ -1,0 +1,374 @@
+setwd("~/expmDerivative/")
+
+##### simulation design (3x3x3)
+# 1) 3 significance proportions: 5 percent, 50 percent, 95 percent
+# 2) 3 gradients: exact, affine-corrected approx, naive approx
+# 3) 3 state space sizes: 15, 30, 45
+
+library(expm)
+
+################################################################################
+# code for simulating data
+
+simCTMC <- function(Q, initialStates, time) {
+  S <- dim(Q)[1]
+  N <- length(initialStates)
+  if(any(initialStates>S)) stop("Initial states cannot be larger than state space.")
+  
+  transMat <- expm(time*Q)
+  
+  output <- rep(0,N)
+  for(n in 1:N) {
+    ntlStt <- rep(0,S)
+    ntlStt[initialStates[n]] <- 1
+    outputProbs <- t(transMat) %*% ntlStt
+    output[n] <- sample(x=1:S,size=1,prob=outputProbs)
+  }
+  return(output)
+}
+
+# # test
+# S <- 15
+# Q       <- matrix(rexp(S^2),S,S)
+# diag(Q) <- 0
+# diag(Q) <- - rowSums(Q)
+# initialStates <- sample(x=1:S,size=100,replace=TRUE)
+# simCTMC(Q,initialStates,1)
+
+lg_lklhd <- function(Q,input,output,time) {
+  if(! length(input)==length(output)) stop("Input and output must have same length.")
+  S <- dim(Q)[1]
+  if(any(input>S)) stop("Input states cannot be larger than state space.")
+  if(any(output>S)) stop("Output states cannot be larger than state space.")
+  N <- length(input)
+  transMat <- expm::expm(time*Q)
+  
+  lklhd <- 0
+  for(n in 1:N) {
+    ntlStt <- rep(0,S)
+    tptStt <- rep(0,S)
+    ntlStt[input[n]] <- 1
+    tptStt[output[n]] <- 1
+    lklhd <- lklhd + log(t(tptStt) %*% t(transMat) %*% ntlStt)
+  }
+  # ntlStts <- matrix(0,N,S)
+  # tptStts <- matrix(0,N,S)
+  # ntlStts[cbind(1:N,input)] <- 1
+  # tptStts[cbind(1:N,output)] <- 1
[…]

### 2023-10-04 code/lowDimSim2.R

@@ -1,0 +1,373 @@
+setwd("~/expmDerivative/")
+
+##### simulation design (3x3x3)
+# 1) 3 significance proportions: 5 percent, 50 percent, 95 percent
+# 2) 3 gradients: exact, affine-corrected approx, naive approx
+# 3) 3 state space sizes: 15, 30, 45
+
+library(expm)
+
+################################################################################
+# code for simulating data
+
+simCTMC <- function(Q, initialStates, time) {
+  S <- dim(Q)[1]
+  N <- length(initialStates)
+  if(any(initialStates>S)) stop("Initial states cannot be larger than state space.")
+  
+  transMat <- expm(time*Q)
+  
+  output <- rep(0,N)
+  for(n in 1:N) {
+    ntlStt <- rep(0,S)
+    ntlStt[initialStates[n]] <- 1
+    outputProbs <- t(transMat) %*% ntlStt
+    output[n] <- sample(x=1:S,size=1,prob=outputProbs)
+  }
+  return(output)
+}
+
+# # test
+# S <- 15
+# Q       <- matrix(rexp(S^2),S,S)
+# diag(Q) <- 0
+# diag(Q) <- - rowSums(Q)
+# initialStates <- sample(x=1:S,size=100,replace=TRUE)
+# simCTMC(Q,initialStates,1)
+
+lg_lklhd <- function(Q,input,output,time) {
+  if(! length(input)==length(output)) stop("Input and output must have same length.")
+  S <- dim(Q)[1]
+  if(any(input>S)) stop("Input states cannot be larger than state space.")
+  if(any(output>S)) stop("Output states cannot be larger than state space.")
+  N <- length(input)
+  transMat <- expm::expm(time*Q)
+  
+  lklhd <- 0
+  for(n in 1:N) {
+    ntlStt <- rep(0,S)
+    tptStt <- rep(0,S)
+    ntlStt[input[n]] <- 1
+    tptStt[output[n]] <- 1
+    lklhd <- lklhd + log(t(tptStt) %*% t(transMat) %*% ntlStt)
+  }
+  # ntlStts <- matrix(0,N,S)
+  # tptStts <- matrix(0,N,S)
+  # ntlStts[cbind(1:N,input)] <- 1
+  # tptStts[cbind(1:N,output)] <- 1
[…]

### 2023-10-04 code/lowDimSim3.R

@@ -303,60 +303,72 @@ hmc <- function(t,S,input,output,method,stepSize=0.01,L=20,maxIts=1000,
 }
 
 # test
+set.seed(1)
 S <- 5
-maxIts <- 2000
-initialStates <- sample(x=1:S,size=5,replace=TRUE)
-Q       <- matrix(rexp(S^2),S,S)
+maxIts <- 100000
+initialStates <- sample(x=1:S,size=20,replace=TRUE)
+Q       <- matrix(exp(rnorm(S^2)),S,S)
 diag(Q) <- 0
 diag(Q) <- - rowSums(Q)
 output <- simCTMC(Q,initialStates,1)
-hmcOut <- hmc(t=1,S=S,input=initialStates,output=output,stepSize=0.1,
-             method="exact",maxIts = maxIts, targetAccept = 0.65, Q_init=Q)
-plot(hmcOut[[2]],type="l")
-chain <- hmcOut[[1]]
-
-variable <- rep(0,maxIts)
-for(i in 1:maxIts) {
-  variable[i] <- chain[[i]][5,1]
-}
-plot(variable,type="l")
-abline(h=log(Q[5,1]),col="red")
-
-# approx
-hmcOut <- hmc(t=1,S=S,input=initialStates,output=output,stepSize=0.1,
-              method="approx",L=8,maxIts = maxIts, targetAccept = 0.65, Q_init=Q)
-plot(hmcOut[[2]],type="l")
-chain <- hmcOut[[1]]
-
-variable <- rep(0,maxIts)
-for(i in 1:maxIts) {
-  variable[i] <- chain[[i]][5,1]
-}
-plot(variable,type="l")
-abline(h=log(Q[5,1]),col="red")
+# hmcOut <- hmc(t=1,S=S,input=initialStates,output=output,stepSize=0.1,
+#              method="exact",maxIts = maxIts, targetAccept = 0.65, Q_init=Q)
+# plot(hmcOut[[2]],type="l")
+# chain <- hmcOut[[1]]
+# 
+# variable <- rep(0,maxIts)
+# for(i in 1:maxIts) {
+#   variable[i] <- chain[[i]][2,3]
+# }
+# plot(variable,type="l")
+# abline(h=log(Q[2,3]),col="red")
+# 
+# # approx
+# hmcOut <- hmc(t=1,S=S,input=initialStates,output=output,stepSize=0.1,
[…]

### 2023-10-05 code/highDimSim.R
bayesian bridge prior not working without local scale parameters

@@ -1,0 +1,391 @@
+setwd("~/expmDerivative/")
+
+##### simulation design (3x3x3)
+# 1) 3 significance proportions: 5 percent, 50 percent, 95 percent
+# 2) 3 gradients: exact, affine-corrected approx, naive approx
+# 3) 3 state space sizes: 15, 30, 45
+
+library(expm)
+
+################################################################################
+# code for simulating data
+
+simCTMC <- function(Q, initialStates, time) {
+  S <- dim(Q)[1]
+  N <- length(initialStates)
+  if(any(initialStates>S)) stop("Initial states cannot be larger than state space.")
+  
+  transMat <- expm(time*Q)
+  
+  output <- rep(0,N)
+  for(n in 1:N) {
+    ntlStt <- rep(0,S)
+    ntlStt[initialStates[n]] <- 1
+    outputProbs <- t(transMat) %*% ntlStt
+    output[n] <- sample(x=1:S,size=1,prob=outputProbs)
+  }
+  return(output)
+}
+
+# # test
+# S <- 15
+# Q       <- matrix(rexp(S^2),S,S)
+# diag(Q) <- 0
+# diag(Q) <- - rowSums(Q)
+# initialStates <- sample(x=1:S,size=100,replace=TRUE)
+# simCTMC(Q,initialStates,1)
+
+lg_lklhd <- function(Q,input,output,time,etQ=NULL) {
+  if(! length(input)==length(output)) stop("Input and output must have same length.")
+  S <- dim(Q)[1]
+  if(any(input>S)) stop("Input states cannot be larger than state space.")
+  if(any(output>S)) stop("Output states cannot be larger than state space.")
+  N <- length(input)
+  if(is.null(etQ)) {
+    transMat <- expm::expm(time*Q)
+  } else {
+    transMat <- etQ
+  }
+
+  lklhd <- 0
+  for(n in 1:N) {
+    ntlStt <- rep(0,S)
+    tptStt <- rep(0,S)
+    ntlStt[input[n]] <- 1
+    tptStt[output[n]] <- 1
+    lklhd <- lklhd + log(t(tptStt) %*% t(transMat) %*% ntlStt)
+  }
+  # ntlStts <- matrix(0,N,S)
[…]

### 2023-10-05 code/analyzeLowDimExp.R
low dim matrix viz

@@ -1,0 +1,88 @@
+setwd("~/expmDerivative/")
+library(ggplot2)
+
+exact <- readRDS("data/lowDimSimExact.rds")
+approx <- readRDS("data/lowDimSimApprox.rds")
+correct <- readRDS("data/lowDimSimCorrected.rds")
+
+plot(exact[[2]],type="l")
+lines(approx[[2]],col="red")
+lines(correct[[2]],col="green")
+
+# flatten lists
+df_ex <- matrix(0,100000,25)
+for (i in 1:100000) {
+  exact[[1]][[i]] <- exp(exact[[1]][[i]])
+  for(j in 1:5) {
+    exact[[1]][[i]][j,j] <- -(exact[[1]][[i]][j,j] - sum(exact[[1]][[i]][j,]))
+  }
+  df_ex[i,] <- as.vector(exact[[1]][[i]])
+}
+df_ex <- data.frame(df_ex)
+df_ex$method <- "exact"
+
+df_ap <- matrix(0,100000,25)
+for (i in 1:100000) {
+  approx[[1]][[i]] <- exp(approx[[1]][[i]])
+  for(j in 1:5) {
+    approx[[1]][[i]][j,j] <- -(approx[[1]][[i]][j,j] - sum(approx[[1]][[i]][j,]))
+  }
+  df_ap[i,] <- as.vector(approx[[1]][[i]])
+}
+df_ap <- data.frame(df_ap)
+df_ap$method <- "approx"
+
+df_co <- matrix(0,100000,25)
+for (i in 1:100000) {
+  correct[[1]][[i]] <- exp(correct[[1]][[i]])
+  for(j in 1:5) {
+    correct[[1]][[i]][j,j] <- -(correct[[1]][[i]][j,j] - sum(correct[[1]][[i]][j,]))
+  }
+  df_co[i,] <- as.vector(correct[[1]][[i]])
+}
+df_co <- data.frame(df_co)
+df_co$method <- "correct"
+
+df <- rbind(df_ap,df_ex,df_co)
+
+df$Gradient <- factor(df$method)
+levels(df$Gradient) <- c("Naive","Corrected","Exact")
+
+df <- reshape2::melt(df,value.name = "value",measure.vars=1:25)
+df$variable <- factor(df$variable)
+levels(df$variable) <- c("1:1","1:2","1:3","1:4","1:5",
+                         "2:1","2:2","2:3","2:4","2:5",
+                         "3:1","3:2","3:3","3:4","3:5",
[…]

### 2023-10-05 code/analyzeLowDimExp.R
manuscript has lowdimviz figure now

@@ -77,7 +77,9 @@ gg <- ggplot(df,aes(x=value,fill=Gradient,color=Gradient)) +
   #ylim(c(0,0.5)) +
   coord_cartesian(ylim=c(0,0.5),xlim=c(0,25)) +
   xlab("Value") + ylab("Density") + #+ ggtitle("Empirical posterior distributions") +
-  theme_minimal()
+  theme_minimal() +
+  theme(legend.margin=margin(0,0,0,0),
+        legend.box.spacing = unit(5, "pt"))
 gg
 
 ggsave(filename="figures/lowDimVis.pdf",plot=gg,device="pdf",width = 7,height=6)

### 2023-10-06 code/highDimSim.R
fixed bug where i was not updating target after updating std dev

@@ -107,20 +107,20 @@ frstRdrCrrctd <- function(t,Q,E) {
 }
 
 lg_prr <- function(Q,tau) {
-  priors <- - abs(log(Q)/tau)^(1/4) #dnorm(log(Q),log = TRUE)
+  priors <- dnorm(log(Q),sd=tau,log = TRUE) #- abs(log(Q)/tau)^(1/4) #dnorm(log(Q),log = TRUE)
   diag(priors) <- 0
   return(sum(priors)) # TODO: add bayesian bridge log prior
 }
 
 lg_prr_grd <- function(Q,tau) {
   # start with gaussian prior
-  # gradient <- - log(Q)
-  # diag(gradient) <- 0
+  gradient <- - log(Q)/tau^2
+  diag(gradient) <- 0
   
   # BB fixing a = 0.25
-  lg_Q <- log(Q)
-  gradient <- - lg_Q / ((4*tau^2) * abs(lg_Q/tau)^(7/4))
-  diag(gradient) <- 0
+  # lg_Q <- log(Q)
+  # gradient <- - lg_Q / ((4*tau^2) * abs(lg_Q/tau)^(7/4))
+  # diag(gradient) <- 0
 
   return(gradient) 
 }
@@ -226,13 +226,11 @@ lg_lklhd_grd <- function(t,Q,input,output,method) {
 # HMC
 target <- function(Q,input,output,time,tau) {
   out <- lg_lklhd(Q,input,output,time) + lg_prr(Q,tau)
-  browser()
   return(out)
 }
 
 grad <- function(t,Q,input,output,method,tau) {
   out <- lg_lklhd_grd(t,Q,input,output,method) + lg_prr_grd(Q,tau)
-  browser()
   return(out)
 }
 
@@ -295,7 +293,6 @@ hmc <- function(t,S,input,output,method,stepSize=0.01,L=20,maxIts=1000,
     
     if (log(u) < currentU + currentK - proposedU - proposedK) {
       lg_Q[[i]]     <- proposalState
-      currentU    <- proposedU
       totalAccept[i] <- 1
       Acceptances = Acceptances + 1
     } else {
@@ -317,10 +314,13 @@ hmc <- function(t,S,input,output,method,stepSize=0.01,L=20,maxIts=1000,
       Acceptances <- 0
     }
     
-    nu <- rgamma(n=1,
-                 shape=1+(S*(S-1))*4,
[…]

### 2023-10-06 code/highDimSim.R
bug seemd to be related to use of abs()

@@ -317,12 +317,13 @@ hmc <- function(t,S,input,output,method,stepSize=0.01,L=20,maxIts=1000,
     # nu <- rgamma(n=1,
     #              shape=1+(S*(S-1))*4,
     #              rate=2+sum(abs(lg_Q[[i]])^(1/4)))
+    #taus[i] <- nu^(-4)
     taus[i] <- 1/sqrt(rgamma(n=1,
                       shape=1+(S*(S-1))*0.5,
-                      rate=2+sum(abs(lg_Q[[i]])^(2))) ) #nu^(-4)
+                      rate=2+sum((lg_Q[[i]])^(2))/2) ) 
     currentU <- - target(lg_Q_to_Q(lg_Q[[i]]),input,output,t,taus[i])
     
-    if (i %% 10 == 0) cat("Iteration ", i,"\n","stepSize: ", stepSize, "\n") 
+    if (i %% 100 == 0) cat("Iteration ", i,"\n","stepSize: ", stepSize, "\n") 
   }
   
   cat("Acceptance rate: ", sum(totalAccept)/(maxIts-1))
@@ -332,7 +333,7 @@ hmc <- function(t,S,input,output,method,stepSize=0.01,L=20,maxIts=1000,
 # test
 #set.seed(1)
 S <- 5
-maxIts <- 2000
+maxIts <- 10000
 initialStates <- sample(x=1:S,size=20,replace=TRUE)
 Q       <- matrix(rnorm(S^2,sd=1),S,S)
 Q       <- exp(Q)
@@ -349,10 +350,10 @@ chain <- hmcOut[[1]]
 
 variable <- rep(0,maxIts)
 for(i in 1:maxIts) {
-  variable[i] <- chain[[i]][2,1]
+  variable[i] <- chain[[i]][3,1]
 }
 plot(variable,type="l")
-abline(h=log(Q[2,1]),col="red")
+abline(h=log(Q[3,1]),col="red")
 
 plot(hmcOut[[3]][2:maxIts])
 # 

### 2023-10-06 code/highDimSim.R
yup that fixed it

@@ -320,7 +320,7 @@ hmc <- function(t,S,input,output,method,stepSize=0.01,L=20,maxIts=1000,
     #taus[i] <- nu^(-4)
     taus[i] <- 1/sqrt(rgamma(n=1,
                       shape=1+(S*(S-1))*0.5,
-                      rate=2+sum((lg_Q[[i]])^(2))/2) ) 
+                      rate=2+sum(abs(lg_Q[[i]])^(2),na.rm = TRUE)/2) ) 
     currentU <- - target(lg_Q_to_Q(lg_Q[[i]]),input,output,t,taus[i])
     
     if (i %% 100 == 0) cat("Iteration ", i,"\n","stepSize: ", stepSize, "\n") 

### 2023-10-07 code/highDimSim.R
collapsed sampler works, slowly

@@ -106,21 +106,21 @@ frstRdrCrrctd <- function(t,Q,E) {
   return(out)
 }
 
-lg_prr <- function(Q,tau) {
-  priors <- dnorm(log(Q),sd=tau,log = TRUE) #- abs(log(Q)/tau)^(1/4) #dnorm(log(Q),log = TRUE)
+lg_prr <- function(Q,tau,alpha) {
+  priors <- - abs(log(Q)/tau)^alpha #dnorm(log(Q),log = TRUE)
   diag(priors) <- 0
   return(sum(priors)) # TODO: add bayesian bridge log prior
 }
 
-lg_prr_grd <- function(Q,tau) {
+lg_prr_grd <- function(Q,tau,alpha) {
   # start with gaussian prior
-  gradient <- - log(Q)/tau^2
-  diag(gradient) <- 0
+  # gradient <- - log(Q)/tau^2
+  # diag(gradient) <- 0
   
   # BB fixing a = 0.25
-  # lg_Q <- log(Q)
-  # gradient <- - lg_Q / ((4*tau^2) * abs(lg_Q/tau)^(7/4))
-  # diag(gradient) <- 0
+  lg_Q <- log(Q)
+  gradient <- -  alpha * lg_Q / (tau^2) * abs(lg_Q/tau)^(alpha-2)
+  diag(gradient) <- 0
 
   return(gradient) 
 }
@@ -224,13 +224,13 @@ lg_lklhd_grd <- function(t,Q,input,output,method) {
 
 ##########################################################
 # HMC
-target <- function(Q,input,output,time,tau) {
-  out <- lg_lklhd(Q,input,output,time) + lg_prr(Q,tau)
+target <- function(Q,input,output,time,tau,alpha) {
+  out <- lg_lklhd(Q,input,output,time) + lg_prr(Q,tau,alpha)
   return(out)
 }
 
-grad <- function(t,Q,input,output,method,tau) {
-  out <- lg_lklhd_grd(t,Q,input,output,method) + lg_prr_grd(Q,tau)
+grad <- function(t,Q,input,output,method,tau,alpha) {
+  out <- lg_lklhd_grd(t,Q,input,output,method) + lg_prr_grd(Q,tau,alpha)
   return(out)
 }
 
@@ -247,7 +247,7 @@ lg_Q_to_Q <- function(lg_Q) {
 }
 
 hmc <- function(t,S,input,output,method,stepSize=0.01,L=20,maxIts=1000,
-                targetAccept=0.8, Q_init=NULL) {
[…]

### 2023-10-25 code/analyzeHighDimResults.R
new figure and remove data thats too big

@@ -1,20 +1,104 @@
 setwd("~/expmDerivative/")
 
 library(readr)
-library(coda)
+library(ggplot2)
 
-df <- read_table("data/highDimSim_30_fourth.txt", col_names = FALSE)
-N <- dim(df)[1]
+df <- read_table(paste0("data/highDimSim_",10,"_",2,".txt"),
+                       col_names = FALSE)
 P <- dim(df)[2] - 1
 df <- df[,1:P]
-df <- df[,as.vector(df[1,]!=0)]
-plot(as.vector(df[1,]),as.vector(colMeans(df)))
-abline(a=0,b=1,col="red")
-cor(as.numeric(df[1,]),as.numeric(colMeans(df)))
+df <- df[,unlist(df[1,]) != 0]
+Truth <- unlist(df[1,])
+df <- df[complete.cases(df),]
+N <- dim(df)[1]
+df2 <- data.frame(Truth=Truth,
+                  Mean=unlist(colMeans(df[ceiling(N/2):N,])),
+                  Dimension=10,
+                  Exponent=2)
+
+  
+  
+for (Dimension in c(10,20,30) ) {
+  for(alpha in c(2,1,"half","fourth")) {
+    df <- read_table(paste0("data/highDimSim_",Dimension,"_",alpha,".txt"),
+                     col_names = FALSE)
+    P <- dim(df)[2] - 1
+    df <- df[,1:P]
+    df <- df[,unlist(df[1,]) != 0]
+    Truth <- unlist(df[1,])
+    df <- df[complete.cases(df),]
+    N <- dim(df)[1]
+    
+    Exponent <- alpha
+    if(Exponent=="half") {
+      Exponent <- 0.5
+    } else if (Exponent=="fourth") {
+      Exponent <- 0.25
+    }
+    
+    df_temp <- data.frame(Truth=Truth,
+                      Mean=unlist(colMeans(df[ceiling(N/2):N,])),
+                      Dimension=Dimension,
+                      Exponent=Exponent)
+    
+    df2 <- rbind(df2,df_temp)
+  }
+}
+
+df <- df2[!duplicated(df2),]
+remove(df2)
+
+df$Dimension <- factor(df$Dimension)
+df$Exponent  <- factor(df$Exponent)
+
+df <- df[sample(1:dim(df)[1]),]
+
[…]

### 2023-10-25 code/highDimSim_30_fourth.R

@@ -120,6 +120,7 @@ lg_prr_grd <- function(Q,tau,alpha) {
   # start with gaussian prior
   # gradient <- - log(Q)/tau^2
   # diag(gradient) <- 0
+  alpha <- 0.5
   
   # BB fixing a = 0.25
   lg_Q <- log(Q)
@@ -380,8 +381,8 @@ tptStts[cbind(1:N,output)] <- 1
 tptStts <- t(tptStts)
 
 #profvis({
-hmcOut <- hmc(t=1,S=S,input=ntlStts,output=tptStts,stepSize=0.001,
-              method="approx",L=8,maxIts = maxIts, targetAccept = 0.65, Q_init=Q,
+hmcOut <- hmc(t=1,S=S,input=ntlStts,output=tptStts,stepSize=0.0001,
+              method="approx",L=8,maxIts = 1000, targetAccept = 0.65, Q_init=Q,
               alpha=0.25)
 #})
 #saveRDS(hmcOut,file="data/lowDimSimApprox.rds")

### 2023-10-25 manuscript/pnas_revision.tex
everything but the simulation
manuscript has lowdimviz figure now
add last figure
begun section

@@
 \begin{theorem}\label{thm:bride:of:frankinstein}
-  Let $\QQ \in \mathcal{S}(d)$ be non-positive, i.e.,  suppose that
+  Let symmetric $\QQ \in \mathcal{S}(d)$ be non-positive, i.e.,  suppose that
   \eqref{eq:ev:d:cond:sym:case} holds.  Take $\QQ^+$ as in
@@
 
-\section{Application: global spread of SARS-CoV-2}
+\section{Empirical Studies}
+
+\begin{figure}[!ht]
+	\centering
+\includegraphics[width=\linewidth]{lowDimVis.pdf}
+\vspace{-2em}
+\caption{All methods agree.}\label{fig:lowDim}
+\end{figure}
+
+\begin{figure*}[!t]
+	\centering
+	\includegraphics[width=\linewidth]{truthMeanFig.pdf}
+	\caption{Frobenius norm errors obtained by first-order approximation $te^{t\QQ}\JJ$ and by affine-corrected first-order approximation
+		$te^{t\QQ}\JJ - t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$
+		(\cref{thm:bride:of:frankinstein}) under increasingly relaxed
+		assumptions.  Within each assumption set, we average over 20
+		independent Monte Carlo simulations of random generator matrices
+		for each dimension. Plot \textbf{a} corresponds to asymmetric
+		generators with off-diagonal elements having independent and
+		identically distributed (iid) standard exponential random
+		variables that correspond to the sub-exponential distribution
+		hypothesis. Plot \textbf{b} drops the identical distribution
+		assumption by allowing each row and column of the generator
+		matrix to additively contribute its own mean---itself given by a
+		standard exponential---to its corresponding exponentially
+		distributed entries. Plot \textbf{c} features rate
+		matrices with iid Cauchy entries truncated to be
+		positive. Empirically, the results of
+		\cref{cor:rig:det:rand:wrapup} extend beyond the symmetric,
+		iid and sub-exponential hypotheses, suggesting scope of future work.}
+	\label{fig:truthMean}
+\end{figure*}
+
+Before applying the naive matrix exponential derivative approximation to the phylogeographic analysis of SARS-CoV-2, we carry out a few targeted studies that illustrate: the empirical performance of the approximate derivative and its affine correction \eqref{eq:better:help:on:sale} (\cref{fig:app}); agreement between CTMC generator matrix empirical posterior distributions generated by surrogate-trajectory Hamiltonian Monte Carlo (HMC) algorithms using approximate derivatives and the truth (\cref{fig:lowDim}); and point estimation of generator matrix element values under different sparsity regimes and different CTMC state space dimensionalities $d$ (\cref{fig:truthMean}).
+
+\section{Application: Global Spread of SARS-CoV-2}
 \label{sec:sars:cov:app}

### 2023-10-25 manuscript/pnas_revision.tex
finished captions

@@
 \vspace{-2em}
-\caption{All methods agree.}\label{fig:lowDim}
+\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) (\cref{sec:Sur:trad:HMC}) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix.
+The near perfect overlap reflects the fact that each algorithm's transition kernel leaves the posterior distribution invariant \cite{glatt2020accept}.
+To generate data, we randomly draw standard normal generator entries once and simulate 20 independent initial/final position pairs from a CTMC with time interval $t=1$.
+We show the true value in red and negate diagonal elements to simplify presentation.
+}\label{fig:lowDim}
 \end{figure}
@@
 	\includegraphics[width=\linewidth]{truthMeanFig.pdf}
-	\caption{Frobenius norm errors obtained by first-order approximation $te^{t\QQ}\JJ$ and by affine-corrected first-order approximation
-		$te^{t\QQ}\JJ - t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$
-		(\cref{thm:bride:of:frankinstein}) under increasingly relaxed
-		assumptions.  Within each assumption set, we average over 20
-		independent Monte Carlo simulations of random generator matrices
-		for each dimension. Plot \textbf{a} corresponds to asymmetric
-		generators with off-diagonal elements having independent and
-		identically distributed (iid) standard exponential random
-		variables that correspond to the sub-exponential distribution
-		hypothesis. Plot \textbf{b} drops the identical distribution
-		assumption by allowing each row and column of the generator
-		matrix to additively contribute its own mean---itself given by a
-		standard exponential---to its corresponding exponentially
-		distributed entries. Plot \textbf{c} features rate
-		matrices with iid Cauchy entries truncated to be
-		positive. Empirically, the results of
-		\cref{cor:rig:det:rand:wrapup} extend beyond the symmetric,
-		iid and sub-exponential hypotheses, suggesting scope of future work.}
+	\caption{Posterior means versus truth for CTMC generator matrix elements under differing sparsity regimes and different dimensionalities $d$ holding the number of observations fixed at 300.
+To affect sparsity levels, we generate generator matrix entries according to the Bayesian bridge distribution \cite{polson2014bayesian} with different exponents ($\alpha \in \{1,1/2,1/4\}$), normalizing by the largest absolute values to ease comparison.
+Smaller $\alpha$ values encode more peaked distributions with heavier tails and thus enforce greater sparsity.
+Plot \textbf{c} reflects the fact that the Bayesian bridge prior with exponent $\alpha=1/4$ helps identify non-null parameters in small sample contexts \cite{magee2023random}.
+With this intuition in mind, we specify such a prior on generator random effects in \cref{sec:sars:cov:app}.}
 	\label{fig:truthMean}

### 2023-10-25 manuscript/pnas_revision.tex
edits to caption

@@
 	\includegraphics[width=\linewidth]{truthMeanFig.pdf}
-	\caption{Posterior means versus truth for CTMC generator matrix elements under differing sparsity regimes and different dimensionalities $d$ holding the number of observations fixed at 300.
+	\caption{Posterior means versus truth for CTMC generator matrix elements within differing sparsity regimes and with different dimensionalities $d$, holding observation count fixed at 300.
+We generate posteriors using surrogate-trajectory Hamiltonian Monte Carlo with the naive matrix exponential derivative.
 To affect sparsity levels, we generate generator matrix entries according to the Bayesian bridge distribution \cite{polson2014bayesian} with different exponents ($\alpha \in \{1,1/2,1/4\}$), normalizing by the largest absolute values to ease comparison.

### 2023-10-25 manuscript/pnas_revision.tex
revision done question mark

@@
 
-\section{Empirical Studies}
-
-\begin{figure}[!ht]
+\begin{figure}[!t]
 	\centering
@@
 \end{figure}
+
+\section{Empirical Studies}
 
@@
 Before applying the naive matrix exponential derivative approximation to the phylogeographic analysis of SARS-CoV-2, we carry out a few targeted studies that illustrate: the empirical performance of the approximate derivative and its affine correction \eqref{eq:better:help:on:sale} (\cref{fig:app}); agreement between CTMC generator matrix empirical posterior distributions generated by surrogate-trajectory Hamiltonian Monte Carlo (HMC) algorithms using approximate derivatives and the truth (\cref{fig:lowDim}); and point estimation of generator matrix element values under different sparsity regimes and different CTMC state space dimensionalities $d$ (\cref{fig:truthMean}).
+
+Here, we fill in remaining simulation details not included in figure captions.
+In the simulations contributing to \cref{fig:app}, we randomly generate new, independent direction matrices $\JJ$ at each time step within each of the 20 independent runs.
+The results are not sensitive to $\JJ$ in general.
+We also note that under \cref{def:sub:exp:RV} the Cauchy distribution is \emph{not} sub-exponential and therefore represents a deviation from core assumptions of \cref{sec:rig:results}.
+Using varying distributions on the elements of generator matrices $\QQ$, the simulations contributing to \cref{fig:lowDim} and \cref{fig:truthMean} both randomly generate $N$ independent initial states $\y^0_1,\dots,\y^0_N$ according to uniform distributions over their respective CTMC state spaces.
+For each of these initial states $\y_0$, we then simulate from the CTMC for time interval $t=1$ to obtain samples $\y^1_1,\dots,\y^1_N$.
+The likelihood then takes the form $\prod_n (\y^{0}_n)^T e^{\QQ} \y^{1}_n$.
+For the simulation contributing to \cref{fig:lowDim}, we specify standard normal priors and sample from the posterior by generating 100,000 iterations from each algorithm.
+For the simulation contributing to \cref{fig:truthMean}, we generate 500,000 MCMC samples using surrogate-trajectory HMC with the naive matrix exponential derivative and calculate the posterior mean of each generator matrix element using the final 100,000 samples.
 

### 2023-10-25 manuscript/revision_letter.tex
everything but the simulation
almost ready

@@
 
+\setcounter{figure}{1}
+\begin{figure}[!ht]
+	\centering
+	\includegraphics[width=0.6\linewidth]{figures/lowDimVis.pdf}
+	\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) (Section \ref{sec:Sur:trad:HMC}) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix.
+The near perfect overlap reflects the fact that each algorithm's transition kernel leaves the posterior distribution invariant \cite{glatt2020accept}.
+To generate data, we randomly draw standard normal generator entries once and simulate 20 independent initial/final position pairs from a CTMC with time interval $t=1$.
+We show the true value in red and negate diagonal elements to simplify presentation.
+}\label{fig:lowDim}
+\end{figure}
+
 1.
@@
 \begin{reply}
-TBD (Simulation related)
-\end{reply}
+  It is important to note that surrogate-trajectory HMC using approximate gradients provides the exact same theoretical guarantees as regular HMC.
+The two algorithms generate transition kernels that leave the targeted posterior distribution invariant.
+\cite{glatt2020accept} provides a deep measure theoretic rationale for why this is the case.
+Perhaps because of these theoretical guarantees, we did not emphasize this in the original manuscript.
+
+  That said, we thank you for pointing out your concern.
+In response, we have added a new figure (Figure \ref{fig:lowDim}) in the new section ``Empirical Studies'' that helps illustrate the fact that all three algorithms sample from the same distribution.
+In order to aid comparison to the exact gradient algorithm (which scales very poorly in state space dimension) we consider a $5$-dimensional state space.
+We have also added a second new figure (Figure \ref{fig:truthMean}) to the same section that illustrates how certain Bayesian bridge priors help identify parameters even in small sample settings.
+Importantly, we use the surrogate-trajectory HMC with the naive derivative in this example.
+Finally, we note that the scalable implementation of the affine corrected approximation into BEAST phylogenetic inference software is a matter of ongoing research since it might prevent the efficient outer product coding of the approximate derivative given in Equation 18 of \citet{magee2023random}.
+\end{reply}
+
+\begin{figure*}[!t]
+	\centering
+	\includegraphics[width=\linewidth]{figures/truthMeanFig.pdf}
+	\caption{Posterior means versus truth for CTMC generator matrix elements within differing sparsity regimes and with different dimensionalities $d$, holding observation count fixed at 300.
+We generate posteriors using surrogate-trajectory Hamiltonian Monte Carlo with the naive matrix exponential derivative.
+To affect sparsity levels, we generate generator matrix entries according to the Bayesian bridge distribution \citep{polson2014bayesian} with different exponents ($\alpha \in \{1,1/2,1/4\}$), normalizing by the largest absolute values to ease comparison.
+Smaller $\alpha$ values encode more peaked distributions with heavier tails and thus enforce greater sparsity.
+Plot \textbf{c} reflects the fact that the Bayesian bridge prior with exponent $\alpha=1/4$ helps identify non-null parameters in small sample contexts \citep{magee2023random}.
+With this intuition in mind, we specify such a prior on generator random effects in Section \ref{sec:sars:cov:app}.}
+	\label{fig:truthMean}
+\end{figure*}
 
@@
 \begin{reply}
-TBD (Simulation related)
-
 Recent work \citep{gotovos2021scaling} suggests that inclusion of auxiliary states can lead to improved model fit even when these additional states do not correspond to observed data.
-In future analyses, we aim to include hundreds of global regions.
-For the time being, we have incorporated this question into the simulation study...
+Inclusion here does not change inference in any appreciable way for the generator elements that do not correspond to these extra states, but it does allow for estimating generator elements that \emph{do} correspond to them.
 \end{reply}
@@
 Figure \ref{fig:randEffs} show posterior means of the exponentiated random effects, which amount to multiplicative deviations from the fixed-effect model.
-As one might expect, only a small number of random effects deviate from 0, but those that do deviate are highly suggestive.
+As one might expect from the Bayesian bridge prior with $\alpha=1/4$ we use (Figure \ref{fig:truthMean}), only a small number of random effects deviate from 0, but those that do deviate are highly suggestive.
 We stop short of interpreting exactly why the transition from the US to Hubei, CN, exhibits largest deviation (a 1.81-fold increase).
@@
 Before responding, let us point out that we have now added a simulation section (similar to that which you specify below) to the manuscript.
-Now, we would like to argue that the decisions we made in structuring our manuscript are well-motivated based on our other work.
+Respectfully, we would like to suggest that the decisions we made in structuring our manuscript are well-motivated based on our other work.
 	\begin{enumerate}
@@
 Again, we thank you for sharing your concerns and hope that the broader context of our work helps explain the positioning of our manuscript.
-Probably more importantly, we have updated our manuscript with a thorough simulation study (see below).
+Probably more importantly, we have updated our manuscript with two empirical studies in response to your valid questions (see below).
 \end{reply}
@@
 
-\begin{reply}
-Simulaiton. TBD.
+\setcounter{figure}{0}
+\begin{figure*}[!t]
+	\centering
+	\includegraphics[width=\linewidth]{figures/appendixFig3.pdf}
+	\caption{Frobenius norm errors obtained by first-order approximation $te^{t\QQ}\JJ$ and by affine-corrected first-order approximation
+		$te^{t\QQ}\JJ - t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$
+		(Theorem \ref{thm:bride:of:frankinstein}) under increasingly relaxed
+		assumptions.  Within each assumption set, we average over 20
+		independent Monte Carlo simulations of random generator matrices
+		for each dimension. Plot \textbf{a} corresponds to asymmetric
+		generators with off-diagonal elements having independent and
+		identically distributed (iid) standard exponential random
+		variables that correspond to the sub-exponential distribution
+		hypothesis. Plot \textbf{b} drops the identical distribution
+		assumption by allowing each row and column of the generator
+		matrix to additively contribute its own mean---itself given by a
+		standard exponential---to its corresponding exponentially
+		distributed entries. Plot \textbf{c} features rate
+		matrices with iid Cauchy entries truncated to be
+		positive. Empirically, the results of
+		Corollary \ref{cor:rig:det:rand:wrapup} extend beyond the symmetric,
+		iid and sub-exponential hypotheses, suggesting scope of future work.}
+	\label{fig:app}
+\end{figure*}
+
+\setcounter{figure}{1}
+\begin{figure}[!ht]
+	\centering
+	\includegraphics[width=0.6\linewidth]{figures/lowDimVis.pdf}
+	\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) (Section \ref{sec:Sur:trad:HMC}) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix.
+The near perfect overlap reflects the fact that each algorithm's transition kernel leaves the posterior distribution invariant \cite{glatt2020accept}.
+To generate data, we randomly draw standard normal generator entries once and simulate 20 independent initial/final position pairs from a CTMC with time interval $t=1$.
+We show the true value in red and negate diagonal elements to simplify presentation.
+}\label{fig:lowDim}
+\end{figure}
+
+\begin{figure*}[!t]
+	\centering
+	\includegraphics[width=\linewidth]{figures/truthMeanFig.pdf}
+	\caption{Posterior means versus truth for CTMC generator matrix elements within differing sparsity regimes and with different dimensionalities $d$, holding observation count fixed at 300.
+We generate posteriors using surrogate-trajectory Hamiltonian Monte Carlo with the naive matrix exponential derivative.
+To affect sparsity levels, we generate generator matrix entries according to the Bayesian bridge distribution \citep{polson2014bayesian} with different exponents ($\alpha \in \{1,1/2,1/4\}$), normalizing by the largest absolute values to ease comparison.
+Smaller $\alpha$ values encode more peaked distributions with heavier tails and thus enforce greater sparsity.
+Plot \textbf{c} reflects the fact that the Bayesian bridge prior with exponent $\alpha=1/4$ helps identify non-null parameters in small sample contexts \citep{magee2023random}.
+With this intuition in mind, we specify such a prior on generator random effects in Section \ref{sec:sars:cov:app}.}
+	\label{fig:truthMean}
+\end{figure*}
+
+\begin{reply}
+We thank you for sharing your concerns.
+In response, we have added a new section, ``Empirical Studies'' to the manuscript.
+The section contains three figures, two of which are new and are intended to respond to your concerns.
+We include the section text and figures in this letter.
+
+\begin{quote}
+	Before applying the naive matrix exponential derivative approximation to the phylogeographic analysis of SARS-CoV-2, we carry out a few targeted studies that illustrate: the empirical performance of the approximate derivative and its affine correction \eqref{eq:better:help:on:sale} (Figure \ref{fig:app}); agreement between CTMC generator matrix empirical posterior distributions generated by surrogate-trajectory Hamiltonian Monte Carlo (HMC) algorithms using approximate derivatives and the truth (Figure \ref{fig:lowDim}); and point estimation of generator matrix element values under different sparsity regimes and different CTMC state space dimensionalities $d$ (Figure \ref{fig:truthMean}).
+
+	Here, we fill in remaining simulation details not included in figure captions.
+In the simulations contributing to Figure \ref{fig:app}, we randomly generate new, independent direction matrices $\JJ$ at each time step within each of the 20 independent runs.
+The results are not sensitive to $\JJ$ in general.
+We also note that under Definition \ref{def:sub:exp:RV} the Cauchy distribution is \emph{not} sub-exponential and therefore represents a deviation from core assumptions of Section \ref{sec:rig:results}.
+Using varying distributions on the elements of generator matrices $\QQ$, the simulations contributing to Figure \ref{fig:lowDim} and Figure \ref{fig:truthMean} both randomly generate $N$ independent initial states $\y^0_1,\dots,\y^0_N$ according to uniform distributions over their respective CTMC state spaces.
+For each of these initial states $\y_0$, we then simulate from the CTMC for time interval $t=1$ to obtain samples $\y^1_1,\dots,\y^1_N$.
+The likelihood then takes the form $\prod_n (\y^{0}_n)^T e^{\QQ} \y^{1}_n$.
+For the simulation contributing to Figure \ref{fig:lowDim}, we specify standard normal priors and sample from the posterior by generating 100,000 iterations from each algorithm.
+For the simulation contributing to Figure \ref{fig:truthMean}, we generate 500,000 MCMC samples using surrogate-trajectory HMC with the naive matrix exponential derivative and calculate the posterior mean of each generator matrix element using the final 100,000 samples.
+\end{quote} 
 \end{reply}
@@
 \begin{reply}
-TBD	
+It is symmetric.  We have amended the Theorem's first sentence to say
+\begin{quote}
+	Let symmetric $\QQ \in \mathcal{S}(d)$ be non-positive, i.e.,  suppose that
+	\eqref{eq:ev:d:cond:sym:case} holds.
+\end{quote}
 \end{reply}

### 2023-10-25 manuscript/revision_letter.tex
right before sending to collabs

@@
 
-To the Editor and Reviewers:
+Dear Editor and Reviewers:
 
@@
 \begin{itemize}
-	\item clarified how the present work relates to our other work that includes thorough simulation studies \citet{magee2023random} and provides strong theoretical guarantees for surrogate-trajectory HMC \citet{glatt2020accept};
-
-	\item implemented an additional simulation study which compares accuracy for multiple algorithms, different model sizes and different signal strength regimes;
+	\item clarified how the present work relates to our other work that includes thorough simulation studies \citep{magee2023random} and provides strong theoretical guarantees for surrogate-trajectory HMC \citep{glatt2020accept};
+
+	\item implemented two additional simulation studies which compare posteriors obtained for multiple algorithms and illustrate parameter identification  for different model sizes and different sparsity regimes;
 
@@
 	Thank you for the opportunity to submit a revised manuscript.
-We have made ample additions to the manuscript in response to the suggestions made by both reviewers.
-Among the many improvements we have made to this manuscript, the simulation study suggested by Reviewer 2 helps the paper smoothly transition between theory and application.
+We have made ample additions to the manuscript in response to the helpful suggestions made by both reviewers.
+Among the many improvements we have made to this manuscript, the simulation studies suggested by Reviewer 2 help the paper smoothly transition between theory and application.
 \end{reply}
@@
 \cite{glatt2020accept} provides a deep measure theoretic rationale for why this is the case.
-Perhaps because of these theoretical guarantees, we did not emphasize this in the original manuscript.
+Perhaps because of these theoretical guarantees, we did not emphasize such a comparison in the original manuscript.
 

### 2023-10-27 code/analyzeHighDimResults.R
updated postmeans fig

@@ -12,7 +12,7 @@ Truth <- unlist(df[1,])
 df <- df[complete.cases(df),]
 N <- dim(df)[1]
 df2 <- data.frame(Truth=Truth,
-                  Mean=unlist(colMeans(df[ceiling(N/2):N,])),
+                  Mean=unlist(colMeans(df[4001:5000,])),
                   Dimension=10,
                   Exponent=2)
 
@@ -37,7 +37,7 @@ for (Dimension in c(10,20,30) ) {
     }
     
     df_temp <- data.frame(Truth=Truth,
-                      Mean=unlist(colMeans(df[ceiling(N/2):N,])),
+                      Mean=unlist(colMeans(df[4001:5000,])),
                       Dimension=Dimension,
                       Exponent=Exponent)
     
@@ -51,6 +51,7 @@ remove(df2)
 df$Dimension <- factor(df$Dimension)
 df$Exponent  <- factor(df$Exponent)
 
+set.seed(1)
 df <- df[sample(1:dim(df)[1]),]
 
 gg <- ggplot(df[df$Exponent==0.25,],aes(x=Truth,y=Mean,color=Dimension)) +

### 2023-10-27 manuscript/revision_letter.tex
marcs suggestions and figure numbering fixes in the letter

@@
 
+1.
+The authors spend 3 out of 7 pages in the main text and three corresponding sections in the Appendix discussing the refined error and the proof, while postpone the application with the affine-corrected error to future work.
+I would expect more than a relatively abstract error comparison figure for this issue.
+For example, a simple numerical simulation similar to the last paragraph in Application to compute the posterior likelihood ratio or posterior density ratio of the corrected and the uncorrected methods, because the authors eventually applied the method within a Bayesian framework.
+
+\begin{reply}
+  It is important to note that surrogate-trajectory HMC using approximate gradients provides the exact same theoretical guarantees as regular HMC.
+The two algorithms generate transition kernels that leave the targeted posterior distribution invariant.
+\cite{glatt2020accept} provides a deep measure theoretic rationale for why this is the case.
+Perhaps because of these theoretical guarantees, we did not emphasize such a comparison in the original manuscript.
+
+  That said, we thank you for pointing out your concern.
+In response, we have added a new figure (Figure \ref{fig:lowDim}) in the new section ``Empirical Studies'' that helps illustrate the fact that all three algorithms sample from the same distribution.
+In order to aid comparison to the exact gradient algorithm (which scales very poorly in state space dimension) we consider a $5$-dimensional state space.
+We have also added a second new figure (Figure \ref{fig:truthMean}) to the same section that illustrates how certain Bayesian bridge priors help identify parameters even in small sample settings.
+Importantly, we use the surrogate-trajectory HMC with the naive derivative in this example.
+\end{reply}
+
 \setcounter{figure}{1}
@@
 \end{figure}
-
-1.
-The authors spend 3 out of 7 pages in the main text and three corresponding sections in the Appendix discussing the refined error and the proof, while postpone the application with the affine-corrected error to future work.
-I would expect more than a relatively abstract error comparison figure for this issue.
-For example, a simple numerical simulation similar to the last paragraph in Application to compute the posterior likelihood ratio or posterior density ratio of the corrected and the uncorrected methods, because the authors eventually applied the method within a Bayesian framework.
-
-\begin{reply}
-  It is important to note that surrogate-trajectory HMC using approximate gradients provides the exact same theoretical guarantees as regular HMC.
-The two algorithms generate transition kernels that leave the targeted posterior distribution invariant.
-\cite{glatt2020accept} provides a deep measure theoretic rationale for why this is the case.
-Perhaps because of these theoretical guarantees, we did not emphasize such a comparison in the original manuscript.
-
-  That said, we thank you for pointing out your concern.
-In response, we have added a new figure (Figure \ref{fig:lowDim}) in the new section ``Empirical Studies'' that helps illustrate the fact that all three algorithms sample from the same distribution.
-In order to aid comparison to the exact gradient algorithm (which scales very poorly in state space dimension) we consider a $5$-dimensional state space.
-We have also added a second new figure (Figure \ref{fig:truthMean}) to the same section that illustrates how certain Bayesian bridge priors help identify parameters even in small sample settings.
-Importantly, we use the surrogate-trajectory HMC with the naive derivative in this example.
-Finally, we note that the scalable implementation of the affine corrected approximation into BEAST phylogenetic inference software is a matter of ongoing research since it might prevent the efficient outer product coding of the approximate derivative given in Equation 18 of \citet{magee2023random}.
-\end{reply}
 
@@
 
-\setcounter{figure}{4}
+\setcounter{figure}{6}
 \begin{figure}[t]
@@
 
-\setcounter{figure}{3}
+\setcounter{figure}{5}
 \begin{figure*}[!t]
@@
 
-In short, our original goal was not to prove validity of the general methodology, which we saw as already established through simulation studies, real-world analyses and measure theoretic arguments.
+In short, our original goal was not to prove validity of surrogate-trajectory HMC, which we saw as already established through simulation studies, real-world analyses and measure theoretic arguments.
 

### 2023-10-30 manuscript/pnas_revision.tex
ready to ship

@@
 
-In \cref{sec:sars:cov:app} we follow these theoretical investigations
+ \cref{sec:emp}, contains simulation studies comparing: accuracy of matrix exponential derivative approximations for different distributional assumptions on the generator matrix; the posterior distributions obtained using surrogate-trajectory Hamiltonian Monte Carlo \cref{sec:Sur:trad:HMC} and traditional Hamiltonian Monte Carlo; and parameter identification under different priors on generator matrix elements.
+
+In \cref{sec:sars:cov:app} we follow these theoretical and empirical investigations
 with an application of the naive, first-order gradient approximation
@@
   \end{align}
-  or we impose that $\QQ \in \mathcal{R}(d) \cap \mathcal{S}(d, \RR)$ as 
+  or we impose that $\QQ \in \mathcal{R}(d) \cap \mathcal{S}(d)$ as 
     \begin{align}
@@
   \eqref{eq:bride:of:frankinstein:1:rm:app},
-  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond
+  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond the
   limitations of \cref{cor:rig:det:rand:wrapup} in several different
@@
 
-\section{Empirical Studies}
+\section{Empirical Studies}\label{sec:emp}
 
@@
 $\ppi=(\pi_1,\dots,\pi_d)=\lim_{t\rightarrow
-  \infty}\widetilde{\ppi}e^{t\QQ}$, for $\widetilde{\ppi}$ for any
+  \infty}\widetilde{\ppi}e^{t\QQ}$, for $\widetilde{\ppi}$ any
 arbitrary probability vector.  Examples of characters are the $d=4$
@@
   $\ttheta\in\mathbb{R}^K$ given observed data $\Y$.}  We proceed by
-considering a `potential energy' of the form
+considering a potential energy of the form
 $U(\ttheta):=-\log \pi(\ttheta)$.  We then select an associated
-potential energy $V(\mom) := | \G^{-1/2} \mom|^2$ for an appropriately
+kinetic energy $V(\mom) := | \G^{-1/2} \mom|^2$ for an appropriately
 chosen symmetric-positive-definite mass matrix $\G$ (which is often

### 2023-10-30 manuscript/revision_letter.tex
ready to ship
really ready to ship

@@
 This is a good question.
-Figure \ref{fig:randEffs} show posterior means of the exponentiated random effects, which amount to multiplicative deviations from the fixed-effect model.
+Figure \ref{fig:randEffs} shows posterior means of the exponentiated random effects, which amount to multiplicative deviations from the fixed-effect model.
 As one might expect from the Bayesian bridge prior with $\alpha=1/4$ we use (Figure \ref{fig:truthMean}), only a small number of random effects deviate from 0, but those that do deviate are highly suggestive.
@@
 From our other previous work with surrogate-trajectory HMC \citep{li2019neural}, we knew this must suggest that the first-order matrix exponential gradient approximation must be pretty good.
+
 As a result, our goal in this paper was to find out what makes the first-order approximation so good and provide a salient example that illustrates the efficacy of the approximation in higher dimensions.
-
 In short, our original goal was not to prove validity of surrogate-trajectory HMC, which we saw as already established through simulation studies, real-world analyses and measure theoretic arguments.

### 2023-10-30 manuscript/pnas_revision.tex
final edits before resubmission

@@
 
- \cref{sec:emp}, contains simulation studies comparing: accuracy of matrix exponential derivative approximations for different distributional assumptions on the generator matrix; the posterior distributions obtained using surrogate-trajectory Hamiltonian Monte Carlo \cref{sec:Sur:trad:HMC} and traditional Hamiltonian Monte Carlo; and parameter identification under different priors on generator matrix elements.
+ \cref{sec:emp} contains simulation studies comparing: accuracy of matrix exponential derivative approximations for different distributional assumptions on the generator matrix; the posterior distributions obtained using surrogate-trajectory Hamiltonian Monte Carlo (\cref{sec:Sur:trad:HMC}) and traditional Hamiltonian Monte Carlo; and parameter identification under different priors on generator matrix elements.
 

### 2023-12-06 manuscript/pnas_revision.tex
changes in response to quality control check at pnas
final manuscript file with appendix and references to appendix removed

--- 2023-10-30 manuscript/pnas_revision.tex
@@
 
-\input{make-edits}
+%\input{make-edits}
 
@@
 % Please include corresponding author, author contribution and author declaration information
-% \authorcontributions{Please provide details of author contributions here.}
+ \authorcontributions{AJH and MAS designed research. GD, NEGH and AJH developed mathematical results. AJH and AFM performed simulations and data analysis.  GD, NEGH and AJH wrote the paper.}
 % \authordeclaration{Please declare any competing interests here.}
@@
 \cite{magee2023random} show that this approximation helps reduce total cost to $\order{Kd^2 + Nd^3}$ and use this speedup within
-surrogate-trajectory Hamiltonian Monte Carlo (see \cref{sec:Sur:trad:HMC}) 
+surrogate-trajectory Hamiltonian Monte Carlo (see Supplement) 
 to obtain a
@@
 
- \cref{sec:emp}, contains simulation studies comparing: accuracy of matrix exponential derivative approximations for different distributional assumptions on the generator matrix; the posterior distributions obtained using surrogate-trajectory Hamiltonian Monte Carlo \cref{sec:Sur:trad:HMC} and traditional Hamiltonian Monte Carlo; and parameter identification under different priors on generator matrix elements.
+ \cref{sec:emp} contains simulation studies comparing: accuracy of matrix exponential derivative approximations for different distributional assumptions on the generator matrix; the posterior distributions obtained using surrogate-trajectory Hamiltonian Monte Carlo and traditional Hamiltonian Monte Carlo; and parameter identification under different priors on generator matrix elements.
 
@@
 \cref{cor:rig:det:rand:wrapup}, the proofs of which appear in the
-supplemental material below.
+Supplement.
 
@@
 These observations lead to the following theorem, the proof of which appears
-in the supplement as \cref{sec:Thm:det:main}.
+in the Supplement.
 \begin{theorem}\label{thm:ass:error}
@@
       \end{align}
-      almost surely. In \eqref{e:lambda_n-1(Q)->-infty_non-symm},
-      $O_{\textnormal{a.s.}}$ is as in
-      \eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
+      almost surely. %In \eqref{e:lambda_n-1(Q)->-infty_non-symm},
+      %$O_{\textnormal{a.s.}}$ is as in
+      %\eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
 \end{theorem}
@@
 	\vspace{-2em}
-	\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) (\cref{sec:Sur:trad:HMC}) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix.
+	\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix.
 The near perfect overlap reflects the fact that each algorithm's transition kernel leaves the posterior distribution invariant \cite{glatt2020accept}.
@@
 \eqref{eq:firstOrder} of the matrix exponential derivative
-\eqref{eq:grad:def} performs remarkably well within surrogate HMC (see
-\cref{sec:Sur:trad:HMC} in the Supplement), achieving an over 30-fold
+\eqref{eq:grad:def} performs remarkably well within surrogate HMC (see Supplement), achieving an over 30-fold
 efficiency gain compared to random-walk Metropolis for a 14-state
@@
 
-\appendix
-
-\section{Proof of \cref{thm:ass:error}}
-\label{sec:Thm:det:main}
-
-Recall that $\Z(t)$ is defined as in \eqref{eq:true:diff}, and
-consider the representation \eqref{eq:ass:procs:0}. We begin by
-decomposing the first term $e^{-s\QQ} \JJ e^{s \QQ}$ inside of the
-integral in \eqref{eq:ass:procs:0} as
-\begin{align}
-  e^{-s\QQ} &\JJ e^{s \QQ} \notag\\
-  =&e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
-     + e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
-                  \notag\\
-      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+}) e^{s\QQ}
-                \notag\\
-       &+e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}) e^{s\QQ}
-                           \notag\\
-  = &e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
-      + (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
-         \label{eq:big:yuck:1}                          
-      \\
-      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
-      %\notag\\
-  + (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}),
-  \notag
-\end{align}
-where we used \eqref{eq:MP:proj:prop} for the second identity.
-Regarding the middle two terms, with our commutativity
-\eqref{eq:GI:com} assumption, we observe that
-\begin{align}
-  (\II &-\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
-         + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
-                \label{eq:big:yuck:2}                          
-                     \\
-  &=  (\II -\QQ^+ \QQ) \JJ \QQ^{+} \QQ e^{s\QQ}
-    + e^{-s\QQ}\QQ \QQ^+ \JJ (\II - \QQ \QQ^{+})
-    \notag\\
-  &= \frac{d}{ds}
-    \left(
-    (\II -\QQ^+ \QQ) \JJ \QQ^{+} e^{s\QQ}
-    - e^{-s\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
-    \right).
- \notag
-\end{align}
-Hence, integrating in time, and then using \eqref{eq:GI:prop},
-\eqref{eq:GI:com} for the first and last term terms on the right hand side, we have
-\begin{align}
-  \int_0^t& \big\{(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
-                             + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})\big\}ds
-                             \notag\\
-   =& (\II -\QQ^+ \QQ) \JJ \QQ^{+}e^{t\QQ}
-      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
-      \notag\\
-    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
-      +  \QQ^+ \JJ (\II - \QQ \QQ^{+})
-      \notag\\
-   =& (\II -\QQ^+ \QQ) \JJ (\QQ^+)^2 \QQ e^{t\QQ}
-      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
-      \notag\\
-    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
-      + \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+})
-             \label{eq:big:yuck:3}                          
-\end{align}
-Finally let us note, regarding the second term on the right hand side
-of \eqref{eq:ass:procs:0}, using again 
-\eqref{eq:MP:proj:prop} and \eqref{eq:GI:prop}, \eqref{eq:GI:com}
-\begin{align}
-  -te^{t\QQ}\JJ 
-  =-te^{t\QQ}\QQ \QQ^+ \JJ  - t (\II - \QQ^+\QQ) \JJ.
-               \label{eq:small:yuck:4}                          
-\end{align}
-
-By combining \eqref{eq:big:yuck:1}--\eqref{eq:small:yuck:4} and
-comparing with \eqref{eq:ass:procs:0}, we now find that
-\begin{align}\label{e:E(t)=integ+sum}
-  \Z(t)\! =\! \! \int_0^t \! \! \! e^{(t-s)\QQ} &\QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds
-           + (\II - \!\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}
-           \notag\\
-           &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ 
-           \notag\\
-           &-\QQ^+ \JJ (\II - \QQ \QQ^{+})- (\II -\QQ^+ \QQ) \JJ \QQ^{+}
-             \notag\\
-         &-t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}.
-\end{align}
-Rearranging \eqref{e:E(t)=integ+sum}, we have 
-\begin{align*}
-\QQ^+ \JJ (\II - \QQ \QQ^{+}) &+(\II -\QQ^+ \QQ) \JJ \QQ^{+}
-   +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
-                                \notag\\
-  &+  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ
-  = \TT_1 + \TT_2,
-\end{align*}
-where 
-\begin{align*}
- \TT_1 :=&  \int_0^t  \! \! e^{(t-s)\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds,\\
-  \TT_2 :=&  (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}\\
-         &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ.
-\end{align*}
-
-We estimate each of $\TT_1$ and $\TT_2$ in turn.  Regarding $\TT_1$,
-using once more \eqref{eq:GI:com}, as well as \eqref{eq:qetq:exp:decay}, we have
-\begin{align*}
-  \| \TT_1\| \leq&
-    \int_0^t \|\QQ e^{(t-s)\QQ}\| \| \QQ^+  \JJ \QQ^{+} \|\| \QQ e^{s\QQ}\|ds
-         \notag\\
-   \leq& C_0^2\int_0^t  e^{-(t-s)\kappa} \| \QQ^+  \JJ \QQ^{+} \| e^{-s\kappa } ds
-                         \notag\\
-   =& C_0^2 t e^{-t\kappa} \| \QQ^+  \JJ  \QQ^{+} \|.
-\end{align*}
-Turning to $\TT_2$, by \eqref{eq:qetq:exp:decay} we have
-\begin{align*}
-  \| \TT_2\|
-  \leq C_0 (&\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
-               + \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
-               \notag\\
-               &+ t\|\QQ^+ \hspace{0.25mm}\JJ\|) e^{- t \kappa }.
-\end{align*}
-Combining these two bound completes the proof.
-
-\section{Proof of \cref{thm:bride:of:frankinstein}}
-
-This result is established from \cref{thm:ass:error} and some basic
-properties $\QQ$ under the given assumptions. First, note that, cf.
-\eqref{eq:ev:d:cond:sym:case}, (\ref{eq:EV:real:ass})
-\begin{equation}\label{e:max_|lambda-j(Q)|=<|lambda-1(Q)|}
-  |\lambda_d(\QQ)| \leq \max_{j =1, \ldots,d_{-}} |\lambda_{j}(\QQ)| \leq |\lambda_1(\QQ)|.
-\end{equation}
-Referring back to
-\eqref{eq:frob:norm}, by \eqref{e:max_|lambda-j(Q)|=<|lambda-1(Q)|} we have the estimate
-\begin{align}\label{eq:frob:sucks}
-  \| \QQ e^{t\QQ} \|_F
-  =&  \left( \sum_{j =1}^{d_{-}}
-     | \lambda_j(\QQ) e^{ t  \lambda_j(\QQ)  } |^2 \right)^{1/2}
-     \notag\\
-  \leq& \sqrt{d_{-}} |\lambda_{1}(\QQ)| e^{-t |\lambda_{d_{-}}(\QQ)|}.
-\end{align}
-Similarly, from \eqref{eq:op:norm} and \eqref{e:max_|lambda-j(Q)|=<|lambda-1(Q)|}, we have that
-\begin{align}\label{eq:op:better}
-  \| \QQ e^{t\QQ} \|_{\op}
-  \leq& \max_{j =1, \ldots,d_{-}}
-        |\lambda_{j}(\QQ)| e^{-t |\lambda_{j}(\QQ)|}
-        \notag\\
-  \leq& |\lambda_{1}(\QQ)| e^{-t |\lambda_{d_{-}}(\QQ)|}.
-\end{align}
-Likewise, we observe that
-\begin{align}\label{eq:thm1:extra:trms:1:F}
-  \lVert \QQ\rVert^2_F = \sum_{k=1}^{d} \lambda_k(\QQ)^2 \, , \quad
-  \lVert \QQ^+ \rVert^2_F = \sum_{k =1}^{d_-}\frac{1}{\lambda_k(\QQ)^2} \, ,
-\end{align}
-and that
-\begin{align}\label{eq:thm1:extra:trms:1:F}
-  \lVert \QQ\rVert^2_{\op} = \lambda_1(\QQ)^2 \, , \quad
-  \lVert \QQ^+ \rVert^2_{\op} = \frac{1}{\lambda_{d_{-}}(\QQ)^2} \,.
-\end{align}
-Let $\UU$ be as in \eqref{eq:MP:inv:sym:case}. Noting, furthermore,
-that $\II - \QQ \QQ^+ = \UU(\II - \II_{d_-})\UU^*$, where $\II_{d_-}$
-is the matrix with $1$'s along the first $d_-$ diagonal elements and
-zero otherwise, we have the bounds
-\begin{align}\label{eq:proj:mat:bnds}
-  \|\II - \QQ  \QQ^+\|_{F} = \sqrt{d - d_{-}},
-  \quad
-  \| \II - \QQ  \QQ^+\|_{\op} = 1.
-\end{align}
-
-Now, for any matrix norm $\|\cdot\|$, \eqref{eq:ass:error} implies
-that
-\begin{align}
-    \| t(\II -\QQ^+ \QQ) &\JJ \QQ \QQ^{+}
-                           +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\|
-                           \notag\\
-  &\leq C(1+ t)e^{-\kappa t}
-    + 2\|   \QQ^+\| \|\JJ\| \|\II - \QQ \QQ^{+}\|
-    \label{eq:ass:error_proof}
-\end{align}    
-for any $t \geq 0$, where $C > 0$ is given by
-\eqref{eq:exp:error:ass:er}. Thus, for the choice of norm
-$\|\cdot\| = \|\cdot\|_{F}$, by \eqref{eq:ass:error_proof},
-\eqref{eq:frob:sucks}, \eqref{eq:thm1:extra:trms:1:F} and
-\eqref{eq:proj:mat:bnds}, we obtain
-\eqref{eq:bride:of:frankinstein:1}.  Similarly, for the choice of norm
-$\|\cdot\| = \|\cdot\|_{\op}$, \eqref{eq:ass:error_proof}, in
-conjunction with \eqref{eq:op:better}, \eqref{eq:thm1:extra:trms:1:F}
-and \eqref{eq:proj:mat:bnds}, yields the bound
-\eqref{eq:bride:of:frankinstein}.  Since
-\eqref{eq:bride:of:frankinstein:1:rm:app} follows immediately from
-\eqref{eq:bride:of:frankinstein:1}, and similarly between
-\eqref{eq:bride:of:frankinstein:0:rm:app} and
-\eqref{eq:bride:of:frankinstein}, the proof is now complete.
-
-\section{Proof of \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}}
-\label{sec:Thm:rand:mat:main}
-
-For use in this section, we introduce a probabilistic version of the
-typical $O$ and $o$ asymptotic notations.  Given collections of random
-variable $\{X_d\}_{d \in \bbN}$, $\{Y_d\}_{d \in \bbN}$, we write
-\begin{equation}\label{e:X=O_a.s.(1)_Y=o_a.s.(1)}
-X_d = O_{\as}(f(d)),\quad Y_d  = o_{\as}(f(d)), 
-\end{equation}
-for some $f: \bbN \to \RR^+$ to mean, respectively, that there exists
-a random variable $C = C(\omega)$, not dependent on $d$, such
-that $|X_d|/f(d) \leq C$ for all $d \in \bbN$ a.s., and also that
-$\lim_{d \rightarrow \infty}Y_d/f(d) = 0$ a.s.
-
-In everything that follows we will make use of the so-called
-\textit{Weyl's inequalities} (see e.g.
-\cite[Theorem 4.3.1, p.\ 239]{horn:johnson:2012}). Namely, let
-${\mathbf A},{\mathbf B} \in {\mathcal S}(d,\bbC)$, each with its own
-eigenvalues listed in increasing order as in our convention
-\eqref{eq:EV:real:ass}. Then, for $i = 1,\hdots,d$,
-\begin{equation}\label{e:Weyl_upper}
-  \lambda_i({\mathbf A}+{\mathbf B}) \leq
-  \lambda_{i+j}({\mathbf A}) + \lambda_{d-j}({\mathbf B}),
-  \quad j = 0,1,\hdots,d-i,
-\end{equation}
-and
-\begin{equation}\label{e:Weyl_lower}
-  \lambda_{i-j+1}({\mathbf A}) + \lambda_j({\mathbf B})
-  \leq \lambda_i({\mathbf A}+{\mathbf B}), \quad j = 1,\hdots,i.
-\end{equation}
-In what follows, we also make use of the fact that, for ${\mathbf A} \in {\mathcal S}(d,\bbC)$,
-\begin{equation}\label{e:eigenvals_of_-A}
-  \lambda_\ell(-{\mathbf A}) = - \lambda_{d-\ell+1}({\mathbf A}),
-  \quad \ell = 1,\hdots,d.
-\end{equation}
-
-With these preliminaries now in hand, we turn to the proof of
-\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.  This result is
-established with the aide of four auxiliary results, \cref{l:||Q||/n=O(1/n^(1/2-eta)},
-\cref{l:max_Yn/n^eps}, \cref{l:||Q_4||=o_P(1)} and \cref{l:lambda-2(Q3)>=1+1/n}, whose
-precise statements and proof are provided immediately afterward.
-%in \cref{sec:aux}.
-\begin{proof}
-  To demonstrate our desired result,
-  \eqref{e:lambda_n-1(Q)->-infty_non-symm}, it suffices to establish
-  the lower bound
-  \begin{equation}\label{e:sigma^2_2(Q)>=_a.s._bound}
-    \sigma^2_2(\QQ) \geq
-    (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
-    + O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)\Big\}
-  \end{equation}
-  as well as the upper bound
-  \begin{equation}\label{e:sigma^2_2(Q)=<_a.s._bound}
-    \sigma^2_d(\QQ) \leq
-    (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
-    + O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)\Big\}.
-  \end{equation}
-  With this in mind, we now decompose $\QQ$ as follows, starting with
-  the case \eqref{e:RM_theorem_general_condition}.  Fix
-  $d \in \bbN \backslash\{1\}$.  Taking $F_X$ to be the distribution
-  defining the elements in \eqref{e:RM_theorem_general_condition}, draw
-  \begin{align}\label{eq:diag:fill:trk}
-    \{\widetilde{q}_{ii}\}_{i=1,\hdots,d}
-    \stackrel{\textnormal{iid}}\sim F_X
-    \text{ independently of } \{q_{ij}\}_{i,j=1,\hdots,d}.
-   \end{align} 
-   Here note carefully that $\widetilde{q}_{ii} \stackrel{d}= q_{ij}$
-   and $\widetilde{q}_{ii} \stackrel{d}\neq q_{ii}$.  Now recast
-   $\QQ = \{q_{ij}\}_{i,j}$ as %{ \scriptsize
-\begin{align}
-  & \begin{pmatrix}
-0 & q_{12}-\mu & \hdots & q_{1d}-\mu \\
-q_{21}-\mu & 0 & \hdots & q_{2d}-\mu \\
-\vdots & \vdots & \ddots & \vdots \\
-q_{d1}-\mu & q_{d2}-\mu & \hdots & 0 \\
-\end{pmatrix}
-+
-\begin{pmatrix}
-q_{11} & \mu & \hdots & \mu \\
-\mu & q_{22} & \hdots & \mu \\
-\vdots & \vdots & \ddots & \vdots \\
-\mu & \mu & \hdots & q_{dd} \\
-\end{pmatrix} \notag \\
-  &=\left\{\!
-    \begin{pmatrix}
-\widetilde{q}_{11}-\mu & q_{12}-\mu & \hdots & q_{1d}-\mu \\
-q_{21}-\mu & \widetilde{q}_{22}-\mu & \hdots & q_{2d}-\mu \\
-\vdots & \vdots & \ddots & \vdots \\
-q_{d1}-\mu & q_{d2}-\mu & \hdots & \widetilde{q}_{dd}-\mu \\
-\end{pmatrix}\right.
-  \label{eq:En:or:Wigner}\\
-& \qquad \quad \quad +
-\left.\begin{pmatrix}
--\widetilde{q}_{11}+\mu & 0 & \hdots & 0 \\
-0 & -\widetilde{q}_{22}+\mu & \hdots & 0 \\
-\vdots & \vdots & \ddots & \vdots \\
-0 & 0 & \hdots & -\widetilde{q}_{dd}+\mu \\
-\end{pmatrix}\!\right\}
-  \notag \\
-  &\quad\;+ (1-d)
-    \left\{ \!
-    \left(\! \!\!\!
-      \begin{array}{*{24}{c@{\hspace{2pt}}}}
-\frac{-q_{11}-(d-1)\mu}{d-1} & 0 & \hdots & 0 \\
-0 & \frac{-q_{22}-(d-1)\mu}{d-1} & \hdots & 0 \\
-\vdots & \vdots & \ddots & \vdots \\
-0 & 0 & \hdots & \frac{-q_{dd}-(d-1)\mu}{d-1}\\
-    \end{array}
-  \! \!  \right)
-  \right.
-  \notag\\ 
-  &\qquad \qquad \qquad \quad+  \left. 
-        \begin{pmatrix}
-\mu & \frac{-\mu}{d-1} & \hdots & \frac{-\mu}{d-1} \\
-\frac{-\mu}{d-1} & \mu & \hdots & \frac{-\mu}{d-1} \\
-\vdots & \vdots & \ddots & \vdots \\
-\frac{-\mu}{d-1} & \frac{-\mu}{d-1} & \hdots & \mu \\
-\end{pmatrix}
-  \!\right\}
-  \label{e:Q_basic_decomp}\\
-  &=: \QQ_1 + \QQ_2 + (1-d) \big\{\QQ_3 +  \QQ_4\big\}
-  =: \RM + (1-d) \QQ_4  \label{e:Q1,Q2,Q3}
-\end{align}
-
-Working from \eqref{e:Q1,Q2,Q3} we start by establishing the first
-bound \eqref{e:sigma^2_2(Q)>=_a.s._bound} under
-\eqref{e:RM_theorem_general_condition}. Noting from
-\cref{l:lambda-2(Q3)>=1+1/n} that $\QQ_4$ is symmetric and positive,
-making use of \eqref{e:Weyl_lower}, first with $i =2, j =1$, then with
-$i=1, j =1$, invoking \eqref{e:eigenvals_of_-A} and finally using that
-$\RM\RM^*$ is symmetric and non-negative, we find
-\begin{align}
-  &\sigma^2_2(\QQ) = \lambda_2(\QQ\QQ^*)
-  \notag\\
- &\geq (d-1)^2 \Big\{ \lambda_{2}(\QQ^2_4) + \lambda_{1}\Big(\frac{\RM\RM^*}{(d-1)^2}
-   - \frac{\QQ_4\RM^*+ \RM\QQ^*_4}{d-1} \Big) \Big\}
-  \notag\\
- &\geq (d-1)^2 \Big\{ \lambda_{2}(\QQ^2_4) + \lambda_{1}\Big(\frac{\RM\RM^*}{(d-1)^2}\Big)
-   +\lambda_{1} \Big(\frac{\QQ_4\RM^*+ \RM\QQ^*_4}{1-d} \Big) \Big\}
-  \notag\\
- &\geq (d-1)^2 \Big\{ \lambda_{2}(\QQ^2_4) 
-   -\lambda_{d} \Big(\frac{\QQ_4\RM^*+ \RM\QQ^*_4}{d-1} \Big) \Big\}.
-   \label{eq:sig:2:lower:setup}
-\end{align}
-Now observe
-\begin{align}
-  |\lambda_j(\AAA)| \leq \sigma_d(\AAA) = \| \AAA\|_{op} \quad
-  \text{ for
-  any $j =1,\ldots, d$},
-  \label{eq:dumb:op:norm:ob}
-\end{align}
-whenever $\AAA$ is symmetric.
-Furthermore,
-\begin{equation}\label{e:lambda_2(Q^2_3)}
-\lambda_{2}(\QQ^2_4) = \lambda^2_{2}(\QQ_4) = \frac{\mu^2d^2}{(d-1)^2},
-\end{equation}
-where the equality follows from
-\eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n} in
-\cref{l:lambda-2(Q3)>=1+1/n}.  Combining \eqref{eq:sig:2:lower:setup},
-\eqref{eq:dumb:op:norm:ob} and \eqref{e:lambda_2(Q^2_3)}, we therefore
-infer
-\begin{equation}\label{e:sigma^2_2(Q)>=_bound}
-  \sigma^2_2(\QQ) \geq (d-1)^2
-  \Big\{ \Big( \frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2 
-- \frac{1}{d-1} \|\QQ_4\RM^*+ \RM\QQ^*_4\|_{\op}\Big\}.
-\end{equation}
-
-However, by  \eqref{e:||Q||/n=O(1/n^(1/2-eta)} in \cref{l:||Q||/n=O(1/n^(1/2-eta)},  \eqref{e:oas:Zd:con} in \cref{l:max_Yn/n^eps}
-and \eqref{e:||Q_4||=o_P(1)} from \cref{l:||Q_4||=o_P(1)}
-we have
-\begin{align}
-  \frac{d}{d-1}  \frac{\|\RM\|_{\op}}{d}
-  \leq& \frac{d}{d-1} \left(\frac{\|\QQ_1\|_{\op}}{d} +
-  \frac{\|\QQ_2\|_{\op}}{d}\right) + \|\QQ_3\|_{\op}
-  \notag\\
-       =& O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big).
-        \label{e:||Q-cal/n||=oP(1)}
-\end{align}
-Hence, by \eqref{e:||Q-cal/n||=oP(1)} and
-\eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n} in
-\cref{l:lambda-2(Q3)>=1+1/n},
-\begin{align}
-  \frac{1}{d-1} &\|\QQ_4\RM^*+ \RM\QQ^*_4\|_{\op}
-  \leq \frac{2d}{d-1} \hspace{0.5mm} \|\QQ_4 \|_{\op} \Big\|\frac{\RM}{d} \Big\|_{\op}
-  \notag\\
- &=  O(1) \cdot O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
-  = O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big).
-  \label{e:(1/n)||Q3_Q*-cal+Q-cal_Q3||}
-\end{align}
-Thus, by \eqref{e:sigma^2_2(Q)>=_bound}, \eqref{e:||Q-cal/n||=oP(1)} and \eqref{e:(1/n)||Q3_Q*-cal+Q-cal_Q3||}, the lower bound \eqref{e:sigma^2_2(Q)>=_a.s._bound} holds.
-
-We now turn to verify \eqref{e:sigma^2_2(Q)=<_a.s._bound}. Here, by
-\eqref{e:Q1,Q2,Q3}, \eqref{e:Weyl_upper} with $j = 0, i = d$, 
-\eqref{e:lambda_2(Q^2_3)} and finally \eqref{eq:dumb:op:norm:ob},
-\begin{align*}
-  &\sigma^2_d(\QQ)
-  = \lambda_d(\QQ \QQ^*)
-  \notag\\
-   &\leq (d-1)^2 \Big\{ \lambda_{d}(\QQ^2_4)
-     + \lambda_{d}\Big(\frac{{\mathbf R}{\mathbf R}^*}{(d-1)^2}\Big) +
-     \lambda_d\Big(\frac{\QQ_4{\mathbf R}^*+ {\mathbf R}\QQ^*_4}{1-d} \Big) \Big\}
-     \notag\\
-   % &\leq (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
+%\appendix  
+%
+%
+%\section{Proof of \cref{thm:ass:error}}
+%\label{sec:Thm:det:main}
+%
+%
+%Recall that $\Z(t)$ is defined as in \eqref{eq:true:diff}, and
+%consider the representation \eqref{eq:ass:procs:0}. We begin by
+%decomposing the first term $e^{-s\QQ} \JJ e^{s \QQ}$ inside of the
+%integral in \eqref{eq:ass:procs:0} as
+%\begin{align}
+%  e^{-s\QQ} &\JJ e^{s \QQ} \notag\\
+%  =&e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
+%     + e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
+%                  \notag\\
+%      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+}) e^{s\QQ}
+%                \notag\\
+%       &+e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}) e^{s\QQ}
+%                           \notag\\
+%  = &e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
+%      + (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
+%         \label{eq:big:yuck:1}                          
+%      \\
+%      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
+%      %\notag\\
+%  + (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}),
+%  \notag
+%\end{align}
+%where we used \eqref{eq:MP:proj:prop} for the second identity.
+%Regarding the middle two terms, with our commutativity
+%\eqref{eq:GI:com} assumption, we observe that
+%\begin{align}
+%  (\II &-\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
+%         + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
+%                \label{eq:big:yuck:2}                          
+%                     \\
+%  &=  (\II -\QQ^+ \QQ) \JJ \QQ^{+} \QQ e^{s\QQ}
+%    + e^{-s\QQ}\QQ \QQ^+ \JJ (\II - \QQ \QQ^{+})
+%    \notag\\
+%  &= \frac{d}{ds}
+%    \left(
+%    (\II -\QQ^+ \QQ) \JJ \QQ^{+} e^{s\QQ}
+%    - e^{-s\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
+%    \right).
+% \notag
+%\end{align}
+%Hence, integrating in time, and then using \eqref{eq:GI:prop},
+%\eqref{eq:GI:com} for the first and last term terms on the right hand side, we have
+%\begin{align}
+%  \int_0^t& \big\{(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
+%                             + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})\big\}ds
+%                             \notag\\
+%   =& (\II -\QQ^+ \QQ) \JJ \QQ^{+}e^{t\QQ}
+%      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
+%      \notag\\
+%    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
+%      +  \QQ^+ \JJ (\II - \QQ \QQ^{+})
+%      \notag\\
+%   =& (\II -\QQ^+ \QQ) \JJ (\QQ^+)^2 \QQ e^{t\QQ}
+%      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
+%      \notag\\
+%    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
+%      + \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+})
+%             \label{eq:big:yuck:3}                          
+%\end{align}
+%Finally let us note, regarding the second term on the right hand side
+%of \eqref{eq:ass:procs:0}, using again 
+%\eqref{eq:MP:proj:prop} and \eqref{eq:GI:prop}, \eqref{eq:GI:com}
+%\begin{align}
+%  -te^{t\QQ}\JJ 
+%  =-te^{t\QQ}\QQ \QQ^+ \JJ  - t (\II - \QQ^+\QQ) \JJ.
+%               \label{eq:small:yuck:4}                          
+%\end{align}
+%
+%By combining \eqref{eq:big:yuck:1}--\eqref{eq:small:yuck:4} and
+%comparing with \eqref{eq:ass:procs:0}, we now find that
+%\begin{align}\label{e:E(t)=integ+sum}
+%  \Z(t)\! =\! \! \int_0^t \! \! \! e^{(t-s)\QQ} &\QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds
+%           + (\II - \!\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}
+%           \notag\\
+%           &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ 
+%           \notag\\
+%           &-\QQ^+ \JJ (\II - \QQ \QQ^{+})- (\II -\QQ^+ \QQ) \JJ \QQ^{+}
+%             \notag\\
+%         &-t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}.
+%\end{align}
+%Rearranging \eqref{e:E(t)=integ+sum}, we have 
+%\begin{align*}
+%\QQ^+ \JJ (\II - \QQ \QQ^{+}) &+(\II -\QQ^+ \QQ) \JJ \QQ^{+}
+%   +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
+%                                \notag\\
+%  &+  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ
+%  = \TT_1 + \TT_2,
+%\end{align*}
+%where 
+%\begin{align*}
+% \TT_1 :=&  \int_0^t  \! \! e^{(t-s)\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds,\\
+%  \TT_2 :=&  (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}\\
+%         &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ.
+%\end{align*}
+%
+%We estimate each of $\TT_1$ and $\TT_2$ in turn.  Regarding $\TT_1$,
+%using once more \eqref{eq:GI:com}, as well as \eqref{eq:qetq:exp:decay}, we have
+%\begin{align*}
+%  \| \TT_1\| \leq&
+%    \int_0^t \|\QQ e^{(t-s)\QQ}\| \| \QQ^+  \JJ \QQ^{+} \|\| \QQ e^{s\QQ}\|ds
+%         \notag\\
+%   \leq& C_0^2\int_0^t  e^{-(t-s)\kappa} \| \QQ^+  \JJ \QQ^{+} \| e^{-s\kappa } ds
+%                         \notag\\
+%   =& C_0^2 t e^{-t\kappa} \| \QQ^+  \JJ  \QQ^{+} \|.
+%\end{align*}
+%Turning to $\TT_2$, by \eqref{eq:qetq:exp:decay} we have
+%\begin{align*}
+%  \| \TT_2\|
+%  \leq C_0 (&\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
+%               + \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
+%               \notag\\
+%               &+ t\|\QQ^+ \hspace{0.25mm}\JJ\|) e^{- t \kappa }.
+%\end{align*}
+%Combining these two bound completes the proof.
+%
+%\section{Proof of \cref{thm:bride:of:frankinstein}}
+%
+%This result is established from \cref{thm:ass:error} and some basic
+%properties $\QQ$ under the given assumptions. First, note that, cf.
+%\eqref{eq:ev:d:cond:sym:case}, (\ref{eq:EV:real:ass})
+%\begin{equation}\label{e:max_|lambda-j(Q)|=<|lambda-1(Q)|}
+%  |\lambda_d(\QQ)| \leq \max_{j =1, \ldots,d_{-}} |\lambda_{j}(\QQ)| \leq |\lambda_1(\QQ)|.
+%\end{equation}
+%Referring back to
+%\eqref{eq:frob:norm}, by \eqref{e:max_|lambda-j(Q)|=<|lambda-1(Q)|} we have the estimate
+%\begin{align}\label{eq:frob:sucks}
+%  \| \QQ e^{t\QQ} \|_F
+%  =&  \left( \sum_{j =1}^{d_{-}}
+%     | \lambda_j(\QQ) e^{ t  \lambda_j(\QQ)  } |^2 \right)^{1/2}
+%     \notag\\
+%  \leq& \sqrt{d_{-}} |\lambda_{1}(\QQ)| e^{-t |\lambda_{d_{-}}(\QQ)|}.
+%\end{align}
+%Similarly, from \eqref{eq:op:norm} and \eqref{e:max_|lambda-j(Q)|=<|lambda-1(Q)|}, we have that
+%\begin{align}\label{eq:op:better}
+%  \| \QQ e^{t\QQ} \|_{\op}
+%  \leq& \max_{j =1, \ldots,d_{-}}
+%        |\lambda_{j}(\QQ)| e^{-t |\lambda_{j}(\QQ)|}
+%        \notag\\
+%  \leq& |\lambda_{1}(\QQ)| e^{-t |\lambda_{d_{-}}(\QQ)|}.
+%\end{align}
+%Likewise, we observe that
+%\begin{align}\label{eq:thm1:extra:trms:1:F}
+%  \lVert \QQ\rVert^2_F = \sum_{k=1}^{d} \lambda_k(\QQ)^2 \, , \quad
+%  \lVert \QQ^+ \rVert^2_F = \sum_{k =1}^{d_-}\frac{1}{\lambda_k(\QQ)^2} \, ,
+%\end{align}
+%and that
+%\begin{align}\label{eq:thm1:extra:trms:1:F}
+%  \lVert \QQ\rVert^2_{\op} = \lambda_1(\QQ)^2 \, , \quad
+%  \lVert \QQ^+ \rVert^2_{\op} = \frac{1}{\lambda_{d_{-}}(\QQ)^2} \,.
+%\end{align}
+%Let $\UU$ be as in \eqref{eq:MP:inv:sym:case}. Noting, furthermore,
+%that $\II - \QQ \QQ^+ = \UU(\II - \II_{d_-})\UU^*$, where $\II_{d_-}$
+%is the matrix with $1$'s along the first $d_-$ diagonal elements and
+%zero otherwise, we have the bounds
+%\begin{align}\label{eq:proj:mat:bnds}
+%  \|\II - \QQ  \QQ^+\|_{F} = \sqrt{d - d_{-}},
+%  \quad
+%  \| \II - \QQ  \QQ^+\|_{\op} = 1.
+%\end{align}
+%
+%Now, for any matrix norm $\|\cdot\|$, \eqref{eq:ass:error} implies
+%that
+%\begin{align}
+%    \| t(\II -\QQ^+ \QQ) &\JJ \QQ \QQ^{+}
+%                           +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\|
+%                           \notag\\
+%  &\leq C(1+ t)e^{-\kappa t}
+%    + 2\|   \QQ^+\| \|\JJ\| \|\II - \QQ \QQ^{+}\|
+%    \label{eq:ass:error_proof}
+%\end{align}    
+%for any $t \geq 0$, where $C > 0$ is given by
+%\eqref{eq:exp:error:ass:er}. Thus, for the choice of norm
+%$\|\cdot\| = \|\cdot\|_{F}$, by \eqref{eq:ass:error_proof},
+%\eqref{eq:frob:sucks}, \eqref{eq:thm1:extra:trms:1:F} and
+%\eqref{eq:proj:mat:bnds}, we obtain
+%\eqref{eq:bride:of:frankinstein:1}.  Similarly, for the choice of norm
+%$\|\cdot\| = \|\cdot\|_{\op}$, \eqref{eq:ass:error_proof}, in
+%conjunction with \eqref{eq:op:better}, \eqref{eq:thm1:extra:trms:1:F}
+%and \eqref{eq:proj:mat:bnds}, yields the bound
+%\eqref{eq:bride:of:frankinstein}.  Since
+%\eqref{eq:bride:of:frankinstein:1:rm:app} follows immediately from
+%\eqref{eq:bride:of:frankinstein:1}, and similarly between
+%\eqref{eq:bride:of:frankinstein:0:rm:app} and
+%\eqref{eq:bride:of:frankinstein}, the proof is now complete.
+%
+%
+%
+%
+%\section{Proof of \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}}
+%\label{sec:Thm:rand:mat:main}
+%
+%For use in this section, we introduce a probabilistic version of the
+%typical $O$ and $o$ asymptotic notations.  Given collections of random
+%variable $\{X_d\}_{d \in \bbN}$, $\{Y_d\}_{d \in \bbN}$, we write
+%\begin{equation}\label{e:X=O_a.s.(1)_Y=o_a.s.(1)}
+%X_d = O_{\as}(f(d)),\quad Y_d  = o_{\as}(f(d)), 
+%\end{equation}
+%for some $f: \bbN \to \RR^+$ to mean, respectively, that there exists
+%a random variable $C = C(\omega)$, not dependent on $d$, such
+%that $|X_d|/f(d) \leq C$ for all $d \in \bbN$ a.s., and also that
+%$\lim_{d \rightarrow \infty}Y_d/f(d) = 0$ a.s.
+%
+%In everything that follows we will make use of the so-called
+%\textit{Weyl's inequalities} (see e.g.
+%\cite[Theorem 4.3.1, p.\ 239]{horn:johnson:2012}). Namely, let
+%${\mathbf A},{\mathbf B} \in {\mathcal S}(d,\bbC)$, each with its own
+%eigenvalues listed in increasing order as in our convention
+%\eqref{eq:EV:real:ass}. Then, for $i = 1,\hdots,d$,
+%\begin{equation}\label{e:Weyl_upper}
+%  \lambda_i({\mathbf A}+{\mathbf B}) \leq
+%  \lambda_{i+j}({\mathbf A}) + \lambda_{d-j}({\mathbf B}),
+%  \quad j = 0,1,\hdots,d-i,
+%\end{equation}
+%and
+%\begin{equation}\label{e:Weyl_lower}
+%  \lambda_{i-j+1}({\mathbf A}) + \lambda_j({\mathbf B})
+%  \leq \lambda_i({\mathbf A}+{\mathbf B}), \quad j = 1,\hdots,i.
+%\end{equation}
+%In what follows, we also make use of the fact that, for ${\mathbf A} \in {\mathcal S}(d,\bbC)$,
+%\begin{equation}\label{e:eigenvals_of_-A}
+%  \lambda_\ell(-{\mathbf A}) = - \lambda_{d-\ell+1}({\mathbf A}),
+%  \quad \ell = 1,\hdots,d.
+%\end{equation}
+%
+%
+%With these preliminaries now in hand, we turn to the proof of
+%\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.  This result is
+%established with the aide of four auxiliary results, \cref{l:||Q||/n=O(1/n^(1/2-eta)},
+%\cref{l:max_Yn/n^eps}, \cref{l:||Q_4||=o_P(1)} and \cref{l:lambda-2(Q3)>=1+1/n}, whose
+%precise statements and proof are provided immediately afterward.
+%%in \cref{sec:aux}.
+%\begin{proof}
+%  To demonstrate our desired result,
+%  \eqref{e:lambda_n-1(Q)->-infty_non-symm}, it suffices to establish
+%  the lower bound
+%  \begin{equation}\label{e:sigma^2_2(Q)>=_a.s._bound}
+%    \sigma^2_2(\QQ) \geq
+%    (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
+%    + O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)\Big\}
+%  \end{equation}
+%  as well as the upper bound
+%  \begin{equation}\label{e:sigma^2_2(Q)=<_a.s._bound}
+%    \sigma^2_d(\QQ) \leq
+%    (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
+%    + O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)\Big\}.
+%  \end{equation}
+%  With this in mind, we now decompose $\QQ$ as follows, starting with
+%  the case \eqref{e:RM_theorem_general_condition}.  Fix
+%  $d \in \bbN \backslash\{1\}$.  Taking $F_X$ to be the distribution
+%  defining the elements in \eqref{e:RM_theorem_general_condition}, draw
+%  \begin{align}\label{eq:diag:fill:trk}
+%    \{\widetilde{q}_{ii}\}_{i=1,\hdots,d}
+%    \stackrel{\textnormal{iid}}\sim F_X
+%    \text{ independently of } \{q_{ij}\}_{i,j=1,\hdots,d}.
+%   \end{align} 
+%   Here note carefully that $\widetilde{q}_{ii} \stackrel{d}= q_{ij}$
+%   and $\widetilde{q}_{ii} \stackrel{d}\neq q_{ii}$.  Now recast
+%   $\QQ = \{q_{ij}\}_{i,j}$ as %{ \scriptsize
+%\begin{align}
+%  & \begin{pmatrix}
+%0 & q_{12}-\mu & \hdots & q_{1d}-\mu \\
+%q_{21}-\mu & 0 & \hdots & q_{2d}-\mu \\
+%\vdots & \vdots & \ddots & \vdots \\
+%q_{d1}-\mu & q_{d2}-\mu & \hdots & 0 \\
+%\end{pmatrix}
+%+
+%\begin{pmatrix}
+%q_{11} & \mu & \hdots & \mu \\
+%\mu & q_{22} & \hdots & \mu \\
+%\vdots & \vdots & \ddots & \vdots \\
+%\mu & \mu & \hdots & q_{dd} \\
+%\end{pmatrix} \notag \\
+%  &=\left\{\!
+%    \begin{pmatrix}
+%\widetilde{q}_{11}-\mu & q_{12}-\mu & \hdots & q_{1d}-\mu \\
+%q_{21}-\mu & \widetilde{q}_{22}-\mu & \hdots & q_{2d}-\mu \\
+%\vdots & \vdots & \ddots & \vdots \\
+%q_{d1}-\mu & q_{d2}-\mu & \hdots & \widetilde{q}_{dd}-\mu \\
+%\end{pmatrix}\right.
+%  \label{eq:En:or:Wigner}\\
+%& \qquad \quad \quad +
+%\left.\begin{pmatrix}
+%-\widetilde{q}_{11}+\mu & 0 & \hdots & 0 \\
+%0 & -\widetilde{q}_{22}+\mu & \hdots & 0 \\
+%\vdots & \vdots & \ddots & \vdots \\
+%0 & 0 & \hdots & -\widetilde{q}_{dd}+\mu \\
+%\end{pmatrix}\!\right\}
+%  \notag \\
+%  &\quad\;+ (1-d)
+%    \left\{ \!
+%    \left(\! \!\!\!
+%      \begin{array}{*{24}{c@{\hspace{2pt}}}}
+%\frac{-q_{11}-(d-1)\mu}{d-1} & 0 & \hdots & 0 \\
+%0 & \frac{-q_{22}-(d-1)\mu}{d-1} & \hdots & 0 \\
+%\vdots & \vdots & \ddots & \vdots \\
+%0 & 0 & \hdots & \frac{-q_{dd}-(d-1)\mu}{d-1}\\
+%    \end{array}
+%  \! \!  \right)
+%  \right.
+%  \notag\\ 
+%  &\qquad \qquad \qquad \quad+  \left. 
+%        \begin{pmatrix}
+%\mu & \frac{-\mu}{d-1} & \hdots & \frac{-\mu}{d-1} \\
+%\frac{-\mu}{d-1} & \mu & \hdots & \frac{-\mu}{d-1} \\
+%\vdots & \vdots & \ddots & \vdots \\
+%\frac{-\mu}{d-1} & \frac{-\mu}{d-1} & \hdots & \mu \\
+%\end{pmatrix}
+%  \!\right\}
+%  \label{e:Q_basic_decomp}\\
+%  &=: \QQ_1 + \QQ_2 + (1-d) \big\{\QQ_3 +  \QQ_4\big\}
+%  =: \RM + (1-d) \QQ_4  \label{e:Q1,Q2,Q3}
+%\end{align}
+%
+%Working from \eqref{e:Q1,Q2,Q3} we start by establishing the first
+%bound \eqref{e:sigma^2_2(Q)>=_a.s._bound} under
+%\eqref{e:RM_theorem_general_condition}. Noting from
+%\cref{l:lambda-2(Q3)>=1+1/n} that $\QQ_4$ is symmetric and positive,
+%making use of \eqref{e:Weyl_lower}, first with $i =2, j =1$, then with
+%$i=1, j =1$, invoking \eqref{e:eigenvals_of_-A} and finally using that
+%$\RM\RM^*$ is symmetric and non-negative, we find
+%\begin{align}
+%  &\sigma^2_2(\QQ) = \lambda_2(\QQ\QQ^*)
+%  \notag\\
+% &\geq (d-1)^2 \Big\{ \lambda_{2}(\QQ^2_4) + \lambda_{1}\Big(\frac{\RM\RM^*}{(d-1)^2}
+%   - \frac{\QQ_4\RM^*+ \RM\QQ^*_4}{d-1} \Big) \Big\}
+%  \notag\\
+% &\geq (d-1)^2 \Big\{ \lambda_{2}(\QQ^2_4) + \lambda_{1}\Big(\frac{\RM\RM^*}{(d-1)^2}\Big)
+%   +\lambda_{1} \Big(\frac{\QQ_4\RM^*+ \RM\QQ^*_4}{1-d} \Big) \Big\}
+%  \notag\\
+% &\geq (d-1)^2 \Big\{ \lambda_{2}(\QQ^2_4) 
+%   -\lambda_{d} \Big(\frac{\QQ_4\RM^*+ \RM\QQ^*_4}{d-1} \Big) \Big\}.
+%   \label{eq:sig:2:lower:setup}
+%\end{align}
+%Now observe
+%\begin{align}
+%  |\lambda_j(\AAA)| \leq \sigma_d(\AAA) = \| \AAA\|_{op} \quad
+%  \text{ for
+%  any $j =1,\ldots, d$},
+%  \label{eq:dumb:op:norm:ob}
+%\end{align}
+%whenever $\AAA$ is symmetric.
+%Furthermore,
+%\begin{equation}\label{e:lambda_2(Q^2_3)}
+%\lambda_{2}(\QQ^2_4) = \lambda^2_{2}(\QQ_4) = \frac{\mu^2d^2}{(d-1)^2},
+%\end{equation}
+%where the equality follows from
+%\eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n} in
+%\cref{l:lambda-2(Q3)>=1+1/n}.  Combining \eqref{eq:sig:2:lower:setup},
+%\eqref{eq:dumb:op:norm:ob} and \eqref{e:lambda_2(Q^2_3)}, we therefore
+%infer
+%\begin{equation}\label{e:sigma^2_2(Q)>=_bound}
+%  \sigma^2_2(\QQ) \geq (d-1)^2
+%  \Big\{ \Big( \frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2 
+%- \frac{1}{d-1} \|\QQ_4\RM^*+ \RM\QQ^*_4\|_{\op}\Big\}.
+%\end{equation}
+%
+%However, by  \eqref{e:||Q||/n=O(1/n^(1/2-eta)} in \cref{l:||Q||/n=O(1/n^(1/2-eta)},  \eqref{e:oas:Zd:con} in \cref{l:max_Yn/n^eps}
+%and \eqref{e:||Q_4||=o_P(1)} from \cref{l:||Q_4||=o_P(1)}
+%we have
+%\begin{align}
+%  \frac{d}{d-1}  \frac{\|\RM\|_{\op}}{d}
+%  \leq& \frac{d}{d-1} \left(\frac{\|\QQ_1\|_{\op}}{d} +
+%  \frac{\|\QQ_2\|_{\op}}{d}\right) + \|\QQ_3\|_{\op}
+%  \notag\\
+%       =& O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big).
+%        \label{e:||Q-cal/n||=oP(1)}
+%\end{align}
+%Hence, by \eqref{e:||Q-cal/n||=oP(1)} and
+%\eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n} in
+%\cref{l:lambda-2(Q3)>=1+1/n},
+%\begin{align}
+%  \frac{1}{d-1} &\|\QQ_4\RM^*+ \RM\QQ^*_4\|_{\op}
+%  \leq \frac{2d}{d-1} \hspace{0.5mm} \|\QQ_4 \|_{\op} \Big\|\frac{\RM}{d} \Big\|_{\op}
+%  \notag\\
+% &=  O(1) \cdot O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
+%  = O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big).
+%  \label{e:(1/n)||Q3_Q*-cal+Q-cal_Q3||}
+%\end{align}
+%Thus, by \eqref{e:sigma^2_2(Q)>=_bound}, \eqref{e:||Q-cal/n||=oP(1)} and \eqref{e:(1/n)||Q3_Q*-cal+Q-cal_Q3||}, the lower bound \eqref{e:sigma^2_2(Q)>=_a.s._bound} holds.
+%
+%We now turn to verify \eqref{e:sigma^2_2(Q)=<_a.s._bound}. Here, by
+%\eqref{e:Q1,Q2,Q3}, \eqref{e:Weyl_upper} with $j = 0, i = d$, 
+%\eqref{e:lambda_2(Q^2_3)} and finally \eqref{eq:dumb:op:norm:ob},
+%\begin{align*}
+%  &\sigma^2_d(\QQ)
+%  = \lambda_d(\QQ \QQ^*)
+%  \notag\\
+%   &\leq (d-1)^2 \Big\{ \lambda_{d}(\QQ^2_4)
 %     + \lambda_{d}\Big(\frac{{\mathbf R}{\mathbf R}^*}{(d-1)^2}\Big) +
@@
 %     \notag\\
-  &\leq (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
-    + \Big\|  \frac{{\mathbf R}}{d-1}\Big\|^2_{\op}
-    + \frac{1}{d-1} \|\QQ_4{\mathbf R}^*+ {\mathbf R}\QQ^*_4\|_{\op}\Big\}.
-\end{align*}
-The upper bound \eqref{e:sigma^2_2(Q)=<_a.s._bound} now follows from
-\eqref{e:||Q-cal/n||=oP(1)} and \eqref{e:(1/n)||Q3_Q*-cal+Q-cal_Q3||}.
-Thus, as a consequence of \eqref{e:sigma^2_2(Q)>=_a.s._bound} and
-\eqref{e:sigma^2_2(Q)=<_a.s._bound},
-\eqref{e:lambda_n-1(Q)->-infty_non-symm} holds under condition
-\eqref{e:RM_theorem_general_condition}, as claimed.  We have
-established the desired result in the first case \eqref{e:RM_theorem_general_condition}.
-
-Now suppose that condition \eqref{e:RM_theorem_symmetry_condition}
-holds. In order to establish \eqref{e:lambda_n-1(Q)->-infty_non-symm},
-it suffices to replace $q_{ij}$, $i < j$, with $q_{ji}$ in expression
-\eqref{eq:En:or:Wigner} and in the definition of $\QQ_3$.  We then
-follow the rest of the argument for proving
-\eqref{e:lambda_n-1(Q)->-infty_non-symm} under condition
-\eqref{e:RM_theorem_general_condition} noting that
-\cref{l:||Q||/n=O(1/n^(1/2-eta)}, \cref{l:||Q_4||=o_P(1)} also hold in
-the symmetric case.  This concludes the proof.
-\end{proof}
-
-% \section{Auxiliary results}
-% \label{sec:aux}
-
-We turn now to the Lemmata \ref{l:||Q||/n=O(1/n^(1/2-eta)},
-\ref{l:max_Yn/n^eps}, \ref{l:||Q_4||=o_P(1)} and
-\ref{l:lambda-2(Q3)>=1+1/n}, which are used in the proof of
-\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.  These results
-correspond to the bounds we use for each of elements in the
-decomposition \eqref{e:Q1,Q2,Q3}. We start off with
-\cref{l:||Q||/n=O(1/n^(1/2-eta)} which essentially packages results
-found in \cite{yin:bai:krishnaiah:1988}, \cite{bai:1999}.
-
-\begin{lemma}\label{l:||Q||/n=O(1/n^(1/2-eta)}
-  Let $\QQ_1 = \QQ_1(d)$, $d \in \bbN \setminus \{1\}$, be the
-  sequence of random matrices defined as in \eqref{eq:En:or:Wigner}.
-  Here we suppose that the off diagonal elements $q_{ij}$ defining
-  $\QQ_1$ are either as in \eqref{e:RM_theorem_general_condition} or
-  as in \eqref{e:RM_theorem_symmetry_condition} and that the diagonal
-  elements $\widetilde{q}_{ii}$ are drawn as in
-  \eqref{eq:diag:fill:trk}.  Then, in either case, we have
-  \begin{equation}\label{e:||Q||/n=O(1/n^(1/2-eta)}
-    \frac{\|\QQ_1\|_{\op}}{d} = O_{\as}\Big( \frac{1}{\sqrt{d}}\Big).
-  \end{equation}
-\end{lemma}
-\begin{proof}
-  To establish \eqref{e:||Q||/n=O(1/n^(1/2-eta)} in the first case,
-  \eqref{e:RM_theorem_general_condition}, we note that
-  $d^{-1}\QQ_1 \QQ^*_1$ forms a sample covariance matrix, where each
-  entry of $\QQ_1$ is centered and has infinitely many moments; cf.
-  \cref{rmk:exp:mom}. Thus, by \cite[Theorem 3.1, p.\
-  517]{yin:bai:krishnaiah:1988},
-\begin{equation}\label{e:|Q-tilde_1|=sqrt(n)OP(1)}
-  \|\QQ_1\|_{\op} = \sqrt{d}\cdot \sqrt{\lambda_{d}\Big(\frac{\QQ_1 \QQ^*_1}{d}\Big)}
-  = \sqrt{d}\cdot \sqrt{4 \sigma^2 + o_{\as}(1)},
-\end{equation}
-which immediately yields \eqref{e:||Q||/n=O(1/n^(1/2-eta)}.
-
-In the second case, where all the off-diagonal elements are determined
-starting from \eqref{e:RM_theorem_symmetry_condition}, $\QQ_1$ is a
-centered (mean-zero) Wigner matrix in the typical nomenclature
-followed by \cite{bai:1999}.  Thus, noting \eqref{eq:op:norm} and
-that, as in the previous case, $\QQ_1$ has infinitely many moments
-\eqref{e:||Q||/n=O(1/n^(1/2-eta)} is thus a direct consequence of
-\cite[Theorem 2.12, p.\ 630]{bai:1999}.
-\end{proof}
-
-We next state and establish \cref{l:max_Yn/n^eps}. Note that the
-assumption of independence among the random variables is not needed
-for this result.  Here and below in what follows we adopt the useful
-notational convention that $o_{\textnormal{a.s.}}(1)$ denotes a random
-variable that vanishes almost surely, as $d\to \infty$ (cf.\
-\eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}).
-\begin{lemma}\label{l:max_Yn/n^eps}
-  Let $\{Y_{d}\}_{d \in \bbN}$ be an identically distributed sequence
-  of sub-exponential variables with mean $\mu$;
-  cf. \cref{def:sub:exp:RV}. Let
-  \begin{align}
-    Z_{d} = \max_{k=1,\hdots,d} |Y_k - \mu|, d \in \bbN.
-    \label{eq:max:sub:exp}
-  \end{align} 
-  Then, for any fixed $\varepsilon_0 > 0$, 
-  \begin{align}\label{e:oas:Zd:con}
-   \frac{Z_d}{d^{\varepsilon_0}} = o_{\as}(1).
-  \end{align}
-   In particular, for the sequence of diagonal matrices
-   $\QQ_2 \equiv \QQ_2(d)$ given as in \eqref{e:Q1,Q2,Q3}, and any
-   $\varepsilon_0 > 0$,
-\begin{align}
-  \frac{\|\QQ_2\|_{op}}{d^{\varepsilon_0}} = o_{\as}(1).
-  \label{e:QQ2=oP(1)}
-\end{align}
-\end{lemma}
-\begin{proof}
-For any $M > 0$,
-\begin{align*}
-  \bbP\Big(\frac{Z_d}{d^{\varepsilon_0}} \geq \frac{M}{d^{\varepsilon_0/2}}\Big)
-  &= \bbP\Big(\bigcup^{d}_{k=1}\Big\{\frac{|Y_k - \mu|}{d^{\varepsilon_0}}
-  \geq \frac{M}{d^{\varepsilon_0/2}}\Big\}\Big) \\
-  &\leq d \cdot\bbP\big(|Y_1 - \mu| \geq d^{\varepsilon_0/2}M\big) 
- \leq  2d\cdot e^{- \frac{d^{\varepsilon_0/2}M}{K}}.
-\end{align*}
-In the first and second inequalities, respectively, we use the
-assumption that the random variables $\{Y_{d}\}_{d \in \bbN}$ are
-identically distributed as well as relation \eqref{e:def_subexp_RV}
-for some suitable value $K > 0$. Therefore,
-$\sum^{\infty}_{d=1}\bbP(Z_d/d^{\varepsilon_0/2} \geq
-M/d^{\varepsilon_0/2})< \infty.$ Since
-$\lim_{d \rightarrow \infty}M/d^{\varepsilon_0/2} = 0$,
-\eqref{e:oas:Zd:con} now follows as a consequence of the
-Borel-Cantelli lemma.
-
-Regarding the second claim observe that, cf. \eqref{eq:diag:fill:trk},
-$\QQ_2$ is a diagonal matrix containing identically distributed
-sub-exponential random variables.  This mean that, taking
-$Y_k :=\tilde{q}_{kk}$, $\|\QQ_2\|$ is of the form
-\eqref{eq:max:sub:exp}. As such the second claim \eqref{e:QQ2=oP(1)}
-now follows from the first \eqref{e:oas:Zd:con}, completing the proof.
-\end{proof}
-
-Turning to our bounds on $\|\QQ_3\|_{op}$, we have the following lemma.
-\begin{lemma}\label{l:||Q_4||=o_P(1)} Let $\QQ_3 \equiv \QQ_3(d)$,
-  $d \in \bbN$, be a sequence of random matrices as in
-  \eqref{e:Q1,Q2,Q3}. Here we assume that the $q_{ii}$ are determined
-  either according to \eqref{e:RM_theorem_general_condition} or
-  according to \eqref{e:RM_theorem_symmetry_condition}. Then, in
-  either of these cases,
-    \begin{equation}\label{e:||Q_4||=o_P(1)}
-      \|\QQ_3\|_{\op} = O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big).
-    \end{equation}
-\end{lemma}
-\begin{proof}
-  Under either \eqref{e:RM_theorem_general_condition} or
-  \eqref{e:RM_theorem_symmetry_condition}, each of the entries along
-  the main diagonal is a renormalized sums of iid sub-exponential
-  random variables i.e.
-  \begin{align}
-    \frac{-q_{ii} - \mu(d-1)}{d-1}
-    = \frac{1}{d-1} \sum_{j \not =i}^d (q_{ij} - \mu)
-  \end{align}
-  While these diagonal elements are not independent under
-  \eqref{e:RM_theorem_symmetry_condition} note carefully
-  that we do not use the independence of the rows of $\QQ_3$
-  in the arguments that follow.
-
-  Now, by Bernstein's inequality,
-  \cite[p.\ 29, Theorem 2.8.1]{vershynin:2018}, there exists a
-  constant $C > 0$ such that, for any $\varepsilon > 0$,
-  \GD{\begin{align}
-        \bbP&\big(\big|-q_{ii}-\mu(d-1)\big| > \varepsilon\big)\notag\\
-            &\leq 2 \cdot \exp\Big\{ - C \min\Big\{
-              \frac{\varepsilon^2}{(d-1)\|X_\mu\|^{2}_{\psi_1}},
-              \frac{\varepsilon}{\|X_\mu\|_{\psi_1}} \Big\}\Big\},
-    \label{e:P(max|-qii-n|/n>eps)}
-\end{align}
-for $i = 1, \ldots, d$ where $X_{\mu}:= X - \mu$ for $X \sim F_X$ and
-the right-hand side of \eqref{e:P(max|-qii-n|/n>eps)} involves the
-sub-exponential norm \eqref{e:sub-exp_norm}. Fix $\delta > 0$, and let
-$\eta:= (2 + \delta)/C$ where $C$ is the constant in the upper
-bound in \eqref{e:P(max|-qii-n|/n>eps)}. Hence, for all
-$d \in \bbN \backslash \{1\}$,
-\begin{align}\label{e:P(|Q3|>eps)}
-  \bbP&\Big( \sqrt{\frac{d-1}{\eta \log d}}
-        \hspace{0.5mm}\|\QQ_3(d)\|_{\op}
-        > \|X_\mu\|_{\psi_1}\Big)\notag\\
-      &= \bbP\Big(
-        \max_{i=1,\hdots,d}\Big|\frac{-q_{ii}-\mu(d-1)}{d-1}\Big|
-        > \|X_\mu\|_{\psi_1}\sqrt{\frac{\eta \log d}{d-1}}\Big) \notag\\
-      &= \bbP\Big(\bigcup^{d}_{i=1}\big\{\big|-q_{ii}-\mu(d-1)\big|
-        > \|X_\mu\|_{\psi_1}\sqrt{\eta (d-1)\log d}\big\}\Big) \notag \\ 
-      &\leq \sum^{d}_{i=1}\bbP\Big(\big|-q_{ii}-\mu(d-1)\big|
-        > \|X_\mu\|_{\psi_1}\sqrt{\eta (d-1)\log d}\Big) \notag \\
-      & \leq  2d \cdot \exp\big\{ -  C \min\big\{ \eta  \log d ,
-        \sqrt{ \eta (d-1) \log d } \big\}\big\}.
-\end{align}}
-In the second inequality in \eqref{e:P(|Q3|>eps)},
-we use \eqref{e:P(max|-qii-n|/n>eps)} with
-$\varepsilon := \|X_\mu\|_{\psi_1} \sqrt{\eta (d-1)\log d}$
-and the fact that $\{q_{ii}\}_{i=1,\hdots,d}$ are identically
-(but not necessarily independently) distributed. Therefore,
-for every $d$ sufficiently large,
-\begin{align*}
-\bbP\Big( \sqrt{\frac{d-1}{\eta \log d}}\|\QQ_3(d)\|_{\op}
-> \|X_\mu\|_{\psi_1}\Big)\leq  \frac{2}{d^{1 + \delta}},
-\end{align*}
-and as a consequence,
-\begin{align*}
-\sum^{\infty}_{d=1}\bbP\big(\|\QQ_3(d)\|_{\op} >
-  \|X_\mu\|_{\psi_1}\sqrt{\frac{\eta \log d}{d-1}}\big)< \infty.
-\end{align*}
-Hence, by the Borel-Cantelli lemma,
-\begin{align*}
-  \bbP\Big(\|\QQ_3(d)\|_{\op} >
-  \|X_\mu\|_{\psi_1}\sqrt{\frac{\eta \log d}{d-1}}
-  \textnormal{ i.o.}\Big)  = 0,
-\end{align*}
-so that the relation \eqref{e:||Q_4||=o_P(1)} holds. The
-proof is complete.
-\end{proof}
-
-Finally we conclude with \cref{l:lambda-2(Q3)>=1+1/n} as follows.
-\begin{lemma}\label{l:lambda-2(Q3)>=1+1/n}
-  Let $\QQ_4 \equiv \QQ_4(d)$, $d \in \bbN$, be the sequence of
-  symmetric matrices defined in \eqref{e:Q1,Q2,Q3}. Then, for any
-  $d \in \bbN \backslash\{1\}$,
-\begin{equation}\label{e:lambda-1(Q3)>=1/n}
-  \lambda_1(\QQ_4) = 0,
-\end{equation}
-and
-\begin{equation}\label{e:lambda-ell(Q3)=1+1/n,ell=2,...,n}
-  \lambda_\ell(\QQ_4) = \GD{\frac{\mu \hspace{0.5mm} d}{d-1} }, \quad
-  \text{ for } \ell = 2,\hdots,d.
-\end{equation}
-\end{lemma}
-\begin{proof}
-  The statement \eqref{e:lambda-1(Q3)>=1/n} is a consequence of the
-  fact that
-  $\QQ_4 \hspace{0.5mm}(1,\hdots,1)^\top = {\mathbf 0} \in \bbR^n$. To
-  establish \eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n}, it suffices to
-  prove that
-\begin{equation}\label{e:lambda-2(Q3)>=1+1/n}
-  \frac{d}{d-1} \leq \GD{\frac{\lambda_2(\QQ_4)}{\mu}
-    \leq \frac{\lambda_d(\QQ_4)}{\mu}} \leq \frac{d}{d-1},
-  \quad d \in \bbN \backslash\{1\}.
-\end{equation}
-Let ${\mathbf 1} \in {\mathcal S}(d,\bbR)$ be a matrix of
-ones and recast
-\begin{equation}\label{e:Q3/mu}
-\frac{1}{\mu}\QQ_4 = \II + \frac{1}{d-1}(\II-{\mathbf 1}). 
-\end{equation}
-By \eqref{e:Q3/mu}, \eqref{e:Weyl_lower} with $i = 2, j=1$
-and \eqref{e:eigenvals_of_-A},
-\begin{align}
-  \frac{\lambda_2(\QQ_4)}{\mu}
-  &\geq \lambda_{1}(\II) + \lambda_{2}\Big(\frac{1}{d-1}(\II-{\mathbf 1})\Big)
-    \notag\\
-  &= \lambda_{1}(\II) - \frac{1}{d-1}\lambda_{d-1}({\mathbf 1}-\II).
-    \label{e:lambda-2(Q3)>=_Weyl_bound}
-\end{align}
-However, by \eqref{e:Weyl_upper}, in the case $i = d-1, j = 1$,
-\begin{equation}\label{e:lambda_n-1(1-I)=-1}
-  \lambda_{d-1}({\mathbf 1}-\II)
-  \leq \lambda_{d-1}({\mathbf 1}) + \lambda_d(-\II) = -1,
-\end{equation}
-where we used that
-\begin{equation}\label{e:rank(1)=1}
-  \lambda_{1}(\mathbf{1}) = \cdots = \lambda_{d-1}({\mathbf 1}) = 0.
-\end{equation}
-This latter claim follows immediately from the fact that
-$\textnormal{rank}({\mathbf 1}) = 1$ and
-$\mathbf{1} \hspace{0.5mm}(1,\hdots,1)^\top= d(1,\hdots,1)^\top$.  The
-first inequality in \eqref{e:lambda-2(Q3)>=1+1/n} is now a consequence
-of \eqref{e:lambda-2(Q3)>=_Weyl_bound} and
-\eqref{e:lambda_n-1(1-I)=-1}.
-
-On the other hand, again by \eqref{e:Q3/mu}, \eqref{e:Weyl_upper}, this
-time with $i = d, j = 0$, and \eqref{e:eigenvals_of_-A},
-\begin{align}
-  \GD{\frac{\lambda_d(\QQ_4)}{\mu}}
-  &\leq \lambda_d(\II)
-         + \lambda_d\Big(\frac{1}{d-1}(\II - {\mathbf 1})\Big)
-    \notag\\
-  &= \lambda_d(\II) - \lambda_1\Big(\frac{1}{d-1}({\mathbf 1}-\II)\Big).
-        \label{e:lambda_n(Q3)=<_upper}
-\end{align}
-However, by \eqref{e:Weyl_lower} and \eqref{e:rank(1)=1},
-\begin{equation}\label{e:lambda_1(1-I)>=_lower}
-\lambda_1({\mathbf 1}-\II) \geq \lambda_1({\mathbf 1}) + \lambda_1(-\II)= -1.
-\end{equation}
-The third inequality in \eqref{e:lambda-2(Q3)>=1+1/n} is now a
-consequence of \eqref{e:lambda_n(Q3)=<_upper} and
-\eqref{e:lambda_1(1-I)>=_lower}. This establishes
-\eqref{e:lambda-2(Q3)>=1+1/n} and, hence,
-\eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n}, completing the proof.
-\end{proof}
-
-\section{Surrogate-trajectory Hamiltonian Monte Carlo}
-\label{sec:Sur:trad:HMC}
-
-\newcommand{\G}{\mathbf{G}}
-
-Hamiltonian Monte Carlo (HMC) \cite{duane1987hybrid,neal2011mcmc} is
-an advanced MCMC procedure that uses numerical approximations to
-Hamiltonian trajectories to generate Metropolis-type
-\cite{metropolis1953equation} proposals far away from the current
-Markov chain state.  On the one hand, this approach to proposal
-generation helps reduce autocorrelation between chain states and is
-particularly helpful within higher-dimensional state spaces.  On the
-other hand, numerical integration of Hamilton's equations requires
-repeated evaluation of the Hamiltonian potential energy's gradient,
-and these repeated floating-point operations may become
-computationally burdensome.
-
-Lets briefly recall this HMC approach to resolve a given density
-$\pi(\cdot)$ of a `target' probability measure of
-interest.\footnote{In the Bayesian inference context, we are typically
-  considering the posterior density function
-  $\pi(\ttheta):=p(\ttheta|\Y)$ for the parameter
-  $\ttheta\in\mathbb{R}^K$ given observed data $\Y$.}  We proceed by
-considering a potential energy of the form
-$U(\ttheta):=-\log \pi(\ttheta)$.  We then select an associated
-kinetic energy $V(\mom) := | \G^{-1/2} \mom|^2$ for an appropriately
-chosen symmetric-positive-definite mass matrix $\G$ (which is often
-taken as the identity for simplicity).  One then observes that the
-Gibbs measure, proportional to $e^{-H(\ttheta, \mom)}$ where
-$H = U +V$, is invariant under the associated Hamiltonian dynamic.
-Here note that the $\ttheta$ marginal of $e^{-H(\ttheta, \mom)}$
-coincides with $\pi$ while the $\mom$ marginal is normally distributed
-as $\mathcal{N}(\Zero,\G)$.
-
-One operationalizes these observations as an algorithmic sampling
-procedure as follows. At each step, given a current sample
-$\ttheta^{(n)}$, one draws $\mom^{(n)} \sim \mathcal{N}(\Zero,\G)$.
-From this initial state
-$(\ttheta(0), \mom(0)) := (\ttheta^{(n)},\mom^{(n)})$ one then
-numerically approximates the associated Hamiltonian dynamics using a
-St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator up to a
-total integration time $\tau >0$ and using integration step size
-$\epsilon > 0$.  In this context, note that a single iteration of this
-integrator takes the form, \cite{leimkuhler2004simulating},
-\begin{align}
-  \mom\left(s + \frac{\epsilon}{2}\right)
-  &:= \mom(s) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(s)),  
-	 \nonumber \\
-  \ttheta(s+\epsilon) &:= \ttheta(s) + \epsilon \, \G^{-1}\mom(s+\frac{\epsilon}{2}), 
-		 \label{eq:leap}\\ 
-  \mom(s + \epsilon)
-  &:= \mom\left(s+\frac{\epsilon}{2}\right)
-    + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(s+\epsilon))  \, .
-	\nonumber
-\end{align}
-In this fashion the proposed new state is given through
-\eqref{eq:leap} by $\ttheta(\tau)$.  To remove bias this procedure can
-be augmented with an acceptance probability of the form
-$\alpha^{(n)} := \exp( H( \ttheta(0), \mom(0)) - H(\ttheta(\tau),
-\mom(\tau)) \wedge 1$.
-
-Different strategies aim to speed up the leapfrog integrator's many
-log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$, as
-these numerical routines often represent the algorithm's computational
-bottleneck.  In model-specific contexts,
-\cite{holbrook2021massive,holbrook2022bayesian,holbrook2022viral}
-yield parallelization strategies, and \cite{ji2020gradients} develops
-dynamic programming techniques, to accelerate the evaluation of
-$\nabla \log \pi (\ttheta)$.
-
-A small body of work considers another approach by replacing
-$\nabla \log \pi (\ttheta)$ with a suitable approximation
-$\widetilde{\nabla} \log \pi (\ttheta)$ and recognizing that the
-modified \eqref{eq:leap} continues to satisfy path reversibility and
-volume preservation, two essential ingredients for well-specified
-HMC. It follows immediately that the resulting `surrogate trajectory
-HMC' continues to sample the correct target distribution $\pi(\cdot)$;
-see further details \cite{neal2011mcmc, glatt2020accept}.
-Nonetheless, the acceptance rates and overall efficiency of such
-samplers may suffer when approximations are poor.
-
-The majority of surrogate HMC methods first obtain a small sample of
-exact gradient evaluations and then use some model to interpolate:
-\cite{zhang2017precomputing} assumes the approximate gradient follows
-a piecewise constant form across a grid;
-\cite{rasmussen2003gaussian,lan2016emulation} construct approximations
-using Gaussian processes; and \cite{zhang2017hamiltonian,li2019neural}
-do the same using neural networks.  But an even simpler approach to
-surrogate HMC may be appropriate when gradients have series
-representations as we can leverage here in \eqref{eq:grad:series}.
-
-\section{Visualizing the posterior mean random effects}\label{sec:randomEffs}
-
-Figure \ref{fig:randEffs} displays posterior means of the exponentiated random effects, which one may interpret as multiplicative deviations from the fixed effects' contributions to the generator matrix.
+%   % &\leq (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
+%   %   + \lambda_{d}\Big(\frac{{\mathbf R}{\mathbf R}^*}{(d-1)^2}\Big) +
+%   %    \lambda_d\Big(\frac{\QQ_4{\mathbf R}^*+ {\mathbf R}\QQ^*_4}{1-d} \Big) \Big\}     
+%   %   \notag\\
+%  &\leq (d-1)^2 \Big\{ \Big(\frac{\mu \hspace{0.5mm} d}{d-1}\Big)^2
+%    + \Big\|  \frac{{\mathbf R}}{d-1}\Big\|^2_{\op}
+%    + \frac{1}{d-1} \|\QQ_4{\mathbf R}^*+ {\mathbf R}\QQ^*_4\|_{\op}\Big\}.
+%\end{align*}
+%The upper bound \eqref{e:sigma^2_2(Q)=<_a.s._bound} now follows from
+%\eqref{e:||Q-cal/n||=oP(1)} and \eqref{e:(1/n)||Q3_Q*-cal+Q-cal_Q3||}.
+%Thus, as a consequence of \eqref{e:sigma^2_2(Q)>=_a.s._bound} and
+%\eqref{e:sigma^2_2(Q)=<_a.s._bound},
+%\eqref{e:lambda_n-1(Q)->-infty_non-symm} holds under condition
+%\eqref{e:RM_theorem_general_condition}, as claimed.  We have
+%established the desired result in the first case \eqref{e:RM_theorem_general_condition}.
+%
+%Now suppose that condition \eqref{e:RM_theorem_symmetry_condition}
+%holds. In order to establish \eqref{e:lambda_n-1(Q)->-infty_non-symm},
+%it suffices to replace $q_{ij}$, $i < j$, with $q_{ji}$ in expression
+%\eqref{eq:En:or:Wigner} and in the definition of $\QQ_3$.  We then
+%follow the rest of the argument for proving
+%\eqref{e:lambda_n-1(Q)->-infty_non-symm} under condition
+%\eqref{e:RM_theorem_general_condition} noting that
+%\cref{l:||Q||/n=O(1/n^(1/2-eta)}, \cref{l:||Q_4||=o_P(1)} also hold in
+%the symmetric case.  This concludes the proof.
+%\end{proof}
+%
+%% \section{Auxiliary results}
+%% \label{sec:aux}
+%
+%
+%We turn now to the Lemmata \ref{l:||Q||/n=O(1/n^(1/2-eta)},
+%\ref{l:max_Yn/n^eps}, \ref{l:||Q_4||=o_P(1)} and
+%\ref{l:lambda-2(Q3)>=1+1/n}, which are used in the proof of
+%\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.  These results
+%correspond to the bounds we use for each of elements in the
+%decomposition \eqref{e:Q1,Q2,Q3}. We start off with
+%\cref{l:||Q||/n=O(1/n^(1/2-eta)} which essentially packages results
+%found in \cite{yin:bai:krishnaiah:1988}, \cite{bai:1999}.
+%
+%\begin{lemma}\label{l:||Q||/n=O(1/n^(1/2-eta)}
+%  Let $\QQ_1 = \QQ_1(d)$, $d \in \bbN \setminus \{1\}$, be the
+%  sequence of random matrices defined as in \eqref{eq:En:or:Wigner}.
+%  Here we suppose that the off diagonal elements $q_{ij}$ defining
+%  $\QQ_1$ are either as in \eqref{e:RM_theorem_general_condition} or
+%  as in \eqref{e:RM_theorem_symmetry_condition} and that the diagonal
+%  elements $\widetilde{q}_{ii}$ are drawn as in
+%  \eqref{eq:diag:fill:trk}.  Then, in either case, we have
+%  \begin{equation}\label{e:||Q||/n=O(1/n^(1/2-eta)}
+%    \frac{\|\QQ_1\|_{\op}}{d} = O_{\as}\Big( \frac{1}{\sqrt{d}}\Big).
+%  \end{equation}
+%\end{lemma}
+%\begin{proof}
+%  To establish \eqref{e:||Q||/n=O(1/n^(1/2-eta)} in the first case,
+%  \eqref{e:RM_theorem_general_condition}, we note that
+%  $d^{-1}\QQ_1 \QQ^*_1$ forms a sample covariance matrix, where each
+%  entry of $\QQ_1$ is centered and has infinitely many moments; cf.
+%  \cref{rmk:exp:mom}. Thus, by \cite[Theorem 3.1, p.\
+%  517]{yin:bai:krishnaiah:1988},
+%\begin{equation}\label{e:|Q-tilde_1|=sqrt(n)OP(1)}
+%  \|\QQ_1\|_{\op} = \sqrt{d}\cdot \sqrt{\lambda_{d}\Big(\frac{\QQ_1 \QQ^*_1}{d}\Big)}
+%  = \sqrt{d}\cdot \sqrt{4 \sigma^2 + o_{\as}(1)},
+%\end{equation}
+%which immediately yields \eqref{e:||Q||/n=O(1/n^(1/2-eta)}.
+%
+%In the second case, where all the off-diagonal elements are determined
+%starting from \eqref{e:RM_theorem_symmetry_condition}, $\QQ_1$ is a
+%centered (mean-zero) Wigner matrix in the typical nomenclature
+%followed by \cite{bai:1999}.  Thus, noting \eqref{eq:op:norm} and
+%that, as in the previous case, $\QQ_1$ has infinitely many moments
+%\eqref{e:||Q||/n=O(1/n^(1/2-eta)} is thus a direct consequence of
+%\cite[Theorem 2.12, p.\ 630]{bai:1999}.
+%\end{proof}
+%
+%
+%
+%
+%We next state and establish \cref{l:max_Yn/n^eps}. Note that the
+%assumption of independence among the random variables is not needed
+%for this result.  Here and below in what follows we adopt the useful
+%notational convention that $o_{\textnormal{a.s.}}(1)$ denotes a random
+%variable that vanishes almost surely, as $d\to \infty$ (cf.\
+%\eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}).
+%\begin{lemma}\label{l:max_Yn/n^eps}
+%  Let $\{Y_{d}\}_{d \in \bbN}$ be an identically distributed sequence
+%  of sub-exponential variables with mean $\mu$;
+%  cf. \cref{def:sub:exp:RV}. Let
+%  \begin{align}
+%    Z_{d} = \max_{k=1,\hdots,d} |Y_k - \mu|, d \in \bbN.
+%    \label{eq:max:sub:exp}
+%  \end{align} 
+%  Then, for any fixed $\varepsilon_0 > 0$, 
+%  \begin{align}\label{e:oas:Zd:con}
+%   \frac{Z_d}{d^{\varepsilon_0}} = o_{\as}(1).
+%  \end{align}
+%   In particular, for the sequence of diagonal matrices
+%   $\QQ_2 \equiv \QQ_2(d)$ given as in \eqref{e:Q1,Q2,Q3}, and any
+%   $\varepsilon_0 > 0$,
+%\begin{align}
+%  \frac{\|\QQ_2\|_{op}}{d^{\varepsilon_0}} = o_{\as}(1).
+%  \label{e:QQ2=oP(1)}
+%\end{align}
+%\end{lemma}
+%\begin{proof}
+%For any $M > 0$,
+%\begin{align*}
+%  \bbP\Big(\frac{Z_d}{d^{\varepsilon_0}} \geq \frac{M}{d^{\varepsilon_0/2}}\Big)
+%  &= \bbP\Big(\bigcup^{d}_{k=1}\Big\{\frac{|Y_k - \mu|}{d^{\varepsilon_0}}
+%  \geq \frac{M}{d^{\varepsilon_0/2}}\Big\}\Big) \\
+%  &\leq d \cdot\bbP\big(|Y_1 - \mu| \geq d^{\varepsilon_0/2}M\big) 
+% \leq  2d\cdot e^{- \frac{d^{\varepsilon_0/2}M}{K}}.
+%\end{align*}
+%In the first and second inequalities, respectively, we use the
+%assumption that the random variables $\{Y_{d}\}_{d \in \bbN}$ are
+%identically distributed as well as relation \eqref{e:def_subexp_RV}
+%for some suitable value $K > 0$. Therefore,
+%$\sum^{\infty}_{d=1}\bbP(Z_d/d^{\varepsilon_0/2} \geq
+%M/d^{\varepsilon_0/2})< \infty.$ Since
+%$\lim_{d \rightarrow \infty}M/d^{\varepsilon_0/2} = 0$,
+%\eqref{e:oas:Zd:con} now follows as a consequence of the
+%Borel-Cantelli lemma.
+%
+%Regarding the second claim observe that, cf. \eqref{eq:diag:fill:trk},
+%$\QQ_2$ is a diagonal matrix containing identically distributed
+%sub-exponential random variables.  This mean that, taking
+%$Y_k :=\tilde{q}_{kk}$, $\|\QQ_2\|$ is of the form
+%\eqref{eq:max:sub:exp}. As such the second claim \eqref{e:QQ2=oP(1)}
+%now follows from the first \eqref{e:oas:Zd:con}, completing the proof.
+%\end{proof}
+%
+%
+%Turning to our bounds on $\|\QQ_3\|_{op}$, we have the following lemma.
+%\begin{lemma}\label{l:||Q_4||=o_P(1)} Let $\QQ_3 \equiv \QQ_3(d)$,
+%  $d \in \bbN$, be a sequence of random matrices as in
+%  \eqref{e:Q1,Q2,Q3}. Here we assume that the $q_{ii}$ are determined
+%  either according to \eqref{e:RM_theorem_general_condition} or
+%  according to \eqref{e:RM_theorem_symmetry_condition}. Then, in
+%  either of these cases,
+%    \begin{equation}\label{e:||Q_4||=o_P(1)}
+%      \|\QQ_3\|_{\op} = O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big).
+%    \end{equation}
+%\end{lemma}
+%\begin{proof}
+%  Under either \eqref{e:RM_theorem_general_condition} or
+%  \eqref{e:RM_theorem_symmetry_condition}, each of the entries along
+%  the main diagonal is a renormalized sums of iid sub-exponential
+%  random variables i.e.
+%  \begin{align}
+%    \frac{-q_{ii} - \mu(d-1)}{d-1}
+%    = \frac{1}{d-1} \sum_{j \not =i}^d (q_{ij} - \mu)
+%  \end{align}
+%  While these diagonal elements are not independent under
+%  \eqref{e:RM_theorem_symmetry_condition} note carefully
+%  that we do not use the independence of the rows of $\QQ_3$
+%  in the arguments that follow.
+%
+%  Now, by Bernstein's inequality,
+%  \cite[p.\ 29, Theorem 2.8.1]{vershynin:2018}, there exists a
+%  constant $C > 0$ such that, for any $\varepsilon > 0$,
+%  \GD{\begin{align}
+%        \bbP&\big(\big|-q_{ii}-\mu(d-1)\big| > \varepsilon\big)\notag\\
+%            &\leq 2 \cdot \exp\Big\{ - C \min\Big\{
+%              \frac{\varepsilon^2}{(d-1)\|X_\mu\|^{2}_{\psi_1}},
+%              \frac{\varepsilon}{\|X_\mu\|_{\psi_1}} \Big\}\Big\},
+%    \label{e:P(max|-qii-n|/n>eps)}
+%\end{align}
+%for $i = 1, \ldots, d$ where $X_{\mu}:= X - \mu$ for $X \sim F_X$ and
+%the right-hand side of \eqref{e:P(max|-qii-n|/n>eps)} involves the
+%sub-exponential norm \eqref{e:sub-exp_norm}. Fix $\delta > 0$, and let
+%$\eta:= (2 + \delta)/C$ where $C$ is the constant in the upper
+%bound in \eqref{e:P(max|-qii-n|/n>eps)}. Hence, for all
+%$d \in \bbN \backslash \{1\}$,
+%\begin{align}\label{e:P(|Q3|>eps)}
+%  \bbP&\Big( \sqrt{\frac{d-1}{\eta \log d}}
+%        \hspace{0.5mm}\|\QQ_3(d)\|_{\op}
+%        > \|X_\mu\|_{\psi_1}\Big)\notag\\
+%      &= \bbP\Big(
+%        \max_{i=1,\hdots,d}\Big|\frac{-q_{ii}-\mu(d-1)}{d-1}\Big|
+%        > \|X_\mu\|_{\psi_1}\sqrt{\frac{\eta \log d}{d-1}}\Big) \notag\\
+%      &= \bbP\Big(\bigcup^{d}_{i=1}\big\{\big|-q_{ii}-\mu(d-1)\big|
+%        > \|X_\mu\|_{\psi_1}\sqrt{\eta (d-1)\log d}\big\}\Big) \notag \\ 
+%      &\leq \sum^{d}_{i=1}\bbP\Big(\big|-q_{ii}-\mu(d-1)\big|
+%        > \|X_\mu\|_{\psi_1}\sqrt{\eta (d-1)\log d}\Big) \notag \\
+%      & \leq  2d \cdot \exp\big\{ -  C \min\big\{ \eta  \log d ,
+%        \sqrt{ \eta (d-1) \log d } \big\}\big\}.
+%\end{align}}
+%In the second inequality in \eqref{e:P(|Q3|>eps)},
+%we use \eqref{e:P(max|-qii-n|/n>eps)} with
+%$\varepsilon := \|X_\mu\|_{\psi_1} \sqrt{\eta (d-1)\log d}$
+%and the fact that $\{q_{ii}\}_{i=1,\hdots,d}$ are identically
+%(but not necessarily independently) distributed. Therefore,
+%for every $d$ sufficiently large,
+%\begin{align*}
+%\bbP\Big( \sqrt{\frac{d-1}{\eta \log d}}\|\QQ_3(d)\|_{\op}
+%> \|X_\mu\|_{\psi_1}\Big)\leq  \frac{2}{d^{1 + \delta}},
+%\end{align*}
+%and as a consequence,
+%\begin{align*}
+%\sum^{\infty}_{d=1}\bbP\big(\|\QQ_3(d)\|_{\op} >
+%  \|X_\mu\|_{\psi_1}\sqrt{\frac{\eta \log d}{d-1}}\big)< \infty.
+%\end{align*}
+%Hence, by the Borel-Cantelli lemma,
+%\begin{align*}
+%  \bbP\Big(\|\QQ_3(d)\|_{\op} >
+%  \|X_\mu\|_{\psi_1}\sqrt{\frac{\eta \log d}{d-1}}
+%  \textnormal{ i.o.}\Big)  = 0,
+%\end{align*}
+%so that the relation \eqref{e:||Q_4||=o_P(1)} holds. The
+%proof is complete.
+%\end{proof}
+%
+%Finally we conclude with \cref{l:lambda-2(Q3)>=1+1/n} as follows.
+%\begin{lemma}\label{l:lambda-2(Q3)>=1+1/n}
+%  Let $\QQ_4 \equiv \QQ_4(d)$, $d \in \bbN$, be the sequence of
+%  symmetric matrices defined in \eqref{e:Q1,Q2,Q3}. Then, for any
+%  $d \in \bbN \backslash\{1\}$,
+%\begin{equation}\label{e:lambda-1(Q3)>=1/n}
+%  \lambda_1(\QQ_4) = 0,
+%\end{equation}
+%and
+%\begin{equation}\label{e:lambda-ell(Q3)=1+1/n,ell=2,...,n}
+%  \lambda_\ell(\QQ_4) = \GD{\frac{\mu \hspace{0.5mm} d}{d-1} }, \quad
+%  \text{ for } \ell = 2,\hdots,d.
+%\end{equation}
+%\end{lemma}
+%\begin{proof}
+%  The statement \eqref{e:lambda-1(Q3)>=1/n} is a consequence of the
+%  fact that
+%  $\QQ_4 \hspace{0.5mm}(1,\hdots,1)^\top = {\mathbf 0} \in \bbR^n$. To
+%  establish \eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n}, it suffices to
+%  prove that
+%\begin{equation}\label{e:lambda-2(Q3)>=1+1/n}
+%  \frac{d}{d-1} \leq \GD{\frac{\lambda_2(\QQ_4)}{\mu}
+%    \leq \frac{\lambda_d(\QQ_4)}{\mu}} \leq \frac{d}{d-1},
+%  \quad d \in \bbN \backslash\{1\}.
+%\end{equation}
+%Let ${\mathbf 1} \in {\mathcal S}(d,\bbR)$ be a matrix of
+%ones and recast
+%\begin{equation}\label{e:Q3/mu}
+%\frac{1}{\mu}\QQ_4 = \II + \frac{1}{d-1}(\II-{\mathbf 1}). 
+%\end{equation}
+%By \eqref{e:Q3/mu}, \eqref{e:Weyl_lower} with $i = 2, j=1$
+%and \eqref{e:eigenvals_of_-A},
+%\begin{align}
+%  \frac{\lambda_2(\QQ_4)}{\mu}
+%  &\geq \lambda_{1}(\II) + \lambda_{2}\Big(\frac{1}{d-1}(\II-{\mathbf 1})\Big)
+%    \notag\\
+%  &= \lambda_{1}(\II) - \frac{1}{d-1}\lambda_{d-1}({\mathbf 1}-\II).
+%    \label{e:lambda-2(Q3)>=_Weyl_bound}
+%\end{align}
+%However, by \eqref{e:Weyl_upper}, in the case $i = d-1, j = 1$,
+%\begin{equation}\label{e:lambda_n-1(1-I)=-1}
+%  \lambda_{d-1}({\mathbf 1}-\II)
+%  \leq \lambda_{d-1}({\mathbf 1}) + \lambda_d(-\II) = -1,
+%\end{equation}
+%where we used that
+%\begin{equation}\label{e:rank(1)=1}
+%  \lambda_{1}(\mathbf{1}) = \cdots = \lambda_{d-1}({\mathbf 1}) = 0.
+%\end{equation}
+%This latter claim follows immediately from the fact that
+%$\textnormal{rank}({\mathbf 1}) = 1$ and
+%$\mathbf{1} \hspace{0.5mm}(1,\hdots,1)^\top= d(1,\hdots,1)^\top$.  The
+%first inequality in \eqref{e:lambda-2(Q3)>=1+1/n} is now a consequence
+%of \eqref{e:lambda-2(Q3)>=_Weyl_bound} and
+%\eqref{e:lambda_n-1(1-I)=-1}.
+%
+%On the other hand, again by \eqref{e:Q3/mu}, \eqref{e:Weyl_upper}, this
+%time with $i = d, j = 0$, and \eqref{e:eigenvals_of_-A},
+%\begin{align}
+%  \GD{\frac{\lambda_d(\QQ_4)}{\mu}}
+%  &\leq \lambda_d(\II)
+%         + \lambda_d\Big(\frac{1}{d-1}(\II - {\mathbf 1})\Big)
+%    \notag\\
+%  &= \lambda_d(\II) - \lambda_1\Big(\frac{1}{d-1}({\mathbf 1}-\II)\Big).
+%        \label{e:lambda_n(Q3)=<_upper}
+%\end{align}
+%However, by \eqref{e:Weyl_lower} and \eqref{e:rank(1)=1},
+%\begin{equation}\label{e:lambda_1(1-I)>=_lower}
+%\lambda_1({\mathbf 1}-\II) \geq \lambda_1({\mathbf 1}) + \lambda_1(-\II)= -1.
+%\end{equation}
+%The third inequality in \eqref{e:lambda-2(Q3)>=1+1/n} is now a
+%consequence of \eqref{e:lambda_n(Q3)=<_upper} and
+%\eqref{e:lambda_1(1-I)>=_lower}. This establishes
+%\eqref{e:lambda-2(Q3)>=1+1/n} and, hence,
+%\eqref{e:lambda-ell(Q3)=1+1/n,ell=2,...,n}, completing the proof.
+%\end{proof}
+%
+%
+%\section{Surrogate-trajectory Hamiltonian Monte Carlo}
+%\label{sec:Sur:trad:HMC}
+%
+%\newcommand{\G}{\mathbf{G}}
+%
+%Hamiltonian Monte Carlo (HMC) \cite{duane1987hybrid,neal2011mcmc} is
+%an advanced MCMC procedure that uses numerical approximations to
+%Hamiltonian trajectories to generate Metropolis-type
+%\cite{metropolis1953equation} proposals far away from the current
+%Markov chain state.  On the one hand, this approach to proposal
+%generation helps reduce autocorrelation between chain states and is
+%particularly helpful within higher-dimensional state spaces.  On the
+%other hand, numerical integration of Hamilton's equations requires
+%repeated evaluation of the Hamiltonian potential energy's gradient,
+%and these repeated floating-point operations may become
+%computationally burdensome.
+%
+%Lets briefly recall this HMC approach to resolve a given density
+%$\pi(\cdot)$ of a `target' probability measure of
+%interest.\footnote{In the Bayesian inference context, we are typically
+%  considering the posterior density function
+%  $\pi(\ttheta):=p(\ttheta|\Y)$ for the parameter
+%  $\ttheta\in\mathbb{R}^K$ given observed data $\Y$.}  We proceed by
+%considering a potential energy of the form
+%$U(\ttheta):=-\log \pi(\ttheta)$.  We then select an associated
+%kinetic energy $V(\mom) := | \G^{-1/2} \mom|^2$ for an appropriately
+%chosen symmetric-positive-definite mass matrix $\G$ (which is often
+%taken as the identity for simplicity).  One then observes that the
+%Gibbs measure, proportional to $e^{-H(\ttheta, \mom)}$ where
+%$H = U +V$, is invariant under the associated Hamiltonian dynamic.
+%Here note that the $\ttheta$ marginal of $e^{-H(\ttheta, \mom)}$
+%coincides with $\pi$ while the $\mom$ marginal is normally distributed
+%as $\mathcal{N}(\Zero,\G)$.
+%
+%One operationalizes these observations as an algorithmic sampling
+%procedure as follows. At each step, given a current sample
+%$\ttheta^{(n)}$, one draws $\mom^{(n)} \sim \mathcal{N}(\Zero,\G)$.
+%From this initial state
+%$(\ttheta(0), \mom(0)) := (\ttheta^{(n)},\mom^{(n)})$ one then
+%numerically approximates the associated Hamiltonian dynamics using a
+%St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator up to a
+%total integration time $\tau >0$ and using integration step size
+%$\epsilon > 0$.  In this context, note that a single iteration of this
+%integrator takes the form, \cite{leimkuhler2004simulating},
+%\begin{align}
+%  \mom\left(s + \frac{\epsilon}{2}\right)
+%  &:= \mom(s) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(s)),  
+%	 \nonumber \\
+%  \ttheta(s+\epsilon) &:= \ttheta(s) + \epsilon \, \G^{-1}\mom(s+\frac{\epsilon}{2}), 
+%		 \label{eq:leap}\\ 
+%  \mom(s + \epsilon)
+%  &:= \mom\left(s+\frac{\epsilon}{2}\right)
+%    + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(s+\epsilon))  \, .
+%	\nonumber
+%\end{align}
+%In this fashion the proposed new state is given through
+%\eqref{eq:leap} by $\ttheta(\tau)$.  To remove bias this procedure can
+%be augmented with an acceptance probability of the form
+%$\alpha^{(n)} := \exp( H( \ttheta(0), \mom(0)) - H(\ttheta(\tau),
+%\mom(\tau)) \wedge 1$.
+%
+%
+%
+%Different strategies aim to speed up the leapfrog integrator's many
+%log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$, as
+%these numerical routines often represent the algorithm's computational
+%bottleneck.  In model-specific contexts,
+%\cite{holbrook2021massive,holbrook2022bayesian,holbrook2022viral}
+%yield parallelization strategies, and \cite{ji2020gradients} develops
+%dynamic programming techniques, to accelerate the evaluation of
+%$\nabla \log \pi (\ttheta)$.
+%
+%A small body of work considers another approach by replacing
+%$\nabla \log \pi (\ttheta)$ with a suitable approximation
+%$\widetilde{\nabla} \log \pi (\ttheta)$ and recognizing that the
+%modified \eqref{eq:leap} continues to satisfy path reversibility and
+%volume preservation, two essential ingredients for well-specified
+%HMC. It follows immediately that the resulting `surrogate trajectory
+%HMC' continues to sample the correct target distribution $\pi(\cdot)$;
+%see further details \cite{neal2011mcmc, glatt2020accept}.
+%Nonetheless, the acceptance rates and overall efficiency of such
+%samplers may suffer when approximations are poor.
+%
+%The majority of surrogate HMC methods first obtain a small sample of
+%exact gradient evaluations and then use some model to interpolate:
+%\cite{zhang2017precomputing} assumes the approximate gradient follows
+%a piecewise constant form across a grid;
+%\cite{rasmussen2003gaussian,lan2016emulation} construct approximations
+%using Gaussian processes; and \cite{zhang2017hamiltonian,li2019neural}
+%do the same using neural networks.  But an even simpler approach to
+%surrogate HMC may be appropriate when gradients have series
+%representations as we can leverage here in \eqref{eq:grad:series}.
+%
+%\section{Visualizing the posterior mean random effects}\label{sec:randomEffs}
+%
+%
+%Figure \ref{fig:randEffs} displays posterior means of the exponentiated random effects, which one may interpret as multiplicative deviations from the fixed effects' contributions to the generator matrix.
 Whereas the vast majority of exponentiated random effects have posterior means close to 1, indicating no deviation from the fixed-effect model, a few exhibit posterior means that are significantly greater than 1.
@@
 This large multiplicative increase agrees with, but goes beyond, the influence that the Hubei asymmetry holds for the entire generator matrix model.
-
-\begin{figure}[t]
-	\centering
-	\includegraphics[width=0.8\linewidth]{figures/randEffects.pdf}
-	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.
+%
+%\setcounter{figure}{0}
+%\renewcommand{\thefigure}{S\arabic{figure}}
+%\begin{figure}[t]
+%	\centering
+%	\includegraphics[width=0.8\linewidth]{figures/randEffects.pdf}
+%	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.
 Notably, we infer a roughly 1.81-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.
 Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}\label{fig:randEffs}
-\end{figure}
-
-%}
-%\showmatmethods{} % Display the Materials and Methods section
+%\end{figure}
+%
+%%}
+%%\showmatmethods{} % Display the Materials and Methods section
+%
 

### 2023-12-18 simulations/src/condenseStates.R
get rid of extraneous things

@@ -1,113 +1,0 @@
-# All 44 locations
-locations_44 <- c("Australia","Belgium","Brazil","Cambodia","Canada","ChinaAnhui","ChinaBeijing","ChinaChongqing","ChinaFujian","ChinaGuangdong","ChinaHenan","ChinaHongKong","ChinaHubei","ChinaHunan","ChinaJiangsu","ChinaJiangxi","ChinaShandong","ChinaSichuan","ChinaYunnan","ChinaZhejiang","Finland","France","Germany","India","Iran","Italy","Japan","Luxembourg","Mexico","Nepal","Netherlands","NewZealand","Nigeria","Portugal","Singapore","SouthKorea","Spain","Sweden","Switzerland","Taiwan","Thailand","UK","USA","Vietnam")
-
-### If making new smaller regions sets, leave Hubei in as its own area, this script is not designed to remove any of the covariates
-
-# A smaller 31 location set with only one for China
-locations_31 <- c(locations_44[1:5],locations_44[13],"ChinaOther",locations_44[21:44])
-map_44_to_31 <- as.list(match(locations_31,locations_44))
-map_44_to_31[[7]] <- which(grepl("China",locations_44) & (!grepl("Hubei",locations_44)))
-
-# China only
-locations_15 <- locations_44[grepl("China",locations_44)]
-map_44_to_15 <- match(locations_15,locations_44)
-
-# The covariates 
-covariates <- c("intraContinent_distances","airTravel","Hubei_assymetry")
-
-### If adding more, make sure that the state space size is unique and the name matches this format exactly
-
-# Assemble
-location_list <- list(locations_31,locations_15)
-map_list <- list(map_44_to_31,map_44_to_15)
-xml_names <- c("real_data_31_state_analysis.xml","real_data_15_state_analysis.xml")
-
-full_xml <- scan("xml/original/real_data_44_state_analysis.xml",sep="\n",what=character(),blank.lines.skip=FALSE)
-
-for (i in 1:length(location_list)) {
-  this_xml <- full_xml
-  
-  nstates <- length(location_list[[i]])
-  
-  # Change DataType
-  dt_open <- grep('<generalDataType id="sampleLoc.dataType">',this_xml)
-  dt_close <- grep('</generalDataType>',this_xml)
-  before <- this_xml[1:dt_open]
-  after <- this_xml[dt_close:length(this_xml)]
[…]

### 2023-12-18 simulations/src/makeAnalysisXMLs.R

@@ -1,70 +1,0 @@
-pibuss_xml <- list.files("xml/piBUSS",full.names=TRUE)
-
-beastjar <- "~/git_repos/beast-mcmc/build/dist/beast.jar"
-beaglepath <- "/usr/local/lib"
-
-for (i in 1:length(pibuss_xml)) {
-  # Simulate
-  system(paste0("java  -Djava.library.path=",beaglepath," -jar ",beastjar," -working -overwrite ",pibuss_xml[i]))
-  
-  # Get XML template for analysis
-  id <- basename(pibuss_xml[i])
-  id <- strsplit(id,"_frac")[[1]][1]
-  if ( id == "nstates_44" ) {
-    template_file <- "xml/original/real_data_44_state_analysis.xml"
-  } else {
-    template_file <- list.files("xml/condensed_original/",full.names=TRUE)
-    template_file <- template_file[grepl(gsub("nstates_","",id),template_file)]
-  }
-  template <- scan(template_file,sep="\n",what=character(),blank.lines.skip=FALSE)
-  
-  # Get simulated data
-  seq_xml <- list.files("xml/piBUSS",full.names=TRUE)
-  seq_xml <- seq_xml[grepl("sequence",seq_xml) & grepl(gsub(".xml","",basename(pibuss_xml[i])),seq_xml)]
-  simulated_data <- scan(seq_xml,sep="\n",what=character(),blank.lines.skip=TRUE)
-  
-  # Make datatype codes upper-case because they come out of piBUSS that way
-  dt_open <- grep('<generalDataType id="sampleLoc.dataType">',template)
-  dt_close <- grep('</generalDataType>',template)
-  for (l in (dt_open+1):(dt_close-1)) {
-    line <- template[l]
-    has_code <- grepl("code",line)
-    has_alias <- grepl("alias",line)
-    line <- strsplit(line,'"')[[1]]
-    if (has_code) {
-      line[grep("code",line)+1] <- toupper(line[grep("code",line)+1])
-    }
-    if (has_alias) {
-      line[grep("state",line)+1] <- toupper(line[grep("state",line)+1])
-    }
-    template[l] <- paste0(line,collapse='"')
-  }
-  
-  # Locate where we need to put simulated data and add it
-  attribute_ends <- grep('</attr>',template)
-  sample_loc_starts <- grep('<attr name="sampleLoc">',template)
[…]

### 2023-12-18 simulations/src/makePiBUSS.R

@@ -1,118 +1,0 @@
-filepaths <- c("xml/original/real_data_44_state_analysis.xml",list.files("xml/condensed_original",full.names=TRUE))
-xmls <- lapply(filepaths,scan,sep="\n",what=character(),blank.lines.skip=FALSE)
-
-# piBUSS block
-pibuss_text <- 
-'	<beagleSequenceSimulator id="simulator" parallel="false" output="XML">
-		<partition from="1" to="1">
-			<treeModel idref="treeModel"/>
-			<glmSubstitutionModel idref="sampleLoc.model"/>
-			<siteModel idref="sampleLoc.siteModel"/>
-			<strictClockBranchRates idref="sampleLoc.branchRates"/>
-			<frequencyModel idref="frequencyModel"/>
-		</partition>
-	</beagleSequenceSimulator>
-
-	<report fileName="sequences.xml">
-		<beagleSequenceSimulator idref="simulator"/>
-	</report>
-'
-
-# Random-effects stuff
-weak_mean <- 0.0
-weak_sd <- 0.1
-
-strong_abs_mean <- log(2.0)
-strong_sd <- 0.1
-
-percent_strong <- c(0.01,0.5,1.0)
-
-cols <- viridis::rocket(2,begin=0.3,end=0.8)
-
-pdf(file="simulating_distributions.pdf",width=4,height=3)
-  par(mai=c(0.8,0.8,0.01,0.01))
-  curve(dlnorm(x,-strong_abs_mean,weak_sd),0,3,1001,add=FALSE,col=cols[2],lwd=2,
-        xlab="Rate-scale random-effect",ylab="Density")
-  curve(dlnorm(x,weak_mean,weak_sd),0,3,1001,add=TRUE,col=cols[1],lwd=2)
-  curve(dlnorm(x,strong_abs_mean,weak_sd),0,3,1001,add=TRUE,col=cols[2],lwd=2)
-dev.off()
-  
-set.seed(42)
-for (i in 1:3) {
-  nstates <- basename(filepaths[i])
-  nstates <- gsub("real_data_","",nstates)
-  nstates <- gsub("_state_analysis.xml","",nstates)
-  nstates <- as.numeric(nstates)
-  refxdim <- 2 * choose(nstates,2)
-  for (j in 1:3) {
-    sim_name <- paste0("nstates_",nstates,"_frac_",percent_strong[j])
-    
-    mean_vec <- rep(weak_mean,refxdim)
[…]
